---
title: "Reactive vs Preemptive Vaccination Strategies"
author: "Jong-Hoon Kim"
bibliography: references.bib
date: "2025-07-05"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---


# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(gridExtra)
library(ggpattern)
library(patchwork)
library(furrr)
library(purrr)
source("R/utils.R")
```

# Introduction

In the face of emerging infectious disease threats, public health officials must decide whether to deploy vaccines preemptively (before an outbreak occurs) or reactively (after an outbreak is detected). This strategic choice carries high stakes for both health outcomes and resource utilization. Preemptive vaccination can avert outbreaks by immunizing at-risk populations in advance, but it incurs upfront costs and risks vaccinating for an outbreak that might never occur. Reactive vaccination, in contrast, targets outbreaks once they are detected, potentially saving doses and avoiding unnecessary campaigns; however, losses accrue before control measures take effect [@Middleton2024; @Azman2019].

This dilemma is complicated by uncertainty. Outbreak risks vary between populations due to heterogeneous social, environmental, and biological factors, yet risk forecasts are imperfect [@Wells2020]. If authorities had perfect information, they could vaccinate high-risk populations in advance. In practice, risk estimates may be noisy, leading to misallocation of limited vaccines and potential failure to prevent major outbreaks [@Colijn2021].

Models from diverse diseases highlight that no one strategy is best across settings. Studies on mpox, hepatitis E, and MERS-CoV suggest that preemptive campaigns can be life-saving if outbreaks are severe or grow quickly [@Quilty2023; @Azman2019; @Wells2020]. Conversely, reactive strategies may use doses more efficiently when supply is constrained or preemptive coverage is uncertain [@Middleton2024].

Despite these insights, no unifying quantitative framework exists to compare reactive and preemptive vaccination under varying outbreak probability, resource availability, and risk uncertainty. Our study addresses this gap by developing an analytic and simulation-based framework for identifying the optimal allocation of vaccine resources — defined as the proportion $\alpha$ used preemptively — across populations with potentially heterogeneous and uncertain risk.

Through analytical derivation and simulation, we characterize the impact of outbreak probability ($p$), reactive efficacy ($r$), vaccine budget ($f$), and targeting accuracy (Spearman rank correlation $
ho$) on strategy performance. Our results offer practical guidance for policymakers deciding how much to invest in preemptive campaigns when vaccine stockpiles are constrained and risk signals are imperfect.

# The Model
## The Single Population Scenario

We evaluate the expected **total cost** of two strategies (preemptive vs reactive vaccination) for **a single population** at risk of an outbreak.

### Parameters

Let:

-   $D$: The full cost of an outbreak (e.g., cases, deaths, DALYs)
-   $C$: The cost of deploying a vaccination campaign
-   $p$: The probability of an outbreak ($0 < p < 1$)
-   $r$: The **relative effectiveness** of a reactive vaccination campaign compared to the preemptive vaccination campaign ($0 < r < 1$)

## Strategy 1: Preemptive Vaccination

-   We vaccinate regardless of whether an outbreak happens
-   Outbreak is fully averted if it would have occurred
-   Always pay vaccination cost $C$
-   Expected outbreak cost is $0$

$$
\text{Cost}_{\text{pre}} = C
$$

## Strategy 2: Reactive Vaccination

-   Vaccinate only if an outbreak occurs (probability $p$)
-   Outbreak is only partially mitigated: remaining cost is $(1 - r)D$
-   Vaccination cost $C$ is also incurred only when outbreak happens

$$
\text{Cost}_{\text{react}} = p \cdot (C + (1 - r)D)
$$

## Comparing the two strategies

We prefer **preemptive vaccination** when:

$$
C < p \cdot (C + (1 - r)D)
$$

Solving for $p$:

$$
p^* = \frac{C}{C + (1 - r)D}
$$

This is the **threshold outbreak probability** above which preemptive vaccination is cost-saving.

### Interpretation

-   As $r \to 1$, reactive vaccination becomes nearly as effective as preemptive -\> threshold $p^*$ increases
-   As the cost of an outbreak compared to the vaccination cost, $D / C$, becomes large -\> threshold $p^*$ decreases

## Symbolic derivation

```{python}
import sympy as sp

# Define symbols
C, D, p, r = sp.symbols('C D p r', positive=True)

# Define cost expressions
cost_pre = C
cost_react = p * (C + (1 - r)*D)

# Solve: when is preemptive better?
ineq = sp.Eq(cost_pre, cost_react)
p_thresh = sp.solve(ineq, p)[0]
p_thresh
```

Expected output: $$
\frac{C}{C + (1 - r)D}
$$

## Visualizing the threshold $p^*$

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)

# Create evaluation grid
grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C,
    cost_react =  p * (C + (1 - r)*D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Threshold lines: where cost_pre == cost_react
threshold_lines <- grid %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

labels_df <- data.frame(
  C = rep(C_vals, each = 2),
  label = rep(c("preemptive", "reactive"), 4),
  r = rep(c(0.05, 0.75), 4),
  p = rep(c(0.95, 0.05), 4)
)

ggplot(grid, aes(x = r, y = p)) +
  geom_line(data = threshold_lines, aes(x = r, y = p_thresh, group = C), 
            color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  scale_x_continuous(limits=c(0,1))+
  scale_fill_manual(values = c("preemptive" = "steelblue", "reactive" = "tomato")) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "vaccination strategy decision with threshold",
    x = expression("relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("outbreak probability" ~ italic(p)),
    fill = "Strategy"
  ) +
  theme_light() +
  theme(legend.position = 'top')
```

## Summary

This analysis shows that **preemptive vaccination** is preferred when:

$$
p > \frac{C}{C + (1 - r)D}
$$

This tradeoff depends jointly on: - **Probability of an outbreak** ($p$), - **Relative effectiveness of a reactive strategy** ($r$), - **Cost of prevention** ($C$), - **Cost of inaction** ($D$)

# Two-population scenario

We assume:

-   Two populations, each has an independent probability $p$ of outbreak
-   Only one vaccination campaign is available (can be used **preemptively** or **reactively** for one population)
-   Key parameters:

| Symbol | Description                                                      |
|----------------------------|--------------------------------------------|
| $X$    | Number of outbreaks ($X=0,1,2$)                                  |
| $p$    | Probability of outbreak in a population                          |
| $C$    | Cost of one vaccination campaign                                 |
| $D$    | Cost of one unmitigated outbreak                                 |
| $r$    | Relative effectiveness of **reactive** vaccination ($0 < r < 1$) |

## Preemptive strategy

We vaccinate **one population in advance**, say population A. No reactive response is used.

### Case 0: No outbreaks ($X = 0$)

-   Probability: $(1 - p)^2$
-   Cost: only unnecessary vaccination of A → $C$

$$
\text{Cost}_{X=0,\text{pre}} = C
$$ \### Case 1: One outbreak ($X = 1$)

-   Probability: $2p(1 - p)$
-   Two subcases (each with prob. $p(1 - p)$):
    -   If outbreak occurs in A (vaccinated): outbreak prevented -\> cost = $C$
    -   If outbreak occurs in B (unvaccinated): outbreak occurs -\> cost = $C + D$

Average cost:

$$
\text{Cost}_{X=1,\text{pre}} = 2p(1 - p)\left(C + \frac{D}{2}\right)
$$ \### Case 2: Two outbreaks ($X = 2$)

-   Probability: $p^2$
-   One outbreak prevented (in vaccinated population), one unmitigated -\> cost = $C + D$

$$
\text{Cost}_{X=2,\text{pre}} = p^2(C + D)
$$

### Total expected cost for the preemptive strategy

$$
\begin{aligned}
\mathbb{E}[\text{Cost}_{\text{pre}}] &= (1 - p)^2 C + 2p(1 - p)\left(C + \frac{D}{2} \right) + p^2(C + D) \\
&= C(1 - p)^2 + 2p(1 - p)C + p(1 - p)D + p^2 C + p^2 D \\
&= C[(1 - p)^2 + 2p(1 - p) + p^2] + D[p(1 - p) + p^2] \\
&= C + pD
\end{aligned}
$$

## Reactive strategy

We wait for outbreaks and respond to **one** if detected.

### Case 0: No outbreaks ($X = 0$)

-   Probability: $(1 - p)^2$
-   No vaccination needed, no outbreak → cost = $0$

$$
\text{Cost}_{X=0,\text{react}} = 0
$$

### Case 1: One outbreak ($X = 1$)

-   Probability: $2p(1 - p)$
-   Detect and respond to outbreak:
    -   Vaccination cost = $C$
    -   Partial mitigation: residual cost = $(1 - r)D$

$$
\text{Cost}_{X=1,\text{react}} = 2p(1 - p)\left(C + (1 - r)D\right)
$$

### Case 2: Two outbreaks ($X = 2$)

-   Probability: $p^2$
-   Only one outbreak can be addressed:
    -   Vaccination cost = $C$
    -   One outbreak mitigated -\> $(1 - r)D$
    -   One outbreak unmitigated -\> $D$

$$
\text{Cost}_{X=2,\text{react}} = p^2\left(C + (1 - r)D + D\right) = p^2(C + (2 - r)D)
$$

### Total Expected Cost – Reactive Strategy

$$
\begin{aligned}
\mathbb{E}[\text{Cost}_{\text{react}}] &= (1 - p)^2 \cdot 0 + 2p(1 - p)\left(C + (1 - r)D\right) + p^2\left(C + (2 - r)D\right) \\
&= 2p(1 - p)\left(C + (1 - r)D\right) + p^2\left(C + (2 - r)D\right)
\end{aligned}
$$

## Symbolic Comparison

Compare:

$$
\mathbb{E}[\text{Cost}_{\text{pre}}] = C + pD
$$

vs.

$$
\mathbb{E}[\text{Cost}_{\text{react}}] = (2p - p^2)C + \left[2p(1 - p)(1 - r) + p^2(2 - r)\right]D
$$

Preemptive strategy is preferred when:

$$
C + pD < (2p - p^2)C + \left[2p(1 - p)(1 - r) + p^2(2 - r)\right]D
$$

```{python}
from sympy import symbols, Eq, solve, simplify

# Define symbols
p, C, D, r = symbols('p C D r', positive=True)

# Total cost of each strategy
cost_pre = C + p * D
cost_react = (2*p - p**2)*C + (2*p*(1 - p)*(1 - r) + p**2*(2 - r)) * D

# Set inequality: preemptive better than reactive
ineq = Eq(cost_pre, cost_react)

# Solve for threshold p
p_thresh = solve(ineq, p)

# Simplify solution
simplified_thresh = [simplify(expr) for expr in p_thresh]

# Output
for i, expr in enumerate(simplified_thresh):
    print(f"Threshold p solution {i+1}:")
    print(expr)
```

```{r}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)

# Create evaluation grid
grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D,
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Threshold lines: where cost_pre == cost_react
threshold_lines <- grid %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

# Labels for each facet
labels_df <- data.frame(
  C = rep(C_vals, each = 2),
  label = rep(c("preemptive", "reactive"), 4),
  r = rep(c(0.05, 0.75), 4),
  p = rep(c(0.95, 0.05), 4)
)

ggplot(grid, aes(x = r, y = p)) +
  geom_line(data = threshold_lines, aes(x = r, y = p_thresh, group = C), 
            color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  scale_x_continuous(limits=c(0,1))+
  scale_fill_manual(values = c("preemptive" = "steelblue", "reactive" = "tomato")) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "vaccination strategy decision with threshold (n=2)",
    x = expression("relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("outbreak probability" ~ italic(p)),
    fill = "Strategy"
  ) +
  theme_light() +
  theme(legend.position = 'top')
```

## Compare single and two population scenarios

```{r}
grid1 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C,
    cost_react =  p * (C + (1 - r)*D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

grid2 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D,
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )


# Combine grids
grid1 <- grid1 %>% mutate(scenario = "Single population")
grid2 <- grid2 %>% mutate(scenario = "Two populations")
grid_all <- bind_rows(grid1, grid2)

# Threshold lines: compute for each scenario
threshold_lines <- grid_all %>%
  group_by(scenario, C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

# Labels for plot interpretation
labels_df <- expand.grid(
  C = C_vals,
  scenario =  c("Single population", "Two populations"),
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.1, 0.6),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

ggplot() +
  geom_line(data = threshold_lines, 
            aes(x = r, y = p_thresh, color = scenario, linetype = scenario, 
                group = interaction(scenario, C)),  linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  scale_color_manual(values = c("Single population" = "steelblue", "Two populations" = "tomato")) +
  scale_linetype_manual(values = c("Single population" = "solid", "Two populations" = "dashed")) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "Vaccination Strategy Thresholds by Scenario and Campaign Cost",
    x = expression("Relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("Outbreak probability" ~ italic(p)),
    color = "Scenario",
    linetype = "Scenario"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

## Multiple populations ($n>2$) scenario

We consider $n$ populations where each has probability $p$ of an outbreak. We can afford to vaccinate a fraction $f$ of them, pre-emptively or reactively.

Let:

-   $n$: number of populations
-   $p$: outbreak probability
-   $f$: fraction of populations we can vaccinate
-   $C$: vaccination cost per population
-   $D$: outbreak cost (if unmitigated)
-   $r$: relative effectiveness of reactive vaccination ($0 < r < 1$)

### Strategy 1: Preemptive Vaccination

We preemptively vaccinate $fn$ populations. If an outbreak occurs in the $(1 - f)n$ unvaccinated group, we pay $D$.

Total expected cost:

$$
\text{Cost}_{\text{pre}} = fn \cdot C + (1 - f)np \cdot D
$$

### Reactive vaccination strategy

## Step 2: Conditioning on $X = x$ outbreaks

For a fixed number of outbreaks $X = x$:

-   If $x \le k$, we can respond to **all** outbreaks.

    Total cost: $$
    \text{Cost}_{x,\text{react}} = x(C + (1 - r)D)
    $$

-   If $x > k$, we can respond to $k$ outbreaks, and $x - k$ go **unmitigated**.

    Mitigated cost: $$
    k(C + (1 - r)D)
    $$

    Unmitigated cost: $$
    (x - k)D
    $$

So the **expected cost when** $X = x$ is:

$$
\text{Cost}_{x,\text{react}} =
\begin{cases}
x(C + (1 - r)D), & \text{if } x \le k \\
k(C + (1 - r)D) + (x - k)D, & \text{if } x > k
\end{cases}
$$

## Step 3: Total Expected Cost

We now compute the total expected cost across all possible $x$:

$$
\mathbb{E}[\text{Cost}_{\text{react}}] =
\sum_{x = 0}^{n} \Pr(X = x) \cdot \text{Cost}_{x,\text{react}}
$$

Split into two parts:

### Case 1: $x \le k$

$$
\sum_{x = 0}^{k}
\binom{n}{x} p^x (1 - p)^{n - x} \cdot x(C + (1 - r)D)
$$

### Case 2: $x > k$

$$
\sum_{x = k+1}^{n}
\binom{n}{x} p^x (1 - p)^{n - x}
\cdot \left[k(C + (1 - r)D) + (x - k)D \right]
$$

This expression does **not simplify into closed form**, but it can be computed numerically or approximated.

### Strategy 2-1: Reactive Vaccination (Approximate)

We reactively vaccinate up to $fn$ populations **after** outbreaks occur.

Expected number of outbreaks: $np$\
We can respond to $\min(np, fn)$ and leave $\max(np - fn, 0)$ unmitigated.

Residual cost per mitigated outbreak = $(1 - r)D$

Expected cost:

$$
\text{Cost}_{\text{react}} =
\min(np, fn)(C + (1 - r)D) + \max(np - fn, 0) \cdot D
$$

## Symbolic Comparison using SymPy

```{python}
import sympy as sp

# Define symbols
n, p, f, C, D, r = sp.symbols('n p f C D r', positive=True)

# Expressions
np_ = n * p
fn_ = f * n

# Preemptive strategy cost
cost_pre = fn_ * C + (1 - f) * np_ * D

# Reactive strategy cost (piecewise via min/max)
covered   = sp.Min(np_, fn_)
uncovered = sp.Max(np_ - fn_, 0)

cost_react = covered * (C + (1 - r) * D) + uncovered * D

# Symbolic difference: reactive - preemptive
cost_diff = sp.simplify(cost_react - cost_pre)
cost_diff
```

## Decision Thresholds for Preemptive vs Reactive (Multiple Populations)

We compute and plot threshold outbreak probability $p^*$ at which preemptive vaccination becomes cheaper than reactive, under various vaccination cost levels $C$ and vaccination capacity levels $f$.

```{r}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.05, 0.1, 0.5)

n <- 1009  # or any fixed value

grid_all <- expand.grid(p = p_vals, r = r_vals, C = C_vals, f = f_vals) %>%
  mutate(
    cost_pre = n * (f * C + (1 - f) * p * D),
    covered = pmin(p, f),
    uncovered = pmax(p - f, 0),
    cost_react = n * (covered * (C + (1 - r) * D) + uncovered * D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Compute threshold lines: p where cost_pre == cost_react
threshold_lines <- grid_all %>%
  group_by(C, f, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))


# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )


# Plot with threshold lines colored by f
ggplot(threshold_lines, aes(x = r, y = p_thresh, color = factor(f), linetype = factor(f))) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("0.05" = "#1b9e77", "0.1" = "#d95f02", "0.5" = "#7570b3")) +
  scale_linetype_manual(values = c("0.05" = "solid", "0.1" = "dashed", "0.5" = "dotdash")) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title = "Threshold Outbreak Probability for Preemptive vs Reactive Strategy",
    subtitle = "Multiple population scenario with varying vaccination capacity f",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Vaccination capacity (f)",
    linetype = "Vaccination capacity (f)"
  ) +
  theme_light() +
  theme(legend.position = "top")
```

```{r}
# Exact cost computation for given parameters
compute_exact_costs <- function(n, p, C, D, r, f) {
  k <- floor(f * n)  # number of populations we can vaccinate

  probs <- dbinom(0:n, size = n, prob = p)

  # Preemptive: vaccinate k populations before outbreaks
  cost_pre <- sum(
    probs * (k * C + (1 - f) * (0:n) * D)  # Expected outbreak cost: only in unvaccinated
  )

  # Reactive: respond to up to k outbreaks after detection
  cost_react <- sum(
    probs * ifelse(
      0:n <= k,
      (0:n) * (C + (1 - r) * D),                    # All outbreaks mitigated
      k * (C + (1 - r) * D) + (0:n - k) * D         # Only k mitigated, rest unmitigated
    )
  )

  list(pre = cost_pre, react = cost_react)
}

find_threshold_p <- function(n, r, C, D, f, p_grid = seq(0.01, 0.99, by = 0.01)) {
  for (p in p_grid) {
    costs <- compute_exact_costs(n, p, C, D, r, f)
    if ((costs$pre - costs$react) < 0) {
      return(p)  # First p where preemptive is cheaper
    }
  }
  return(NA)
}
```

```{r}
# Parameters
n <- 20  # try 2 to compare with two-pop figure
r_vals <- seq(0.01, 0.99, by = 0.01)
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.05, 0.1, 0.5)
D <- 1

# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

# Grid of all thresholds
thresholds <- expand.grid(C = C_vals, f = f_vals, r = r_vals) %>%
  rowwise() %>%
  mutate(
    p_thresh = find_threshold_p(n = n, r = r, C = C, D = D, f = f)
  ) %>%
  ungroup()


ggplot(thresholds, aes(x = r, y = p_thresh, color = factor(f), linetype = factor(f))) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("0.05" = "#1b9e77", "0.1" = "#d95f02", "0.5" = "#7570b3")) +
  scale_linetype_manual(values = c("0.05" = "solid", "0.1" = "dashed", "0.5" = "dotdash")) +
  labs(
    title = "Exact Threshold Outbreak Probability for Preemptive vs Reactive",
    subtitle = paste("Based on Binomial outbreak distribution with n =", n),
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Vaccination capacity (f)",
    linetype = "Vaccination capacity (f)"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

```{r}
library(ggplot2)
library(dplyr)

# --- Parameters ---
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.4)  # exact overlay for f = 0.5
n <- 20            # exact enumeration for n = 2

# --- Function: exact expected cost given outbreak count ---
compute_exact_costs <- function(n, p, C, D, r, f) {
  k <- floor(f * n)
  x_vals <- 0:n
  probs <- dbinom(x_vals, size = n, prob = p)

  # Preemptive: vaccinate k units, pay outbreak cost only for (1 - f)
  cost_pre <- sum(probs * (k * C + (1 - f) * x_vals * D))

  # Reactive: cover up to k outbreaks after detection
  cost_react <- sum(probs * ifelse(
    x_vals <= k,
    x_vals * (C + (1 - r) * D),
    k * (C + (1 - r) * D) + (x_vals - k) * D
  ))

  list(pre = cost_pre, react = cost_react)
}

# --- Function: Find p* threshold where preemptive becomes better ---
find_threshold_p <- function(n, r, C, D, f, p_grid = seq(0.01, 0.99, by = 0.01)) {
  for (p in p_grid) {
    costs <- compute_exact_costs(n, p, C, D, r, f)
    if ((costs$pre - costs$react) < 0) {
      return(p)
    }
  }
  return(NA)
}

# --- Compute threshold p* for each (r, C) using exact method ---
threshold_exact <- expand.grid(C = C_vals, f = f_vals, r = r_vals) %>%
  rowwise() %>%
  mutate(
    p_thresh = find_threshold_p(n = n, r = r, C = C, D = D, f = f)
  ) %>%
  ungroup() %>%
  mutate(method = "Exact (n = 2)", f = factor(f))

# --- Load two-population approximation for comparison ---
grid2 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D
  )

threshold_two_pop <- grid2 %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh)) %>%
  mutate(method = "Two population approx", f = factor(0.5))

# --- Combine and plot ---
threshold_combined <- bind_rows(threshold_exact, threshold_two_pop)

# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

ggplot(threshold_combined, aes(x = r, y = p_thresh, color = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("Exact (n = 2)" = "#1b9e77", "Two population approx" = "#d95f02")) +
  scale_linetype_manual(values = c("Exact (n = 2)" = "solid", "Two population approx" = "dashed")) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title = "Threshold Outbreak Probability for Preemptive vs Reactive (f = 0.5)",
    subtitle = "Comparison between exact enumeration for n = 2 and two-population approximation",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Method",
    linetype = "Method"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

## Asymptotic Limit: Preemptive vs Reactive

As the number of populations $n \to \infty$ and vaccination capacity fraction $f \to 0$ (i.e., resources are extremely limited), the expected costs for preemptive and reactive strategies simplify.

Let:

-   $p$: probability of an outbreak per population,
-   $r$: relative effectiveness of reactive vaccination,
-   $C$: cost of one vaccination campaign,
-   $D$: cost of one unmitigated outbreak,
-   $fn$: number of populations we can afford to vaccinate (preemptively or reactively)

### Preemptive Expected Cost

Total cost:

$$
\text{Cost}_{\text{pre}} = fn \cdot C + (1 - f)np \cdot D
$$

As $f \to 0$, this becomes:

$$
\text{Cost}_{\text{pre}} \approx npD
$$

### Reactive Expected Cost

Using the expression:

$$
\text{Cost}_{\text{react}} = \min(fn, np)(C + (1 - r)D) + \max(np - fn, 0) \cdot D
$$

If $fn \ll np$:

$$
\text{Cost}_{\text{react}} \approx fn(C + (1 - r)D) + (np - fn)D
$$

### Cost Difference

Compute:

$$
\begin{aligned}
\Delta &= \text{Cost}_{\text{pre}} - \text{Cost}_{\text{react}} \\
&= fnC + (1 - f)npD - \left[fn(C + (1 - r)D) + (np - fn)D\right] \\
&= fnD(r - p)
\end{aligned}
$$

### Result

-   If $r < p$: preemptive is cheaper ($\Delta < 0$)
-   If $r > p$: reactive is cheaper ($\Delta > 0$)

Thus, in the limit:

$$
p > r \quad \Rightarrow \quad \text{preemptive is preferred}
$$

This corresponds to the decision threshold line $p^* = r$ as $f \to 0$ and $n \to \infty$.

```{r}
# Overlay limit line p = r
ggplot(threshold_combined, aes(x = r, y = p_thresh, color = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.7, label = "p = r (limit)", angle = 45, hjust = -0.1, vjust = 1.2, size = 3.5) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("Exact (n = 2)" = "#1b9e77", "Two population approx" = "#d95f02")) +
  scale_linetype_manual(values = c("Exact (n = 2)" = "solid", "Two population approx" = "dashed")) +
  labs(
    title = "Threshold Outbreak Probability with Asymptotic Line p = r",
    subtitle = "Black dotted line shows p = r decision boundary as n → ∞ and f → 0",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Method",
    linetype = "Method"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

# Optimizing Hybrid Vaccination Strategy

## Problem

> Given a limited vaccine supply, how should we split our efforts between **preemptive** and **reactive** vaccination strategies to **minimize total expected cost**?

We define:

-   $\alpha \in [0, 1]$: fraction of vaccination resources used **preemptively**
-   $1 - \alpha$: fraction used **reactively**
-   $f$: total fraction of the population that can be vaccinated
-   $n$: total number of populations
-   $p$: probability of outbreak in each population
-   $C$: cost of one vaccination campaign
-   $D$: cost of one unmitigated outbreak
-   $r$: relative effectiveness of reactive vaccination to the preemptive strategy

## Expected Cost Formula

The total expected cost of a hybrid strategy that uses $\alpha$ of the vaccination capacity preemptively is:

$$
\text{Cost}(\alpha) =
\underbrace{\alpha f n C}_{\text{preemptive vaccination}} +
\underbrace{\min(x, k_{\text{react}})(C + (1 - r)D)}_{\text{reactive cost}} +
\underbrace{\max(x - k_{\text{react}}, 0) D}_{\text{unmitigated outbreaks}}
$$

Where: - $x = (1 - \alpha f)np$: expected number of outbreaks in unvaccinated group - $k_{\text{react}} = (1 - \alpha)fn$: number of reactive campaigns available

### Visualizing Cost of Hybrid Strategy

We explore how the total expected cost varies with the fraction $\alpha$ of vaccine budget used preemptively.

```{r}
library(ggplot2)
library(dplyr)

# Parameters (adjustable)
f <- 0.5     # total vaccination capacity
n <- 100     # number of populations
p <- 0.3     # outbreak probability per population
C <- 1       # cost of a vaccination campaign
D <- 5       # cost of an unmitigated outbreak
r <- 0.4     # relative effectiveness of reactive vaccination

# Alpha grid: preemptive share
alpha_vals <- seq(0, 1, by = 0.01)

# Compute total cost for each alpha
cost_df <- data.frame(alpha = alpha_vals) %>%
  mutate(
    k_pre = alpha * f * n,
    k_react = (1 - alpha) * f * n,
    x = (1 - alpha * f) * n * p,
    covered = pmin(x, k_react),
    uncovered = pmax(x - k_react, 0),
    cost = k_pre * C + covered * (C + (1 - r) * D) + uncovered * D
  )
```

Plot
```{r}
ggplot(cost_df, aes(x = alpha, y = cost)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(data = cost_df %>% slice_min(cost), aes(x = alpha, y = cost), 
             color = "firebrick", size = 3) +
  annotate("text", 
           x = cost_df$alpha[which.min(cost_df$cost)], 
           y = min(cost_df$cost), 
           label = "Minimum", 
           hjust = -0.2, vjust = -1, color = "firebrick") +
  labs(
    title = expression("Expected Total Cost of Hybrid Strategy vs " ~ alpha),
    subtitle = paste0("Parameters: p = ", p, ", f = ", f, ", n = ", n, 
                      ", C = ", C, ", D = ", D, ", r = ", r),
    x = expression("Fraction of vaccine budget used preemptively" ~ (alpha)),
    y = "Expected total cost"
  ) +
  theme_light()
```

Optimum $\alpha$

```{r}
D <- 1
p_vals <- seq(0.05, 0.95, by = 0.05)
r_vals <- seq(0.05, 0.95, by = 0.05)
C_vals <- c(0.1, 0.5, 1,  2)
f_vals <- c(0.1, 0.2, 0.5)
n_vals <- 100

#Optimal $\alpha$ for each parameter set

grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals, f = f_vals, n = n_vals)

grid$alpha_opt <- 
  map_dbl(1:nrow(grid), function(i) {
    row <- grid[i, ]
    optimize_alpha(
      risk_profile = rep(row$p, row$n),
      C = row$C,
      D = D,
      r = row$r,
      f = row$f,
      method = "optimize")$alpha_opt})
```

Plot

```{r}
grid <- grid %>%  mutate(DC_ratio = D / C)

grid_plot <- grid %>% filter(f == 0.1)

ggplot(grid_plot, aes(x = r, y = p, fill = alpha_opt)) +
  geom_tile() +
  facet_wrap(~DC_ratio, labeller = label_bquote(D/C == .(DC_ratio)), nrow = 2) +
  scale_fill_viridis_c(option = "D", alpha=0.7, name = expression("Optimal" ~ alpha^"*")) +
  labs(title = expression("Optimal Preemptive Allocation (" * alpha^"*" * ") for Budget " * italic(f) * " = 0.25"),
    subtitle = "Faceted by cost ratio (D/C)",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Mean outbreak probability" ~ italic(p))) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


## Heterogeneous risk

### Beta distribution

Alternative parameterization of $\text{Beta}(a,b)$ is using mean $p$ and the sample size (or concentration)

$$a=p\kappa$$, 
$$b=(1-p)\kappa$$
$$\text{Var}(X) = \frac{p(1-p)}{1+\kappa}$$


Find optimal $\alpha$


```{r}
# --- Grid parameters ---
D <- 1
p_vals <- seq(0.05, 0.95, by = 0.05)
r_vals <- seq(0.05, 0.95, by = 0.05)
C_vals <- c(0.1, 0.5, 1,  2)
f_vals <- c(0.1, 0.2, 0.5)
concentration <- 20
n <- 100
set.seed(42)

grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals, f = f_vals)

risk_profiles <- pmap(grid, function(p, r, C, f) {
  a <- p * concentration
  b <- (1 - p) * concentration
  generate_risk_profile(n=n, dist="beta", params=list(a=a, b=b), seed=42)
})
```

Parallel session although overhead is larger than the performance gain in this case
```{r}
plan(multisession, workers = parallel::detectCores() - 1) 
risk_profiles2 <- future_pmap(grid, function(p, r, C, f) {
  a <- p * concentration
  b <- (1 - p) * concentration
  generate_risk_profile(n=n, dist="beta", params=list(a=a, b=b), seed=42)})

plan(sequential)  # Reset to sequential execution
# > all.equal(risk_profiles, risk_profiles2)
# [1] TRUE
```

Optimal $\alpha$ for each parameter set
```{r}
grid$alpha_opt <- map2_dbl(risk_profiles, seq_len(nrow(grid)), function(profile, i) {
  row <- grid[i, ]
  optimize_alpha(
    risk_profile = profile,
    C = row$C,
    D = D,
    r = row$r,
    f = row$f,
    method = "optimize"
  )$alpha_opt
})

grid <- grid %>%  mutate(DC_ratio = D / C)
```

```{r}
grid_plot <- grid %>% filter(f == 0.1)

ggplot(grid_plot, aes(x = r, y = p, fill = alpha_opt)) +
  geom_tile() +
  facet_wrap(~DC_ratio, labeller = label_bquote(D/C == .(DC_ratio)), nrow = 2) +
  scale_fill_viridis_c(option = "D", alpha=0.7, name = expression("Optimal" ~ alpha^"*")) +
  labs(title = expression("Optimal Preemptive Allocation (" * alpha^"*" * ") for Budget " * italic(f) * " = 0.25"),
    subtitle = "Faceted by cost ratio (D/C)",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Mean outbreak probability" ~ italic(p))) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Imperfect assessment

### Generating noisy ranks with a specified correlation to the original sequence

We aim to generate a noisy version of a standardized signal vector `p_z` such that the new vector `z_hat` has a target **Pearson correlation** $r$ with `p_z`.


The following formula is used to construct the noisy vector:

```{r}
z_hat <- r * p_z + sqrt(1 - r^2) * noise
```

where:

- `p_z` is a normalized version of the original signal (mean 0, variance 1)
- `noise` is an independent standard normal vector
- `rho` is the desired correlation between `z_hat` and `p_z`


This is a linear combination of two independent standard normal vectors:

$$
Z_{\text{hat}} = r \cdot Z_{\text{signal}} + \sqrt{1 - r^2} \cdot Z_{\text{noise}}
$$

where:

- $Z_{\text{signal}} \sim \mathcal{N}(0, 1)$
- $Z_{\text{noise}} \sim \mathcal{N}(0, 1)$
- $\operatorname{Cov}(Z_{\text{signal}}, Z_{\text{noise}}) = 0$

The variance of the resulting vector is:

$$
\begin{align}
\operatorname{Var}(Z_{\text{hat}}) &= r^2 \cdot \operatorname{Var}(Z_{\text{signal}}) + (1 - r^2) \cdot \operatorname{Var}(Z_{\text{noise}}) \\
&= r^2 + (1 - r^2) = 1
\end{align}
$$

The covariance between `z_hat` and `p_z` is:

$$
\operatorname{Cov}(Z_{\text{hat}}, Z_{\text{signal}}) = r \cdot \operatorname{Var}(Z_{\text{signal}}) = r
$$

Therefore, the Pearson correlation is:

$$
\operatorname{Cor}(Z_{\text{hat}}, Z_{\text{signal}}) = \frac{\operatorname{Cov}(Z_{\text{hat}}, Z_{\text{signal}})}{\sqrt{\operatorname{Var}(Z_{\text{hat}}) \cdot \operatorname{Var}(Z_{\text{signal}})}} = \frac{r}{1} = r
$$

Once we know how to generate a sequence with a specified Pearson's $r$, given a sequence, we can also generate a sequence with desired Spearman's $\rho$ because Spearman's $\rho$ is a Pearson's $r$ applied to ranked sequence. Similarly, using the relationship between Spearman's $\rho$, we can generate a sequence with a desired Kendall's $tau$.   

#### test the function

```{r}
set.seed(12)
p <- rbeta(1e4, 2, 5)
target_rho <- 0.8
target_tau <- 0.6
noisy1 <- generate_noisy_rank(p, target_rho = target_rho)
# Spearman correlation should match
cor(rank(p), rank(noisy1), method = "spearman")  # ~ target_rho
noisy2 <- generate_noisy_rank(p, target_tau = target_tau)
cor(noisy2, p, method = "kendall")    # ~0.6
```


```{r}
# Example
p <- runif(1000)  # Random risk scores
noisy <- generate_noisy_rank(p, target_tau = 0.6, seed = 12)
cor(p, noisy, method = "kendall")  # Should be approximately 0.6
```


```{r}
rho <- 0.8
p_hat <- generate_noisy_rank_with_rho(p, target_rho = rho)
target_order <- order(p_hat, decreasing = TRUE)
p_sorted <- p[target_order]  # true risks sorted by noisy estimates

```

```{r}
compute_total_cost_given_alpha_imperfect <- function(alpha, p_true, rho, C, D, r, f = 1) {
  n <- length(p_true)
  
  # 1. Generate imperfect estimates
  p_hat <- generate_noisy_rank_with_rho(p_true, target_rho = rho)
  
  # 2. Rank by estimated risk, apply to true risks
  target_order <- order(p_hat, decreasing = TRUE)
  p_sorted <- p_true[target_order]
  
  # 3. Vaccinate top-k
  k <- floor(alpha * f * n)
  
  # 4. Compute cost
  cost_preemptive <- k * C
  cost_reactive <- sum(p_sorted[(k + 1):n]) * (1 - r) * D
  
  total_cost <- cost_preemptive + cost_reactive
  return(total_cost)
}

```



```{r}
n <- 100
p_true <- rbeta(n, 2, 5)

# Parameters
C <- 1
D <- 10
r <- 0.6
f <- 0.1

# Run for rho = 1 (perfect targeting)
res_perfect <- optimize_alpha(
  risk_profile = p_true,
  C = C,
  D = D,
  r = r,
  f = f,
  rho = NULL,           # NULL → use perfect targeting
  method = "optimize"
)

# Run for rho = 0.6 (imperfect targeting)
res_imperfect <- optimize_alpha(
  risk_profile = p_true,
  C = C,
  D = D,
  r = r,
  f = f,
  rho = 0.6,
  method = "optimize"
)

# Show results
cat("Perfect targeting (rho = 1):\n")
cat(sprintf("  alpha* = %.3f\n  total cost = %.3f\n\n", res_perfect$alpha_opt, res_perfect$cost_opt))

cat("Imperfect targeting (rho = 0.6):\n")
cat(sprintf("  alpha* = %.3f\n  total cost = %.3f\n", res_imperfect$alpha_opt, res_imperfect$cost_opt))

```

```{r}
p_vals <- seq(0.3, 0.5, 0.7)
r_vals <- c(0.2, 0.5, 0.8)
f_vals <- c(0.1)
C_vals <- c(0.3, 1, 2, 5)
rho_vals <- seq(0.1, 0.9, by=0.05)
D <- 2
n <- 100
concentration <- 20
set.seed(42)

# Full parameter grid excluding rho
grid_core <- expand.grid(p = p_vals, r = r_vals, f = f_vals, C = C_vals)

# Add D/C ratio
grid_core <- grid_core %>% mutate(DC_ratio = D / C)

# Generate true risk profiles (same for all rhos in the same scenario)
risk_profiles <- pmap(
  grid_core %>% select(p, r, f, C),  # only use needed columns
  function(p, r, f, C) {
    a <- p * concentration
    b <- (1 - p) * concentration
    generate_risk_profile(n = n, dist = "beta", params = list(a = a, b = b))
  }
)
```

```{r}
grid_core$alpha_ref <- map2_dbl(risk_profiles, seq_len(nrow(grid_core)), function(profile, i) {
  row <- grid_core[i, ]
  optimize_alpha(
    risk_profile = profile,
    C = row$C,
    D = D,
    r = row$r,
    f = row$f,
    rho = NULL  # perfect targeting
  )$alpha_opt
})
```

```{r}
# Expand the grid to include each rho
grid_with_rho <- grid_core %>%
  slice(rep(1:n(), each = length(rho_vals))) %>%
  mutate(rho = rep(rho_vals, times = nrow(grid_core)))

# Repeat risk profiles accordingly
risk_profiles_rho <- rep(risk_profiles, each = length(rho_vals))

# Compute alpha for each rho
grid_with_rho$alpha_rho <- map2_dbl(risk_profiles_rho, seq_len(nrow(grid_with_rho)), function(profile, i) {
  row <- grid_with_rho[i, ]
  optimize_alpha(
    risk_profile = profile,
    C = row$C,
    D = D,
    r = row$r,
    f = row$f,
    rho = row$rho
  )$alpha_opt
})

# Compute Δalpha = alpha_rho - alpha_ref
grid_with_rho$alpha_ref <- rep(grid_core$alpha_ref, each = length(rho_vals))
grid_with_rho$delta_alpha <- grid_with_rho$alpha_rho - grid_with_rho$alpha_ref

```

```{r}
ggplot(grid_with_rho, aes(x = rho, y = delta_alpha, group = interaction(p, r), color = p)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(f ~ DC_ratio, labeller = label_bquote(rows = f == .(f), cols = D/C == .(DC_ratio))) +
  scale_color_viridis_c(name = "True\nmean p", option = "D") +
  labs(
    title = expression("Change in Optimal Preemptive Allocation (" * Delta * alpha^"*" * ") with Imperfect Risk Assessment"),
    subtitle = expression(Delta * alpha^"*" * " = " * alpha^"*_{rho} - " * alpha^"*_{perfect}"),
    x = expression("Spearman rank correlation (" * rho * ")"),
    y = expression(Delta * alpha^"*")
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")

```

```{r}
p_vals <- c(0.3, 0.55, 0.8)
r_vals <- c(0.3, 0.55, 0.8)
f_vals <- c(0.1)
C_vals <- c(1, 2, 5, 10)
rho_vals <- seq(0.1, 0.9, by=0.1)
D <- 10
n <- 1000
concentration <- 20
n_sim <- 200

# Base grid (1 row per simulation and scenario excluding rho)
grid_core <- expand.grid(
  sim = 1:n_sim,
  p = p_vals,
  r = r_vals,
  C = C_vals,
  f = f_vals
) %>%
  mutate(DC_ratio = D / C)

# Expand over rho
grid_rho <- grid_core %>%
  slice(rep(1:n(), each = length(rho_vals))) %>%
  mutate(rho = rep(rho_vals, times = nrow(grid_core)))

# Generate unique risk profiles once
grid_core_unique <- distinct(grid_core, p, r, C, f, .keep_all = TRUE)
risk_profiles_unique <- pmap(
  grid_core_unique %>% select(p, r, C, f),
  function(p, r, C, f) {
    a <- p * concentration
    b <- (1 - p) * concentration
    replicate(n_sim, generate_risk_profile(n = n, dist = "beta", params = list(a = a, b = b)))
  }
)


simulate_alpha_diff <- function(p, r, C, f, rho, sim, risk_profiles_unique, grid_core_unique) {
  profile_idx <- which(grid_core_unique$p == p & grid_core_unique$r == r & 
                       grid_core_unique$C == C & grid_core_unique$f == f)
  p_true <- risk_profiles_unique[[profile_idx]][, sim]
  alpha_ref <- optimize_alpha(
    risk_profile = p_true, C = C, D = D, r = r, f = f, rho = NULL, method = "grid", by = 0.05
  )$alpha_opt
  alpha_rho <- optimize_alpha(
    risk_profile = p_true, C = C, D = D, r = r, f = f, rho = rho, method = "grid", by = 0.05
  )$alpha_opt
  tibble(p, r, C, f, rho, sim, alpha_ref, alpha_rho, delta_alpha = alpha_rho - alpha_ref)
}
# run scenarios
results <- pmap_dfr(
  grid_rho %>% select(p, r, C, f, rho, sim),
  simulate_alpha_diff,
  risk_profiles_unique = risk_profiles_unique,
  grid_core_unique = grid_core_unique
)

library(furrr)
plan(multisession, workers = parallel::detectCores() - 1)  # Use all but one core
# results <- future_pmap_dfr(
#   grid_rho %>% select(p, r, C, f, rho, sim),
#   simulate_alpha_diff,
#   .options = furrr_options(seed = TRUE)  # Ensure reproducible random numbers
# )

results <- future_pmap_dfr(
  grid_rho %>% select(p, r, C, f, rho, sim),
  simulate_alpha_diff,
  risk_profiles_unique = risk_profiles_unique,
  grid_core_unique = grid_core_unique,
  .options = furrr_options(seed = TRUE) 
)
plan(sequential)  # Reset to sequential execution

# Function to simulate one scenario

# simulate_alpha_diff <- function(p, r, C, f, rho, sim) {
#   a <- p * concentration
#   b <- (1 - p) * concentration
#   p_true <- generate_risk_profile(n = n, dist = "beta", params = list(a = a, b = b))
# 
#   alpha_ref <- optimize_alpha(
#     risk_profile = p_true,
#     C = C,
#     D = D,
#     r = r,
#     f = f,
#     rho = NULL,
#     method = "optimize"
#   )$alpha_opt
# 
#   alpha_rho <- optimize_alpha(
#     risk_profile = p_true,
#     C = C,
#     D = D,
#     r = r,
#     f = f,
#     rho = rho,
#     method = "optimize"
#   )$alpha_opt
# 
#   tibble(p, r, C, f, rho, sim, alpha_ref, alpha_rho, delta_alpha = alpha_rho - alpha_ref)
# }
# 
# # Run all simulations
# results <- pmap_dfr(
#   grid_rho %>% select(p, r, C, f, rho, sim),  # only pass needed columns
#   simulate_alpha_diff
# )
```

```{r}
results <- results %>%
  mutate(DC_ratio = D / C)

results %>% filter(r == 0.3) %>% 
  ggplot(aes(x = rho, y = delta_alpha)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  stat_summary(fun = median, geom = "line", color = "black", linewidth = 1) +
  stat_summary(fun.data = ~quantile(., probs = c(0.25, 0.75)) %>%
                 {data.frame(y = median(.), ymin = .[1], ymax = .[2])},
               geom = "ribbon", alpha = 0.2, fill = "gray60") +
  facet_grid(p ~ DC_ratio, labeller = label_bquote(rows = p == .(p), cols = D/C == .(round(DC_ratio, 1)))) +
  labs(
    title = expression("Effect of Imperfect Risk Ranking on Optimal Preemptive Allocation"),
    subtitle = expression(Delta * alpha^"*" * " = " * alpha[rho]^"*" * " - " * alpha[perfect]^"*"),
    x = expression("Spearman rank correlation (" * rho * ")"),
    y = expression(Delta * alpha^"*")
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

```

$alpha$

```{r}
results %>% filter(rho == 0.9) %>% 
  ggplot(aes(x = r, y = alpha_ref)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  stat_summary(fun = median, geom = "line", color = "black", linewidth = 1) +
  stat_summary(fun.data = ~quantile(., probs = c(0.25, 0.75)) %>%
                 {data.frame(y = median(.), ymin = .[1], ymax = .[2])},
               geom = "ribbon", alpha = 0.2, fill = "gray60") +
  facet_grid(p ~ DC_ratio, labeller = label_bquote(rows = p == .(p), cols = D/C == .(round(DC_ratio, 1)))) +
  labs(
    title = expression("Effect of Imperfect Risk Ranking on Optimal Preemptive Allocation"),
    subtitle = expression(Delta * alpha^"*" * " = " * alpha[rho]^"*" * " - " * alpha[perfect]^"*"),
    x = expression("Spearman rank correlation (" * rho * ")"),
    y = expression(Delta * alpha^"*")
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

```

# Future work
1. Incidence rate rather than the presence or absence of the outbreak (or in a way always outbreak)
2. It is critical how we measure the cost of the outbreak / Cost-benefit or cost-effectiveness 
3. The ratio is also a function of the outbreak size.
