---
title: "Reactive vs Preemptive Vaccination Strategies"
author: "Jong-Hoon Kim"
date: "2025-07-05"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---

# The Single Population Scenario

We evaluate the expected **total cost** of two strategies (preemptive vs reactive vaccination) for **a single population** at risk of an outbreak.

## Parameters

Let:

- $D$: The full cost of an outbreak (e.g., cases, deaths, DALYs)
- $C$: The cost of deploying a vaccination campaign
- $p$: The probability of an outbreak ($0 < p < 1$)
- $r$: The **relative effectiveness** of a reactive vaccination campaign compared to the preemptive vaccination campaign ($0 < r < 1$)

## Strategy 1: Preemptive Vaccination

-   We vaccinate regardless of whether an outbreak happens
-   Outbreak is fully averted if it would have occurred
-   Always pay vaccination cost $C$
-   Expected outbreak cost is $0$

$$
\text{Cost}_{\text{pre}} = C
$$

## Strategy 2: Reactive Vaccination

-   Vaccinate only if an outbreak occurs (probability $p$)
-   Outbreak is only partially mitigated: remaining cost is $(1 - r)D$
-   Vaccination cost $C$ is also incurred only when outbreak happens

$$
\text{Cost}_{\text{react}} = p \cdot (C + (1 - r)D)
$$

## Comparing the two strategies

We prefer **preemptive vaccination** when:

$$
C < p \cdot (C + (1 - r)D)
$$

Solving for $p$:

$$
p^* = \frac{C}{C + (1 - r)D}
$$

This is the **threshold outbreak probability** above which preemptive vaccination is cost-saving.

### Interpretation

-   As $r \to 1$, reactive vaccination becomes nearly as effective as preemptive -> threshold $p^*$ increases 
-   As the cost of an outbreak compared to the vaccination cost, $D / C$, becomes large -> threshold $p^*$ decreases

## Symbolic derivation 

```{python}
import sympy as sp

# Define symbols
C, D, p, r = sp.symbols('C D p r', positive=True)

# Define cost expressions
cost_pre = C
cost_react = p * (C + (1 - r)*D)

# Solve: when is preemptive better?
ineq = sp.Eq(cost_pre, cost_react)
p_thresh = sp.solve(ineq, p)[0]
p_thresh
```

Expected output:
$$
\frac{C}{C + (1 - r)D}
$$

## Visualizing the threshold $p^*$

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)

# Create evaluation grid
grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C,
    cost_react =  p * (C + (1 - r)*D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Threshold lines: where cost_pre == cost_react
threshold_lines <- grid %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

labels_df <- data.frame(
  C = rep(C_vals, each = 2),
  label = rep(c("preemptive", "reactive"), 4),
  r = rep(c(0.05, 0.75), 4),
  p = rep(c(0.95, 0.05), 4)
)

ggplot(grid, aes(x = r, y = p)) +
  geom_line(data = threshold_lines, aes(x = r, y = p_thresh, group = C), 
            color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  scale_x_continuous(limits=c(0,1))+
  scale_fill_manual(values = c("preemptive" = "steelblue", "reactive" = "tomato")) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "vaccination strategy decision with threshold",
    x = expression("relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("outbreak probability" ~ italic(p)),
    fill = "Strategy"
  ) +
  theme_light() +
  theme(legend.position = 'top')
```

## Summary

This analysis shows that **preemptive vaccination** is preferred when:

$$
p > \frac{C}{C + (1 - r)D}
$$

This tradeoff depends jointly on: - **Probability of an outbreak** ($p$), - **Relative effectiveness of a reactive strategy** ($r$), - **Cost of prevention** ($C$), - **Cost of inaction** ($D$)

# Two-population scenario

We assume:

-   Two populations, each has an independent probability $p$ of outbreak
-   Only one vaccination campaign is available (can be used **preemptively** or **reactively** for one population)
-   Key parameters:

| Symbol | Description                                                      |
|----------------------------|--------------------------------------------|
| $X$    | Number of outbreaks ($X=0,1,2$)                                  |
| $p$    | Probability of outbreak in a population                          |
| $C$    | Cost of one vaccination campaign                                 |
| $D$    | Cost of one unmitigated outbreak                                 |
| $r$    | Relative effectiveness of **reactive** vaccination ($0 < r < 1$) |

## Preemptive strategy

We vaccinate **one population in advance**, say population A. No reactive response is used.

### Case 0: No outbreaks ($X = 0$)

-   Probability: $(1 - p)^2$
-   Cost: only unnecessary vaccination of A → $C$

$$
\text{Cost}_{X=0,\text{pre}} = C
$$
### Case 1: One outbreak ($X = 1$)

-   Probability: $2p(1 - p)$
-   Two subcases (each with prob. $p(1 - p)$):
    -   If outbreak occurs in A (vaccinated): outbreak prevented -> cost = $C$
    -   If outbreak occurs in B (unvaccinated): outbreak occurs -> cost = $C + D$

Average cost:

$$
\text{Cost}_{X=1,\text{pre}} = 2p(1 - p)\left(C + \frac{D}{2}\right)
$$
### Case 2: Two outbreaks ($X = 2$)

-   Probability: $p^2$
-   One outbreak prevented (in vaccinated population), one unmitigated -> cost = $C + D$

$$
\text{Cost}_{X=2,\text{pre}} = p^2(C + D)
$$


### Total expected cost for the preemptive strategy

$$
\begin{aligned}
\mathbb{E}[\text{Cost}_{\text{pre}}] &= (1 - p)^2 C + 2p(1 - p)\left(C + \frac{D}{2} \right) + p^2(C + D) \\
&= C(1 - p)^2 + 2p(1 - p)C + p(1 - p)D + p^2 C + p^2 D \\
&= C[(1 - p)^2 + 2p(1 - p) + p^2] + D[p(1 - p) + p^2] \\
&= C + pD
\end{aligned}
$$

## Reactive strategy

We wait for outbreaks and respond to **one** if detected.

### Case 0: No outbreaks ($X = 0$)

-   Probability: $(1 - p)^2$
-   No vaccination needed, no outbreak → cost = $0$

$$
\text{Cost}_{X=0,\text{react}} = 0
$$



### Case 1: One outbreak ($X = 1$)

-   Probability: $2p(1 - p)$
-   Detect and respond to outbreak:
    -   Vaccination cost = $C$
    -   Partial mitigation: residual cost = $(1 - r)D$

$$
\text{Cost}_{X=1,\text{react}} = 2p(1 - p)\left(C + (1 - r)D\right)
$$

### Case 2: Two outbreaks ($X = 2$)

-   Probability: $p^2$
-   Only one outbreak can be addressed:
    -   Vaccination cost = $C$
    -   One outbreak mitigated -> $(1 - r)D$
    -   One outbreak unmitigated -> $D$

$$
\text{Cost}_{X=2,\text{react}} = p^2\left(C + (1 - r)D + D\right) = p^2(C + (2 - r)D)
$$

### Total Expected Cost – Reactive Strategy

$$
\begin{aligned}
\mathbb{E}[\text{Cost}_{\text{react}}] &= (1 - p)^2 \cdot 0 + 2p(1 - p)\left(C + (1 - r)D\right) + p^2\left(C + (2 - r)D\right) \\
&= 2p(1 - p)\left(C + (1 - r)D\right) + p^2\left(C + (2 - r)D\right)
\end{aligned}
$$


## Symbolic Comparison

Compare:

$$
\mathbb{E}[\text{Cost}_{\text{pre}}] = C + pD
$$

vs.

$$
\mathbb{E}[\text{Cost}_{\text{react}}] = (2p - p^2)C + \left[2p(1 - p)(1 - r) + p^2(2 - r)\right]D
$$

Preemptive strategy is preferred when:

$$
C + pD < (2p - p^2)C + \left[2p(1 - p)(1 - r) + p^2(2 - r)\right]D
$$

```{python}
from sympy import symbols, Eq, solve, simplify

# Define symbols
p, C, D, r = symbols('p C D r', positive=True)

# Total cost of each strategy
cost_pre = C + p * D
cost_react = (2*p - p**2)*C + (2*p*(1 - p)*(1 - r) + p**2*(2 - r)) * D

# Set inequality: preemptive better than reactive
ineq = Eq(cost_pre, cost_react)

# Solve for threshold p
p_thresh = solve(ineq, p)

# Simplify solution
simplified_thresh = [simplify(expr) for expr in p_thresh]

# Output
for i, expr in enumerate(simplified_thresh):
    print(f"Threshold p solution {i+1}:")
    print(expr)
```

```{r}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)

# Create evaluation grid
grid <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D,
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Threshold lines: where cost_pre == cost_react
threshold_lines <- grid %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

# Labels for each facet
labels_df <- data.frame(
  C = rep(C_vals, each = 2),
  label = rep(c("preemptive", "reactive"), 4),
  r = rep(c(0.05, 0.75), 4),
  p = rep(c(0.95, 0.05), 4)
)

ggplot(grid, aes(x = r, y = p)) +
  geom_line(data = threshold_lines, aes(x = r, y = p_thresh, group = C), 
            color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  scale_x_continuous(limits=c(0,1))+
  scale_fill_manual(values = c("preemptive" = "steelblue", "reactive" = "tomato")) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "vaccination strategy decision with threshold (n=2)",
    x = expression("relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("outbreak probability" ~ italic(p)),
    fill = "Strategy"
  ) +
  theme_light() +
  theme(legend.position = 'top')
```

## Compare single and two population scenarios

```{r}
grid1 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C,
    cost_react =  p * (C + (1 - r)*D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

grid2 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    # Total cost for each strategy
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D,
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )


# Combine grids
grid1 <- grid1 %>% mutate(scenario = "Single population")
grid2 <- grid2 %>% mutate(scenario = "Two populations")
grid_all <- bind_rows(grid1, grid2)

# Threshold lines: compute for each scenario
threshold_lines <- grid_all %>%
  group_by(scenario, C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))

# Labels for plot interpretation
labels_df <- expand.grid(
  C = C_vals,
  scenario =  c("Single population", "Two populations"),
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.1, 0.6),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

ggplot() +
  geom_line(data = threshold_lines, 
            aes(x = r, y = p_thresh, color = scenario, linetype = scenario, 
                group = interaction(scenario, C)),  linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  scale_color_manual(values = c("Single population" = "steelblue", "Two populations" = "tomato")) +
  scale_linetype_manual(values = c("Single population" = "solid", "Two populations" = "dashed")) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~C, labeller = label_bquote(D/C == .(D/C))) +
  labs(
    title = "Vaccination Strategy Thresholds by Scenario and Campaign Cost",
    x = expression("Relative effectiveness of reactive strategy" ~ italic(r)),
    y = expression("Outbreak probability" ~ italic(p)),
    color = "Scenario",
    linetype = "Scenario"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

## Multiple populations ($n>2$) scenario

We consider $n$ populations where each has probability $p$ of an outbreak. We can afford to vaccinate a fraction $f$ of them, pre-emptively or reactively.

Let:

-   $n$: number of populations
-   $p$: outbreak probability
-   $f$: fraction of populations we can vaccinate
-   $C$: vaccination cost per population
-   $D$: outbreak cost (if unmitigated)
-   $r$: relative effectiveness of reactive vaccination ($0 < r < 1$)

### Strategy 1: Preemptive Vaccination

We preemptively vaccinate $fn$ populations. If an outbreak occurs in the $(1 - f)n$ unvaccinated group, we pay $D$.

Total expected cost:

$$
\text{Cost}_{\text{pre}} = fn \cdot C + (1 - f)np \cdot D
$$

### Reactive vaccination strategy

## Step 2: Conditioning on $X = x$ outbreaks

For a fixed number of outbreaks $X = x$:

- If $x \le k$, we can respond to **all** outbreaks.

  Total cost:
  $$
  \text{Cost}_{x,\text{react}} = x(C + (1 - r)D)
  $$

- If $x > k$, we can respond to **$k$ outbreaks**, and $x - k$ go **unmitigated**.

  Mitigated cost:
  $$
  k(C + (1 - r)D)
  $$

  Unmitigated cost:
  $$
  (x - k)D
  $$

So the **expected cost when $X = x$** is:

$$
\text{Cost}_{x,\text{react}} =
\begin{cases}
x(C + (1 - r)D), & \text{if } x \le k \\
k(C + (1 - r)D) + (x - k)D, & \text{if } x > k
\end{cases}
$$

## Step 3: Total Expected Cost

We now compute the total expected cost across all possible $x$:

$$
\mathbb{E}[\text{Cost}_{\text{react}}] =
\sum_{x = 0}^{n} \Pr(X = x) \cdot \text{Cost}_{x,\text{react}}
$$

Split into two parts:

### Case 1: $x \le k$
$$
\sum_{x = 0}^{k}
\binom{n}{x} p^x (1 - p)^{n - x} \cdot x(C + (1 - r)D)
$$

### Case 2: $x > k$
$$
\sum_{x = k+1}^{n}
\binom{n}{x} p^x (1 - p)^{n - x}
\cdot \left[k(C + (1 - r)D) + (x - k)D \right]
$$

This expression does **not simplify into closed form**, but it can be computed numerically or approximated.

### Strategy 2-1: Reactive Vaccination (Approximate)

We reactively vaccinate up to $fn$ populations **after** outbreaks occur.

Expected number of outbreaks: $np$\
We can respond to $\min(np, fn)$ and leave $\max(np - fn, 0)$ unmitigated.

Residual cost per mitigated outbreak = $(1 - r)D$

Expected cost:

$$
\text{Cost}_{\text{react}} =
\min(np, fn)(C + (1 - r)D) + \max(np - fn, 0) \cdot D
$$


## Symbolic Comparison using SymPy

```{python}
import sympy as sp

# Define symbols
n, p, f, C, D, r = sp.symbols('n p f C D r', positive=True)

# Expressions
np_ = n * p
fn_ = f * n

# Preemptive strategy cost
cost_pre = fn_ * C + (1 - f) * np_ * D

# Reactive strategy cost (piecewise via min/max)
covered   = sp.Min(np_, fn_)
uncovered = sp.Max(np_ - fn_, 0)

cost_react = covered * (C + (1 - r) * D) + uncovered * D

# Symbolic difference: reactive - preemptive
cost_diff = sp.simplify(cost_react - cost_pre)
cost_diff
```

## Decision Thresholds for Preemptive vs Reactive (Multiple Populations)

We compute and plot threshold outbreak probability $p^*$ at which preemptive vaccination becomes cheaper than reactive, under various vaccination cost levels $C$ and vaccination capacity levels $f$.

```{r}
library(ggplot2)
library(dplyr)

# Parameters
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.05, 0.1, 0.5)

n <- 1009  # or any fixed value

grid_all <- expand.grid(p = p_vals, r = r_vals, C = C_vals, f = f_vals) %>%
  mutate(
    cost_pre = n * (f * C + (1 - f) * p * D),
    covered = pmin(p, f),
    uncovered = pmax(p - f, 0),
    cost_react = n * (covered * (C + (1 - r) * D) + uncovered * D),
    decision = ifelse(cost_pre < cost_react, "preemptive", "reactive")
  )

# Compute threshold lines: p where cost_pre == cost_react
threshold_lines <- grid_all %>%
  group_by(C, f, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh))


# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )


# Plot with threshold lines colored by f
ggplot(threshold_lines, aes(x = r, y = p_thresh, color = factor(f), linetype = factor(f))) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("0.05" = "#1b9e77", "0.1" = "#d95f02", "0.5" = "#7570b3")) +
  scale_linetype_manual(values = c("0.05" = "solid", "0.1" = "dashed", "0.5" = "dotdash")) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title = "Threshold Outbreak Probability for Preemptive vs Reactive Strategy",
    subtitle = "Multiple population scenario with varying vaccination capacity f",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Vaccination capacity (f)",
    linetype = "Vaccination capacity (f)"
  ) +
  theme_light() +
  theme(legend.position = "top")
```

```{r}
# Exact cost computation for given parameters
compute_exact_costs <- function(n, p, C, D, r, f) {
  k <- floor(f * n)  # number of populations we can vaccinate

  probs <- dbinom(0:n, size = n, prob = p)

  # Preemptive: vaccinate k populations before outbreaks
  cost_pre <- sum(
    probs * (k * C + (1 - f) * (0:n) * D)  # Expected outbreak cost: only in unvaccinated
  )

  # Reactive: respond to up to k outbreaks after detection
  cost_react <- sum(
    probs * ifelse(
      0:n <= k,
      (0:n) * (C + (1 - r) * D),                    # All outbreaks mitigated
      k * (C + (1 - r) * D) + (0:n - k) * D         # Only k mitigated, rest unmitigated
    )
  )

  list(pre = cost_pre, react = cost_react)
}

find_threshold_p <- function(n, r, C, D, f, p_grid = seq(0.01, 0.99, by = 0.01)) {
  for (p in p_grid) {
    costs <- compute_exact_costs(n, p, C, D, r, f)
    if ((costs$pre - costs$react) < 0) {
      return(p)  # First p where preemptive is cheaper
    }
  }
  return(NA)
}
```

```{r}
# Parameters
n <- 20  # try 2 to compare with two-pop figure
r_vals <- seq(0.01, 0.99, by = 0.01)
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.05, 0.1, 0.5)
D <- 1

# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

# Grid of all thresholds
thresholds <- expand.grid(C = C_vals, f = f_vals, r = r_vals) %>%
  rowwise() %>%
  mutate(
    p_thresh = find_threshold_p(n = n, r = r, C = C, D = D, f = f)
  ) %>%
  ungroup()


ggplot(thresholds, aes(x = r, y = p_thresh, color = factor(f), linetype = factor(f))) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("0.05" = "#1b9e77", "0.1" = "#d95f02", "0.5" = "#7570b3")) +
  scale_linetype_manual(values = c("0.05" = "solid", "0.1" = "dashed", "0.5" = "dotdash")) +
  labs(
    title = "Exact Threshold Outbreak Probability for Preemptive vs Reactive",
    subtitle = paste("Based on Binomial outbreak distribution with n =", n),
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Vaccination capacity (f)",
    linetype = "Vaccination capacity (f)"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

```{r}
library(ggplot2)
library(dplyr)

# --- Parameters ---
r_vals <- seq(0.01, 0.99, by = 0.01)
p_vals <- seq(0.01, 0.99, by = 0.01)
D <- 1
C_vals <- c(0.3, 1, 2, 5)
f_vals <- c(0.4)  # exact overlay for f = 0.5
n <- 20            # exact enumeration for n = 2

# --- Function: exact expected cost given outbreak count ---
compute_exact_costs <- function(n, p, C, D, r, f) {
  k <- floor(f * n)
  x_vals <- 0:n
  probs <- dbinom(x_vals, size = n, prob = p)

  # Preemptive: vaccinate k units, pay outbreak cost only for (1 - f)
  cost_pre <- sum(probs * (k * C + (1 - f) * x_vals * D))

  # Reactive: cover up to k outbreaks after detection
  cost_react <- sum(probs * ifelse(
    x_vals <= k,
    x_vals * (C + (1 - r) * D),
    k * (C + (1 - r) * D) + (x_vals - k) * D
  ))

  list(pre = cost_pre, react = cost_react)
}

# --- Function: Find p* threshold where preemptive becomes better ---
find_threshold_p <- function(n, r, C, D, f, p_grid = seq(0.01, 0.99, by = 0.01)) {
  for (p in p_grid) {
    costs <- compute_exact_costs(n, p, C, D, r, f)
    if ((costs$pre - costs$react) < 0) {
      return(p)
    }
  }
  return(NA)
}

# --- Compute threshold p* for each (r, C) using exact method ---
threshold_exact <- expand.grid(C = C_vals, f = f_vals, r = r_vals) %>%
  rowwise() %>%
  mutate(
    p_thresh = find_threshold_p(n = n, r = r, C = C, D = D, f = f)
  ) %>%
  ungroup() %>%
  mutate(method = "Exact (n = 2)", f = factor(f))

# --- Load two-population approximation for comparison ---
grid2 <- expand.grid(p = p_vals, r = r_vals, C = C_vals) %>%
  mutate(
    cost_pre = C + p * D,
    cost_react = (2 * p - p^2) * C + (2 * p * (1 - p) * (1 - r) + p^2 * (2 - r)) * D
  )

threshold_two_pop <- grid2 %>%
  group_by(C, r) %>%
  summarise(
    p_thresh = approx(cost_pre - cost_react, p, xout = 0)$y,
    .groups = "drop"
  ) %>%
  filter(!is.na(p_thresh)) %>%
  mutate(method = "Two population approx", f = factor(0.5))

# --- Combine and plot ---
threshold_combined <- bind_rows(threshold_exact, threshold_two_pop)

# Dynamic label positioning based on actual threshold ranges
labels_df <- expand.grid(
  C = C_vals,
  f = f_vals,
  label = c("preemptive", "reactive")
) %>%
  mutate(
    r = ifelse(label == "preemptive", 0.05, 0.75),
    p = ifelse(label == "preemptive", 0.95, 0.05)  # Different p positions
  )

ggplot(threshold_combined, aes(x = r, y = p_thresh, color = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("Exact (n = 2)" = "#1b9e77", "Two population approx" = "#d95f02")) +
  scale_linetype_manual(values = c("Exact (n = 2)" = "solid", "Two population approx" = "dashed")) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title = "Threshold Outbreak Probability for Preemptive vs Reactive (f = 0.5)",
    subtitle = "Comparison between exact enumeration for n = 2 and two-population approximation",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Method",
    linetype = "Method"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

## Asymptotic Limit: Preemptive vs Reactive

As the number of populations $n \to \infty$ and vaccination capacity fraction $f \to 0$ (i.e., resources are extremely limited), the expected costs for preemptive and reactive strategies simplify.

Let:

-   $p$: probability of an outbreak per population,
-   $r$: relative effectiveness of reactive vaccination,
-   $C$: cost of one vaccination campaign,
-   $D$: cost of one unmitigated outbreak,
-   $fn$: number of populations we can afford to vaccinate (preemptively or reactively)



### Preemptive Expected Cost

Total cost:

$$
\text{Cost}_{\text{pre}} = fn \cdot C + (1 - f)np \cdot D
$$

As $f \to 0$, this becomes:

$$
\text{Cost}_{\text{pre}} \approx npD
$$


### Reactive Expected Cost

Using the expression:

$$
\text{Cost}_{\text{react}} = \min(fn, np)(C + (1 - r)D) + \max(np - fn, 0) \cdot D
$$

If $fn \ll np$:

$$
\text{Cost}_{\text{react}} \approx fn(C + (1 - r)D) + (np - fn)D
$$



### Cost Difference

Compute:

$$
\begin{aligned}
\Delta &= \text{Cost}_{\text{pre}} - \text{Cost}_{\text{react}} \\
&= fnC + (1 - f)npD - \left[fn(C + (1 - r)D) + (np - fn)D\right] \\
&= fnD(r - p)
\end{aligned}
$$



### Result

-   If $r < p$: preemptive is cheaper ($\Delta < 0$)
-   If $r > p$: reactive is cheaper ($\Delta > 0$)

Thus, in the limit:

$$
p > r \quad \Rightarrow \quad \text{preemptive is preferred}
$$

This corresponds to the decision threshold line $p^* = r$ as $f \to 0$ and $n \to \infty$.

```{r}
# Overlay limit line p = r
ggplot(threshold_combined, aes(x = r, y = p_thresh, color = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_text(data = labels_df, aes(x = r, y = p, label = label),
            hjust = 0, color = "black", size = 4, inherit.aes = FALSE) +
  scale_y_continuous(limits=c(0,1))+
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.7, label = "p = r (limit)", angle = 45, hjust = -0.1, vjust = 1.2, size = 3.5) +
  facet_wrap(~C, labeller = label_bquote(C == .(C))) +
  scale_color_manual(values = c("Exact (n = 2)" = "#1b9e77", "Two population approx" = "#d95f02")) +
  scale_linetype_manual(values = c("Exact (n = 2)" = "solid", "Two population approx" = "dashed")) +
  labs(
    title = "Threshold Outbreak Probability with Asymptotic Line p = r",
    subtitle = "Black dotted line shows p = r decision boundary as n → ∞ and f → 0",
    x = expression("Relative effectiveness of reactive vaccination" ~ italic(r)),
    y = expression("Threshold outbreak probability" ~ italic(p^"*")),
    color = "Method",
    linetype = "Method"
  ) +
  theme_light() +
  theme(legend.position = "top")

```

# Optimizing Hybrid Vaccination Strategy

## Problem

> Given a limited vaccine supply, how should we split our efforts between **preemptive** and **reactive** vaccination strategies to **minimize total expected cost**?

We define:

-   $\alpha \in [0, 1]$: fraction of vaccination resources used **preemptively**
-   $1 - \alpha$: fraction used **reactively**
-   $f$: total fraction of the population that can be vaccinated
-   $n$: total number of populations
-   $p$: probability of outbreak in each population
-   $C$: cost of one vaccination campaign
-   $D$: cost of one unmitigated outbreak
-   $r$: relative effectiveness of reactive vaccination to the preemptive strategy


##  Expected Cost Formula

The total expected cost of a hybrid strategy that uses $\alpha$ of the vaccination capacity preemptively is:

$$
\text{Cost}(\alpha) =
\underbrace{\alpha f n C}_{\text{preemptive vaccination}} +
\underbrace{\min(x, k_{\text{react}})(C + (1 - r)D)}_{\text{reactive cost}} +
\underbrace{\max(x - k_{\text{react}}, 0) D}_{\text{unmitigated outbreaks}}
$$

Where: - $x = (1 - \alpha f)np$: expected number of outbreaks in unvaccinated group - $k_{\text{react}} = (1 - \alpha)fn$: number of reactive campaigns available


## Visualizing Cost of Hybrid Strategy

We explore how the total expected cost varies with the fraction $\alpha$ of vaccine budget used preemptively.

```{r}
library(ggplot2)
library(dplyr)

# Parameters (adjustable)
f <- 0.5     # total vaccination capacity
n <- 100     # number of populations
p <- 0.3     # outbreak probability per population
C <- 1       # cost of a vaccination campaign
D <- 5       # cost of an unmitigated outbreak
r <- 0.4     # relative effectiveness of reactive vaccination

# Alpha grid: preemptive share
alpha_vals <- seq(0, 1, by = 0.01)

# Compute total cost for each alpha
cost_df <- data.frame(alpha = alpha_vals) %>%
  mutate(
    k_pre = alpha * f * n,
    k_react = (1 - alpha) * f * n,
    x = (1 - alpha * f) * n * p,
    covered = pmin(x, k_react),
    uncovered = pmax(x - k_react, 0),
    cost = k_pre * C + covered * (C + (1 - r) * D) + uncovered * D
  )
```

```{r}
ggplot(cost_df, aes(x = alpha, y = cost)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(data = cost_df %>% slice_min(cost), aes(x = alpha, y = cost), 
             color = "firebrick", size = 3) +
  annotate("text", 
           x = cost_df$alpha[which.min(cost_df$cost)], 
           y = min(cost_df$cost), 
           label = "Minimum", 
           hjust = -0.2, vjust = -1, color = "firebrick") +
  labs(
    title = expression("Expected Total Cost of Hybrid Strategy vs " ~ alpha),
    subtitle = paste0("Parameters: p = ", p, ", f = ", f, ", n = ", n, 
                      ", C = ", C, ", D = ", D, ", r = ", r),
    x = expression("Fraction of vaccine budget used preemptively" ~ (alpha)),
    y = "Expected total cost"
  ) +
  theme_light()
```

## Cost Function Derivation for Parametric Risk Distributions

Define:
- $p_i$: true outbreak probability in region $i$
- $q(k_{\text{pre}}) = \sum_{i \in \text{top-}k_{\text{pre}}} p_i$: expected number of outbreaks covered by preemptive vaccination
- $x = \sum_{i \notin \text{top-}k_{\text{pre}}} p_i = np - q(k_{\text{pre}})$: expected number of outbreaks among the non-preemptively vaccinated regions


### Case 1: Uniform$(a, b)$
- $p_i \sim \text{Uniform}(a, b)$
- $\mathbb{E}[p_i] = \frac{a + b}{2}$
- Expected total outbreaks: $np = n \cdot \frac{a + b}{2}$
- Expected $q(k)$: $q(k_{\text{pre}}) = k_{\text{pre}} \cdot \frac{a + b}{2}$
- $x = np - q(k_{\text{pre}}) = (n - k_{\text{pre}}) \cdot \frac{a + b}{2}$

---

### Case 2: Beta$(a, b)$
- $p_i \sim \text{Beta}(a, b)$
- $\mathbb{E}[p_i] = \frac{a}{a + b}$
- $np = n \cdot \frac{a}{a + b}$
- $q(k_{\text{pre}}) = k_{\text{pre}} \cdot \frac{a}{a + b}$
- $x = (n - k_{\text{pre}}) \cdot \frac{a}{a + b}$

---

### Case 3: Truncated Normal$(\mu, \sigma^2)$ on $[0,1]$
- $p_i \sim \text{TruncNorm}(\mu, \sigma^2, [0,1])$
- $\mathbb{E}[p_i] = \mu_{\text{trunc}} = \text{mean of truncated normal}$
- $np = n \cdot \mu_{\text{trunc}}$
- $q(k_{\text{pre}}) = k_{\text{pre}} \cdot \mu_{\text{trunc}}$
- $x = (n - k_{\text{pre}}) \cdot \mu_{\text{trunc}}$

---

### R Code for All Cases
```{r}
# Common parameters
n <- 100
f <- 0.5
C <- 1
D <- 5
r <- 0.4
alpha_vals <- seq(0, 1, by = 0.01)

# Risk distribution parameters
uniform_ab <- c(0.1, 0.5)
beta_ab <- c(2, 5)
mu <- 0.3
sigma <- 0.15

# Truncated normal mean (approximation via simulation)
get_truncnorm_mean <- function(mu, sigma, n = 1e6) {
  a <- (0 - mu) / sigma
  b <- (1 - mu) / sigma
  mean(rtruncnorm(n, a = 0, b = 1, mean = mu, sd = sigma))
}

mu_trunc <- get_truncnorm_mean(mu, sigma)

# Cost computation function
eval_cost <- function(alpha, n, f, p_mean, C, D, r) {
  k_pre <- floor(alpha * f * n)
  k_react <- floor((1 - alpha) * f * n)
  x <- (n - k_pre) * p_mean
  cost <- k_pre * C + min(x, k_react) * (C + (1 - r) * D) + max(0, x - k_react) * D
  return(cost)
}

costs_uniform <- sapply(alpha_vals, eval_cost, n = n, f = f, p_mean = mean(uniform_ab), C = C, D = D, r = r)
costs_beta    <- sapply(alpha_vals, eval_cost, n = n, f = f, p_mean = beta_ab[1] / sum(beta_ab), C = C, D = D, r = r)
costs_trunc   <- sapply(alpha_vals, eval_cost, n = n, f = f, p_mean = mu_trunc, C = C, D = D, r = r)

# Plot
library(ggplot2)
df <- data.frame(
  alpha = rep(alpha_vals, 3),
  cost = c(costs_uniform, costs_beta, costs_trunc),
  dist = rep(c("Uniform(0.1, 0.5)", "Beta(2,5)", "TruncNorm(0.3,0.15)"), each = length(alpha_vals))
)

ggplot(df, aes(x = alpha, y = cost, color = dist)) +
  geom_line(size = 1) +
  labs(x = expression(alpha), y = "Expected Total Cost",
       title = "Cost vs. Preemptive Allocation (alpha)",
       color = "Risk Distribution") +
  theme_minimal()
```

