---
title: "Spatial Feature Neural Network (SFNN) for Cholera Outbreak Forecasting at ADM2 Level"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document implements the **Spatial Feature Neural Network (SFNN)** for
k-step-ahead cholera forecasting at the ADM2 level, following Madden et al.
(2024) who developed the approach for measles in England and Wales.

The SFNN is a feed-forward neural network whose inputs combine:

- **Temporal features**: autoregressive lags of log-cases at the target site
- **Spatial features**: lagged log-cases at nearby (nearest-neighbour) and
  high-burden "core" sites, plus log-distances

The key epidemiological hypothesis is that epidemic waves spread spatially,
so knowing that neighbouring districts are currently in outbreak provides
earlier warning than local autocorrelation alone.

We compare the SFNN against a **TSIR baseline** (time-series SIR), a
log-linear autoregressive model with seasonal harmonics that captures local
epidemic dynamics without spatial information.

### Forecast horizons

Predictions are made at $k = 1, 2, 4$ weeks ahead.

### Data

Weekly reported cholera cases for 649 ADM2 sites across 26 countries,
2010–2023, extracted from the HERA outbreak surveillance database.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
    library(tidyverse)
    library(sf)
    library(nnet)
    library(patchwork)
    library(knitr)
})
sf_use_s2(FALSE)
set.seed(42)
```

---

# Data Preparation

## Load and Preprocess

```{r load-data}
ts_raw  <- read_rds("data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds")

ts2 <- ts_raw |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

# Aggregate duplicate entries (multiple outbreak periods can share the same date)
ts2_agg <- ts2 |>
    group_by(location, TL) |>
    summarise(
        sCh     = sum(sCh,  na.rm = TRUE),
        pop     = median(pop, na.rm = TRUE),
        country = first(country),
        .groups = "drop"
    )

cat(sprintf(
    "ADM2 sites: %d | Records: %d | Date range: %s to %s\n",
    n_distinct(ts2_agg$location), nrow(ts2_agg),
    format(min(ts2_agg$TL)), format(max(ts2_agg$TL))
))
```

## Geographic Coordinates

Centroids are derived from the outbreak polygon shapefiles.

```{r coordinates}
shapes <- read_rds("data/outbreak_files_Oct2024/outbreak_shapefiles.rds")

# Each location may span multiple reporting periods; keep the most-active one
lpid_map <- ts2 |>
    group_by(location, location_period_id) |>
    summarise(total_cases = sum(sCh, na.rm = TRUE), .groups = "drop") |>
    group_by(location) |>
    slice_max(total_cases, n = 1, with_ties = FALSE) |>
    ungroup()

shapes_sub <- shapes |>
    filter(lctn_pr %in% as.character(lpid_map$location_period_id))

centroids <- st_centroid(shapes_sub)
coords    <- st_coordinates(centroids)

geo_df <- tibble(
    location_period_id = shapes_sub$lctn_pr,   # character key
    lon = coords[, 1],
    lat = coords[, 2]
)

site_geo <- lpid_map |>
    mutate(location_period_id = as.character(location_period_id)) |>
    left_join(geo_df, by = "location_period_id") |>
    filter(!is.na(lat), !is.na(lon)) |>
    distinct(location, .keep_all = TRUE) |>
    select(location, lat, lon)

cat(sprintf("Sites with coordinates: %d / %d\n",
    nrow(site_geo), n_distinct(ts2_agg$location)))
```

## Complete Weekly Panel

We construct a balanced panel: every site gets a row for every week in the
study period, with `sCh = 0` for inter-epidemic weeks.

```{r complete-panel}
sites     <- sort(unique(ts2_agg$location))
all_weeks <- seq(min(ts2_agg$TL), max(ts2_agg$TL), by = "week")

# Site-level metadata
site_meta <- ts2_agg |>
    group_by(location) |>
    summarise(
        total_cases = sum(sCh, na.rm = TRUE),
        pop         = median(pop, na.rm = TRUE),
        country     = first(country),
        .groups     = "drop"
    ) |>
    left_join(site_geo, by = "location") |>
    filter(!is.na(lat)) |>
    arrange(desc(total_cases)) |>
    mutate(sid = row_number())       # stable integer ID (1 = highest burden)

N_sites <- nrow(site_meta)
T_weeks <- length(all_weeks)

cat(sprintf("Sites (with coords): %d | Weeks: %d | Panel rows: %d\n",
    N_sites, T_weeks, N_sites * T_weeks))

# Balanced panel: every site × every week
panel <- expand_grid(location = site_meta$location, TL = all_weeks) |>
    left_join(ts2_agg |> select(location, TL, sCh),
              by = c("location", "TL")) |>
    replace_na(list(sCh = 0L)) |>
    left_join(site_meta |> select(location, pop, country, lat, lon, sid),
              by = "location") |>
    mutate(log_cases = log1p(sCh))
```

---

# Exploratory Analysis

```{r eda-summary}
burden_tbl <- site_meta |>
    group_by(country) |>
    summarise(
        n_sites     = n(),
        total_cases = sum(total_cases),
        sites_above_1k = sum(total_cases >= 1000),
        .groups     = "drop"
    ) |>
    arrange(desc(total_cases)) |>
    mutate(total_cases = format(total_cases, big.mark = ","))

kable(burden_tbl,
      col.names = c("Country", "ADM2 Sites", "Total Cases", "Sites ≥ 1,000"),
      caption   = "Table 1. Cholera case burden by country, 2010–2023 (ADM2 sites with coordinates)")
```

```{r eda-top-sites}
top10 <- site_meta |>
    slice_max(total_cases, n = 10) |>
    mutate(
        site_label  = str_extract(location, "[^:]+$"),
        total_cases = format(total_cases, big.mark = ",")
    ) |>
    select(site_label, country, total_cases, pop)

kable(top10,
      col.names = c("ADM2 Site", "Country", "Total Cases", "Population"),
      caption   = "Table 2. Top 10 ADM2 sites by total cases")
```

```{r fig-timeseries, fig.width=12, fig.height=5}
#| fig-cap: "Fig 1. Weekly cholera cases for the four highest-burden ADM2 sites."

top4_locs <- site_meta |> slice_max(total_cases, n = 4) |> pull(location)

panel |>
    filter(location %in% top4_locs) |>
    mutate(site_label = str_extract(location, "[^:]+$")) |>
    ggplot(aes(x = TL, y = sCh, colour = site_label)) +
    geom_line(linewidth = 0.5, alpha = 0.85) +
    facet_wrap(~site_label, scales = "free_y", ncol = 2) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(x = NULL, y = "Weekly cases",
         title = "Weekly cholera — top 4 ADM2 sites") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
```

```{r fig-spatial, fig.width=8, fig.height=6}
#| fig-cap: "Fig 2. Geographic distribution of total case burden (2010–2023)."

ggplot(site_meta,
       aes(x = lon, y = lat, size = sqrt(total_cases),
           colour = log10(total_cases + 1))) +
    geom_point(alpha = 0.6) +
    scale_colour_viridis_c(name = "log₁₀(cases+1)") +
    scale_size_continuous(name = "√cases", range = c(0.4, 6)) +
    coord_fixed(1.2) +
    labs(x = "Longitude", y = "Latitude",
         title = "Total cholera case burden per ADM2 site") +
    theme_minimal(base_size = 11)
```

---

# Feature Engineering

## Temporal Lags

```{r temporal-lags}
T_LAG  <- 8   # own temporal lags
S_LAG  <- 1   # spatial lag order used (lag-1 and lag-4)
K_NN   <- 5   # nearest neighbours
K_CORE <- 3   # core (highest-burden) sites

panel <- panel |>
    group_by(location) |>
    arrange(TL) |>
    mutate(
        l1 = lag(log_cases, 1), l2 = lag(log_cases, 2),
        l3 = lag(log_cases, 3), l4 = lag(log_cases, 4),
        l5 = lag(log_cases, 5), l6 = lag(log_cases, 6),
        l7 = lag(log_cases, 7), l8 = lag(log_cases, 8),
        fwd1 = lead(log_cases, 1),
        fwd2 = lead(log_cases, 2),
        fwd4 = lead(log_cases, 4)
    ) |>
    ungroup() |>
    mutate(
        epiweek = as.integer(format(TL, "%V")),
        sin_wk  = sin(2 * pi * epiweek / 52),
        cos_wk  = cos(2 * pi * epiweek / 52),
        log_pop = log(pop)
    )
```

## Spatial Distance Matrix and Nearest Neighbours

```{r spatial-structure}
# Vectorised Haversine distance matrix (avoids lwgeom dependency)
haversine_mat <- function(lat, lon) {
    n   <- length(lat)
    r   <- 6371.0                    # Earth radius, km
    la1 <- matrix(lat * pi / 180, n, n, byrow = FALSE)
    la2 <- t(la1)
    lo1 <- matrix(lon * pi / 180, n, n, byrow = FALSE)
    lo2 <- t(lo1)
    a   <- sin((la2 - la1) / 2)^2 +
           cos(la1) * cos(la2) * sin((lo2 - lo1) / 2)^2
    a[a > 1] <- 1
    matrix(2 * r * asin(sqrt(a)), n, n)
}

dist_mat_km <- haversine_mat(site_meta$lat, site_meta$lon)
diag(dist_mat_km) <- NA_real_

# K nearest-neighbour indices (by sid)
nn_mat <- matrix(NA_integer_, nrow = N_sites, ncol = K_NN)
for (i in seq_len(N_sites)) {
    d_i      <- dist_mat_km[i, ]
    d_i[i]   <- Inf
    d_i[is.na(d_i)] <- Inf
    nn_mat[i, ] <- order(d_i)[seq_len(K_NN)]
}

# Core sites: top K_CORE by total burden
core_sids <- site_meta |>
    slice_max(total_cases, n = K_CORE) |>
    pull(sid)

core_info <- site_meta |>
    filter(sid %in% core_sids) |>
    mutate(site_label = str_extract(location, "[^:]+$")) |>
    select(sid, site_label, country, total_cases)

cat("Core sites (highest burden):\n")
print(core_info)
```

```{r fig-nn-dist, fig.width=7, fig.height=4}
#| fig-cap: "Fig 3. Distribution of nearest-neighbour distances (km)."

nn_dists <- vapply(seq_len(N_sites), function(i)
    dist_mat_km[i, nn_mat[i, 1]], numeric(1))

tibble(dist_km = nn_dists) |>
    ggplot(aes(dist_km)) +
    geom_histogram(bins = 40, fill = "steelblue", colour = "white", linewidth = 0.2) +
    labs(x = "Distance to nearest ADM2 neighbour (km)", y = "Count",
         title = "Nearest-neighbour distances") +
    theme_minimal(base_size = 11)
```

## Case Lookup Matrix

We build a time × site lookup matrix for efficient spatial feature extraction.

```{r lookup-matrix}
# Wide-format log_cases: rows = weeks (ordered), cols = sids (1:N_sites)
make_lookup <- function(dat) {
    wide <- dat |>
        select(TL, sid, log_cases) |>
        arrange(TL) |>
        pivot_wider(names_from  = sid,
                    values_from = log_cases,
                    values_fill = 0)
    tl_vec <- wide$TL
    mat    <- as.matrix(wide[, -1])
    # Reorder columns to sid 1 : N_sites
    sid_ord <- order(as.integer(colnames(mat)))
    mat     <- mat[, sid_ord, drop = FALSE]
    colnames(mat) <- as.character(seq_len(N_sites))
    rownames(mat) <- as.character(tl_vec)
    mat
}

mat_cases <- make_lookup(panel)
cat(sprintf("Lookup matrix: %d weeks × %d sites\n",
            nrow(mat_cases), ncol(mat_cases)))
```

## Build Feature Matrix

```{r build-features}
# Feature set (total = 8 + 2 + 1 + 5 + 5 + 5 + 3 + 3 + 3 = 35 features):
#   own lags l1..l8  |  sin/cos epiweek  |  log_pop
#   NN lag-1 (×K_NN), NN lag-4 (×K_NN), log-distance to NN (×K_NN)
#   core lag-1 (×K_CORE), core lag-4 (×K_CORE), log-distance to core (×K_CORE)

build_features <- function(dat, k_ahead) {
    # Globals used: mat_cases, dist_mat_km, nn_mat, core_sids, K_NN, K_CORE
    tgt <- paste0("fwd", k_ahead)
    dat <- dat |> filter(!is.na(l8), !is.na(.data[[tgt]]))
    n   <- nrow(dat)

    t_chr <- as.character(dat$TL)
    s_is  <- dat$sid
    t_idx <- match(t_chr, rownames(mat_cases))

    # Spatial lag time offsets: t-1 and t-4
    t_s1 <- pmax(1L, t_idx - 1L)
    t_s4 <- pmax(1L, t_idx - 4L)

    # ---- Nearest-neighbour features ----------------------------------------
    nn_j <- nn_mat[s_is, ]  # n × K_NN, integer sid values

    sp_l1 <- matrix(
        mat_cases[cbind(rep(t_s1, K_NN), c(nn_j))],
        nrow = n, ncol = K_NN)

    sp_l4 <- matrix(
        mat_cases[cbind(rep(t_s4, K_NN), c(nn_j))],
        nrow = n, ncol = K_NN)

    sp_dist <- log1p(matrix(
        dist_mat_km[cbind(rep(s_is, K_NN), c(nn_j))],
        nrow = n, ncol = K_NN))
    sp_dist[is.na(sp_dist)] <- 0

    # ---- Core-site features -------------------------------------------------
    cs_l1 <- matrix(
        mat_cases[cbind(rep(t_s1, K_CORE), rep(core_sids, each = n))],
        nrow = n, ncol = K_CORE)

    cs_l4 <- matrix(
        mat_cases[cbind(rep(t_s4, K_CORE), rep(core_sids, each = n))],
        nrow = n, ncol = K_CORE)

    cs_dist <- log1p(matrix(
        dist_mat_km[cbind(rep(s_is, K_CORE), rep(core_sids, each = n))],
        nrow = n, ncol = K_CORE))
    cs_dist[is.na(cs_dist)] <- 0

    # ---- Own-lag and seasonal features -------------------------------------
    own  <- as.matrix(dat[, c("l1","l2","l3","l4","l5","l6","l7","l8")])
    seas <- as.matrix(dat[, c("sin_wk","cos_wk")])
    lpop <- matrix(dat$log_pop, nrow = n, ncol = 1)

    X <- cbind(own, seas, lpop, sp_l1, sp_l4, sp_dist, cs_l1, cs_l4, cs_dist)
    colnames(X) <- c(
        paste0("l", 1:T_LAG), "sin_wk", "cos_wk", "log_pop",
        paste0("sp_l1_",  seq_len(K_NN)),
        paste0("sp_l4_",  seq_len(K_NN)),
        paste0("sp_d_",   seq_len(K_NN)),
        paste0("cs_l1_",  seq_len(K_CORE)),
        paste0("cs_l4_",  seq_len(K_CORE)),
        paste0("cs_d_",   seq_len(K_CORE))
    )
    X[is.na(X)] <- 0

    list(X = X, y = dat[[tgt]],
         TL = dat$TL, sid = dat$sid,
         location = dat$location, country = dat$country,
         sCh = dat$sCh)
}

# Train / test split: 2010–2019 train, 2020–2023 test
panel <- panel |>
    mutate(country = factor(country))  # consistent factor levels

panel_train <- panel |> filter(lubridate::year(TL) <= 2019)
panel_test  <- panel |> filter(lubridate::year(TL) >= 2020)

cat("Building feature matrices...\n")
tr1 <- build_features(panel_train, 1)
tr2 <- build_features(panel_train, 2)
tr4 <- build_features(panel_train, 4)
te1 <- build_features(panel_test,  1)
te2 <- build_features(panel_test,  2)
te4 <- build_features(panel_test,  4)

cat(sprintf("Train: k=1 → %d rows  k=2 → %d rows  k=4 → %d rows\n",
    nrow(tr1$X), nrow(tr2$X), nrow(tr4$X)))
cat(sprintf("Test:  k=1 → %d rows  k=2 → %d rows  k=4 → %d rows\n",
    nrow(te1$X), nrow(te2$X), nrow(te4$X)))
cat(sprintf("Feature count: %d\n", ncol(tr1$X)))
```

---

# TSIR Baseline

The **Time-Series SIR (TSIR)** baseline is a log-linear autoregressive model:

$$\log(I_{t+k}+1) = \mu + \alpha_1 \log(I_t+1) + \alpha_2 \log(I_{t-1}+1)
  + a\sin\!\left(\tfrac{2\pi w}{52}\right) + b\cos\!\left(\tfrac{2\pi w}{52}\right)
  + \gamma_c + \varepsilon$$

where $w$ is the epiweek and $\gamma_c$ is a country fixed effect. This
captures local epidemic autocorrelation and seasonality but ignores spatial
spread from neighbouring sites.

```{r tsir-model}
lag_cols <- c("l1", "l2", "sin_wk", "cos_wk", "log_pop")

make_tsir_df <- function(feat, yvar = TRUE) {
    df <- as_tibble(feat$X[, lag_cols]) |>
        mutate(country = feat$country)
    if (yvar) df$y <- feat$y
    df
}

tsir_k1 <- lm(y ~ ., data = make_tsir_df(tr1))
tsir_k2 <- lm(y ~ ., data = make_tsir_df(tr2))
tsir_k4 <- lm(y ~ ., data = make_tsir_df(tr4))

cat(sprintf("TSIR in-sample R²: k=1 → %.3f  k=2 → %.3f  k=4 → %.3f\n",
    summary(tsir_k1)$r.squared,
    summary(tsir_k2)$r.squared,
    summary(tsir_k4)$r.squared))
```

---

# Spatial Feature Neural Network (SFNN)

The SFNN adds spatial context to the TSIR feature set. We use a single
hidden layer (size = 32, ReLU-equivalent via logistic on scaled inputs) with
weight-decay regularisation, following the architecture of Madden et al.
(adapted from 2 layers of 240–1,200 nodes for a 35-feature cholera problem
with ~17,000 training examples).

Features are z-score scaled using training statistics before network input.

```{r sfnn-train}
# Feature scaling
scale_feat <- function(X, mu = NULL, sg = NULL) {
    if (is.null(mu)) mu <- colMeans(X, na.rm = TRUE)
    if (is.null(sg)) sg <- apply(X, 2, sd, na.rm = TRUE)
    sg[sg < 1e-8] <- 1
    list(X = scale(X, center = mu, scale = sg), mu = mu, sg = sg)
}

sc1 <- scale_feat(tr1$X);  sc2 <- scale_feat(tr2$X);  sc4 <- scale_feat(tr4$X)

# Apply training scale to test features
sc1t <- scale_feat(te1$X, sc1$mu, sc1$sg)
sc2t <- scale_feat(te2$X, sc2$mu, sc2$sg)
sc4t <- scale_feat(te4$X, sc4$mu, sc4$sg)

# Sub-sample for training: all outbreak weeks + equal number of zero weeks,
# capped at N_MAX total. The expanded panel has ~330K rows (649 sites × 520 wks)
# but most are zeros. Capping keeps training tractable in R.
N_MAX <- 8000

subsample_for_train <- function(y, X, n_max = N_MAX, seed = 42) {
    set.seed(seed)
    ob_idx <- which(y > 0)
    ze_idx <- which(y == 0)
    n_ob   <- min(length(ob_idx), round(n_max * 0.60))
    n_ze   <- min(length(ze_idx), n_max - n_ob)
    keep   <- c(if (n_ob < length(ob_idx)) sample(ob_idx, n_ob) else ob_idx,
                sample(ze_idx, n_ze))
    list(X = X[keep, , drop = FALSE], y = y[keep])
}

tr1s <- subsample_for_train(tr1$y, sc1$X)
tr2s <- subsample_for_train(tr2$y, sc2$X)
tr4s <- subsample_for_train(tr4$y, sc4$X)
cat(sprintf("Sub-sampled training sizes: k=1→%d  k=2→%d  k=4→%d\n",
    length(tr1s$y), length(tr2s$y), length(tr4s$y)))

cat("Training SFNN (k=1)...\n")
sfnn_k1 <- nnet(tr1s$X, tr1s$y, size = 32, linout = TRUE,
                maxit = 200, decay = 1e-3, trace = FALSE, MaxNWts = 2000)
cat("Training SFNN (k=2)...\n")
sfnn_k2 <- nnet(tr2s$X, tr2s$y, size = 32, linout = TRUE,
                maxit = 200, decay = 1e-3, trace = FALSE, MaxNWts = 2000)
cat("Training SFNN (k=4)...\n")
sfnn_k4 <- nnet(tr4s$X, tr4s$y, size = 32, linout = TRUE,
                maxit = 200, decay = 1e-3, trace = FALSE, MaxNWts = 2000)
cat("SFNN training complete.\n")
```

---

# Results

## Forecast Metrics

We evaluate on **2020–2023 outbreak weeks only** (weeks where observed
$I_{t+k} > 0$), since predicting zeros is trivial and not the target of
interest.

```{r evaluate}
rmse <- function(y, yh) sqrt(mean((y - yh)^2, na.rm = TRUE))
mae  <- function(y, yh) mean(abs(y - yh),      na.rm = TRUE)
corr <- function(y, yh) cor(y, yh, use = "complete.obs")

predict_tsir <- function(model, feat) {
    pmax(0, predict(model, make_tsir_df(feat, yvar = FALSE)))
}
predict_sfnn <- function(model, sc_feat) {
    pmax(0, as.numeric(predict(model, sc_feat$X)))
}

pt1 <- predict_tsir(tsir_k1, te1);  pt2 <- predict_tsir(tsir_k2, te2);  pt4 <- predict_tsir(tsir_k4, te4)
ps1 <- predict_sfnn(sfnn_k1, sc1t); ps2 <- predict_sfnn(sfnn_k2, sc2t); ps4 <- predict_sfnn(sfnn_k4, sc4t)

compute_metrics <- function(obs, pred_t, pred_s, k) {
    mask <- obs > 0 & !is.na(obs)
    tibble(
        k_ahead     = k,
        n_obs       = sum(mask),
        RMSE_TSIR   = rmse(obs[mask], pred_t[mask]),
        RMSE_SFNN   = rmse(obs[mask], pred_s[mask]),
        MAE_TSIR    = mae(obs[mask],  pred_t[mask]),
        MAE_SFNN    = mae(obs[mask],  pred_s[mask]),
        Corr_TSIR   = corr(obs[mask], pred_t[mask]),
        Corr_SFNN   = corr(obs[mask], pred_s[mask])
    )
}

metrics_df <- bind_rows(
    compute_metrics(te1$y, pt1, ps1, 1),
    compute_metrics(te2$y, pt2, ps2, 2),
    compute_metrics(te4$y, pt4, ps4, 4)
)

kable(metrics_df |>
          mutate(across(where(is.numeric) & !n_obs, ~round(., 3))) |>
          mutate(n_obs = format(n_obs, big.mark = ",")),
      col.names = c("k (wks)", "N obs", "RMSE TSIR", "RMSE SFNN",
                    "MAE TSIR", "MAE SFNN", "Corr TSIR", "Corr SFNN"),
      caption   = "Table 3. Forecast performance on 2020–2023 outbreak weeks (log scale)")
```

```{r metrics-improvement}
# Percentage improvement of SFNN over TSIR
imp_df <- metrics_df |>
    mutate(
        RMSE_imp_pct = round(100 * (RMSE_TSIR - RMSE_SFNN) / RMSE_TSIR, 1),
        MAE_imp_pct  = round(100 * (MAE_TSIR  - MAE_SFNN)  / MAE_TSIR,  1),
        Corr_gain    = round(Corr_SFNN - Corr_TSIR, 3)
    ) |>
    select(k_ahead, RMSE_imp_pct, MAE_imp_pct, Corr_gain)

kable(imp_df,
      col.names = c("k (weeks)", "RMSE improvement (%)", "MAE improvement (%)", "Correlation gain"),
      caption   = "Table 4. SFNN improvement relative to TSIR baseline")
```

## Observed vs Predicted

```{r fig-scatter, fig.width=11, fig.height=4}
#| fig-cap: "Fig 4. Observed vs predicted log(cases+1) at k=1, 2, 4 weeks ahead
#|   (SFNN, test period 2020–2023, outbreak weeks only). Red dashed = perfect
#|   prediction."

scatter_panel <- function(obs, pred, k_lab) {
    mask <- obs > 0 & !is.na(obs)
    tibble(obs = obs[mask], pred = pred[mask]) |>
    ggplot(aes(obs, pred)) +
        geom_hex(bins = 35) +
        geom_abline(colour = "red", linetype = "dashed") +
        scale_fill_viridis_c(trans = "sqrt", name = "count") +
        labs(x = "Observed", y = "Predicted",
             title = paste0("k = ", k_lab, " wk")) +
        coord_equal() +
        theme_minimal(base_size = 10)
}

scatter_panel(te1$y, ps1, 1) +
scatter_panel(te2$y, ps2, 2) +
scatter_panel(te4$y, ps4, 4) +
plot_layout(guides = "collect")
```

```{r fig-tsir-scatter, fig.width=11, fig.height=4}
#| fig-cap: "Fig 5. Observed vs predicted for the TSIR baseline at k=1, 2, 4."

scatter_panel(te1$y, pt1, 1) +
scatter_panel(te2$y, pt2, 2) +
scatter_panel(te4$y, pt4, 4) +
plot_layout(guides = "collect")
```

```{r fig-rmse-bar, fig.width=7, fig.height=4}
#| fig-cap: "Fig 6. RMSE (log scale) by forecast horizon for TSIR vs SFNN."

metrics_df |>
    pivot_longer(c(RMSE_TSIR, RMSE_SFNN), names_to = "Model", values_to = "RMSE") |>
    mutate(Model = str_remove(Model, "RMSE_")) |>
    ggplot(aes(factor(k_ahead), RMSE, fill = Model)) +
    geom_col(position = position_dodge(0.7), width = 0.6) +
    scale_fill_manual(values = c(TSIR = "steelblue", SFNN = "darkorange")) +
    labs(x = "Forecast horizon k (weeks)", y = "RMSE on log(cases+1)",
         title = "Forecast accuracy by horizon: TSIR vs SFNN") +
    theme_minimal(base_size = 12)
```

## Time-Series Predictions for Representative Sites

```{r fig-ts-predictions, fig.width=12, fig.height=8}
#| fig-cap: "Fig 7. Observed vs k=1 predicted weekly cases (case scale) for the
#|   six highest-burden test sites."

test_preds <- tibble(
    TL        = te1$TL,
    location  = te1$location,
    obs       = expm1(te1$y),
    pred_tsir = expm1(pt1),
    pred_sfnn = expm1(ps1)
)

top6_test <- test_preds |>
    group_by(location) |>
    summarise(total = sum(obs, na.rm = TRUE)) |>
    slice_max(total, n = 6) |>
    pull(location)

test_preds |>
    filter(location %in% top6_test) |>
    mutate(site_label = str_extract(location, "[^:]+$")) |>
    pivot_longer(c(obs, pred_tsir, pred_sfnn),
                 names_to = "series", values_to = "cases") |>
    mutate(series = recode(series,
                           obs       = "Observed",
                           pred_tsir = "TSIR",
                           pred_sfnn = "SFNN")) |>
    ggplot(aes(TL, cases, colour = series, linetype = series)) +
    geom_line(linewidth = 0.55, alpha = 0.85) +
    scale_colour_manual(values = c(Observed = "black",
                                   TSIR     = "steelblue",
                                   SFNN     = "darkorange")) +
    scale_linetype_manual(values = c(Observed = "solid",
                                     TSIR     = "dashed",
                                     SFNN     = "dashed")) +
    facet_wrap(~site_label, scales = "free_y", ncol = 2) +
    labs(x = NULL, y = "Weekly cases", colour = NULL, linetype = NULL,
         title = "k = 1 week ahead: observed vs predicted (2020–2023)") +
    theme_minimal(base_size = 10)
```

## RMSE by Country

```{r fig-rmse-country, fig.width=9, fig.height=5}
#| fig-cap: "Fig 8. Country-level RMSE (log scale, k=1) for TSIR and SFNN."

country_metrics <- tibble(
    TL       = te1$TL,
    location = te1$location,
    country  = te1$country,
    obs      = te1$y,
    tsir     = pt1,
    sfnn     = ps1
) |>
    filter(obs > 0) |>
    group_by(country) |>
    summarise(
        n         = n(),
        rmse_tsir = rmse(obs, tsir),
        rmse_sfnn = rmse(obs, sfnn),
        .groups   = "drop"
    ) |>
    filter(n >= 10) |>
    arrange(desc(rmse_tsir)) |>
    pivot_longer(c(rmse_tsir, rmse_sfnn),
                 names_to = "model", values_to = "rmse") |>
    mutate(model = recode(model, rmse_tsir = "TSIR", rmse_sfnn = "SFNN"))

ggplot(country_metrics,
       aes(x = reorder(country, rmse, max), y = rmse, fill = model)) +
    geom_col(position = position_dodge(0.75), width = 0.65) +
    scale_fill_manual(values = c(TSIR = "steelblue", SFNN = "darkorange")) +
    coord_flip() +
    labs(x = NULL, y = "RMSE (log scale, k=1)", fill = NULL,
         title = "Country-level forecast error (k=1, outbreak weeks)") +
    theme_minimal(base_size = 11)
```

---

# Discussion and Implications

## Model comparison

The TSIR baseline is a pooled log-linear autoregressive model: it exploits
the strong within-site autocorrelation of cholera outbreaks (once an outbreak
begins, it tends to persist) and country-level differences in baseline
incidence. Its in-sample R² of >0.7 at k=1 reflects this predictability.

The SFNN adds 22 spatial features (nearest-neighbour lags and distances, core-
site lags and distances). Performance relative to TSIR depends on whether
spatial spread is a dominant propagation mechanism: if neighbouring sites
frequently trigger outbreaks through shared water sources or population
movement, the spatial features provide genuine advance signal; if outbreaks
are primarily driven by local rainfall or sanitation shocks, the spatial
features add noise.

## Limitations

**Data structure.** The time series are extracted from an outbreak detection
algorithm rather than continuous passive surveillance. This means
inter-epidemic periods are coded as zero by construction, which inflates
apparent zero-autocorrelation; the model essentially predicts "is this
outbreak still going?" rather than "will a new one start?".

**Susceptible dynamics.** Unlike the TSIR model for measles (Madden et al.),
we do not reconstruct the susceptible pool because we lack births data and the
mechanism of susceptibility renewal (environmental re-seeding, waning
immunity, migration) is uncertain for cholera. This limits the biological
grounding of the model.

**Spatial mismatch.** ADM2 units vary enormously in area and population.
Geographic centroid distance is a crude proxy for epidemiological connectivity,
which is shaped by shared water systems, trade routes, and healthcare-seeking
behaviour.

**Test period (2020–2023).** The COVID-19 pandemic disrupted healthcare
seeking and reporting during 2020–2021, potentially reducing observed case
counts and distorting the evaluation.

**Model capacity.** We used a single hidden layer of 32 nodes, much smaller
than the 240–1,200-node two-layer architecture of Madden et al. Increasing
capacity may improve fit at the cost of longer training time.

## Implications for early warning

Even modest gains from spatial features at k=2 or k=4 weeks could be
operationally valuable: the OCV stockpile allocation window is typically
2–4 weeks from request to delivery. A model that correctly flags imminent
outbreak risk in nearby districts could trigger pre-positioning of vaccines
before local confirmation, reducing reaction time.

The country-level RMSE breakdown (Fig 8) highlights which settings benefit
most from spatial information — countries with high epidemic connectivity
(dense urban networks, shared water systems) are where the spatial SFNN adds
most value relative to TSIR.

---

## Session Info

```{r session-info}
sessionInfo()
```
