---
title: "ADM2 Cholera Incidence Rate (2010–2014)"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document selects ADM2 sites from the outbreak surveillance data and
computes the mean annual cholera incidence rate (cases per 100,000 person-years)
for each site over the 2010–2014 period. Population is derived by summing the
WorldPop density raster over each ADM2 polygon.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(scales)
library(knitr)
```

# Data

## Outbreak surveillance

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)

# Restrict to ADM2 level (location string has exactly 3 '::' separators,
# i.e. Continent::Country::ADM1::ADM2).
ts2 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

cat("ADM2 time-series rows:", nrow(ts2), "\n")
cat("Unique ADM2 locations:", n_distinct(ts2$location), "\n")
cat("Columns:", paste(names(ts2), collapse = ", "), "\n")
```

## Shapefiles

```{r load-shapefiles}
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Keep the most recent location_period_id per location (see risk_prediction_simple.qmd
# for the full multi-period analysis justifying this choice: 7 locations have >1
# period ID; 4 have identical polygons, 3 have minor-to-moderate revisions; in
# all cases the most recent boundary is appropriate for analysis).
loc_id_map <- ts2 |>
    distinct(location_period_id, location) |>
    group_by(location) |>
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) |>
    ungroup()

shp_admin2 <- shp_all |>
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id")) |>
    filter(!is.na(location), str_count(location, "::") == 3L) |>
    st_make_valid()

# Flag and remove degenerate sliver polygons before population extraction.
# A valid ADM2 health district cannot be < 1 km²; polygons that small are
# geometry artefacts that yield nonsensical population estimates (e.g. Moba
# Health District has a 0.098 km² sliver giving pop = 318 despite ~1400 cases).
shp_admin2$area_km2 <- as.numeric(st_area(shp_admin2)) / 1e6
slivers <- shp_admin2 |> st_drop_geometry() |>
    filter(area_km2 < 1) |>
    select(location, area_km2)
if (nrow(slivers) > 0) {
    cat("Removing", nrow(slivers), "sliver polygon(s) (area < 1 km²):\n")
    print(slivers)
}
shp_admin2 <- shp_admin2 |> filter(area_km2 >= 1)

stopifnot(!anyDuplicated(shp_admin2$location))
cat("ADM2 polygons after sliver removal:", nrow(shp_admin2), "\n")
```

## Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)

# The raster stores population density (persons per km²) in WGS84 at 1/6° × 1/6°
# resolution. The geodetic pixel area at the African mid-latitude (~1.3°N) is:
# (1/6 × 111.32)² × cos(1.3°) ≈ 344 km². Summing density values over a polygon
# and multiplying by POP_SCALE gives total head-count.
POP_SCALE <- 344  # pixel area: (1/6 * 111.32)^2 * cos(1.3 * pi/180) ≈ 344 km²

shp_pop  <- shp_admin2 |> st_transform(crs(pop_raster))
pop_vals <- exactextractr::exact_extract(pop_raster, shp_pop,
                                         fun = "sum", progress = FALSE)
shp_admin2$pop_sum <- pop_vals * POP_SCALE

cat("Total population:", format(round(sum(shp_admin2$pop_sum, na.rm = TRUE)),
                                big.mark = ","), "\n")
```

# ADM2 Site Selection

Select ADM2 sites that have at least one recorded outbreak **starting** in the
2010–2014 period. Sites with no outbreak record are retained as well (zero
cases), to keep the denominator consistent across sites and allow comparison
of high- and low-incidence areas.

```{r select-sites}
STUDY_YEARS <- 2010:2014

# Collapse time-series rows to one row per outbreak event.
# sCh: suspected cholera cases reported during each time-step.
outbreak_events <- ts2 |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(
        year_start  = year(min(TL)),
        year_end    = year(max(TR)),
        total_cases = sum(sCh, na.rm = TRUE),
        .groups     = "drop"
    )

# Sum cases by (location, year) using year_start as the index year.
# Using year_start ensures each outbreak is counted only once, in the year
# it first emerged, avoiding double-counting for multi-year outbreaks.
cases_annual <- outbreak_events |>
    filter(year_start %in% STUDY_YEARS) |>
    group_by(location, year = year_start) |>
    summarise(cases = sum(total_cases, na.rm = TRUE), .groups = "drop")

# Complete the panel: fill zeros for location-years with no outbreak.
all_locs <- unique(ts2$location)
cases_panel <- cases_annual |>
    right_join(expand_grid(location = all_locs, year = STUDY_YEARS),
               by = c("location", "year")) |>
    replace_na(list(cases = 0))

cat("Location-year observations:", nrow(cases_panel), "\n")
cat("Locations with ≥1 case in 2010-2014:",
    n_distinct(cases_panel$location[cases_panel$cases > 0]), "\n")
```

# Incidence Rate Computation

Incidence rate = total cases over 2010–2014 divided by (population × 5 years),
expressed per 100,000 person-years.

```{r compute-incidence}
# Sum cases over 2010-2014 for each location.
cases_total <- cases_panel |>
    group_by(location) |>
    summarise(
        total_cases   = sum(cases),
        n_years       = n_distinct(year),   # should be 5 for all locations
        .groups       = "drop"
    )

# Join population and compute incidence rate.
pop_lookup <- shp_admin2 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

incidence_df <- cases_total |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(
        # person-years = population × number of years observed
        person_years    = pop_sum * n_years,
        # incidence rate per 100,000 person-years
        incidence_rate  = total_cases / person_years * 1e5
    ) |>
    arrange(desc(incidence_rate))

cat("Sites with computable incidence rate:", nrow(incidence_df), "\n")
cat(sprintf("Incidence rate range: %.2f – %.2f per 100,000 person-years\n",
            min(incidence_df$incidence_rate),
            max(incidence_df$incidence_rate)))
```

# Results

## Summary table (top 20 sites by incidence rate)

```{r table-top20}
incidence_df |>
    slice_head(n = 20) |>
    dplyr::select(location, total_cases, pop_sum, person_years, incidence_rate) |>
    mutate(pop_sum      = round(pop_sum),
           person_years = round(person_years)) |>
    knitr::kable(
        digits  = c(0, 0, 0, 0, 2),
        col.names = c("Location", "Total cases", "Population",
                      "Person-years", "Rate per 100k py"),
        caption = "Top 20 ADM2 sites by mean annual cholera incidence rate (2010–2014)"
    )
```

## Distribution of incidence rates

```{r fig-incidence-dist, fig.width=10, fig.height=4}
#| fig-cap: "Left: histogram of incidence rates across all ADM2 sites (log10 scale,
#|   excluding zero-case sites). Right: empirical CDF. Both exclude sites with
#|   zero total cases."

df_nonzero <- incidence_df |> filter(total_cases > 0)

p_hist <- ggplot(df_nonzero, aes(x = incidence_rate)) +
    geom_histogram(bins = 40, fill = "steelblue", colour = "white", linewidth = 0.2) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Number of ADM2 sites",
         title = sprintf("Distribution across %d sites with ≥1 case",
                         nrow(df_nonzero))) +
    theme_minimal(base_size = 11)

p_cdf <- ggplot(df_nonzero |> arrange(incidence_rate) |>
                    mutate(ecdf = seq_len(n()) / n()),
                aes(x = incidence_rate, y = ecdf)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Cumulative proportion of sites",
         title = "ECDF of incidence rates") +
    theme_minimal(base_size = 11)

library(patchwork)
p_hist + p_cdf
```

## Map of incidence rates

```{r fig-incidence-map, fig.width=10, fig.height=8}
#| fig-cap: "Mean annual cholera incidence rate (2010–2014) per ADM2. Grey =
#|   sub-Saharan Africa country boundaries. Sites with zero cases are shown in
#|   light grey."

library(rnaturalearthdata)

ssa_sf <- sf::st_as_sf(rnaturalearthdata::countries50) |>
    dplyr::filter(subregion %in% c("Eastern Africa", "Middle Africa",
                                    "Southern Africa", "Western Africa")) |>
    sf::st_make_valid() |>
    sf::st_transform(sf::st_crs(shp_admin2))

shp_rate <- shp_admin2 |>
    left_join(incidence_df |> dplyr::select(location, incidence_rate),
              by = "location") |>
    replace_na(list(incidence_rate = 0))

ssa_bbox <- sf::st_bbox(ssa_sf)

ggplot() +
    geom_sf(data = ssa_sf,  fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = shp_rate, aes(fill = incidence_rate), colour = NA) +
    scale_fill_viridis_c(
        option    = "magma", direction = -1,
        trans     = "log1p",
        breaks    = c(0, 1, 10, 100, 1000),
        labels    = c("0", "1", "10", "100", "1000"),
        name      = "Rate\n(per 100k py)"
    ) +
    coord_sf(xlim = ssa_bbox[c("xmin", "xmax")],
             ylim = ssa_bbox[c("ymin", "ymax")]) +
    labs(title    = "Cholera incidence rate 2010–2014 (ADM2)",
         subtitle = "Cases attributed to year_start; population from WorldPop 2017 raster") +
    theme_minimal(base_size = 11) +
    theme(panel.grid  = element_blank(),
          axis.text   = element_blank(),
          axis.ticks  = element_blank(),
          legend.position = "right")
```

## Full incidence table

```{r table-full}
incidence_df |>
    dplyr::select(location, total_cases, pop_sum, incidence_rate) |>
    mutate(pop_sum        = round(pop_sum),
           incidence_rate = round(incidence_rate, 2)) |>
    DT::datatable(
        colnames  = c("Location", "Total cases (2010–2014)",
                      "Population", "Rate per 100k py"),
        caption   = "Cholera incidence rates for all ADM2 sites (2010–2014)",
        rownames  = FALSE,
        filter    = "top",
        options   = list(pageLength = 25, scrollX = TRUE)
    )
```

# Vaccine Impact Scenarios

## Setup

Scenarios vary two parameters:

- **ρ** (targeting accuracy): Spearman rank correlation between the vaccination
  priority score and true case burden. ρ = 1 = oracle (rank by true cases),
  ρ = 0 = random allocation.
- **f** (coverage): fraction of the total population vaccinated, from 0.1 to 0.5.

A **two-dose OCV regimen** is assumed (doses = 2 × vaccinated population).
Pre-emptive effectiveness **ν = 0.90**.

Cases averted are computed directly from observed case counts: for each
vaccinated location, cases averted = observed cases × ν. A **3-year case
total** (2010–2012) is used, reflecting the assumed duration of population-level
protection from a two-dose regimen.

```{r vaccine-setup}
NU        <- 0.90   # pre-emptive vaccine effectiveness (nu)
DOSES_PER <- 2L     # doses per vaccinated person
VAX_YEARS <- 2010:2012  # 3-year protection window

rho_values <- seq(0, 1, by = 0.1)
f_values   <- seq(0.1, 0.5, by = 0.1)

# 3-year observed case totals per location (used directly — no attack rate needed).
cases_3yr <- cases_panel |>
    filter(year %in% VAX_YEARS) |>
    group_by(location) |>
    summarise(cases_3yr = sum(cases), .groups = "drop")

sim_vax <- incidence_df |>
    left_join(cases_3yr, by = "location") |>
    replace_na(list(cases_3yr = 0))

total_pop     <- sum(sim_vax$pop_sum)
baseline_cases <- sum(sim_vax$cases_3yr)

cat(sprintf("Total population:     %s\n", format(round(total_pop), big.mark = ",")))
cat(sprintf("Baseline cases (%d-%d): %s\n",
            min(VAX_YEARS), max(VAX_YEARS),
            format(baseline_cases, big.mark = ",")))

# Gaussian copula: generate a priority score with a target Spearman rho to
# the true 3-year case burden. rho_P = 2 sin(pi * rho_S / 6) converts the
# target Spearman rho to the equivalent Pearson correlation.
make_rank_score <- function(truth, target_rho, seed = 42) {
    set.seed(seed)
    n       <- length(truth)
    z_truth <- qnorm(rank(truth, ties.method = "random") / (n + 1))
    z_noise <- rnorm(n)
    rho_p   <- 2 * sin(pi * target_rho / 6)
    pnorm(rho_p * z_truth + sqrt(1 - rho_p^2) * z_noise)
}
```

## Simulation

```{r vaccine-sim}
results <- expand_grid(rho = rho_values, f = f_values) |>
    pmap_dfr(function(rho, f) {
        # Priority score at target rho relative to true 3-year case burden
        score <- make_rank_score(sim_vax$cases_3yr, target_rho = rho)

        # Rank locations by score (highest risk first), fill budget f * P.
        # No partial-location vaccination: include only fully covered locations.
        ord     <- order(score, decreasing = TRUE)
        cum_pop <- cumsum(sim_vax$pop_sum[ord])
        n_vax   <- sum(cum_pop <= f * total_pop)
        vax_idx <- ord[seq_len(n_vax)]

        pop_vaccinated <- sum(sim_vax$pop_sum[vax_idx])
        doses          <- pop_vaccinated * DOSES_PER
        # Cases averted = observed cases in vaccinated locations × ν
        cases_averted  <- sum(sim_vax$cases_3yr[vax_idx]) * NU

        tibble(
            rho                        = rho,
            f                          = f,
            pop_vaccinated             = pop_vaccinated,
            doses                      = doses,
            cases_averted              = cases_averted,
            cases_per_1k_doses         = cases_averted / doses * 1000,
            doses_per_case_averted     = if (cases_averted > 0) doses / cases_averted else NA_real_
        )
    })
```

## Results table

```{r table-vaccine}
results |>
    mutate(
        pop_vaccinated = format(round(pop_vaccinated), big.mark = ","),
        doses          = format(round(doses),          big.mark = ","),
        cases_averted  = round(cases_averted, 1)
    ) |>
    knitr::kable(
        align   = c("r","r","r","r","r","r","r"),
        digits  = c(1, 1, 0, 0, 1, 2, 1),
        col.names = c("rho", "f", "Pop vaccinated", "Doses",
                      "Cases averted", "Cases averted / 1k doses",
                      "Doses / case averted"),
        caption = sprintf(
            "Vaccine impact scenarios | nu = %.2f, %d-dose regimen | 3-year cases (%d-%d)",
            NU, DOSES_PER, min(VAX_YEARS), max(VAX_YEARS))
    )
```

## Figures

```{r fig-vaccine-cases, fig.width=10, fig.height=5}
#| fig-cap: "Cases averted (left) and doses per case averted (right) by targeting
#|   accuracy (rho) and coverage fraction (f). rho = 1 = oracle; rho = 0 = random."

p_cases <- ggplot(results, aes(x = rho, y = cases_averted,
                                colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_y_continuous(labels = comma) +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = sprintf("Cases averted (%d-%d)", min(VAX_YEARS), max(VAX_YEARS)),
         title = "Cases averted") +
    theme_minimal(base_size = 11)

p_doses <- ggplot(results, aes(x = rho, y = doses_per_case_averted,
                                colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_y_continuous(labels = comma) +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = "Doses per case averted",
         title = "Doses required per case averted") +
    theme_minimal(base_size = 11)

p_cases + p_doses +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

```{r fig-vaccine-efficiency, fig.width=8, fig.height=5}
#| fig-cap: "Cases averted per 1,000 doses. Higher rho (better targeting) and
#|   lower f (more selective) yield greater efficiency."

ggplot(results, aes(x = rho, y = cases_per_1k_doses,
                    colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = "Cases averted per 1,000 doses",
         title = "Vaccine programme efficiency") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Optimal Pre-emptive Fraction via Incidence Rate

## Revised cost model

We define the cost ratio at the **individual level**:

$$
R = \frac{c_i}{c_v},
$$

where $c_i$ is the cost of illness per infected person and $c_v$ is the cost of
vaccination per person. The population-level costs are then:

$$
C_I = a \cdot c_i \cdot N = a \cdot R \cdot c_v \cdot N,
\qquad
C_V = C \cdot c_v \cdot N,
$$

with $a$ = attack rate (fraction of the population infected during an outbreak),
$C$ = vaccine coverage proportion (fraction of the population vaccinated), and
$N$ = population size. Normalising by $c_v \cdot N$ (i.e. the cost of vaccinating
every person once), the single-population expected costs are:

$$
c_\text{pre}^{(1)} = C + p\,a\,(1-\nu)\,R, \qquad
c_\text{react}^{(1)} = p\bigl[C + a\,(1-r)\,R\bigr],
$$

where $p$ is the outbreak probability. Setting
$c_\text{pre}^{(1)} = c_\text{react}^{(1)}$ gives the critical probability:

$$
p_\text{crit}^{(1)} = \frac{C}{C + a\,R\,(\nu - r)}.
$$

Compared with the original formulation (where $R = a c_i / c_v C$ absorbed both
$a$ and $C$), the individual-level definition keeps $C$ and $a$ visible,
making it straightforward to assess how campaign coverage and attack rate jointly
determine strategy boundaries.

## Translation to incidence rates

With incidence data, we characterise each location $i$ by its **mean annual
incidence rate** $\lambda_i$ (cases per person per year). Over a planning horizon
of $T$ years, the expected cumulative proportion of the population infected is:

$$
q_i = \lambda_i \cdot T,
$$

which plays exactly the role of $p_i \cdot a$ in the outbreak-probability
framework (expected illness per person per period). The individual-level cost
ratio $R = c_i / c_v$ is unchanged. Replacing $p \cdot a$ with $q_i$ and
normalising by $c_v \cdot N$, the multi-location cost equations become:

$$
c_\text{pre}(\alpha) = f_\text{pre}\bigl[C + \bar{q}_\text{pre}(1-\nu)\,R\bigr],
$$

$$
c_\text{react}(\alpha) = \begin{cases}
  q_\text{rem}\bigl[C + (1-r)\,R\bigr] & \text{if } f_\text{react} \ge q_\text{rem}, \\
  f_\text{react}\bigl[C + (1-r)\,R\bigr] + (q_\text{rem} - f_\text{react})\,R
  & \text{otherwise,}
\end{cases}
$$

where $f_\text{pre} = \alpha f$, $f_\text{react} = (1-\alpha)f$,
$\bar{q}_\text{pre}$ is the population-weighted mean $q_i$ in the pre-emptively
vaccinated group, and $q_\text{rem} = \bar{q} - f_\text{pre}\bar{q}_\text{pre}$.
The term $C$ appears in every vaccination-cost component; locations that receive
neither pre-emptive nor reactive vaccination incur only the illness cost $q_i \cdot R$.

The between-location distribution of $\lambda_i$ is modelled as
**Gamma**$(k, k/\mu)$, the natural continuous analogue of the Beta$(1,\theta)$
used in the outbreak-probability framework.

## Gamma-Poisson fit to 2010–2012 data

We model the 3-year case count at each location as:

$$
Y_i \sim \text{Poisson}(\lambda_i \cdot E_i), \qquad \lambda_i \sim \text{Gamma}(k,\, k/\mu),
$$

where $E_i = N_i \cdot 3$ is the person-years exposure. Marginalising over
$\lambda_i$ gives the **Negative Binomial** distribution with mean $\mu \cdot E_i$
and size (overdispersion) parameter $k$. We fit $\mu$ and $k$ by maximum likelihood.

```{r gamma-poisson-fit}
library(MASS)

# Restrict to 2010-2012 (3-year horizon matching the vaccine impact section)
FIT_YEARS <- 2010:2012
T_HORIZON <- length(FIT_YEARS)

cases_fit <- cases_panel |>
    filter(year %in% FIT_YEARS) |>
    group_by(location) |>
    summarise(cases_3yr = sum(cases), .groups = "drop") |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0)

# Exposure in person-years for each location
cases_fit <- cases_fit |>
    mutate(
        exposure  = pop_sum * T_HORIZON,   # person-years
        # Naive incidence rate (cases / person-years)
        rate_obs  = cases_3yr / exposure
    )

# Fit Negative Binomial via MASS::glm.nb (log link, log(exposure) as offset).
# glm.nb estimates mu (mean rate) and theta (= size = shape k of the
# Gamma mixing distribution). Larger theta → less overdispersion.
nb_fit <- MASS::glm.nb(
    cases_3yr ~ 1 + offset(log(exposure)),
    data = cases_fit
)

k_hat  <- nb_fit$theta                          # Gamma shape parameter
mu_hat <- exp(coef(nb_fit)[["(Intercept)"]])    # mean incidence rate (per py)

cat(sprintf("Gamma-Poisson fit (2010-2012):\n"))
cat(sprintf("  Mean rate mu  = %.4e cases/person-year\n", mu_hat))
cat(sprintf("  Shape k       = %.3f\n", k_hat))
cat(sprintf("  Mean rate * 1e5 = %.2f per 100,000 py\n", mu_hat * 1e5))
```

```{r fig-gamma-poisson, fig.width=10, fig.height=4}
#| fig-cap: "Left: observed vs Gamma-Poisson fitted distribution of incidence
#|   rates (2010–2012, log10 scale). Right: quantile-quantile plot of observed
#|   vs NB-predicted quantiles."

# Observed distribution
df_obs <- cases_fit |> filter(cases_3yr > 0)

# Fitted Gamma density for lambda (rate per person-year) on log10 scale
lambda_grid <- 10^seq(-8, -1, length.out = 400)
gamma_dens  <- dgamma(lambda_grid,
                      shape = k_hat,
                      rate  = k_hat / mu_hat)
# Convert density to log10-rate axis: d(log10 lambda) = lambda * log(10) * d(lambda)
df_gamma <- tibble(lambda = lambda_grid,
                   dens   = gamma_dens * lambda_grid * log(10))

p_hist2 <- ggplot(df_obs, aes(x = rate_obs)) +
    geom_histogram(aes(y = after_stat(density * x * log(10))),
                   bins = 35, fill = "steelblue", colour = "white", linewidth = 0.2) +
    geom_line(data = df_gamma, aes(x = lambda, y = dens),
              colour = "firebrick", linewidth = 1) +
    scale_x_log10(labels = function(x) format(x * 1e5, big.mark = ",", digits = 3)) +
    labs(x = "Incidence rate (per 100,000 py, log scale)",
         y = "Density (log10 scale)",
         title = sprintf("Gamma(%0.2f, %0.4e) fit", k_hat, k_hat / mu_hat),
         subtitle = sprintf("Red: fitted Gamma; n = %d sites with ≥1 case",
                            nrow(df_obs))) +
    theme_minimal(base_size = 11)

# Q-Q against fitted Gamma
n_sites  <- nrow(cases_fit)
probs    <- (seq_len(n_sites) - 0.5) / n_sites
q_obs    <- sort(cases_fit$rate_obs)
q_gamma  <- qgamma(probs, shape = k_hat, rate = k_hat / mu_hat)

p_qq <- ggplot(tibble(obs = q_obs, fitted = q_gamma),
               aes(x = fitted, y = obs)) +
    geom_point(size = 1, alpha = 0.5, colour = "steelblue") +
    geom_abline(slope = 1, intercept = 0, colour = "firebrick") +
    scale_x_log10(labels = function(x) format(x * 1e5, digits = 2)) +
    scale_y_log10(labels = function(x) format(x * 1e5, digits = 2)) +
    labs(x = "Gamma quantiles (per 100,000 py, log scale)",
         y = "Observed quantiles (per 100,000 py, log scale)",
         title = "Q-Q: observed vs fitted Gamma") +
    theme_minimal(base_size = 11)

p_hist2 + p_qq
```

## Optimal pre-emptive fraction $\alpha^*$

We minimise total expected cost $c(\alpha)$ over $\alpha \in [0,1]$ for each
combination of:

- Coverage fraction $f \in \{0.1,\; 0.3,\; 0.5\}$
- Targeting accuracy $\rho \in \{0,\; 0.5,\; 1\}$

For $\rho < 1$, imperfect targeting is implemented via the Gaussian copula:
synthetic priority scores are generated with Spearman rank correlation $\rho$ to
the true rate $\lambda_i$, and locations are ranked by score.

```{r alpha-star-setup}
# Cost parameters (illustrative; can be varied in sensitivity analyses)
R_VAL    <- 10   # c_i / c_v: individual-level cost ratio (illness cost per
                  # infected person divided by vaccination cost per person)
C_COV    <- 0.9  # vaccine coverage proportion C (fraction of pop vaccinated
                  # within each targeted location)
R_REACT  <- 0.5  # reactive effectiveness r
NU_VAL   <- 0.90 # pre-emptive effectiveness nu
T_PLAN   <- 3L   # planning horizon (years); matches VAX_YEARS length

# q_i = lambda_i * T: expected cumulative incidence per person over T years.
# Replaces p_i * a in the outbreak-probability framework.
cases_fit <- cases_fit |>
    mutate(q_i = rate_obs * T_PLAN)

# ------------------------------------------------------------------
# Cost function: total expected cost c(alpha), normalised by c_v * N,
# for a given set of location-level q_i values ranked by a priority score.
#
# alpha : pre-emptive fraction in [0,1]
# f     : vaccination capacity (fraction of total population coverable)
# q     : expected cumulative incidence per person (q_i = lambda_i * T)
# pop   : population sizes
# score : priority score (higher = vaccinate first)
# R     : individual-level cost ratio c_i / c_v
# r     : reactive effectiveness
# nu    : pre-emptive effectiveness
# C_cov : vaccine coverage proportion within each vaccinated location
# ------------------------------------------------------------------
cost_incidence <- function(alpha, f, q, pop, score, R, r, nu = 0.9, C_cov = 0.9) {
    P          <- sum(pop)
    f_pre_cap  <- alpha * f * P        # pre-emptive population budget

    # Rank by score (highest first) and fill pre-emptive budget
    ord        <- order(score, decreasing = TRUE)
    cum_pop    <- cumsum(pop[ord])
    n_pre      <- sum(cum_pop <= f_pre_cap)
    pre_idx    <- ord[seq_len(max(n_pre, 0L))]

    # Pre-emptive cost: C_cov (vaccination) + q_bar_pre*(1-nu)*R (residual illness)
    if (length(pre_idx) > 0) {
        q_bar_pre <- weighted.mean(q[pre_idx], pop[pre_idx])
    } else {
        q_bar_pre <- 0
    }
    f_pre_used <- sum(pop[pre_idx]) / P
    c_pre      <- f_pre_used * (C_cov + q_bar_pre * (1 - nu) * R)

    # Expected remaining illness burden in unvaccinated locations
    q_rem   <- weighted.mean(q, pop) - f_pre_used * q_bar_pre
    f_react <- (1 - alpha) * f

    # Reactive cost: C_cov + (1-r)*R per reactively covered unit;
    # uncovered remainder pays full illness cost R
    if (f_react >= q_rem) {
        c_react <- q_rem * (C_cov + (1 - r) * R)
    } else {
        c_react <- f_react * (C_cov + (1 - r) * R) + (q_rem - f_react) * R
    }

    c_pre + c_react
}
```

```{r alpha-star-compute}
f_grid   <- c(0.1, 0.3, 0.5)
rho_grid <- c(0.0, 0.5, 1.0)
alpha_grid <- seq(0, 1, length.out = 201)

alpha_star_results <- expand_grid(f = f_grid, rho = rho_grid) |>
    pmap_dfr(function(f, rho) {
        # Generate priority score with target Spearman rho to true lambda
        score <- make_rank_score(cases_fit$rate_obs, target_rho = rho)

        # Evaluate cost across alpha grid
        costs <- sapply(alpha_grid, function(a) {
            cost_incidence(
                alpha = a,
                f     = f,
                q     = cases_fit$q_i,
                pop   = cases_fit$pop_sum,
                score = score,
                R     = R_VAL,
                r     = R_REACT,
                nu    = NU_VAL,
                C_cov = C_COV
            )
        })

        alpha_opt <- alpha_grid[which.min(costs)]
        cost_opt  <- min(costs)

        tibble(
            f         = f,
            rho       = rho,
            alpha_star = alpha_opt,
            cost_min  = cost_opt,
            # full cost curve for plotting
            cost_curve = list(tibble(alpha = alpha_grid, cost = costs))
        )
    })

alpha_star_results |>
    dplyr::select(f, rho, alpha_star, cost_min) |>
    knitr::kable(
        digits    = c(1, 1, 3, 4),
        col.names = c("f", "rho", "alpha*", "min cost"),
        caption   = sprintf(
            "Optimal pre-emptive fraction alpha* | R = %g, r = %g, C = %g, nu = %g, T = %d yr",
            R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN)
    )
```

```{r fig-alpha-star, fig.width=11, fig.height=9}
#| fig-cap: "Expected cost c(alpha) as a function of the pre-emptive fraction
#|   alpha for each combination of f (columns) and rho (rows). The vertical
#|   dashed line marks the optimal alpha*."

cost_curves <- alpha_star_results |>
    dplyr::select(f, rho, cost_curve) |>
    unnest(cost_curve)

opt_points <- alpha_star_results |>
    dplyr::select(f, rho, alpha_star, cost_min)

ggplot(cost_curves, aes(x = alpha, y = cost)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    geom_vline(data = opt_points,
               aes(xintercept = alpha_star),
               linetype = "dashed", colour = "firebrick", linewidth = 0.8) +
    geom_point(data = opt_points,
               aes(x = alpha_star, y = cost_min),
               colour = "firebrick", size = 2.5) +
    facet_grid(
        rows = vars(paste0("rho = ", rho)),
        cols = vars(paste0("f = ", f))
    ) +
    labs(
        x     = expression("Pre-emptive fraction " ~ alpha),
        y     = expression("Normalised expected cost " ~ italic(c)(alpha)),
        title = sprintf("Optimal alpha* | R = %g, r = %g, C = %g, nu = %g, T = %d yr",
                        R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN)
    ) +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold"))
```

```{r fig-alpha-star-summary, fig.width=8, fig.height=4}
#| fig-cap: "Optimal pre-emptive fraction alpha* for each (f, rho) combination."

ggplot(alpha_star_results,
       aes(x = rho, y = alpha_star,
           colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 3) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_x_continuous(breaks = rho_grid) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        x     = expression(rho ~ "(targeting accuracy)"),
        y     = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
        title = sprintf("alpha* vs targeting accuracy | R = %g, r = %g, C = %g, nu = %g",
                        R_VAL, R_REACT, C_COV, NU_VAL)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```
