---
title: "ADM2 Cholera Incidence Rate (2010–2014)"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document selects ADM2 sites from the outbreak surveillance data and
computes the mean annual cholera incidence rate (cases per 100,000 person-years)
for each site over the 2010–2014 period. Population is derived by summing the
WorldPop density raster over each ADM2 polygon.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(scales)
library(knitr)
```

# Data

## Outbreak surveillance

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)

# Restrict to ADM2 level (location string has exactly 3 '::' separators,
# i.e. Continent::Country::ADM1::ADM2).
ts2 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

cat("ADM2 time-series rows:", nrow(ts2), "\n")
cat("Unique ADM2 locations:", n_distinct(ts2$location), "\n")
cat("Columns:", paste(names(ts2), collapse = ", "), "\n")
```

## Shapefiles

```{r load-shapefiles}
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Keep the most recent location_period_id per location (see risk_prediction_simple.qmd
# for the full multi-period analysis justifying this choice: 7 locations have >1
# period ID; 4 have identical polygons, 3 have minor-to-moderate revisions; in
# all cases the most recent boundary is appropriate for analysis).
loc_id_map <- ts2 |>
    distinct(location_period_id, location) |>
    group_by(location) |>
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) |>
    ungroup()

shp_admin2 <- shp_all |>
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id")) |>
    filter(!is.na(location), str_count(location, "::") == 3L) |>
    st_make_valid()

# Flag and remove degenerate sliver polygons before population extraction.
# A valid ADM2 health district cannot be < 1 km²; polygons that small are
# geometry artefacts that yield nonsensical population estimates (e.g. Moba
# Health District has a 0.098 km² sliver giving pop = 318 despite ~1400 cases).
shp_admin2$area_km2 <- as.numeric(st_area(shp_admin2)) / 1e6
slivers <- shp_admin2 |> st_drop_geometry() |>
    filter(area_km2 < 1) |>
    select(location, area_km2)
if (nrow(slivers) > 0) {
    cat("Removing", nrow(slivers), "sliver polygon(s) (area < 1 km²):\n")
    print(slivers)
}
shp_admin2 <- shp_admin2 |> filter(area_km2 >= 1)

stopifnot(!anyDuplicated(shp_admin2$location))
cat("ADM2 polygons after sliver removal:", nrow(shp_admin2), "\n")
```

## Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)

# The raster stores population density (persons per km²) in WGS84 at 1/6° × 1/6°
# resolution. The geodetic pixel area at the African mid-latitude (~1.3°N) is:
# (1/6 × 111.32)² × cos(1.3°) ≈ 344 km². Summing density values over a polygon
# and multiplying by POP_SCALE gives total head-count.
POP_SCALE <- 344  # pixel area: (1/6 * 111.32)^2 * cos(1.3 * pi/180) ≈ 344 km²

shp_pop  <- shp_admin2 |> st_transform(crs(pop_raster))
pop_vals <- exactextractr::exact_extract(pop_raster, shp_pop,
                                         fun = "sum", progress = FALSE)
shp_admin2$pop_sum <- pop_vals * POP_SCALE

cat("Total population:", format(round(sum(shp_admin2$pop_sum, na.rm = TRUE)),
                                big.mark = ","), "\n")
```

# ADM2 Site Selection

Select ADM2 sites that have at least one recorded outbreak **starting** in the
2010–2014 period. Sites with no outbreak record are retained as well (zero
cases), to keep the denominator consistent across sites and allow comparison
of high- and low-incidence areas.

```{r select-sites}
STUDY_YEARS <- 2010:2014

# Collapse time-series rows to one row per outbreak event.
# sCh: suspected cholera cases reported during each time-step.
outbreak_events <- ts2 |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(
        year_start  = year(min(TL)),
        year_end    = year(max(TR)),
        total_cases = sum(sCh, na.rm = TRUE),
        .groups     = "drop"
    )

# Sum cases by (location, year) using year_start as the index year.
# Using year_start ensures each outbreak is counted only once, in the year
# it first emerged, avoiding double-counting for multi-year outbreaks.
cases_annual <- outbreak_events |>
    filter(year_start %in% STUDY_YEARS) |>
    group_by(location, year = year_start) |>
    summarise(cases = sum(total_cases, na.rm = TRUE), .groups = "drop")

# Complete the panel: fill zeros for location-years with no outbreak.
all_locs <- unique(ts2$location)
cases_panel <- cases_annual |>
    right_join(expand_grid(location = all_locs, year = STUDY_YEARS),
               by = c("location", "year")) |>
    replace_na(list(cases = 0))

cat("Location-year observations:", nrow(cases_panel), "\n")
cat("Locations with ≥1 case in 2010-2014:",
    n_distinct(cases_panel$location[cases_panel$cases > 0]), "\n")
```

# Incidence Rate Computation

Incidence rate = total cases over 2010–2014 divided by (population × 5 years),
expressed per 100,000 person-years.

```{r compute-incidence}
# Sum cases over 2010-2014 for each location.
cases_total <- cases_panel |>
    group_by(location) |>
    summarise(
        total_cases   = sum(cases),
        n_years       = n_distinct(year),   # should be 5 for all locations
        .groups       = "drop"
    )

# Join population and compute incidence rate.
pop_lookup <- shp_admin2 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

incidence_df <- cases_total |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(
        # person-years = population × number of years observed
        person_years    = pop_sum * n_years,
        # incidence rate per 100,000 person-years
        incidence_rate  = total_cases / person_years * 1e5
    ) |>
    arrange(desc(incidence_rate))

cat("Sites with computable incidence rate:", nrow(incidence_df), "\n")
cat(sprintf("Incidence rate range: %.2f – %.2f per 100,000 person-years\n",
            min(incidence_df$incidence_rate),
            max(incidence_df$incidence_rate)))
```

# Results

## Summary table (top 20 sites by incidence rate)

```{r table-top20}
incidence_df |>
    slice_head(n = 20) |>
    dplyr::select(location, total_cases, pop_sum, person_years, incidence_rate) |>
    mutate(pop_sum      = round(pop_sum),
           person_years = round(person_years)) |>
    knitr::kable(
        digits  = c(0, 0, 0, 0, 2),
        col.names = c("Location", "Total cases", "Population",
                      "Person-years", "Rate per 100k py"),
        caption = "Top 20 ADM2 sites by mean annual cholera incidence rate (2010–2014)"
    )
```

## Distribution of incidence rates

```{r fig-incidence-dist, fig.width=10, fig.height=4}
#| fig-cap: "Left: histogram of incidence rates across all ADM2 sites (log10 scale,
#|   excluding zero-case sites). Right: empirical CDF. Both exclude sites with
#|   zero total cases."

df_nonzero <- incidence_df |> filter(total_cases > 0)

p_hist <- ggplot(df_nonzero, aes(x = incidence_rate)) +
    geom_histogram(bins = 40, fill = "steelblue", colour = "white", linewidth = 0.2) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Number of ADM2 sites",
         title = sprintf("Distribution across %d sites with ≥1 case",
                         nrow(df_nonzero))) +
    theme_minimal(base_size = 11)

p_cdf <- ggplot(df_nonzero |> arrange(incidence_rate) |>
                    mutate(ecdf = seq_len(n()) / n()),
                aes(x = incidence_rate, y = ecdf)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Cumulative proportion of sites",
         title = "ECDF of incidence rates") +
    theme_minimal(base_size = 11)

library(patchwork)
p_hist + p_cdf
```

## Map of incidence rates

```{r fig-incidence-map, fig.width=10, fig.height=8}
#| fig-cap: "Mean annual cholera incidence rate (2010–2014) per ADM2. Grey =
#|   sub-Saharan Africa country boundaries. Sites with zero cases are shown in
#|   light grey."

library(rnaturalearthdata)

ssa_sf <- sf::st_as_sf(rnaturalearthdata::countries50) |>
    dplyr::filter(subregion %in% c("Eastern Africa", "Middle Africa",
                                    "Southern Africa", "Western Africa")) |>
    sf::st_make_valid() |>
    sf::st_transform(sf::st_crs(shp_admin2))

shp_rate <- shp_admin2 |>
    left_join(incidence_df |> dplyr::select(location, incidence_rate),
              by = "location") |>
    replace_na(list(incidence_rate = 0))

ssa_bbox <- sf::st_bbox(ssa_sf)

ggplot() +
    geom_sf(data = ssa_sf,  fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = shp_rate, aes(fill = incidence_rate), colour = NA) +
    scale_fill_viridis_c(
        option    = "magma", direction = -1,
        trans     = "log1p",
        breaks    = c(0, 1, 10, 100, 1000),
        labels    = c("0", "1", "10", "100", "1000"),
        name      = "Rate\n(per 100k py)"
    ) +
    coord_sf(xlim = ssa_bbox[c("xmin", "xmax")],
             ylim = ssa_bbox[c("ymin", "ymax")]) +
    labs(title    = "Cholera incidence rate 2010–2014 (ADM2)",
         subtitle = "Cases attributed to year_start; population from WorldPop 2017 raster") +
    theme_minimal(base_size = 11) +
    theme(panel.grid  = element_blank(),
          axis.text   = element_blank(),
          axis.ticks  = element_blank(),
          legend.position = "right")
```

## Full incidence table

```{r table-full}
incidence_df |>
    dplyr::select(location, total_cases, pop_sum, incidence_rate) |>
    mutate(pop_sum = round(pop_sum)) |>
    knitr::kable(
        digits    = c(0, 0, 0, 2),
        col.names = c("Location", "Total cases (2010–2014)",
                      "Population", "Rate per 100k py"),
        caption   = "Cholera incidence rates for all ADM2 sites (2010–2014)"
    )
```
