---
title: "ADM2 Cholera Incidence Rate (2010–2019, Hurdle NB Fit)"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document selects ADM2 sites from the outbreak surveillance data and
computes the mean annual cholera incidence rate (cases per 100,000 person-years)
for each site over the 2010–2019 period. Population is derived by summing the
WorldPop density raster over each ADM2 polygon.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(scales)
library(knitr)
```

# Data

## Outbreak surveillance

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)

# Restrict to ADM2 level (location string has exactly 3 '::' separators,
# i.e. Continent::Country::ADM1::ADM2).
ts2 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

cat("ADM2 time-series rows:", nrow(ts2), "\n")
cat("Unique ADM2 locations:", n_distinct(ts2$location), "\n")
cat("Columns:", paste(names(ts2), collapse = ", "), "\n")
```

## Shapefiles

```{r load-shapefiles}
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Keep the most recent location_period_id per location (see risk_prediction_simple.qmd
# for the full multi-period analysis justifying this choice: 7 locations have >1
# period ID; 4 have identical polygons, 3 have minor-to-moderate revisions; in
# all cases the most recent boundary is appropriate for analysis).
loc_id_map <- ts2 |>
    distinct(location_period_id, location) |>
    group_by(location) |>
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) |>
    ungroup()

shp_admin2 <- shp_all |>
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id")) |>
    filter(!is.na(location), str_count(location, "::") == 3L) |>
    st_make_valid()

# Flag and remove degenerate sliver polygons before population extraction.
# A valid ADM2 health district cannot be < 1 km²; polygons that small are
# geometry artefacts that yield nonsensical population estimates (e.g. Moba
# Health District has a 0.098 km² sliver giving pop = 318 despite ~1400 cases).
shp_admin2$area_km2 <- as.numeric(st_area(shp_admin2)) / 1e6
slivers <- shp_admin2 |> st_drop_geometry() |>
    filter(area_km2 < 1) |>
    select(location, area_km2)
if (nrow(slivers) > 0) {
    cat("Removing", nrow(slivers), "sliver polygon(s) (area < 1 km²):\n")
    print(slivers)
}
shp_admin2 <- shp_admin2 |> filter(area_km2 >= 1)

stopifnot(!anyDuplicated(shp_admin2$location))
cat("ADM2 polygons after sliver removal:", nrow(shp_admin2), "\n")
```

## Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)

# The raster stores population density (persons per km²) in WGS84 at 1/6° × 1/6°
# resolution. The geodetic pixel area at the African mid-latitude (~1.3°N) is:
# (1/6 × 111.32)² × cos(1.3°) ≈ 344 km². Summing density values over a polygon
# and multiplying by POP_SCALE gives total head-count.
POP_SCALE <- 344  # pixel area: (1/6 * 111.32)^2 * cos(1.3 * pi/180) ≈ 344 km²

shp_pop  <- shp_admin2 |> st_transform(crs(pop_raster))
pop_vals <- exactextractr::exact_extract(pop_raster, shp_pop,
                                         fun = "sum", progress = FALSE)
shp_admin2$pop_sum <- pop_vals * POP_SCALE

cat("Total population:", format(round(sum(shp_admin2$pop_sum, na.rm = TRUE)),
                                big.mark = ","), "\n")
```

# ADM2 Site Selection

Select ADM2 sites that have at least one recorded outbreak **starting** in the
2010–2019 period. Sites with no outbreak record are retained as well (zero
cases), to keep the denominator consistent across sites and allow comparison
of high- and low-incidence areas.

```{r select-sites}
STUDY_YEARS <- 2010:2019

# Collapse time-series rows to one row per outbreak event.
# sCh: suspected cholera cases reported during each time-step.
outbreak_events <- ts2 |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(
        year_start  = year(min(TL)),
        year_end    = year(max(TR)),
        total_cases = sum(sCh, na.rm = TRUE),
        .groups     = "drop"
    )

# Sum cases by (location, year) using year_start as the index year.
# Using year_start ensures each outbreak is counted only once, in the year
# it first emerged, avoiding double-counting for multi-year outbreaks.
cases_annual <- outbreak_events |>
    filter(year_start %in% STUDY_YEARS) |>
    group_by(location, year = year_start) |>
    summarise(cases = sum(total_cases, na.rm = TRUE), .groups = "drop")

# Complete the panel: fill zeros for location-years with no outbreak.
all_locs <- unique(ts2$location)
cases_panel <- cases_annual |>
    right_join(expand_grid(location = all_locs, year = STUDY_YEARS),
               by = c("location", "year")) |>
    replace_na(list(cases = 0))

cat("Location-year observations:", nrow(cases_panel), "\n")
cat("Locations with ≥1 case in 2010-2019:",
    n_distinct(cases_panel$location[cases_panel$cases > 0]), "\n")
```

# Incidence Rate Computation

Incidence rate = total cases over 2010–2019 divided by (population × 10 years),
expressed per 100,000 person-years.

```{r compute-incidence}
# Sum cases over 2010-2019 for each location.
cases_total <- cases_panel |>
    group_by(location) |>
    summarise(
        total_cases   = sum(cases),
        n_years       = n_distinct(year),   # should be 5 for all locations
        .groups       = "drop"
    )

# Join population and compute incidence rate.
pop_lookup <- shp_admin2 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

incidence_df <- cases_total |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(
        # person-years = population × number of years observed
        person_years    = pop_sum * n_years,
        # incidence rate per 100,000 person-years
        incidence_rate  = total_cases / person_years * 1e5
    ) |>
    arrange(desc(incidence_rate))

cat("Sites with computable incidence rate:", nrow(incidence_df), "\n")
cat(sprintf("Incidence rate range: %.2f – %.2f per 100,000 person-years\n",
            min(incidence_df$incidence_rate),
            max(incidence_df$incidence_rate)))
```

# Results

## Summary table (top 20 sites by incidence rate)

```{r table-top20}
incidence_df |>
    slice_head(n = 20) |>
    dplyr::select(location, total_cases, pop_sum, person_years, incidence_rate) |>
    mutate(pop_sum      = round(pop_sum),
           person_years = round(person_years)) |>
    knitr::kable(
        digits  = c(0, 0, 0, 0, 2),
        col.names = c("Location", "Total cases", "Population",
                      "Person-years", "Rate per 100k py"),
        caption = "Top 20 ADM2 sites by mean annual cholera incidence rate (2010–2019)"
    )
```

## Distribution of incidence rates

```{r fig-incidence-dist, fig.width=10, fig.height=4}
#| fig-cap: "Left: histogram of incidence rates across all ADM2 sites (log10 scale,
#|   excluding zero-case sites). Right: empirical CDF. Both exclude sites with
#|   zero total cases."

df_nonzero <- incidence_df |> filter(total_cases > 0)

p_hist <- ggplot(df_nonzero, aes(x = incidence_rate)) +
    geom_histogram(bins = 40, fill = "steelblue", colour = "white", linewidth = 0.2) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Number of ADM2 sites",
         title = sprintf("Distribution across %d sites with ≥1 case",
                         nrow(df_nonzero))) +
    theme_minimal(base_size = 11)

p_cdf <- ggplot(df_nonzero |> arrange(incidence_rate) |>
                    mutate(ecdf = seq_len(n()) / n()),
                aes(x = incidence_rate, y = ecdf)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    scale_x_log10(labels = comma) +
    labs(x = "Incidence rate per 100,000 py (log scale)",
         y = "Cumulative proportion of sites",
         title = "ECDF of incidence rates") +
    theme_minimal(base_size = 11)

library(patchwork)
p_hist + p_cdf
```

## Map of incidence rates

```{r fig-incidence-map, fig.width=10, fig.height=8}
#| fig-cap: "Mean annual cholera incidence rate (2010–2019) per ADM2. Grey =
#|   sub-Saharan Africa country boundaries. Sites with zero cases are shown in
#|   light grey."

library(rnaturalearthdata)

ssa_sf <- sf::st_as_sf(rnaturalearthdata::countries50) |>
    dplyr::filter(subregion %in% c("Eastern Africa", "Middle Africa",
                                    "Southern Africa", "Western Africa")) |>
    sf::st_make_valid() |>
    sf::st_transform(sf::st_crs(shp_admin2))

shp_rate <- shp_admin2 |>
    left_join(incidence_df |> dplyr::select(location, incidence_rate),
              by = "location") |>
    replace_na(list(incidence_rate = 0))

ssa_bbox <- sf::st_bbox(ssa_sf)

ggplot() +
    geom_sf(data = ssa_sf,  fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = shp_rate, aes(fill = incidence_rate), colour = NA) +
    scale_fill_viridis_c(
        option    = "magma", direction = -1,
        trans     = "log1p",
        breaks    = c(0, 1, 10, 100, 1000),
        labels    = c("0", "1", "10", "100", "1000"),
        name      = "Rate\n(per 100k py)"
    ) +
    coord_sf(xlim = ssa_bbox[c("xmin", "xmax")],
             ylim = ssa_bbox[c("ymin", "ymax")]) +
    labs(title    = "Cholera incidence rate 2010–2019 (ADM2)",
         subtitle = "Cases attributed to year_start; population from WorldPop 2017 raster") +
    theme_minimal(base_size = 11) +
    theme(panel.grid  = element_blank(),
          axis.text   = element_blank(),
          axis.ticks  = element_blank(),
          legend.position = "right")
```

## Full incidence table

```{r table-full}
incidence_df |>
    dplyr::select(location, total_cases, pop_sum, incidence_rate) |>
    mutate(pop_sum        = round(pop_sum),
           incidence_rate = round(incidence_rate, 2)) |>
    DT::datatable(
        colnames  = c("Location", "Total cases (2010–2019)",
                      "Population", "Rate per 100k py"),
        caption   = "Cholera incidence rates for all ADM2 sites (2010–2019)",
        rownames  = FALSE,
        filter    = "top",
        options   = list(pageLength = 25, scrollX = TRUE)
    )
```

# Vaccine Impact Scenarios

## Setup

Scenarios vary two parameters:

- **ρ** (targeting accuracy): Spearman rank correlation between the vaccination
  priority score and true case burden. ρ = 1 = oracle (rank by true cases),
  ρ = 0 = random allocation.
- **f** (coverage): fraction of the total population vaccinated, from 0.1 to 0.5.

A **two-dose OCV regimen** is assumed (doses = 2 × vaccinated population).
Pre-emptive effectiveness **ν = 0.90**.

Cases averted are computed directly from observed case counts: for each
vaccinated location, cases averted = observed cases × ν. A **3-year case
total** (2010–2012) is used, reflecting the assumed duration of population-level
protection from a two-dose regimen.

```{r vaccine-setup}
NU        <- 0.90   # pre-emptive vaccine effectiveness (nu)
DOSES_PER <- 2L     # doses per vaccinated person
VAX_YEARS <- 2010:2012  # 3-year protection window

rho_values <- seq(0, 1, by = 0.1)
f_values   <- seq(0.1, 0.5, by = 0.1)

# 3-year observed case totals per location (used directly — no attack rate needed).
cases_3yr <- cases_panel |>
    filter(year %in% VAX_YEARS) |>
    group_by(location) |>
    summarise(cases_3yr = sum(cases), .groups = "drop")

sim_vax <- incidence_df |>
    left_join(cases_3yr, by = "location") |>
    replace_na(list(cases_3yr = 0))

total_pop     <- sum(sim_vax$pop_sum)
baseline_cases <- sum(sim_vax$cases_3yr)

cat(sprintf("Total population:     %s\n", format(round(total_pop), big.mark = ",")))
cat(sprintf("Baseline cases (%d-%d): %s\n",
            min(VAX_YEARS), max(VAX_YEARS),
            format(baseline_cases, big.mark = ",")))

# Gaussian copula: generate a priority score with a target Spearman rho to
# the true 3-year case burden. rho_P = 2 sin(pi * rho_S / 6) converts the
# target Spearman rho to the equivalent Pearson correlation.
make_rank_score <- function(truth, target_rho, seed = 42) {
    set.seed(seed)
    n       <- length(truth)
    z_truth <- qnorm(rank(truth, ties.method = "random") / (n + 1))
    z_noise <- rnorm(n)
    rho_p   <- 2 * sin(pi * target_rho / 6)
    pnorm(rho_p * z_truth + sqrt(1 - rho_p^2) * z_noise)
}
```

## Simulation

```{r vaccine-sim}
results <- expand_grid(rho = rho_values, f = f_values) |>
    pmap_dfr(function(rho, f) {
        # Priority score at target rho relative to true 3-year case burden
        score <- make_rank_score(sim_vax$cases_3yr, target_rho = rho)

        # Rank locations by score (highest risk first), fill budget f * P.
        # No partial-location vaccination: include only fully covered locations.
        ord     <- order(score, decreasing = TRUE)
        cum_pop <- cumsum(sim_vax$pop_sum[ord])
        n_vax   <- sum(cum_pop <= f * total_pop)
        vax_idx <- ord[seq_len(n_vax)]

        pop_vaccinated <- sum(sim_vax$pop_sum[vax_idx])
        doses          <- pop_vaccinated * DOSES_PER
        # Cases averted = observed cases in vaccinated locations × ν
        cases_averted  <- sum(sim_vax$cases_3yr[vax_idx]) * NU

        tibble(
            rho                        = rho,
            f                          = f,
            pop_vaccinated             = pop_vaccinated,
            doses                      = doses,
            cases_averted              = cases_averted,
            cases_per_1k_doses         = cases_averted / doses * 1000,
            doses_per_case_averted     = if (cases_averted > 0) doses / cases_averted else NA_real_
        )
    })
```

## Results table

```{r table-vaccine}
results |>
    mutate(
        pop_vaccinated = format(round(pop_vaccinated), big.mark = ","),
        doses          = format(round(doses),          big.mark = ","),
        cases_averted  = round(cases_averted, 1)
    ) |>
    knitr::kable(
        align   = c("r","r","r","r","r","r","r"),
        digits  = c(1, 1, 0, 0, 1, 2, 1),
        col.names = c("rho", "f", "Pop vaccinated", "Doses",
                      "Cases averted", "Cases averted / 1k doses",
                      "Doses / case averted"),
        caption = sprintf(
            "Vaccine impact scenarios | nu = %.2f, %d-dose regimen | 3-year cases (%d-%d)",
            NU, DOSES_PER, min(VAX_YEARS), max(VAX_YEARS))
    )
```

## Figures

```{r fig-vaccine-cases, fig.width=10, fig.height=5}
#| fig-cap: "Cases averted (left) and doses per case averted (right) by targeting
#|   accuracy (rho) and coverage fraction (f). rho = 1 = oracle; rho = 0 = random."

p_cases <- ggplot(results, aes(x = rho, y = cases_averted,
                                colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_y_continuous(labels = comma) +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = sprintf("Cases averted (%d-%d)", min(VAX_YEARS), max(VAX_YEARS)),
         title = "Cases averted") +
    theme_minimal(base_size = 11)

p_doses <- ggplot(results, aes(x = rho, y = doses_per_case_averted,
                                colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_y_continuous(labels = comma) +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = "Doses per case averted",
         title = "Doses required per case averted") +
    theme_minimal(base_size = 11)

p_cases + p_doses +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

```{r fig-vaccine-efficiency, fig.width=8, fig.height=5}
#| fig-cap: "Cases averted per 1,000 doses. Higher rho (better targeting) and
#|   lower f (more selective) yield greater efficiency."

ggplot(results, aes(x = rho, y = cases_per_1k_doses,
                    colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    labs(x = expression(rho ~ "(targeting accuracy)"),
         y = "Cases averted per 1,000 doses",
         title = "Vaccine programme efficiency") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Optimal Pre-emptive Fraction via Incidence Rate

## Revised cost model

We define the cost ratio at the **individual level**:

$$
R = \frac{c_i}{c_v},
$$

where $c_i$ is the cost of illness per infected person and $c_v$ is the cost of
vaccination per person. The population-level costs are then:

$$
C_I = a \cdot c_i \cdot N = a \cdot R \cdot c_v \cdot N,
\qquad
C_V = C \cdot c_v \cdot N,
$$

with $a$ = attack rate (fraction of the population infected during an outbreak),
$C$ = vaccine coverage proportion (fraction of the population vaccinated), and
$N$ = population size. Normalising by $c_v \cdot N$ (i.e. the cost of vaccinating
every person once), the single-population expected costs are:

$$
c_\text{pre}^{(1)} = C + p\,a\,(1-\nu)\,R, \qquad
c_\text{react}^{(1)} = p\bigl[C + a\,(1-r)\,R\bigr],
$$

where $p$ is the outbreak probability. Setting
$c_\text{pre}^{(1)} = c_\text{react}^{(1)}$ gives the critical probability:

$$
p_\text{crit}^{(1)} = \frac{C}{C + a\,R\,(\nu - r)}.
$$

Compared with the original formulation (where $R = a c_i / c_v C$ absorbed both
$a$ and $C$), the individual-level definition keeps $C$ and $a$ visible,
making it straightforward to assess how campaign coverage and attack rate jointly
determine strategy boundaries.

## Translation to incidence rates

With incidence data, we characterise each location $i$ by its **mean annual
incidence rate** $\lambda_i$ (cases per person per year). Over a planning horizon
of $T$ years, the expected cumulative proportion of the population infected is:

$$
q_i = \lambda_i \cdot T,
$$

which plays exactly the role of $p_i \cdot a$ in the outbreak-probability
framework (expected illness per person per period). The individual-level cost
ratio $R = c_i / c_v$ is unchanged.

### Reactive campaign threshold

A reactive campaign at location $i$ (not pre-emptively vaccinated) is
economically justified only if it reduces expected cost relative to no
intervention:

$$
C + q_i(1-r)R \;<\; q_i R
\quad\Longrightarrow\quad
q_i \;>\; \frac{C}{r\,R} \;\equiv\; q^*.
$$

Locations with $q_i \le q^*$ are left without any intervention under the
reactive strategy: the vaccination overhead $C$ exceeds the illness cost saved.
Equivalently, in rate terms the threshold is
$\lambda^* = q^*/T = C/(r\,R\,T)$.

### Cost equations

Replacing $p \cdot a$ with $q_i$ and normalising by $c_v \cdot N$, the costs
for a mixed strategy with pre-emptive fraction $\alpha$ are:

$$
c_\text{pre}(\alpha) = f_\text{pre}\bigl[C + \bar{q}_\text{pre}(1-\nu)\,R\bigr],
$$

$$
c_\text{react}(\alpha) = \sum_{i \in \mathcal{T}} w_i
\begin{cases}
  C + q_i(1-r)\,R & \text{if reactive budget available,} \\
  q_i\,R          & \text{otherwise,}
\end{cases}
\;+\; \sum_{i \notin \mathcal{T}} w_i\, q_i\,R,
$$

where $\mathcal{T} = \{i \notin \text{pre-emptive} : q_i > q^*\}$ is the set of
locations that trigger a reactive campaign, $w_i = N_i / N$ are population
weights, and the reactive budget covers $f_\text{react} = (1-\alpha)f$ of the
total population. Triggered locations are prioritised by descending $q_i$ when
the reactive budget is limited (since the campaign responds to observed burden,
not a predicted score).

The between-location distribution of $\lambda_i$ is modelled as
**Gamma**$(k, k/\mu)$, the natural continuous analogue of the Beta$(1,\theta)$
used in the outbreak-probability framework.

## Hurdle Negative Binomial fit to 2010–2019 data

A single Negative Binomial fitted to all sites collapses toward $k \approx 0$
because it conflates structural zeros with the at-risk distribution. A two-part
Gamma on positive sites alone still misses the tails (S-curve in the Q-Q),
indicating the continuous part is not Gamma-shaped.

A **hurdle Negative Binomial** model addresses both issues simultaneously:

$$
P(Y_i = 0) = 1 - \pi, \qquad
P(Y_i = y \mid Y_i > 0) = \frac{\text{NB}(y;\,\mu_i,\,k)}{1 - \text{NB}(0;\,\mu_i,\,k)}, \quad y = 1,2,\ldots
$$

where $\mu_i = \mu \cdot E_i$ (with $E_i = N_i T$ person-years) and $k$ is the
NB dispersion parameter. The logistic part models the probability $\pi$ of
observing any cases; the zero-truncated NB part models the count distribution
at sites that do have cases, giving it a free shape that can accommodate
heavier tails than the Gamma.

```{r gamma-poisson-fit}
library(MASS)
library(pscl)

# Use 2010-2019 (10-year horizon) to substantially reduce structural zeros.
FIT_YEARS <- 2010:2019
T_HORIZON <- length(FIT_YEARS)

cases_fit <- cases_panel |>
    filter(year %in% FIT_YEARS) |>
    group_by(location) |>
    summarise(cases_10yr = sum(cases), .groups = "drop") |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0)

# Exposure in person-years for each location
cases_fit <- cases_fit |>
    mutate(
        exposure  = pop_sum * T_HORIZON,   # person-years
        rate_obs  = cases_10yr / exposure, # observed incidence rate (per py)
        cases_int = as.integer(cases_10yr)
    )

# ── Hurdle NB ─────────────────────────────────────────────────────────────────
# Logistic component: P(Y > 0) = pi
# Zero-truncated NB component: Y | Y > 0 ~ TruncNB(mu_i = mu * E_i, size = k)
fit_hurdle <- pscl::hurdle(
    cases_int ~ 1 + offset(log(exposure)),
    data      = cases_fit,
    dist      = "negbin",
    zero.dist = "binomial"
)

mu_hat  <- exp(coef(fit_hurdle, "count")[["(Intercept)"]])  # rate per py
k_hat   <- fit_hurdle$theta                                  # NB dispersion

# pi = P(Y > 0) taken from the empirical proportion. The hurdle's logistic
# component collapses numerically when zeros are rare (~4%): the count model
# absorbs all variation and the binomial intercept diverges to -Inf.
# With 620/646 observed positives the empirical MLE is reliable.
pi_hat  <- mean(cases_fit$rate_obs > 0)

cat(sprintf("Hurdle NB fit (2010-2019):\n"))
cat(sprintf("  pi (P[site has cases]) = %.3f  (%d / %d sites, empirical)\n",
            pi_hat, sum(cases_fit$rate_obs > 0), nrow(cases_fit)))
cat(sprintf("  NB dispersion k        = %.3f\n", k_hat))
cat(sprintf("  Mean rate mu           = %.4e per py  (%.2f per 100k py)\n",
            mu_hat, mu_hat * 1e5))

# Cache cases_fit for standalone diagnostic scripts
saveRDS(cases_fit, "data/cases_fit_cache.rds")

# Helper: draw from zero-truncated NB via rejection sampling.
# mu_vec: site-level mean COUNTS (= mu_rate * exposure_i).
draw_trunc_nb <- function(mu_vec, size, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    counts   <- rnbinom(length(mu_vec), mu = mu_vec, size = size)
    zero_idx <- which(counts == 0)
    while (length(zero_idx) > 0) {
        counts[zero_idx] <- rnbinom(length(zero_idx),
                                    mu   = mu_vec[zero_idx],
                                    size = size)
        zero_idx <- which(counts[zero_idx] == 0) |>
                    (\(x) zero_idx[x])()
    }
    counts
}
```

```{r fig-gamma-poisson, fig.width=10, fig.height=4}
#| fig-cap: "Left: observed incidence rate distribution at positive sites vs
#|   simulation-based predictive density from the hurdle NB (2010–2019, log10
#|   scale). Right: predictive Q-Q — observed quantiles vs simulated quantiles
#|   from the hurdle NB using site-specific exposures (positive sites only)."

# Positive-rate sites only
df_obs <- cases_fit |> filter(rate_obs > 0)
n_pos  <- nrow(df_obs)

# Simulate rates from hurdle NB: draw truncated-NB counts using each site's
# exposure, then convert to rate. Repeat 200 times and pool for a smooth
# predictive density.
set.seed(42)
N_SIM_DIAG <- 200
sim_rates_all <- replicate(N_SIM_DIAG, {
    cnt <- draw_trunc_nb(mu_hat * df_obs$exposure, size = k_hat)
    cnt / df_obs$exposure
}, simplify = FALSE) |> unlist()

df_sim <- tibble(rate = sim_rates_all)

# ── Left: histogram + predictive KDE ─────────────────────────────────────────
p_hist2 <- ggplot() +
    geom_histogram(data = df_obs, aes(x = rate_obs, y = after_stat(density * x * log(10))),
                   bins = 35, fill = "steelblue", colour = "white", linewidth = 0.2) +
    geom_density(data = df_sim, aes(x = rate, y = after_stat(density * x * log(10))),
                 colour = "firebrick", linewidth = 1, adjust = 1.2) +
    scale_x_log10(labels = function(x) format(x * 1e5, big.mark = ",", digits = 3)) +
    labs(x = "Incidence rate (per 100,000 py, log scale)",
         y = "Density (log10 scale)",
         title = sprintf("Hurdle NB (k=%.2f, mu=%.2e, pi=%.2f)", k_hat, mu_hat, pi_hat),
         subtitle = sprintf("Red: predictive density; n = %d / %d sites with ≥1 case",
                            n_pos, nrow(cases_fit))) +
    theme_minimal(base_size = 11)

# ── Right: predictive Q-Q (one simulation, same n as observed) ───────────────
set.seed(99)
sim_rates_qq <- draw_trunc_nb(mu_hat * df_obs$exposure, size = k_hat) / df_obs$exposure

p_qq <- ggplot(tibble(obs = sort(df_obs$rate_obs), sim = sort(sim_rates_qq)),
               aes(x = sim, y = obs)) +
    geom_point(size = 1.5, alpha = 0.6, colour = "steelblue") +
    geom_abline(slope = 1, intercept = 0, colour = "firebrick") +
    scale_x_log10(labels = function(x) format(x * 1e5, digits = 2)) +
    scale_y_log10(labels = function(x) format(x * 1e5, digits = 2)) +
    labs(x = "Simulated quantiles — Hurdle NB (per 100,000 py, log scale)",
         y = "Observed quantiles (per 100,000 py, log scale)",
         title = sprintf("Predictive Q-Q: Hurdle NB (n = %d positive sites)", n_pos)) +
    theme_minimal(base_size = 11)

p_hist2 + p_qq
```

## Optimal pre-emptive fraction $\alpha^*$

We minimise total expected cost $c(\alpha)$ over $\alpha \in [0,1]$ for each
combination of:

- Coverage fraction $f \in \{0.1,\; 0.3,\; 0.5\}$
- Targeting accuracy $\rho \in \{0,\; 0.5,\; 1\}$

For $\rho < 1$, imperfect targeting is implemented via the Gaussian copula:
synthetic priority scores are generated with Spearman rank correlation $\rho$ to
the true rate $\lambda_i$, and locations are ranked by score.

```{r alpha-star-setup}
# Cost parameters (illustrative; can be varied in sensitivity analyses)
R_VAL    <- 500  # c_i / c_v: individual-level cost ratio (illness cost per
                  # infected person divided by vaccination cost per person)
C_COV    <- 0.9  # vaccine coverage proportion C (fraction of pop vaccinated
                  # within each targeted location)
R_REACT  <- 0.5  # reactive effectiveness r
NU_VAL   <- 0.90 # pre-emptive effectiveness nu
T_PLAN   <- 3L   # planning horizon (years); matches VAX_YEARS length

# q_i = lambda_i * T: expected cumulative incidence per person over T years.
# Replaces p_i * a in the outbreak-probability framework.
cases_fit <- cases_fit |>
    mutate(q_i = rate_obs * T_PLAN)

# Reactive threshold: top X% of observed incidence rates (baseline X = 25).
# Only the top-X% highest-incidence locations qualify for a reactive campaign.
# This replaces the analytic threshold q* = C_cov / (r * R) with an empirical
# percentile-based cutoff derived directly from the observed distribution.
Q_THRESH_PCT <- 25   # baseline: top 25% triggers reactive campaign
q_star <- quantile(cases_fit$q_i, 1 - Q_THRESH_PCT / 100)

cat(sprintf("Reactive threshold (top %d%% of q_i): q* = %.5f  (%.1f per 100,000 py over %d yr)\n",
            Q_THRESH_PCT, q_star, q_star / T_PLAN * 1e5, T_PLAN))
cat(sprintf("Locations above threshold: %d / %d (%.0f%%)\n",
            sum(cases_fit$q_i > q_star), nrow(cases_fit), Q_THRESH_PCT))

# ------------------------------------------------------------------
# Cost function: total expected cost c(alpha), normalised by c_v * N.
#
# alpha : pre-emptive fraction in [0,1]
# f     : vaccination capacity (fraction of total population coverable)
# q     : expected cumulative incidence per person (q_i = lambda_i * T)
# pop   : population sizes
# score : priority score for pre-emptive targeting (higher = vaccinate first)
# R     : individual-level cost ratio c_i / c_v
# r     : reactive effectiveness
# nu    : pre-emptive effectiveness
# C_cov : vaccine coverage proportion within each vaccinated location
#
# Reactive campaign threshold: q* = C_cov / (r * R).
# A reactive campaign at location i is only deployed if q_i > q*, i.e.,
# the illness burden saved exceeds the overhead cost C_cov.
# When the reactive budget is limited, triggered locations are covered in
# descending order of q_i (most severe first), since reactive responds to
# observed burden rather than a predicted score.
# ------------------------------------------------------------------
cost_incidence <- function(alpha, f, q, pop, score, R, r, nu = 0.9, C_cov = 0.9, q_star = NULL) {
    P         <- sum(pop)
    all_idx   <- seq_along(q)

    # ── Pre-emptive allocation (ranked by score) ──────────────────────────
    f_pre_cap <- alpha * f * P
    ord_pre   <- order(score, decreasing = TRUE)
    n_pre     <- sum(cumsum(pop[ord_pre]) <= f_pre_cap)
    pre_idx   <- ord_pre[seq_len(max(n_pre, 0L))]

    if (length(pre_idx) > 0) {
        q_bar_pre <- weighted.mean(q[pre_idx], pop[pre_idx])
    } else {
        q_bar_pre <- 0
    }
    f_pre_used <- sum(pop[pre_idx]) / P
    c_pre      <- f_pre_used * (C_cov + q_bar_pre * (1 - nu) * R)

    # ── Reactive allocation (remaining locations, threshold q*) ───────────
    rem_idx <- setdiff(all_idx, pre_idx)

    # Threshold: reactive campaign only triggered where q_i > q*.
    # If q_star is not supplied, fall back to the analytic threshold C_cov/(r*R).
    if (is.null(q_star)) q_star <- C_cov / (r * R)
    triggered_idx <- rem_idx[q[rem_idx] > q_star]
    no_action_idx <- rem_idx[q[rem_idx] <= q_star]

    # No-action locations: full illness cost, no campaign
    c_no_action <- sum(pop[no_action_idx] * q[no_action_idx] * R) / P

    # Triggered locations: cover in descending q_i order until budget exhausted
    f_react_cap <- (1 - alpha) * f * P
    if (length(triggered_idx) > 0) {
        ord_react  <- triggered_idx[order(q[triggered_idx], decreasing = TRUE)]
        n_react    <- sum(cumsum(pop[ord_react]) <= f_react_cap)
        react_idx  <- ord_react[seq_len(max(n_react, 0L))]
        missed_idx <- ord_react[seq_len(length(ord_react)) > n_react]

        c_react  <- sum(pop[react_idx]  * (C_cov + q[react_idx]  * (1 - r) * R)) / P
        c_missed <- sum(pop[missed_idx] *                q[missed_idx] * R)       / P
    } else {
        c_react  <- 0
        c_missed <- 0
    }

    c_pre + c_react + c_missed + c_no_action
}
```

```{r alpha-star-compute}
f_grid   <- c(0.1, 0.3, 0.5)
rho_grid <- c(0.0, 0.5, 1.0)
alpha_grid <- seq(0, 1, length.out = 201)

alpha_star_results <- expand_grid(f = f_grid, rho = rho_grid) |>
    pmap_dfr(function(f, rho) {
        # Generate priority score with target Spearman rho to true lambda
        score <- make_rank_score(cases_fit$rate_obs, target_rho = rho)

        # Evaluate cost across alpha grid
        costs <- sapply(alpha_grid, function(a) {
            cost_incidence(
                alpha  = a,
                f      = f,
                q      = cases_fit$q_i,
                pop    = cases_fit$pop_sum,
                score  = score,
                R      = R_VAL,
                r      = R_REACT,
                nu     = NU_VAL,
                C_cov  = C_COV,
                q_star = q_star
            )
        })

        alpha_opt <- alpha_grid[which.min(costs)]
        cost_opt  <- min(costs)

        tibble(
            f         = f,
            rho       = rho,
            alpha_star = alpha_opt,
            cost_min  = cost_opt,
            # full cost curve for plotting
            cost_curve = list(tibble(alpha = alpha_grid, cost = costs))
        )
    })

alpha_star_results |>
    dplyr::select(f, rho, alpha_star, cost_min) |>
    knitr::kable(
        digits    = c(1, 1, 3, 4),
        col.names = c("f", "rho", "alpha*", "min cost"),
        caption   = sprintf(
            "Optimal pre-emptive fraction alpha* | R = %g, r = %g, C = %g, nu = %g, T = %d yr | threshold = top %d%% of incidence",
            R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN, Q_THRESH_PCT)
    )
```

```{r fig-alpha-star, fig.width=11, fig.height=9}
#| fig-cap: "Expected cost c(alpha) as a function of the pre-emptive fraction
#|   alpha for each combination of f (columns) and rho (rows). The vertical
#|   dashed line marks the optimal alpha*."

cost_curves <- alpha_star_results |>
    dplyr::select(f, rho, cost_curve) |>
    unnest(cost_curve)

opt_points <- alpha_star_results |>
    dplyr::select(f, rho, alpha_star, cost_min)

ggplot(cost_curves, aes(x = alpha, y = cost)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    geom_vline(data = opt_points,
               aes(xintercept = alpha_star),
               linetype = "dashed", colour = "firebrick", linewidth = 0.8) +
    geom_point(data = opt_points,
               aes(x = alpha_star, y = cost_min),
               colour = "firebrick", size = 2.5) +
    facet_grid(
        rows = vars(paste0("rho = ", rho)),
        cols = vars(paste0("f = ", f))
    ) +
    labs(
        x     = expression("Pre-emptive fraction " ~ alpha),
        y     = expression("Normalised expected cost " ~ italic(c)(alpha)),
        title = sprintf("Optimal alpha* | R = %g, r = %g, C = %g, nu = %g, T = %d yr | threshold = top %d%%",
                        R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN, Q_THRESH_PCT)
    ) +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold"))
```

```{r fig-alpha-star-summary, fig.width=8, fig.height=4}
#| fig-cap: "Optimal pre-emptive fraction alpha* for each (f, rho) combination."

ggplot(alpha_star_results,
       aes(x = rho, y = alpha_star,
           colour = factor(f), group = factor(f))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 3) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_x_continuous(breaks = rho_grid) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        x     = expression(rho ~ "(targeting accuracy)"),
        y     = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
        title = sprintf("alpha* vs targeting accuracy | R = %g, r = %g, C = %g, nu = %g | threshold = top %d%%",
                        R_VAL, R_REACT, C_COV, NU_VAL, Q_THRESH_PCT)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Monte Carlo $\alpha^*$ from Two-Part Gamma Model

Rather than evaluating $\alpha^*$ on the single observed incidence landscape, we
propagate distributional uncertainty by **drawing replicate landscapes from the
fitted hurdle NB model** and computing the optimal pre-emptive fraction for each:

1. For each Monte Carlo iteration $s = 1, \ldots, S$, draw site-level counts
   from the hurdle NB using each site's observed exposure $E_i$:
   $$
   \tilde Y_i^{(s)} \sim
   \begin{cases} 0 & \text{w.p. } 1-\pi, \\ \text{TruncNB}(\mu \cdot E_i,\; k) & \text{w.p. } \pi, \end{cases}
   $$
   then set $\tilde\lambda_i^{(s)} = \tilde Y_i^{(s)} / E_i$ and
   $\tilde q_i^{(s)} = \tilde\lambda_i^{(s)} \cdot T$.
2. The reactive threshold $q^*$ is the empirical $(1-X/100)$-th quantile of
   $\tilde q_i^{(s)}$ (top $X = 25\%$ rule applied to the simulated landscape).
3. The targeting score is regenerated from $\tilde\lambda_i^{(s)}$ via the
   Gaussian copula at the target $\rho$, reflecting that the decision-maker's
   information is correlated with the realised landscape.
4. $\alpha^*$ is the minimiser of $c(\alpha)$ over a grid $[0,1]$.

The output is a distribution over $\alpha^*$ across $S = 500$ replications.

```{r mc-alpha-star}
N_MC <- 500
n_sites_fit <- nrow(cases_fit)

mc_results <- expand_grid(f = f_grid, rho = rho_grid) |>
    pmap_dfr(function(f, rho) {
        alpha_mc <- numeric(N_MC)
        for (s in seq_len(N_MC)) {
            set.seed(s)
            # Draw site rates from hurdle NB:
            #   zero sites : Bernoulli(1 - pi_hat)
            #   positive sites: zero-truncated NB count / exposure
            is_pos   <- rbinom(n_sites_fit, 1, pi_hat) == 1L
            counts_s <- integer(n_sites_fit)
            if (any(is_pos)) {
                counts_s[is_pos] <- draw_trunc_nb(
                    mu_vec = mu_hat * cases_fit$exposure[is_pos],
                    size   = k_hat
                )
            }
            lambda_s <- counts_s / cases_fit$exposure   # rate per py
            q_s      <- lambda_s * T_PLAN

            # Empirical top-X% threshold on simulated landscape
            q_thr_s <- quantile(q_s, 1 - Q_THRESH_PCT / 100)

            # Targeting score correlated with simulated rates at target rho
            score_s <- make_rank_score(lambda_s, target_rho = rho)

            costs_s <- sapply(alpha_grid, function(a) {
                cost_incidence(
                    alpha  = a,
                    f      = f,
                    q      = q_s,
                    pop    = cases_fit$pop_sum,
                    score  = score_s,
                    R      = R_VAL,
                    r      = R_REACT,
                    nu     = NU_VAL,
                    C_cov  = C_COV,
                    q_star = q_thr_s
                )
            })
            alpha_mc[s] <- alpha_grid[which.min(costs_s)]
        }

        tibble(
            f              = f,
            rho            = rho,
            alpha_mean     = mean(alpha_mc),
            alpha_median   = median(alpha_mc),
            alpha_lo       = quantile(alpha_mc, 0.025),
            alpha_hi       = quantile(alpha_mc, 0.975),
            draws          = list(alpha_mc)
        )
    })

mc_results |>
    dplyr::select(f, rho, alpha_mean, alpha_median, alpha_lo, alpha_hi) |>
    knitr::kable(
        digits    = c(1, 1, 3, 3, 3, 3),
        col.names = c("f", "rho", "mean alpha*", "median alpha*",
                      "2.5%", "97.5%"),
        caption   = sprintf(
            "MC alpha* distribution (S = %d) | R = %g, r = %g, C = %g, nu = %g, T = %d yr | threshold = top %d%%",
            N_MC, R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN, Q_THRESH_PCT)
    )
```

## Distribution of $\alpha^*$ across MC replications

```{r fig-mc-alpha-dist, fig.width=11, fig.height=7}
#| fig-cap: "Distribution of optimal pre-emptive fraction alpha* across 500 Monte
#|   Carlo draws from the two-part Gamma model. Each panel is one (f, rho)
#|   combination. The vertical dashed line is the observed-data alpha* from the
#|   grid search above; the red dot and bar show the MC median and 95% interval."

# Observed alpha* for comparison
obs_alpha <- alpha_star_results |>
    dplyr::select(f, rho, alpha_star_obs = alpha_star)

mc_long <- mc_results |>
    dplyr::select(f, rho, draws) |>
    unnest(draws) |>
    rename(alpha_star = draws)

mc_summary <- mc_results |>
    dplyr::select(f, rho, alpha_median, alpha_lo, alpha_hi) |>
    left_join(obs_alpha, by = c("f", "rho"))

ggplot(mc_long, aes(x = alpha_star)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white",
                   linewidth = 0.2, aes(y = after_stat(density))) +
    geom_vline(data = mc_summary,
               aes(xintercept = alpha_star_obs),
               linetype = "dashed", colour = "grey30", linewidth = 0.8) +
    geom_point(data = mc_summary,
               aes(x = alpha_median, y = 0),
               colour = "firebrick", size = 3) +
    geom_errorbarh(data = mc_summary,
                   aes(xmin = alpha_lo, xmax = alpha_hi, x = alpha_median, y = 0),
                   colour = "firebrick", height = 0.3, linewidth = 0.9) +
    facet_grid(
        rows = vars(paste0("rho = ", rho)),
        cols = vars(paste0("f = ", f))
    ) +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
        x     = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
        y     = "Density",
        title = sprintf("MC distribution of alpha* | R = %g, r = %g, C = %g, nu = %g | S = %d",
                        R_VAL, R_REACT, C_COV, NU_VAL, N_MC),
        subtitle = "Dashed = observed-data alpha*; red point + bar = MC median [95% CI]"
    ) +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold"))
```

## MC $\alpha^*$ summary: mean ± 95% CI

```{r fig-mc-alpha-summary, fig.width=8, fig.height=4}
#| fig-cap: "MC mean alpha* with 95% interval vs targeting accuracy rho, for each
#|   coverage fraction f. Points are means; ribbons are 2.5–97.5 percentiles.
#|   Dashed lines show observed-data alpha* for comparison."

mc_plot <- mc_results |>
    left_join(obs_alpha, by = c("f", "rho"))

ggplot(mc_plot, aes(x = rho, colour = factor(f), fill = factor(f),
                     group = factor(f))) +
    geom_ribbon(aes(ymin = alpha_lo, ymax = alpha_hi), alpha = 0.15,
                colour = NA) +
    geom_line(aes(y = alpha_mean), linewidth = 0.9) +
    geom_point(aes(y = alpha_mean), size = 3) +
    geom_line(aes(y = alpha_star_obs), linetype = "dashed", linewidth = 0.6) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_fill_viridis_d(option = "plasma", end = 0.85, name = "f") +
    scale_x_continuous(breaks = rho_grid) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        x        = expression(rho ~ "(targeting accuracy)"),
        y        = expression("alpha*"),
        title    = sprintf("MC alpha* (mean ± 95%% CI) | R = %g, r = %g, C = %g, nu = %g | threshold = top %d%%",
                           R_VAL, R_REACT, C_COV, NU_VAL, Q_THRESH_PCT),
        subtitle = "Solid = MC mean; ribbon = 95%% interval; dashed = observed-data alpha*"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Sensitivity to Reactive Campaign Threshold

We treat $q^*$ as the empirical quantile at the $(1 - X/100)$-th percentile of
the observed $q_i$ distribution, so exactly the top $X\%$ of locations qualify
for a reactive campaign. The baseline is $X = 25$. Here we vary $X$ across
$\{5, 10, 15, 20, 25, 30, 40, 50\}$ to show how $\alpha^*$ responds to a
stricter or more permissive threshold.

```{r threshold-sensitivity}
# Threshold percentiles to sweep (X = top-X% triggers reactive campaign)
thresh_pcts <- c(5, 10, 15, 20, 25, 30, 40, 50)

thresh_results <- expand_grid(
    thresh_pct = thresh_pcts,
    f          = f_grid,
    rho        = rho_grid
) |>
    pmap_dfr(function(thresh_pct, f, rho) {
        q_thr  <- quantile(cases_fit$q_i, 1 - thresh_pct / 100)
        score  <- make_rank_score(cases_fit$rate_obs, target_rho = rho)

        costs <- sapply(alpha_grid, function(a) {
            cost_incidence(
                alpha  = a,
                f      = f,
                q      = cases_fit$q_i,
                pop    = cases_fit$pop_sum,
                score  = score,
                R      = R_VAL,
                r      = R_REACT,
                nu     = NU_VAL,
                C_cov  = C_COV,
                q_star = q_thr
            )
        })

        tibble(
            thresh_pct = thresh_pct,
            f          = f,
            rho        = rho,
            alpha_star = alpha_grid[which.min(costs)],
            cost_min   = min(costs),
            q_star_val = q_thr
        )
    })
```

## alpha* vs threshold percentile

```{r fig-thresh-alpha, fig.width=11, fig.height=5}
#| fig-cap: "Optimal pre-emptive fraction alpha* as a function of the reactive
#|   campaign threshold percentile X (top X% of locations qualify for reactive
#|   campaign). Columns: coverage fraction f. Colours: targeting accuracy rho.
#|   Dashed vertical line marks the baseline X = 25."

ggplot(thresh_results,
       aes(x = thresh_pct, y = alpha_star,
           colour = factor(rho), group = factor(rho))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = 25, linetype = "dashed", colour = "grey40") +
    facet_wrap(~ paste0("f = ", f), nrow = 1) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = expression(rho)) +
    scale_x_continuous(breaks = thresh_pcts) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        x     = "Threshold percentile X  (top X% triggers reactive campaign)",
        y     = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
        title = sprintf("Sensitivity of alpha* to reactive threshold | R = %g, r = %g, C = %g, nu = %g",
                        R_VAL, R_REACT, C_COV, NU_VAL)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

## Minimum cost vs threshold percentile

```{r fig-thresh-cost, fig.width=11, fig.height=5}
#| fig-cap: "Minimum normalised expected cost at alpha* as a function of the
#|   reactive threshold percentile X. Lower is better. Columns: coverage f.
#|   Colours: targeting accuracy rho. Dashed line marks baseline X = 25."

ggplot(thresh_results,
       aes(x = thresh_pct, y = cost_min,
           colour = factor(rho), group = factor(rho))) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = 25, linetype = "dashed", colour = "grey40") +
    facet_wrap(~ paste0("f = ", f), nrow = 1) +
    scale_colour_viridis_d(option = "plasma", end = 0.85, name = expression(rho)) +
    scale_x_continuous(breaks = thresh_pcts) +
    labs(
        x     = "Threshold percentile X  (top X% triggers reactive campaign)",
        y     = expression("Minimum normalised expected cost at " ~ alpha^"*"),
        title = sprintf("Cost sensitivity to reactive threshold | R = %g, r = %g, C = %g, nu = %g",
                        R_VAL, R_REACT, C_COV, NU_VAL)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

## Summary table

```{r table-thresh}
thresh_results |>
    mutate(
        q_star_per100k = round(q_star_val / T_PLAN * 1e5, 1)
    ) |>
    dplyr::select(thresh_pct, f, rho, alpha_star, cost_min, q_star_per100k) |>
    knitr::kable(
        digits    = c(0, 1, 1, 3, 4, 1),
        col.names = c("Top X%", "f", "rho", "alpha*", "min cost",
                      "q* (per 100k py)"),
        caption   = sprintf(
            "alpha* across threshold percentiles | R = %g, r = %g, C = %g, nu = %g, T = %d yr",
            R_VAL, R_REACT, C_COV, NU_VAL, T_PLAN)
    )
```
