---
title: "Annual Outbreak Prediction Model"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document develops an annual outbreak prediction model for admin-2 units
using cholera surveillance data from 2010--2017. The model is evaluated via
**expanding-window temporal cross-validation** and **spatial cross-validation**,
and the final predictions (for 2018) are saved for use in the vaccine impact
simulation.

**Design**:

| Fold  | Feature years | Outcome year | Purpose    |
|-------|--------------|--------------|------------|
| 1     | 2010--2014   | 2015         | Validation |
| 2     | 2011--2015   | 2016         | Validation |
| 3     | 2012--2016   | 2017         | Validation |
| Final | 2013--2017   | 2018         | Prediction |

Five classifiers are compared: logistic regression, elastic net, random forest,
XGBoost, and GAM.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(patchwork)
library(pROC)
library(scales)
library(knitr)
library(mgcv)         # GAM
library(glmnet)       # elastic net / LASSO / ridge
library(ranger)       # fast random forest
library(xgboost)      # gradient boosted trees
library(PRROC)        # precision-recall AUC
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```


## Data Loading

### Outbreak surveillance data

```{r load-outbreak-data}
# Raw surveillance data: each row = (location, time interval TL-TR, outbreak_number)
# Location format: "REGION::COUNTRY::ADMIN1::ADMIN2" (3 separators = admin2)
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)
# Keep admin2 (adm=2) and admin3 (adm=3) only
ts23 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm %in% c(2, 3))

# Collapse time-series rows into one row per outbreak event, extracting
# start/end years, duration, case counts, and a unique ID
outbreak_events <- ts23 |>
    group_by(location, adm, outbreak_number) |>
    summarise(
        year_start      = year(min(TL)),
        year_end        = year(max(TR)),
        duration_weeks  = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
        total_cases     = sum(sCh, na.rm = TRUE),
        total_deaths    = sum(deaths, na.rm = TRUE),
        peak_weekly_sCh = max(sCh, na.rm = TRUE),
        id_outbreak     = paste(first(location), min(TL), max(TR), sep = "_"),
        .groups         = "drop"
    )

all_years <- 2010:2023
all_locs <- unique(ts23$location)

# Build a yearly outbreak summary per location: for each (location, year),
# aggregate outbreak counts, total cases, maximum duration, etc.
# Assigns each outbreak to its END year (observable after conclusion).
outbreak_year_summary <- outbreak_events |>
    filter(year_end %in% all_years) |>
    group_by(location, year = year_end) |>
    summarise(
        outbreak_count     = n(),
        total_cases        = sum(total_cases, na.rm = TRUE),
        max_duration_days  = max(duration_weeks * 7, na.rm = TRUE),
        mean_duration_days = mean(duration_weeks * 7, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(outbreak_count = 0L, total_cases = 0,
                    max_duration_days = 0, mean_duration_days = 0))

# loc_year_full: complete (location × year) panel with n_active = number of
# outbreaks ACTIVE (ongoing) in that year. Used for the outcome variable.
loc_year <- outbreak_events %>%
    mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) %>%
    unnest(years) %>%
    rename(year = years) %>%
    filter(year %in% all_years) %>%
    count(location, year, name = "n_active")

loc_year_full <- expand_grid(location = all_locs, year = all_years) %>%
    left_join(loc_year, by = c("location", "year")) %>%
    replace_na(list(n_active = 0L))

# loc_year_end_full: complete (location × year) panel with n_ended = number of
# outbreaks that ENDED in that year. Used for feature construction (hist_count)
# because an outbreak's end year marks when it became observable.
loc_year_end_full <- outbreak_events %>%
    filter(year_end %in% all_years) %>%
    count(location, year_end, name = "n_ended") %>%
    rename(year = year_end) %>%
    right_join(expand_grid(location = all_locs, year = all_years),
        by = c("location", "year")
    ) %>%
    replace_na(list(n_ended = 0L))

cat("Surveillance locations:", length(all_locs), "\n")
cat("Year range:", range(all_years), "\n")
cat("Total outbreak events:", nrow(outbreak_events), "\n")
cat("Total outbreak-end-years:", sum(loc_year_end_full$n_ended), "\n")
```

### Shapefiles and spatial neighbours

```{r load-shapefiles}
# Shapefiles are keyed by location_period_id; we need to map them to
# the human-readable location strings (e.g. "AFR::COD::Kinshasa::...").
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Build a lookup: for each location, keep the most recent location_period_id
# (a location can have multiple period IDs if boundaries changed over time)
loc_id_map <- ts23 %>%
    distinct(location_period_id, location) %>%
    group_by(location) %>%
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) %>%
    ungroup()

# Join shapefile geometries to location names via the ID map
shp_named <- shp_all %>%
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id"))

# Keep only admin2/3 polygons (3 or 4 "::" separators in location string)
shp_admin23 <- shp_named %>%
    filter(!is.na(location), str_count(location, "::") %in% c(3, 4))

cat("ADM2/3 polygons:", nrow(shp_admin23), "\n")

# Build a spatial neighbour list: for each polygon, find all neighbours
# whose centroid is within DIST_LIMIT_KM km (geodesic distance).
# Used for the nbr_hist feature (spatial contagion signal).
DIST_LIMIT_KM <- 50
centroids_sf <- shp_admin23 %>%
    st_make_valid() %>%
    st_point_on_surface()

# Pairwise geodesic distance matrix (meters); convert to km
dist_mat_sp <- st_distance(centroids_sf) |> units::drop_units()
dist_mat_sp <- dist_mat_sp / 1000 # m -> km
rownames(dist_mat_sp) <- colnames(dist_mat_sp) <- shp_admin23$location
diag(dist_mat_sp) <- Inf

# For each location, store neighbours within the distance limit (sorted)
nbr_list <- setNames(
    lapply(shp_admin23$location, function(loc) {
        d <- dist_mat_sp[loc, ]
        names(sort(d[d <= DIST_LIMIT_KM]))
    }),
    shp_admin23$location
)

# Extract centroid coordinates for GAM spatial smooth
centroid_coords <- centroids_sf |>
    st_coordinates() |>
    as_tibble() |>
    rename(lon = X, lat = Y) |>
    mutate(location = shp_admin23$location)
```

### Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)  # convert RasterLayer → SpatRaster
cat("Raster dimensions:", dim(pop_raster), "\n")
cat("Resolution:", res(pop_raster), "degrees (~20 km)\n")
```


## Population Extraction

Aggregate the 20 km population raster to ADM2 polygons via zonal statistics.

The raster stores **mean population per ~1 km² source cell**, having been
aggregated from the WorldPop ~1 km PPP (People Per Pixel) product to 20 km
resolution by averaging. Each 20 km pixel spans approximately
$(0.167° \times 111\;\text{km/°})^2 \approx 344\;\text{km}^2$ at the equator,
i.e.\ roughly 344 underlying 1 km² source cells. Because `exact_extract(fun =
"sum")` sums these per-km² means, the result must be multiplied by the pixel
area (~344 km²) to recover actual population counts. Verification: the
continent-wide raster sum ($\approx$ 3.65 M) $\times$ 344 $\approx$ 1.26
billion, matching Africa's 2017 population.

```{r population-extraction}
POP_SCALE <- 344 # pixel area in km² ≈ (0.167° × 111 km)²

shp_pop <- shp_admin23 |>
    st_make_valid() |>
    st_transform(crs(pop_raster))

# Sum mean-population values within each polygon, then scale to actual counts
pop_by_adm2 <- exactextractr::exact_extract(pop_raster, shp_pop,
    fun = "sum",
    progress = FALSE
)
pop_by_adm2 <- pop_by_adm2 * POP_SCALE

shp_admin23$pop_sum <- pop_by_adm2

cat(
    "Polygons with population > 0:",
    sum(!is.na(pop_by_adm2) & pop_by_adm2 > 0), "of", nrow(shp_admin23), "\n"
)
cat("Total population:", format(round(sum(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n")
cat(
    "Range:", format(round(min(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "-",
    format(round(max(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n"
)
```


## Annual Outbreak Prediction Model

### Overview

We predict outbreak occurrence one year ahead for admin-2 units using an
**expanding window** design. Only outbreak surveillance data are used
(no environmental or demographic covariates). Data are available for
2010--2017. The year 2018 is the true prediction target.

| Fold  | Feature years | Outcome year | Purpose    |
|-------|--------------|--------------|------------|
| 1     | 2010--2014   | 2015         | Validation |
| 2     | 2011--2015   | 2016         | Validation |
| 3     | 2012--2016   | 2017         | Validation |
| Final | 2013--2017   | 2018         | Prediction |

Within each temporal fold, 5-fold spatial cross-validation tests
generalization across regions. Five classifiers are compared: logistic
regression, elastic net, random forest, XGBoost, and GAM.

### Feature engineering

```{r build-features}
T_WINDOW <- 5L  # number of feature years (rolling window)

# Build features for a given feature window ending at year_end.
# Uses the most recent T_WINDOW years only (e.g. T=5: 2013-2017 for 2018).
# Returns one row per location with same-unit and neighbor features.
build_features <- function(year_end, T = T_WINDOW, nbr_map = nbr_list,
                           coord_lookup = centroid_coords) {
    feat_yrs <- (year_end - T + 1L):year_end
    n_yrs <- length(feat_yrs)

    # Per-location summary across the feature window
    loc_summary <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            n_outbreak_years    = sum(outbreak_count > 0),
            prop_outbreak_years = n_outbreak_years / n_yrs,
            total_outbreaks     = sum(outbreak_count),
            mean_cases_per_year = mean(total_cases),
            max_cases_any_year  = max(total_cases),
            mean_duration       = mean(mean_duration_days),
            max_duration        = max(max_duration_days),
            .groups = "drop"
        )

    # Recent-year features
    recent_1yr <- outbreak_year_summary |>
        filter(year == year_end) |>
        transmute(location, recent_outbreak = as.integer(outbreak_count > 0))

    recent_2yr_yrs <- intersect(c(year_end - 1L, year_end), feat_yrs)
    recent_2yr <- outbreak_year_summary |>
        filter(year %in% recent_2yr_yrs) |>
        group_by(location) |>
        summarise(recent_2yr_outbreaks = sum(outbreak_count > 0), .groups = "drop")

    # Trend features: slope of outbreak occurrence and cases on year
    trend_df <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            trend = if (n() >= 2)
                coef(lm(as.integer(outbreak_count > 0) ~ year))[2] else 0,
            trend_cases = if (n() >= 2)
                coef(lm(total_cases ~ year))[2] else 0,
            .groups = "drop"
        )

    # Years since last outbreak
    last_ob <- outbreak_year_summary |>
        filter(year %in% feat_yrs, outbreak_count > 0) |>
        group_by(location) |>
        summarise(last_ob_year = max(year), .groups = "drop")

    yrs_since <- tibble(location = all_locs) |>
        left_join(last_ob, by = "location") |>
        mutate(years_since_last_outbreak = ifelse(
            is.na(last_ob_year), n_yrs + 1L, as.integer(year_end - last_ob_year)
        )) |>
        dplyr::select(location, years_since_last_outbreak)

    # Combine same-unit features
    feat <- tibble(location = all_locs) |>
        left_join(loc_summary, by = "location") |>
        left_join(recent_1yr, by = "location") |>
        left_join(recent_2yr, by = "location") |>
        left_join(trend_df, by = "location") |>
        left_join(yrs_since, by = "location") |>
        replace_na(list(
            n_outbreak_years = 0L, prop_outbreak_years = 0,
            total_outbreaks = 0L, mean_cases_per_year = 0,
            max_cases_any_year = 0, mean_duration = 0, max_duration = 0,
            recent_outbreak = 0L, recent_2yr_outbreaks = 0L,
            trend = 0, trend_cases = 0
        ))

    # Neighbor features: helper to compute a summary stat across neighbours
    make_nbr_feat <- function(values, locs, nbr_map, fun = mean) {
        sv <- setNames(values, locs)
        map_dbl(locs, function(loc) {
            nbrs <- nbr_map[[loc]]
            if (is.null(nbrs) || length(nbrs) == 0L) return(0)
            nbrs <- nbrs[nbrs %in% names(sv)]
            if (length(nbrs) == 0L) return(0)
            fun(sv[nbrs])
        })
    }

    feat <- feat |>
        mutate(
            nbr_prop_outbreak = make_nbr_feat(
                prop_outbreak_years, location, nbr_map),
            nbr_mean_cases = make_nbr_feat(
                mean_cases_per_year, location, nbr_map),
            nbr_recent_outbreak_prop = make_nbr_feat(
                recent_outbreak, location, nbr_map),
            nbr_recent_2yr_prop = make_nbr_feat(
                as.numeric(recent_2yr_outbreaks > 0), location, nbr_map),
            nbr_max_cases = make_nbr_feat(
                max_cases_any_year, location, nbr_map, fun = max),
            nbr_trend = make_nbr_feat(trend, location, nbr_map)
        )

    if (!is.null(coord_lookup)) {
        feat <- feat |> left_join(coord_lookup, by = "location")
    }
    feat
}

# Define feature sets
SAME_UNIT_FEATURES <- c(
    "n_outbreak_years", "prop_outbreak_years", "total_outbreaks",
    "mean_cases_per_year", "max_cases_any_year",
    "mean_duration", "max_duration",
    "recent_outbreak", "recent_2yr_outbreaks",
    "trend", "trend_cases", "years_since_last_outbreak"
)
NBR_FEATURES <- c(
    "nbr_prop_outbreak", "nbr_mean_cases",
    "nbr_recent_outbreak_prop", "nbr_recent_2yr_prop",
    "nbr_max_cases", "nbr_trend"
)
ALL_FEATURES <- c(SAME_UNIT_FEATURES, NBR_FEATURES)
cat("Features:", length(ALL_FEATURES),
    "(", length(SAME_UNIT_FEATURES), "same-unit +",
    length(NBR_FEATURES), "neighbour )\n")
```

### Model fitting helpers

```{r model-helpers}
# Fit all 5 models on training data, predict on test data.
# Returns a tibble with location, obs, and prediction columns.
fit_and_predict <- function(train, test, features = ALL_FEATURES) {
    X_train <- as.matrix(train[, features])
    y_train <- train$obs
    X_test <- as.matrix(test[, features])
    preds <- tibble(location = test$location, obs = test$obs)

    # 1. Logistic regression
    lr_fit <- glm(obs ~ ., data = train[, c("obs", features)], family = binomial)
    preds$Logistic <- predict(lr_fit, newdata = test, type = "response")

    # 2. Elastic net (choose best alpha from {0, 0.5, 1} via internal CV)
    best_enet <- NULL; best_auc <- -Inf
    for (a in c(0, 0.5, 1)) {
        cv_fit <- tryCatch(
            cv.glmnet(X_train, y_train, family = "binomial",
                      alpha = a, nfolds = 5, type.measure = "auc"),
            error = function(e) NULL
        )
        if (!is.null(cv_fit) && max(cv_fit$cvm) > best_auc) {
            best_auc <- max(cv_fit$cvm); best_enet <- cv_fit
        }
    }
    preds$`Elastic Net` <- as.numeric(
        predict(best_enet, newx = X_test, s = "lambda.min", type = "response"))

    # 3. Random forest (ranger)
    set.seed(42)
    rf_fit <- ranger(
        obs ~ ., data = train[, c("obs", features)] |> mutate(obs = factor(obs)),
        num.trees = 500, probability = TRUE,
        min.node.size = 5, mtry = max(1, floor(sqrt(length(features))))
    )
    preds$`Random Forest` <- predict(rf_fit, data = test)$predictions[, "1"]

    # 4. XGBoost
    set.seed(42)
    dtrain <- xgb.DMatrix(data = X_train, label = y_train)
    dtest <- xgb.DMatrix(data = X_test)
    xgb_params <- list(objective = "binary:logistic", eval_metric = "auc",
                       max_depth = 4, eta = 0.1, subsample = 0.8,
                       colsample_bytree = 0.8)
    xgb_cv <- xgb.cv(params = xgb_params, data = dtrain, nrounds = 300,
                      nfold = min(5, nrow(train)),
                      early_stopping_rounds = 20, verbose = 0)
    best_nrounds <- xgb_cv$best_iteration
    if (is.null(best_nrounds) || best_nrounds == 0) best_nrounds <- 50
    xgb_fit <- xgb.train(params = xgb_params, data = dtrain,
                         nrounds = best_nrounds, verbose = 0)
    preds$XGBoost <- predict(xgb_fit, dtest)

    # 5. GAM with smooth terms on key continuous features
    gam_feats <- c("n_outbreak_years", "mean_cases_per_year",
                   "recent_2yr_outbreaks", "years_since_last_outbreak",
                   "nbr_prop_outbreak", "nbr_mean_cases")
    gam_formula <- as.formula(paste(
        "obs ~", paste0("s(", gam_feats, ", k = 5)", collapse = " + ")))
    gam_fit <- tryCatch(
        gam(gam_formula, data = train, family = binomial, method = "REML"),
        error = function(e) {
            gam(obs ~ s(n_outbreak_years, k = 5) + s(nbr_prop_outbreak, k = 5),
                data = train, family = binomial, method = "REML")
        }
    )
    preds$GAM <- as.numeric(predict(gam_fit, newdata = test, type = "response"))

    preds
}

MODEL_NAMES <- c("Logistic", "Elastic Net", "Random Forest", "XGBoost", "GAM")

# Compute evaluation metrics for a single (obs, pred) pair
eval_metrics <- function(obs, pred) {
    roc_obj <- pROC::roc(obs, pred, quiet = TRUE)
    auc_roc <- as.numeric(pROC::auc(roc_obj))
    pr_obj <- PRROC::pr.curve(scores.class0 = pred[obs == 1],
                              scores.class1 = pred[obs == 0])
    auc_pr <- pr_obj$auc.integral
    brier <- mean((pred - obs)^2)
    rho   <- cor(pred, obs, method = "spearman")
    coords <- pROC::coords(roc_obj, "best", best.method = "youden", ret = "all")
    tibble(rho = rho, AUC_ROC = auc_roc, AUC_PR = auc_pr, Brier = brier,
           Sensitivity = coords$sensitivity[1],
           Specificity = coords$specificity[1],
           Threshold = coords$threshold[1])
}
```

### Expanding-window temporal cross-validation

```{r temporal-cv}
# Temporal folds: T=5 rolling feature window ending at train_end
# Fold 1: features 2010-2014 → outcome 2015
# Fold 2: features 2011-2015 → outcome 2016
# Fold 3: features 2012-2016 → outcome 2017
temporal_folds <- tibble(
    fold = 1:3,
    train_end = 2014:2016,
    outcome_year = 2015:2017
)

# Build feature + outcome data for each fold
fold_data <- map(1:nrow(temporal_folds), function(i) {
    fe <- temporal_folds[i, ]
    feat <- build_features(fe$train_end)
    outcome <- loc_year_full |>
        filter(year == fe$outcome_year) |>
        group_by(location) |>
        summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")
    feat |>
        left_join(outcome, by = "location") |>
        replace_na(list(obs = 0L))
})
names(fold_data) <- paste0("fold", 1:3)

for (i in 1:3) {
    fe <- temporal_folds[i, ]
    d <- fold_data[[i]]
    cat(sprintf("Fold %d: features [2010,%d] → outcome %d | n=%d, prev=%.1f%%\n",
                i, fe$train_end, fe$outcome_year, nrow(d), 100 * mean(d$obs)))
}

# Final prediction data: features from 2010-2017
final_features <- build_features(2017)
```

### Temporal fold evaluation

For each fold, we train on **all preceding folds** and test on the current
fold. Fold 1 has no prior data, so it is evaluated via spatial CV only
(see next section).

```{r temporal-eval}
# Folds 2 and 3: train on preceding folds, test on current
temporal_results <- map_dfr(2:3, function(i) {
    # Expanding training set: pool all folds before i
    train <- bind_rows(fold_data[1:(i - 1)])
    test <- fold_data[[i]]

    preds <- fit_and_predict(train, test)

    map_dfr(MODEL_NAMES, function(m) {
        metrics <- eval_metrics(preds$obs, preds[[m]])
        metrics |> mutate(
            model = m, fold = i,
            train_end = temporal_folds$train_end[i],
            outcome_year = temporal_folds$outcome_year[i]
        )
    })
})

temporal_results |>
    dplyr::select(fold, outcome_year, model, rho, AUC_ROC, AUC_PR, Brier,
                  Sensitivity, Specificity) |>
    arrange(fold, desc(AUC_ROC)) |>
    knitr::kable(digits = 3,
        caption = "Temporal CV: performance by fold and model")
```

### Spatial cross-validation within each temporal fold

```{r spatial-cv}
set.seed(42)

spatial_cv_results <- map_dfr(1:3, function(i) {
    d <- fold_data[[i]]
    n_folds_sp <- 5

    # Assign spatial folds (random partition of locations)
    locs <- unique(d$location)
    sp_fold_ids <- sample(rep(1:n_folds_sp, length.out = length(locs)))
    loc_folds <- tibble(location = locs, sp_fold = sp_fold_ids)
    d <- d |> left_join(loc_folds, by = "location")

    map_dfr(1:n_folds_sp, function(sf) {
        train_sp <- d |> filter(sp_fold != sf)
        test_sp <- d |> filter(sp_fold == sf)

        preds <- fit_and_predict(train_sp, test_sp)

        map_dfr(MODEL_NAMES, function(m) {
            metrics <- eval_metrics(preds$obs, preds[[m]])
            metrics |> mutate(model = m, temporal_fold = i, spatial_fold = sf)
        })
    })
})

# Summary: mean across spatial folds, per temporal fold and model
spatial_cv_summary <- spatial_cv_results |>
    group_by(temporal_fold, model) |>
    summarise(across(c(rho, AUC_ROC, AUC_PR, Brier, Sensitivity, Specificity),
                     list(mean = mean, sd = sd)), .groups = "drop")

spatial_cv_summary |>
    dplyr::select(temporal_fold, model, rho_mean, rho_sd,
                  AUC_ROC_mean, AUC_ROC_sd, AUC_PR_mean, Brier_mean) |>
    arrange(temporal_fold, desc(AUC_ROC_mean)) |>
    knitr::kable(digits = 3,
        caption = "Spatial CV within each temporal fold (mean ± SD across 5 folds)")
```

### ROC curves

```{r fig-roc-curves, fig.width=12, fig.height=4}
# Re-generate predictions for ROC plotting (folds 2-3 temporal)
roc_plot_data <- map_dfr(2:3, function(i) {
    train <- bind_rows(fold_data[1:(i - 1)])
    test <- fold_data[[i]]
    preds <- fit_and_predict(train, test)
    preds |> mutate(fold = i)
})

roc_plots <- map(2:3, function(i) {
    d <- roc_plot_data |> filter(fold == i)
    roc_list <- map(MODEL_NAMES, function(m) {
        pROC::roc(d$obs, d[[m]], quiet = TRUE)
    })
    names(roc_list) <- MODEL_NAMES
    ggroc(roc_list, legacy.axes = TRUE) +
        labs(title = sprintf("Fold %d (outcome %d)", i, 2014 + i),
             x = "1 - Specificity", y = "Sensitivity", colour = "Model") +
        theme_minimal(base_size = 10) +
        theme(legend.position = if (i == 3) "right" else "none")
})
wrap_plots(roc_plots, nrow = 1)
```

### Variable importance

```{r fig-variable-importance, fig.width=12, fig.height=6}
# Train final models on all fold data for variable importance
all_train <- bind_rows(fold_data)

# Random Forest importance
set.seed(42)
rf_final <- ranger(
    obs ~ ., data = all_train[, c("obs", ALL_FEATURES)] |> mutate(obs = factor(obs)),
    num.trees = 500, probability = TRUE, importance = "permutation",
    min.node.size = 5
)
rf_imp <- tibble(feature = names(rf_final$variable.importance),
                 importance = rf_final$variable.importance) |>
    arrange(desc(importance))

# XGBoost importance
X_all <- as.matrix(all_train[, ALL_FEATURES])
y_all <- all_train$obs
dtrain_all <- xgb.DMatrix(data = X_all, label = y_all)
xgb_cv_all <- xgb.cv(
    params = list(objective = "binary:logistic", eval_metric = "auc",
                  max_depth = 4, eta = 0.1, subsample = 0.8,
                  colsample_bytree = 0.8),
    data = dtrain_all, nrounds = 300, nfold = 5,
    early_stopping_rounds = 20, verbose = 0
)
xgb_final <- xgb.train(
    params = list(objective = "binary:logistic", max_depth = 4, eta = 0.1,
                  subsample = 0.8, colsample_bytree = 0.8),
    data = dtrain_all,
    nrounds = if (is.null(xgb_cv_all$best_iteration) || xgb_cv_all$best_iteration == 0)
        50 else xgb_cv_all$best_iteration,
    verbose = 0
)
xgb_imp_raw <- xgb.importance(model = xgb_final)
xgb_imp <- tibble(feature = xgb_imp_raw$Feature, importance = xgb_imp_raw$Gain)

p_rf <- ggplot(rf_imp |> head(15),
               aes(x = reorder(feature, importance), y = importance)) +
    geom_col(fill = "steelblue") + coord_flip() +
    labs(title = "Random Forest", x = NULL, y = "Permutation importance") +
    theme_minimal(base_size = 10)

p_xgb <- ggplot(xgb_imp |> head(15),
                aes(x = reorder(feature, importance), y = importance)) +
    geom_col(fill = "darkorange") + coord_flip() +
    labs(title = "XGBoost", x = NULL, y = "Gain importance") +
    theme_minimal(base_size = 10)

p_rf + p_xgb + plot_annotation(title = "Variable importance (trained on all folds)")
```

### Calibration plots

```{r fig-calibration, fig.width=14, fig.height=4}
# Calibration: predicted prob vs observed proportion (decile bins)
calibration_data <- roc_plot_data |>
    pivot_longer(cols = all_of(MODEL_NAMES), names_to = "model",
                 values_to = "predicted") |>
    mutate(bin = ntile(predicted, 10)) |>
    group_by(model, fold, bin) |>
    summarise(
        pred_mean = mean(predicted),
        obs_mean = mean(obs),
        n = n(), .groups = "drop"
    )

ggplot(calibration_data, aes(x = pred_mean, y = obs_mean, colour = model)) +
    geom_point(size = 2) +
    geom_line() +
    geom_abline(linetype = "dashed", colour = "grey50") +
    facet_wrap(~ paste("Fold", fold), nrow = 1) +
    labs(x = "Mean predicted probability", y = "Observed proportion",
         title = "Calibration plots (decile bins)", colour = "Model") +
    theme_minimal(base_size = 10) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1))
```

### Final model: train on all data, predict 2018

```{r final-model}
# Train on all 3 folds (2010-2014→2015, 2010-2015→2016, 2010-2016→2017)
all_train <- bind_rows(fold_data)

# Final predictions for 2018 using features from 2010-2017
final_preds <- fit_and_predict(all_train, final_features)

# Attach predictions to final_features for downstream use
pred_data <- final_features |>
    left_join(final_preds |> dplyr::select(location, all_of(MODEL_NAMES)),
              by = "location")

# Map model names to strategy names for downstream pipeline
pred_data <- pred_data |>
    rename(
        pred_prob = Logistic,
        pred_rf   = `Random Forest`,
        pred_gam  = GAM
    )
# Keep Elastic Net and XGBoost as additional columns
pred_data <- pred_data |>
    rename(pred_enet = `Elastic Net`, pred_xgb = XGBoost)

# Also store hist_count for backward compatibility (Count-ranked strategy)
# hist_count = n_outbreak_years in the new feature set
pred_data <- pred_data |>
    mutate(hist_count = n_outbreak_years)

cat("Final predictions for 2018: n =", nrow(pred_data), "locations\n")
cat("Prevalence (2018):", round(100 * mean(final_preds$obs), 1), "%\n")
```

## Save Predictions

Save model predictions and supporting data for use in `vaccine_impact_sim.qmd`.

```{r save-predictions}
# Pre-compute pop_lookup so vaccine_impact_sim.qmd doesn't need shapefiles
pop_lookup <- shp_admin23 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

saveRDS(
    list(
        pred_data     = pred_data,
        loc_year_full = loc_year_full,
        shp_admin23   = shp_admin23,
        nbr_list      = nbr_list,
        pop_lookup    = pop_lookup
    ),
    "data/outbreak_predictions.rds"
)
cat("Saved data/outbreak_predictions.rds\n")
cat("  pred_data:     ", nrow(pred_data), "locations\n")
cat("  loc_year_full: ", nrow(loc_year_full), "rows\n")
cat("  shp_admin23:   ", nrow(shp_admin23), "polygons\n")
cat("  nbr_list:      ", length(nbr_list), "entries\n")
```
