---
title: "Annual Outbreak Prediction Model"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document develops an annual outbreak prediction model for admin-2 units
using cholera surveillance data from 2010--2023. Models are evaluated via
**rolling-window temporal cross-validation** (train on fold i−1, test on fold i)
and **spatial cross-validation** (5 random spatial partitions within each fold).
Two outcome horizons are compared:

- **k=1**: predict any outbreak in the next single year (9 validation folds)
- **k=3**: predict any outbreak in the next 3-year window (7 validation folds)

**k=1 design (T=5 rolling feature window)**:

| Fold  | Feature years | Outcome year | Purpose    |
|-------|--------------|--------------|------------|
| 1     | 2010--2014   | 2015         | Validation |
| 2     | 2011--2015   | 2016         | Validation |
| 3     | 2012--2016   | 2017         | Validation |
| 4     | 2013--2017   | 2018         | Validation |
| 5     | 2014--2018   | 2019         | Validation |
| 6     | 2015--2019   | 2020         | Validation |
| 7     | 2016--2020   | 2021         | Validation |
| 8     | 2017--2021   | 2022         | Validation |
| 9     | 2018--2022   | 2023         | Validation |
| Final | 2019--2023   | 2024         | Prediction |

**k=3 design (T=5 rolling feature window)**:

| Fold  | Feature years | Outcome window   | Purpose    |
|-------|--------------|-----------------|------------|
| 1     | 2010--2014   | 2015--2017       | Validation |
| 2     | 2011--2015   | 2016--2018       | Validation |
| 3     | 2012--2016   | 2017--2019       | Validation |
| 4     | 2013--2017   | 2018--2020       | Validation |
| 5     | 2014--2018   | 2019--2021       | Validation |
| 6     | 2015--2019   | 2020--2022       | Validation |
| 7     | 2016--2020   | 2021--2023       | Validation |
| Final | 2017--2021   | 2022--2024       | Prediction |

Five classifiers are compared: logistic regression, elastic net, random forest, XGBoost, and GAM.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(patchwork)
library(pROC)
library(scales)
library(knitr)
library(mgcv)         # GAM
library(glmnet)       # elastic net / LASSO / ridge
library(ranger)       # fast random forest
library(xgboost)      # gradient boosted trees
library(PRROC)        # precision-recall AUC
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```


## Data Loading

### Outbreak surveillance data

```{r load-outbreak-data}
# Raw surveillance data: each row = (location, time interval TL-TR, outbreak_number)
# Location format: "REGION::COUNTRY::ADMIN1::ADMIN2" (3 separators = admin2)
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)
# Keep admin2 (adm=2) and admin3 (adm=3) only
ts23 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm %in% c(2, 3))

# Collapse time-series rows into one row per outbreak event, extracting
# start/end years, duration, case counts, and a unique ID
outbreak_events <- ts23 |>
    group_by(location, adm, outbreak_number) |>
    summarise(
        year_start      = year(min(TL)),
        year_end        = year(max(TR)),
        duration_weeks  = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
        total_cases     = sum(sCh, na.rm = TRUE),
        total_deaths    = sum(deaths, na.rm = TRUE),
        peak_weekly_sCh = max(sCh, na.rm = TRUE),
        id_outbreak     = paste(first(location), min(TL), max(TR), sep = "_"),
        .groups         = "drop"
    )

all_years <- 2010:2023
all_locs <- unique(ts23$location)

# Build a yearly outbreak summary per location: for each (location, year),
# aggregate outbreak counts, total cases, maximum duration, etc.
# Assigns each outbreak to its END year (observable after conclusion).
outbreak_year_summary <- outbreak_events |>
    filter(year_end %in% all_years) |>
    group_by(location, year = year_end) |>
    summarise(
        outbreak_count     = n(),
        total_cases        = sum(total_cases, na.rm = TRUE),
        max_duration_days  = max(duration_weeks * 7, na.rm = TRUE),
        mean_duration_days = mean(duration_weeks * 7, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(outbreak_count = 0L, total_cases = 0,
                    max_duration_days = 0, mean_duration_days = 0))

# loc_year_full: complete (location × year) panel with n_active = number of
# outbreaks ACTIVE (ongoing) in that year. Used for the outcome variable.
loc_year <- outbreak_events %>%
    mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) %>%
    unnest(years) %>%
    rename(year = years) %>%
    filter(year %in% all_years) %>%
    count(location, year, name = "n_active")

loc_year_full <- expand_grid(location = all_locs, year = all_years) %>%
    left_join(loc_year, by = c("location", "year")) %>%
    replace_na(list(n_active = 0L))

# loc_year_end_full: complete (location × year) panel with n_ended = number of
# outbreaks that ENDED in that year. Used for feature construction (hist_count)
# because an outbreak's end year marks when it became observable.
loc_year_end_full <- outbreak_events %>%
    filter(year_end %in% all_years) %>%
    count(location, year_end, name = "n_ended") %>%
    rename(year = year_end) %>%
    right_join(expand_grid(location = all_locs, year = all_years),
        by = c("location", "year")
    ) %>%
    replace_na(list(n_ended = 0L))

cat("Surveillance locations:", length(all_locs), "\n")
cat("Year range:", range(all_years), "\n")
cat("Total outbreak events:", nrow(outbreak_events), "\n")
cat("Total outbreak-end-years:", sum(loc_year_end_full$n_ended), "\n")
```

### Shapefiles and spatial neighbours

```{r load-shapefiles}
# Shapefiles are keyed by location_period_id; we need to map them to
# the human-readable location strings (e.g. "AFR::COD::Kinshasa::...").
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Build a lookup: for each location, keep the most recent location_period_id
# (a location can have multiple period IDs if boundaries changed over time)
loc_id_map <- ts23 %>%
    distinct(location_period_id, location) %>%
    group_by(location) %>%
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) %>%
    ungroup()

# Join shapefile geometries to location names via the ID map
shp_named <- shp_all %>%
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id"))

# Keep only admin2/3 polygons (3 or 4 "::" separators in location string)
shp_admin23 <- shp_named %>%
    filter(!is.na(location), str_count(location, "::") %in% c(3, 4))

cat("ADM2/3 polygons:", nrow(shp_admin23), "\n")

# Build a spatial neighbour list: for each polygon, find all neighbours
# whose centroid is within DIST_LIMIT_KM km (geodesic distance).
# Used for the nbr_hist feature (spatial contagion signal).
DIST_LIMIT_KM <- 50
centroids_sf <- shp_admin23 %>%
    st_make_valid() %>%
    st_point_on_surface()

# Pairwise geodesic distance matrix (meters); convert to km
dist_mat_sp <- st_distance(centroids_sf) |> units::drop_units()
dist_mat_sp <- dist_mat_sp / 1000 # m -> km
rownames(dist_mat_sp) <- colnames(dist_mat_sp) <- shp_admin23$location
diag(dist_mat_sp) <- Inf

# For each location, store neighbours within the distance limit (sorted)
nbr_list <- setNames(
    lapply(shp_admin23$location, function(loc) {
        d <- dist_mat_sp[loc, ]
        names(sort(d[d <= DIST_LIMIT_KM]))
    }),
    shp_admin23$location
)

# Extract centroid coordinates for GAM spatial smooth
centroid_coords <- centroids_sf |>
    st_coordinates() |>
    as_tibble() |>
    rename(lon = X, lat = Y) |>
    mutate(location = shp_admin23$location)
```

### Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)  # convert RasterLayer → SpatRaster
cat("Raster dimensions:", dim(pop_raster), "\n")
cat("Resolution:", res(pop_raster), "degrees (~20 km)\n")
```


## Population Extraction

Aggregate the 20 km population raster to ADM2 polygons via zonal statistics.

The raster stores **mean population per ~1 km² source cell**, having been
aggregated from the WorldPop ~1 km PPP (People Per Pixel) product to 20 km
resolution by averaging. Each 20 km pixel spans approximately
$(0.167° \times 111\;\text{km/°})^2 \approx 344\;\text{km}^2$ at the equator,
i.e.\ roughly 344 underlying 1 km² source cells. Because `exact_extract(fun =
"sum")` sums these per-km² means, the result must be multiplied by the pixel
area (~344 km²) to recover actual population counts. Verification: the
continent-wide raster sum ($\approx$ 3.65 M) $\times$ 344 $\approx$ 1.26
billion, matching Africa's 2017 population.

```{r population-extraction}
POP_SCALE <- 344 # pixel area in km² ≈ (0.167° × 111 km)²

shp_pop <- shp_admin23 |>
    st_make_valid() |>
    st_transform(crs(pop_raster))

# Sum mean-population values within each polygon, then scale to actual counts
pop_by_adm2 <- exactextractr::exact_extract(pop_raster, shp_pop,
    fun = "sum",
    progress = FALSE
)
pop_by_adm2 <- pop_by_adm2 * POP_SCALE

shp_admin23$pop_sum <- pop_by_adm2

cat(
    "Polygons with population > 0:",
    sum(!is.na(pop_by_adm2) & pop_by_adm2 > 0), "of", nrow(shp_admin23), "\n"
)
cat("Total population:", format(round(sum(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n")
cat(
    "Range:", format(round(min(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "-",
    format(round(max(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n"
)
```


## Annual Outbreak Prediction Model

### Overview

We predict outbreak occurrence ahead for admin-2 units using a **rolling-window**
design. Only outbreak surveillance data are used (no environmental covariates).
Data cover 2010--2023. Two outcome horizons are evaluated: **k=1** (next year)
and **k=3** (any outbreak in the next 3 years). See the Overview section for
full fold tables. **Rolling training**: each fold trained on fold i−1 only.
Five classifiers: logistic regression, elastic net, random forest, XGBoost, GAM.

### Feature engineering

```{r build-features}
T_WINDOW <- 5L  # number of feature years (rolling window)

# Build features for a given feature window ending at year_end.
# Uses the most recent T_WINDOW years only (e.g. T=5: 2013-2017 for 2018).
# Returns one row per location with same-unit and neighbor features.
build_features <- function(year_end, T = T_WINDOW, nbr_map = nbr_list,
                           coord_lookup = centroid_coords) {
    feat_yrs <- (year_end - T + 1L):year_end
    n_yrs <- length(feat_yrs)

    # Per-location summary across the feature window
    loc_summary <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            n_outbreak_years    = sum(outbreak_count > 0),
            prop_outbreak_years = n_outbreak_years / n_yrs,
            total_outbreaks     = sum(outbreak_count),
            mean_cases_per_year = mean(total_cases),
            max_cases_any_year  = max(total_cases),
            mean_duration       = mean(mean_duration_days),
            max_duration        = max(max_duration_days),
            .groups = "drop"
        )

    # Recent-year features
    recent_1yr <- outbreak_year_summary |>
        filter(year == year_end) |>
        transmute(location, recent_outbreak = as.integer(outbreak_count > 0))

    recent_2yr_yrs <- intersect(c(year_end - 1L, year_end), feat_yrs)
    recent_2yr <- outbreak_year_summary |>
        filter(year %in% recent_2yr_yrs) |>
        group_by(location) |>
        summarise(recent_2yr_outbreaks = sum(outbreak_count > 0), .groups = "drop")

    # Trend features: slope of outbreak occurrence and cases on year
    trend_df <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            trend = if (n() >= 2)
                coef(lm(as.integer(outbreak_count > 0) ~ year))[2] else 0,
            trend_cases = if (n() >= 2)
                coef(lm(total_cases ~ year))[2] else 0,
            .groups = "drop"
        )

    # Years since last outbreak
    last_ob <- outbreak_year_summary |>
        filter(year %in% feat_yrs, outbreak_count > 0) |>
        group_by(location) |>
        summarise(last_ob_year = max(year), .groups = "drop")

    yrs_since <- tibble(location = all_locs) |>
        left_join(last_ob, by = "location") |>
        mutate(years_since_last_outbreak = ifelse(
            is.na(last_ob_year), n_yrs + 1L, as.integer(year_end - last_ob_year)
        )) |>
        dplyr::select(location, years_since_last_outbreak)

    # Combine same-unit features
    feat <- tibble(location = all_locs) |>
        left_join(loc_summary, by = "location") |>
        left_join(recent_1yr, by = "location") |>
        left_join(recent_2yr, by = "location") |>
        left_join(trend_df, by = "location") |>
        left_join(yrs_since, by = "location") |>
        replace_na(list(
            n_outbreak_years = 0L, prop_outbreak_years = 0,
            total_outbreaks = 0L, mean_cases_per_year = 0,
            max_cases_any_year = 0, mean_duration = 0, max_duration = 0,
            recent_outbreak = 0L, recent_2yr_outbreaks = 0L,
            trend = 0, trend_cases = 0
        ))

    # Neighbor features: helper to compute a summary stat across neighbours
    make_nbr_feat <- function(values, locs, nbr_map, fun = mean) {
        sv <- setNames(values, locs)
        map_dbl(locs, function(loc) {
            nbrs <- nbr_map[[loc]]
            if (is.null(nbrs) || length(nbrs) == 0L) return(0)
            nbrs <- nbrs[nbrs %in% names(sv)]
            if (length(nbrs) == 0L) return(0)
            fun(sv[nbrs])
        })
    }

    feat <- feat |>
        mutate(
            nbr_prop_outbreak = make_nbr_feat(
                prop_outbreak_years, location, nbr_map),
            nbr_mean_cases = make_nbr_feat(
                mean_cases_per_year, location, nbr_map),
            nbr_recent_outbreak_prop = make_nbr_feat(
                recent_outbreak, location, nbr_map),
            nbr_recent_2yr_prop = make_nbr_feat(
                as.numeric(recent_2yr_outbreaks > 0), location, nbr_map),
            nbr_max_cases = make_nbr_feat(
                max_cases_any_year, location, nbr_map, fun = max),
            nbr_trend = make_nbr_feat(trend, location, nbr_map)
        )

    if (!is.null(coord_lookup)) {
        feat <- feat |> left_join(coord_lookup, by = "location")
    }
    feat
}

# Define feature sets
SAME_UNIT_FEATURES <- c(
    "n_outbreak_years", "prop_outbreak_years", "total_outbreaks",
    "mean_cases_per_year", "max_cases_any_year",
    "mean_duration", "max_duration",
    "recent_outbreak", "recent_2yr_outbreaks",
    "trend", "trend_cases", "years_since_last_outbreak"
)
NBR_FEATURES <- c(
    "nbr_prop_outbreak", "nbr_mean_cases",
    "nbr_recent_outbreak_prop", "nbr_recent_2yr_prop",
    "nbr_max_cases", "nbr_trend"
)
ALL_FEATURES <- c(SAME_UNIT_FEATURES, NBR_FEATURES)
cat("Features:", length(ALL_FEATURES),
    "(", length(SAME_UNIT_FEATURES), "same-unit +",
    length(NBR_FEATURES), "neighbour )\n")
```

### Model fitting helpers

```{r model-helpers}
# Fit all 5 models on training data, predict on test data.
# Returns a tibble with location, obs, and prediction columns.
fit_and_predict <- function(train, test, features = ALL_FEATURES) {
    X_train <- as.matrix(train[, features])
    y_train <- train$obs
    X_test <- as.matrix(test[, features])
    preds <- tibble(location = test$location, obs = test$obs)

    # 1. Logistic regression
    lr_fit <- glm(obs ~ ., data = train[, c("obs", features)], family = binomial)
    preds$Logistic <- predict(lr_fit, newdata = test, type = "response")

    # 2. Elastic net (choose best alpha from {0, 0.5, 1} via internal CV)
    best_enet <- NULL; best_auc <- -Inf
    for (a in c(0, 0.5, 1)) {
        cv_fit <- tryCatch(
            cv.glmnet(X_train, y_train, family = "binomial",
                      alpha = a, nfolds = 5, type.measure = "auc"),
            error = function(e) NULL
        )
        if (!is.null(cv_fit) && max(cv_fit$cvm) > best_auc) {
            best_auc <- max(cv_fit$cvm); best_enet <- cv_fit
        }
    }
    preds$`Elastic Net` <- if (!is.null(best_enet)) {
        as.numeric(predict(best_enet, newx = X_test, s = "lambda.min", type = "response"))
    } else {
        # Fallback to logistic regression when cv.glmnet fails (e.g., tiny folds)
        as.numeric(predict(lr_fit, newdata = test, type = "response"))
    }

    # 3. Random forest (ranger)
    set.seed(42)
    rf_fit <- ranger(
        obs ~ ., data = train[, c("obs", features)] |> mutate(obs = factor(obs)),
        num.trees = 500, probability = TRUE,
        min.node.size = 5, mtry = max(1, floor(sqrt(length(features))))
    )
    rf_preds <- predict(rf_fit, data = test)$predictions
    preds$`Random Forest` <- if ("1" %in% colnames(rf_preds)) rf_preds[, "1"] else rep(0, nrow(test))

    # 4. XGBoost
    set.seed(42)
    dtrain <- xgb.DMatrix(data = X_train, label = y_train)
    dtest <- xgb.DMatrix(data = X_test)
    xgb_params <- list(objective = "binary:logistic", eval_metric = "auc",
                       max_depth = 4, eta = 0.1, subsample = 0.8,
                       colsample_bytree = 0.8)
    xgb_cv <- tryCatch(
        xgb.cv(params = xgb_params, data = dtrain, nrounds = 300,
               nfold = min(5, nrow(train)),
               early_stopping_rounds = 20, verbose = 0),
        error = function(e) NULL
    )
    best_nrounds <- if (!is.null(xgb_cv)) xgb_cv$best_iteration else NULL
    if (is.null(best_nrounds) || is.na(best_nrounds) || best_nrounds == 0) best_nrounds <- 50
    xgb_fit <- xgb.train(params = xgb_params, data = dtrain,
                         nrounds = best_nrounds, verbose = 0)
    preds$XGBoost <- predict(xgb_fit, dtest)

    # 5. GAM with smooth terms on key continuous features
    # k is capped at (unique values - 1) to avoid "too few unique values" errors
    gam_feats <- c("n_outbreak_years", "mean_cases_per_year",
                   "recent_2yr_outbreaks", "years_since_last_outbreak",
                   "nbr_prop_outbreak", "nbr_mean_cases")
    safe_k <- function(feat, dat, kmax = 5L)
        max(3L, min(kmax, length(unique(dat[[feat]])) - 1L))
    gam_terms <- paste0("s(", gam_feats, ", k = ",
                        sapply(gam_feats, safe_k, dat = train), ")",
                        collapse = " + ")
    gam_fit <- tryCatch(
        gam(as.formula(paste("obs ~", gam_terms)), data = train,
            family = binomial, method = "REML"),
        error = function(e) {
            k2 <- safe_k("n_outbreak_years", train, 3L)
            k3 <- safe_k("nbr_prop_outbreak", train, 3L)
            tryCatch(
                gam(as.formula(sprintf("obs ~ s(n_outbreak_years, k=%d) + s(nbr_prop_outbreak, k=%d)", k2, k3)),
                    data = train, family = binomial, method = "REML"),
                error = function(e2)
                    glm(obs ~ n_outbreak_years + nbr_prop_outbreak,
                        data = train, family = binomial)
            )
        }
    )
    preds$GAM <- as.numeric(predict(gam_fit, newdata = test, type = "response"))

    preds
}

MODEL_NAMES <- c("Logistic", "Elastic Net", "Random Forest", "XGBoost", "GAM")

# Compute evaluation metrics for a single (obs, pred) pair
eval_metrics <- function(obs, pred) {
    roc_obj <- pROC::roc(obs, pred, quiet = TRUE)
    auc_roc <- as.numeric(pROC::auc(roc_obj))
    pr_obj <- PRROC::pr.curve(scores.class0 = pred[obs == 1],
                              scores.class1 = pred[obs == 0])
    auc_pr <- pr_obj$auc.integral
    brier <- mean((pred - obs)^2)
    rho   <- cor(pred, obs, method = "spearman")
    coords <- pROC::coords(roc_obj, "best", best.method = "youden", ret = "all")
    tibble(rho = rho, AUC_ROC = auc_roc, AUC_PR = auc_pr, Brier = brier,
           Sensitivity = coords$sensitivity[1],
           Specificity = coords$specificity[1],
           Threshold = coords$threshold[1])
}
```

### Rolling-window temporal cross-validation

```{r temporal-cv}
K_VALUES <- c(1L, 3L)

# Build fold table for a given k: train_end runs from 2014 to
# max(all_years) - k (last fold whose outcome is fully in the data).
# Final prediction uses train_end = max(all_years) - k + 1 (one step ahead,
# so outcome window extends one year beyond the data).
make_temporal_folds <- function(k) {
    train_ends <- 2014:(max(all_years) - k)
    tibble(
        fold          = seq_along(train_ends),
        k             = k,
        train_end     = train_ends,
        outcome_start = train_ends + 1L,
        outcome_end   = train_ends + k
    )
}

temporal_folds_list <- setNames(map(K_VALUES, make_temporal_folds),
                                paste0("k", K_VALUES))

# Build feature + outcome data for every fold of every k
make_fold_data <- function(tf) {
    fd <- map(1:nrow(tf), function(i) {
        fe <- tf[i, ]
        feat <- build_features(fe$train_end)
        outcome <- loc_year_full |>
            filter(year %in% (fe$outcome_start:fe$outcome_end)) |>
            group_by(location) |>
            summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")
        feat |>
            left_join(outcome, by = "location") |>
            replace_na(list(obs = 0L))
    })
    names(fd) <- paste0("fold", seq_along(fd))
    fd
}

fold_data_list <- setNames(map(temporal_folds_list, make_fold_data),
                           paste0("k", K_VALUES))

# Print fold summaries
for (k in K_VALUES) {
    tf <- temporal_folds_list[[paste0("k", k)]]
    cat(sprintf("\n--- k = %d (%d folds) ---\n", k, nrow(tf)))
    for (i in 1:nrow(tf)) {
        fe <- tf[i, ]
        d  <- fold_data_list[[paste0("k", k)]][[i]]
        cat(sprintf(
            "  Fold %d: features [%d,%d] → outcome [%d,%d] | n=%d, prev=%.1f%%\n",
            i, fe$train_end - T_WINDOW + 1L, fe$train_end,
            fe$outcome_start, fe$outcome_end, nrow(d), 100 * mean(d$obs)
        ))
    }
}

# Backward-compatible aliases for k=1 (used by spatial-cv and variable-importance)
temporal_folds <- temporal_folds_list$k1
fold_data      <- fold_data_list$k1
N_FOLDS        <- nrow(temporal_folds)   # 9 folds for k=1

# Final-prediction feature sets (one T-window beyond the last complete fold)
final_features_list <- setNames(
    map(K_VALUES, function(k) build_features(max(all_years) - k + 1L)),
    paste0("k", K_VALUES)
)
# k=1: build_features(2023) → feat_yrs 2019-2023 → predict 2024
# k=3: build_features(2021) → feat_yrs 2017-2021 → predict 2022-2024
final_features <- final_features_list$k1
```

### Temporal fold evaluation

For each fold i (2–9), the model is trained on **fold i−1 only** (rolling
window) and tested on fold i. This ensures the training data always comes from
the immediately preceding year's T=5 feature window, which is most analogous to
how the model would be used in practice. Fold 1 has no preceding fold, so it is
evaluated via spatial CV only (see next section).

Predictions are computed once here and reused for ROC curves and calibration
plots below.

```{r temporal-predictions}
# Rolling training for every k: train on fold i-1, test on fold i.
# Results used by temporal-eval, ROC, and calibration chunks.
temporal_pred_list_all <- setNames(map(K_VALUES, function(kv) {
    fd <- fold_data_list[[paste0("k", kv)]]
    nf <- length(fd)
    preds <- map(2:nf, function(i) {
        fit_and_predict(fd[[i - 1]], fd[[i]]) |> mutate(fold = i, k = kv)
    })
    names(preds) <- paste0("fold", 2:nf)
    preds
}), paste0("k", K_VALUES))

temporal_pred_df_all <- bind_rows(
    map(K_VALUES, function(kv) bind_rows(temporal_pred_list_all[[paste0("k", kv)]]))
)

# Backward-compatible aliases for k=1
temporal_pred_list <- temporal_pred_list_all$k1
temporal_pred_df   <- temporal_pred_df_all |> filter(k == 1L)
```

```{r temporal-eval}
temporal_results <- map_dfr(K_VALUES, function(kv) {
    tf <- temporal_folds_list[[paste0("k", kv)]]
    preds_k <- temporal_pred_list_all[[paste0("k", kv)]]
    nf <- nrow(tf)
    map_dfr(2:nf, function(i) {
        preds <- preds_k[[paste0("fold", i)]]
        fe <- tf[i, ]
        map_dfr(MODEL_NAMES, function(m) {
            eval_metrics(preds$obs, preds[[m]]) |>
                mutate(model = m, fold = i, k = kv,
                       outcome = if (kv == 1L) as.character(fe$outcome_start)
                                 else sprintf("%d-%d", fe$outcome_start, fe$outcome_end))
        })
    })
})

temporal_results |>
    dplyr::select(k, fold, outcome, model, rho, AUC_ROC, AUC_PR, Brier,
                  Sensitivity, Specificity) |>
    arrange(k, fold, desc(AUC_ROC)) |>
    knitr::kable(digits = 3,
        caption = "Rolling temporal CV: train on fold i-1, test on fold i (k=1 and k=3)")
```

### Spatial cross-validation within each temporal fold

```{r spatial-cv}
set.seed(42)

spatial_cv_results <- map_dfr(1:N_FOLDS, function(i) {
    d <- fold_data[[i]]
    n_folds_sp <- 5

    # Assign spatial folds (random partition of locations)
    locs <- unique(d$location)
    sp_fold_ids <- sample(rep(1:n_folds_sp, length.out = length(locs)))
    loc_folds <- tibble(location = locs, sp_fold = sp_fold_ids)
    d <- d |> left_join(loc_folds, by = "location")

    map_dfr(1:n_folds_sp, function(sf) {
        train_sp <- d |> filter(sp_fold != sf)
        test_sp <- d |> filter(sp_fold == sf)

        preds <- fit_and_predict(train_sp, test_sp)

        map_dfr(MODEL_NAMES, function(m) {
            metrics <- eval_metrics(preds$obs, preds[[m]])
            metrics |> mutate(model = m, temporal_fold = i, spatial_fold = sf)
        })
    })
})

# Summary: mean across spatial folds, per temporal fold and model
spatial_cv_summary <- spatial_cv_results |>
    group_by(temporal_fold, model) |>
    summarise(across(c(rho, AUC_ROC, AUC_PR, Brier, Sensitivity, Specificity),
                     list(mean = mean, sd = sd)), .groups = "drop")

spatial_cv_summary |>
    dplyr::select(temporal_fold, model, rho_mean, rho_sd,
                  AUC_ROC_mean, AUC_ROC_sd, AUC_PR_mean, Brier_mean) |>
    arrange(temporal_fold, desc(AUC_ROC_mean)) |>
    knitr::kable(digits = 3,
        caption = "Spatial CV within each temporal fold (mean ± SD across 5 folds)")
```

### ROC curves

```{r fig-roc-curves, fig.width=16, fig.height=14}
#| fig-cap: "ROC curves for all temporal folds (rolling CV: train on fold i−1,
#|   test on fold i). Top rows: k=1 (predict one year ahead, folds 2–9).
#|   Bottom rows: k=3 (predict any outbreak over a 3-year window, folds 2–7).
#|   Each staircase step corresponds to one decision threshold."
make_roc_plots <- function(kv) {
    tf <- temporal_folds_list[[paste0("k", kv)]]
    preds_k <- temporal_pred_list_all[[paste0("k", kv)]]
    nf <- nrow(tf)
    fold_ids <- 2:nf
    roc_df <- map_dfr(fold_ids, function(i) {
        d <- preds_k[[paste0("fold", i)]]
        map_dfr(MODEL_NAMES, function(m) {
            roc_obj <- pROC::roc(d$obs, d[[m]], quiet = TRUE)
            fe <- tf[i, ]
            tibble(
                fpr     = 1 - roc_obj$specificities,
                tpr     = roc_obj$sensitivities,
                model   = m,
                fold    = i,
                outcome = if (kv == 1L) as.character(fe$outcome_start)
                          else sprintf("%d-%d", fe$outcome_start, fe$outcome_end),
                auc     = as.numeric(pROC::auc(roc_obj))
            )
        })
    })
    map(fold_ids, function(i) {
        d <- roc_df |> filter(fold == i)
        out_lbl <- d$outcome[1]
        ggplot(d, aes(x = fpr, y = tpr, colour = model)) +
            geom_step(direction = "hv", linewidth = 0.5) +
            geom_abline(linetype = "dashed", colour = "grey60") +
            labs(title = sprintf("k=%d fold %d\n(outcome %s)", kv, i, out_lbl),
                 x = "1 - Specificity", y = "Sensitivity", colour = "Model") +
            coord_equal() +
            theme_minimal(base_size = 9) +
            theme(plot.title = element_text(size = 8))
    })
}

plots_k1 <- make_roc_plots(1L)   # 8 panels (folds 2-9)
plots_k3 <- make_roc_plots(3L)   # 6 panels (folds 2-7)

(wrap_plots(plots_k1, nrow = 2) + plot_layout(guides = "collect") &
    theme(legend.position = "bottom")) /
(wrap_plots(plots_k3, nrow = 2) + plot_layout(guides = "collect") &
    theme(legend.position = "bottom"))
```

### Variable importance

```{r compute-variable-importance}
# Train RF and XGBoost on each temporal fold independently; extract importance.
# RF: permutation importance. XGBoost: gain.
# Importance is normalised to [0,1] within each fold × model.
imp_by_fold <- map_dfr(1:N_FOLDS, function(i) {
    d <- fold_data[[i]]
    fe <- temporal_folds[i, ]
    fold_label <- sprintf("Fold %d (%d-%d\u2192%d)",
                          i, fe$train_end - T_WINDOW + 1L, fe$train_end,
                          fe$outcome_start)

    # Random Forest
    set.seed(42)
    rf_fit <- ranger(
        obs ~ ., data = d[, c("obs", ALL_FEATURES)] |> mutate(obs = factor(obs)),
        num.trees = 500, probability = TRUE, importance = "permutation",
        min.node.size = 5
    )
    rf_imp <- tibble(
        feature    = names(rf_fit$variable.importance),
        importance = rf_fit$variable.importance,
        model      = "Random Forest",
        fold       = fold_label,
        fold_num   = i
    )

    # XGBoost
    X <- as.matrix(d[, ALL_FEATURES])
    y <- d$obs
    dtrain <- xgb.DMatrix(data = X, label = y)
    xgb_cv_f <- tryCatch(
        xgb.cv(
            params = list(objective = "binary:logistic", eval_metric = "auc",
                          max_depth = 4, eta = 0.1, subsample = 0.8,
                          colsample_bytree = 0.8),
            data = dtrain, nrounds = 300,
            nfold = min(5, nrow(d)),
            early_stopping_rounds = 20, verbose = 0
        ),
        error = function(e) NULL
    )
    best_nr_f <- if (!is.null(xgb_cv_f)) xgb_cv_f$best_iteration else NULL
    if (is.null(best_nr_f) || is.na(best_nr_f) || best_nr_f == 0) best_nr_f <- 50
    xgb_fit <- xgb.train(
        params = list(objective = "binary:logistic", max_depth = 4, eta = 0.1,
                      subsample = 0.8, colsample_bytree = 0.8),
        data = dtrain,
        nrounds = best_nr_f,
        verbose = 0
    )
    xgb_raw <- xgb.importance(model = xgb_fit)
    xgb_imp <- tibble(
        feature    = xgb_raw$Feature,
        importance = xgb_raw$Gain,
        model      = "XGBoost",
        fold       = fold_label,
        fold_num   = i
    )

    bind_rows(rf_imp, xgb_imp)
}) |>
    # Normalise to [0,1] within each fold × model
    group_by(fold, model) |>
    mutate(importance = importance / max(importance, na.rm = TRUE)) |>
    ungroup()

# Feature order: by mean importance across all folds and models
feat_order <- imp_by_fold |>
    group_by(feature) |>
    summarise(mean_imp = mean(importance, na.rm = TRUE), .groups = "drop") |>
    arrange(mean_imp) |>
    pull(feature)

imp_by_fold <- imp_by_fold |>
    mutate(feature = factor(feature, levels = feat_order))
```

#### Aggregated across all temporal folds

```{r fig-variable-importance-agg, fig.width=10, fig.height=6}
#| fig-cap: "Variable importance aggregated across all nine temporal folds (k=1).
#|   Points show the mean normalised importance; error bars show ±1 SD across folds.
#|   Features are ordered by mean importance across both models. Random Forest
#|   uses permutation importance; XGBoost uses gain."
imp_summary <- imp_by_fold |>
    group_by(feature, model) |>
    summarise(mean_imp = mean(importance), sd_imp = sd(importance),
              .groups = "drop")

ggplot(imp_summary, aes(x = mean_imp, y = feature, colour = model)) +
    geom_errorbarh(aes(xmin = pmax(mean_imp - sd_imp, 0),
                       xmax = mean_imp + sd_imp),
                   height = 0.3, position = position_dodge(0.6)) +
    geom_point(size = 2.5, position = position_dodge(0.6)) +
    scale_colour_manual(values = c("Random Forest" = "steelblue",
                                   "XGBoost" = "darkorange")) +
    labs(x = "Mean normalised importance (±1 SD across 9 folds)",
         y = NULL, colour = "Model",
         title = "Variable importance aggregated across temporal folds") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

#### By temporal fold

```{r fig-variable-importance-fold, fig.width=14, fig.height=14}
#| fig-cap: "Variable importance by temporal fold. Each panel shows one fold
#|   trained independently on that fold's data. Importance values are normalised
#|   to [0, 1] within each fold and model. Features are ordered by mean
#|   importance across all folds."
ggplot(imp_by_fold, aes(x = importance, y = feature, fill = model)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_wrap(~ fold, nrow = 3) +
    scale_fill_manual(values = c("Random Forest" = "steelblue",
                                 "XGBoost" = "darkorange")) +
    labs(x = "Normalised importance", y = NULL, fill = "Model",
         title = "Variable importance by temporal fold") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
```

### Calibration plots

```{r fig-calibration, fig.width=16, fig.height=4}
#| fig-cap: "Calibration plots for temporal folds 2–9 (k=1). Locations are binned into
#|   10 deciles by predicted probability; points show the mean predicted
#|   probability vs. observed outbreak proportion within each bin. The dashed
#|   diagonal is perfect calibration. Points above the diagonal indicate
#|   under-prediction; points below indicate over-prediction."
# Calibration: predicted prob vs observed proportion (decile bins)
# Uses temporal_pred_df from the temporal-predictions chunk (folds 2:N_FOLDS)
calibration_data <- temporal_pred_df |>
    pivot_longer(cols = all_of(MODEL_NAMES), names_to = "model",
                 values_to = "predicted") |>
    mutate(bin = ntile(predicted, 10)) |>
    group_by(model, fold, bin) |>
    summarise(
        pred_mean = mean(predicted),
        obs_mean = mean(obs),
        n = n(), .groups = "drop"
    )

ggplot(calibration_data, aes(x = pred_mean, y = obs_mean, colour = model)) +
    geom_point(size = 2) +
    geom_line() +
    geom_abline(linetype = "dashed", colour = "grey50") +
    facet_wrap(~ paste("Fold", fold), nrow = 2) +
    labs(x = "Mean predicted probability", y = "Observed proportion",
         title = "Calibration plots (decile bins)", colour = "Model") +
    theme_minimal(base_size = 10) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1))
```

### Final model: train on all data, predict 2024 (k=1) and 2022–2024 (k=3)

Two final models are trained:

- **k=1**: trained on all 9 validation folds (outcomes 2015–2023), predicts 2024.
  Features: `build_features(2023)` → `feat_yrs = 2019:2023`.
- **k=3**: trained on all 7 validation folds (outcome windows 2015–2023),
  predicts whether any outbreak occurs in 2022–2024.
  Features: `build_features(2021)` → `feat_yrs = 2017:2021`.

No future outcome data enter either model — these are prospective forecasts.

```{r final-model}
# k=1: train on all 9 folds (outcomes 2015-2023), predict 2024
# k=3: train on all 7 folds (outcomes 2015-2023 3-year windows), predict 2022-2024

make_final_pred <- function(kv) {
    fd <- fold_data_list[[paste0("k", kv)]]
    ff <- final_features_list[[paste0("k", kv)]]
    all_tr <- bind_rows(fd)
    fp <- fit_and_predict(all_tr, ff)
    ff |>
        left_join(fp |> dplyr::select(location, all_of(MODEL_NAMES)), by = "location") |>
        rename(pred_prob = Logistic, pred_enet = `Elastic Net`,
               pred_rf = `Random Forest`, pred_xgb = XGBoost, pred_gam = GAM) |>
        mutate(hist_count = n_outbreak_years, k = kv)
}

pred_data_k1 <- make_final_pred(1L)
pred_data_k3 <- make_final_pred(3L)

# Backward-compatible alias: pred_data = k=1 predictions
pred_data <- pred_data_k1

cat("k=1: final predictions for 2024: n =", nrow(pred_data_k1), "locations\n")
cat("  Fold 9 (outcome 2023) prevalence:",
    round(100 * mean(fold_data_list$k1[["fold9"]]$obs), 1), "%\n")
cat("k=3: final predictions for 2022-2024: n =", nrow(pred_data_k3), "locations\n")
cat("  Fold 7 (outcome 2021-2023) prevalence:",
    round(100 * mean(fold_data_list$k3[["fold7"]]$obs), 1), "%\n")
```

## Save Predictions

Save model predictions and supporting data for use in `vaccine_impact_sim.qmd`.

```{r save-predictions}
# Pre-compute pop_lookup so vaccine_impact_sim.qmd doesn't need shapefiles
pop_lookup <- shp_admin23 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

saveRDS(
    list(
        pred_data     = pred_data,      # k=1 predictions (2024 forecast)
        pred_data_k3  = pred_data_k3,   # k=3 predictions (2022-2024 forecast)
        loc_year_full = loc_year_full,
        shp_admin23   = shp_admin23,
        nbr_list      = nbr_list,
        pop_lookup    = pop_lookup
    ),
    "data/outbreak_predictions.rds"
)
cat("Saved data/outbreak_predictions.rds\n")
cat("  pred_data (k=1):  ", nrow(pred_data), "locations\n")
cat("  pred_data_k3:     ", nrow(pred_data_k3), "locations\n")
cat("  loc_year_full:    ", nrow(loc_year_full), "rows\n")
cat("  shp_admin23:      ", nrow(shp_admin23), "polygons\n")
cat("  nbr_list:         ", length(nbr_list), "entries\n")
```

# Beta-Binomial Risk Model and Optimal Alpha (2018 / 2018–2020)

We focus on **fold 4** (training features 2013–2017, T=5), which predicts:

- **k=1**: outbreak in 2018 (single year)
- **k=3**: any outbreak in 2018–2020 (3-year window)

Two vaccination regimens are compared:

- **Single-dose OCV** (k=1): protects within the 1-year horizon; cost per person =
  $C_{V,1}$, so $R_1 = C_I / C_{V,1}$.
- **Two-dose OCV** (k=3): longer-lasting protection needed for a 3-year horizon;
  cost per person = $2 C_{V,1}$, so $R_2 = C_I / (2 C_{V,1}) = R_1 / 2$.

Optimal $\alpha^*$ is computed two ways:

1. **Analytical** (Beta-Binomial, perfect targeting): derived from `cost_beta()` with
   Beta$(a,b)$ parameters fitted to the 2013–2017 training window.
2. **Rho-adjusted** (Monte Carlo with model scores): same MC sweep as in
   `vaccine_impact_sim.qmd` but using fold 4 model predictions as targeting ranks,
   which captures actual rho. The gap between (1) and (2) reflects the penalty of
   imperfect targeting.

```{r bb-alpha-helpers}
# --- Beta-Binomial MLE (same as vaccine_impact_sim.qmd) ---
fit_betabinom_op <- function(k_obs, n_trials) {
    ll <- function(par, k, n) {
        a <- exp(par[1]); b <- exp(par[2])
        sum(lchoose(n, k) + lbeta(k + a, n - k + b) - lbeta(a, b))
    }
    p_init <- mean(k_obs) / n_trials
    v_init <- var(k_obs / n_trials)
    a_init <- max(p_init * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    b_init <- max((1 - p_init) * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    fit <- optim(par = log(c(a_init, b_init)), fn = ll, k = k_obs, n = n_trials,
                 method = "Nelder-Mead", control = list(fnscale = -1, maxit = 5000))
    list(a = exp(fit$par[1]), b = exp(fit$par[2]),
         p = exp(fit$par[1]) / (exp(fit$par[1]) + exp(fit$par[2])),
         convergence = fit$convergence)
}

# --- Analytical cost under Beta(a,b) risk distribution (perfect targeting) ---
mean_risk_pre_op <- function(f_pre, a, b) {
    if (f_pre <= 0) return(0)
    if (f_pre >= 1) return(a / (a + b))
    q <- qbeta(1 - f_pre, a, b)
    (a / ((a + b) * f_pre)) * (1 - pbeta(q, a + 1, b))
}

cost_beta_op <- function(alpha, f, a, b, R, r, nu = 1) {
    alpha <- pmax(0, pmin(1, alpha))
    p <- a / (a + b)
    f_pre <- alpha * f; f_react_cap <- (1 - alpha) * f
    p_bar_pre <- vapply(f_pre, function(fp) mean_risk_pre_op(fp, a, b), numeric(1))
    E_pre <- f_pre * p_bar_pre
    c_pre_val <- f_pre + E_pre * (1 - nu) * R
    E_rem <- p - E_pre
    is_rich <- f_react_cap >= E_rem
    c_react_val <- ifelse(is_rich,
        E_rem * (1 + (1 - r) * R),
        f_react_cap * (1 + (1 - r) * R) + (E_rem - f_react_cap) * R)
    c_pre_val + c_react_val
}

get_alpha_star_beta_op <- function(R, f, a, b, r, nu = 1) {
    optim(par = 0.5, fn = cost_beta_op, f = f, a = a, b = b, R = R, r = r, nu = nu,
          method = "L-BFGS-B", lower = 0, upper = 1)$par
}

# --- MC cost sweep (imperfect targeting via model scores) ---
alpha_star_mc_op <- function(f, pop, obs_rate, score, R, r, AR,
                             nu = 1, n_mc = 300, grid_len = 51, seed = 42) {
    n <- length(pop); P_total <- sum(pop)
    alpha_grid <- seq(0, 1, length.out = grid_len)
    pre_budget_grid <- alpha_grid * f * P_total
    react_budget_grid <- (1 - alpha_grid) * f * P_total
    ranked <- order(score, decreasing = TRUE)
    costs <- sapply(seq_along(alpha_grid), function(gi) {
        pre_budget <- pre_budget_grid[gi]
        react_budget <- react_budget_grid[gi]
        cum_pop <- cumsum(pop[ranked])
        n_pre <- sum(cum_pop <= pre_budget)
        pre_idx <- if (n_pre > 0) ranked[seq_len(n_pre)] else integer(0)
        mc_costs <- numeric(n_mc)
        for (s in seq_len(n_mc)) {
            set.seed(seed + s)
            Y <- rbinom(n, 1, obs_rate)
            pop_pre <- sum(pop[pre_idx])
            burden_pre_res <- sum(Y[pre_idx] * pop[pre_idx]) * (1 - nu)
            rem_idx <- setdiff(seq_len(n), pre_idx)
            ob_rem <- rem_idx[Y[rem_idx] == 1]
            pop_react <- 0; burden_react_res <- 0; burden_uncov <- 0
            if (length(ob_rem) > 0 && react_budget > 0) {
                ob_ord <- ob_rem[order(pop[ob_rem], decreasing = TRUE)]
                n_react <- sum(cumsum(pop[ob_ord]) <= react_budget)
                if (n_react > 0) {
                    ri <- ob_ord[seq_len(n_react)]
                    pop_react <- sum(pop[ri])
                    burden_react_res <- pop_react * (1 - r)
                    if (n_react < length(ob_ord))
                        burden_uncov <- sum(pop[ob_ord[(n_react+1):length(ob_ord)]])
                } else burden_uncov <- sum(pop[ob_rem])
            } else if (length(ob_rem) > 0) burden_uncov <- sum(pop[ob_rem])
            mc_costs[s] <- (pop_pre + pop_react +
                R * AR * (burden_pre_res + burden_react_res + burden_uncov)) / P_total
        }
        mean(mc_costs)
    })
    list(alpha_star = alpha_grid[which.min(costs)],
         cost_star = min(costs), alpha_grid = alpha_grid, cost_grid = costs)
}
```

```{r bb-fit-fold4}
FOLD_IDX <- 4L   # fold of interest: train_end = 2017

# Training feature window for fold 4: 2013-2017 (T=5)
fe4_k1 <- temporal_folds_list$k1[FOLD_IDX, ]
feat_yrs_f4 <- (fe4_k1$train_end - T_WINDOW + 1L):fe4_k1$train_end  # 2013:2017

# Count outbreak-years per location in the training window (n_trials = T_WINDOW)
bb_train_df <- loc_year_end_full |>
    filter(year %in% feat_yrs_f4) |>
    group_by(location) |>
    summarise(k_obs = sum(n_ended > 0), .groups = "drop")

bb_fit_train <- fit_betabinom_op(bb_train_df$k_obs, T_WINDOW)
cat(sprintf("BB fit to training window [%d-%d] (T=%d):\n",
            min(feat_yrs_f4), max(feat_yrs_f4), T_WINDOW))
cat(sprintf("  Beta(%.3f, %.3f), mean p = %.3f, convergence = %d\n",
            bb_fit_train$a, bb_fit_train$b, bb_fit_train$p, bb_fit_train$convergence))

# Fold 4 test data: observed outcomes in 2018 (k=1) and 2018-2020 (k=3)
fold4_k1 <- fold_data_list$k1[[FOLD_IDX]]  # obs = outbreak in 2018
fold4_k3 <- fold_data_list$k3[[FOLD_IDX]]  # obs = any outbreak in 2018-2020

# Attach fold 4 predictions (trained on fold 3, tested on fold 4)
preds4_k1 <- temporal_pred_list_all$k1[[paste0("fold", FOLD_IDX)]]
preds4_k3 <- temporal_pred_list_all$k3[[paste0("fold", FOLD_IDX)]]

cat(sprintf("\nFold 4 k=1 (outcome 2018): n=%d, prevalence=%.1f%%\n",
            nrow(fold4_k1), 100 * mean(fold4_k1$obs)))
cat(sprintf("Fold 4 k=3 (outcome 2018-2020): n=%d, prevalence=%.1f%%\n",
            nrow(fold4_k3), 100 * mean(fold4_k3$obs)))

# Population lookup for MC (use pop_lookup from shapefiles)
pop_f4 <- pop_lookup |> rename(pop_sum = pop_sum)

# Compute obs_rate per location for MC: use empirical outbreak frequency in
# training window as each location's probability
obs_rate_train <- bb_train_df |>
    mutate(obs_rate = k_obs / T_WINDOW) |>
    dplyr::select(location, obs_rate)
```

```{r alpha-star-fold4}
# Vaccine scenario parameters
f_val_op  <- 0.5    # coverage fraction
r_val_op  <- 0.6    # reactive effectiveness
AR_op     <- 0.005  # attack rate
R_base    <- 500    # C_I / C_V (single dose)
nu_1dose  <- 1.0    # single-dose pre-emptive VE within 1 year (fully prevents)
nu_2dose  <- 1.0    # two-dose pre-emptive VE within 3 years (fully prevents)
# Two-dose costs 2× per person → R_2dose = R_base / 2
R_1dose   <- R_base
R_2dose   <- R_base / 2L

# --- 1. Analytical alpha* (Beta-Binomial, perfect targeting) ---
a_star_k1_ew <- get_alpha_star_beta_op(R_1dose, f_val_op,
                                        bb_fit_train$a, bb_fit_train$b,
                                        r_val_op, nu_1dose)
a_star_k3_ew <- get_alpha_star_beta_op(R_2dose, f_val_op,
                                        bb_fit_train$a, bb_fit_train$b,
                                        r_val_op, nu_2dose)

cat(sprintf("Analytical alpha* (perfect targeting):\n"))
cat(sprintf("  k=1 single-dose (R=%.0f, nu=%.1f): alpha* = %.3f\n",
            R_1dose, nu_1dose, a_star_k1_ew))
cat(sprintf("  k=3 two-dose   (R=%.0f, nu=%.1f): alpha* = %.3f\n",
            R_2dose, nu_2dose, a_star_k3_ew))

# --- 2. Rho-adjusted alpha* (MC with fold 4 model predictions) ---
# For each k, use best-rho model from fold 4 as the targeting score
rho4_k1 <- map_dbl(MODEL_NAMES, function(m) {
    tryCatch(cor(preds4_k1[[m]], preds4_k1$obs, method = "spearman"),
             error = function(e) NA_real_)
}) |> setNames(MODEL_NAMES)
rho4_k3 <- map_dbl(MODEL_NAMES, function(m) {
    tryCatch(cor(preds4_k3[[m]], preds4_k3$obs, method = "spearman"),
             error = function(e) NA_real_)
}) |> setNames(MODEL_NAMES)

best_model_k1 <- names(which.max(rho4_k1))
best_model_k3 <- names(which.max(rho4_k3))
cat(sprintf("\nFold 4 rho by model:\n  k=1: %s\n  k=3: %s\n",
            paste(sprintf("%s=%.3f", MODEL_NAMES, rho4_k1), collapse=", "),
            paste(sprintf("%s=%.3f", MODEL_NAMES, rho4_k3), collapse=", ")))
cat(sprintf("Best model k=1: %s (rho=%.3f)\n", best_model_k1, rho4_k1[best_model_k1]))
cat(sprintf("Best model k=3: %s (rho=%.3f)\n", best_model_k3, rho4_k3[best_model_k3]))

# Build MC input: join predictions with pop and obs_rate
build_mc_df <- function(preds, best_model) {
    preds |>
        left_join(pop_f4, by = "location") |>
        left_join(obs_rate_train, by = "location") |>
        filter(!is.na(pop_sum), pop_sum > 0) |>
        replace_na(list(obs_rate = 0))
}
mc_df_k1 <- build_mc_df(preds4_k1, best_model_k1)
mc_df_k3 <- build_mc_df(preds4_k3, best_model_k3)

# MC alpha* for k=1 (single-dose)
mc_k1 <- alpha_star_mc_op(f_val_op, mc_df_k1$pop_sum, mc_df_k1$obs_rate,
                           mc_df_k1[[best_model_k1]],
                           R_1dose, r_val_op, AR_op, nu_1dose, n_mc = 200)
# MC alpha* for k=3 (two-dose: R = R_2dose)
mc_k3 <- alpha_star_mc_op(f_val_op, mc_df_k3$pop_sum, mc_df_k3$obs_rate,
                           mc_df_k3[[best_model_k3]],
                           R_2dose, r_val_op, AR_op, nu_2dose, n_mc = 200)

cat(sprintf("\nRho-adjusted alpha* (MC, best model per k):\n"))
cat(sprintf("  k=1 single-dose (%s, rho=%.3f): alpha* = %.3f\n",
            best_model_k1, rho4_k1[best_model_k1], mc_k1$alpha_star))
cat(sprintf("  k=3 two-dose   (%s, rho=%.3f): alpha* = %.3f\n",
            best_model_k3, rho4_k3[best_model_k3], mc_k3$alpha_star))

# Summary table across all models
alpha_summary <- map_dfr(MODEL_NAMES, function(m) {
    rho_k1 <- rho4_k1[m]; rho_k3 <- rho4_k3[m]
    if (is.na(rho_k1) || is.na(rho_k3)) return(NULL)
    mc1 <- tryCatch(
        alpha_star_mc_op(f_val_op, mc_df_k1$pop_sum, mc_df_k1$obs_rate,
                         mc_df_k1[[m]], R_1dose, r_val_op, AR_op, nu_1dose, n_mc=100),
        error = function(e) list(alpha_star = NA_real_))
    mc3 <- tryCatch(
        alpha_star_mc_op(f_val_op, mc_df_k3$pop_sum, mc_df_k3$obs_rate,
                         mc_df_k3[[m]], R_2dose, r_val_op, AR_op, nu_2dose, n_mc=100),
        error = function(e) list(alpha_star = NA_real_))
    tibble(model = m,
           rho_k1 = rho_k1, rho_k3 = rho_k3,
           alpha_star_mc_k1 = mc1$alpha_star,
           alpha_star_mc_k3 = mc3$alpha_star)
})

alpha_summary |>
    add_row(model = "Analytical (rho=1)",
            rho_k1 = 1, rho_k3 = 1,
            alpha_star_mc_k1 = a_star_k1_ew,
            alpha_star_mc_k3 = a_star_k3_ew) |>
    knitr::kable(
        digits = 3,
        col.names = c("Model", "\u03c1 (k=1)", "\u03c1 (k=3)",
                      "\u03b1* k=1 single-dose", "\u03b1* k=3 two-dose"),
        caption = sprintf(
            "Optimal pre-emptive fraction by model (f=%.1f, r=%.1f, R\u2081=%.0f, R\u2082=%.0f)",
            f_val_op, r_val_op, R_1dose, R_2dose)
    )
```

```{r fig-alpha-cost-curves, fig.width=12, fig.height=5}
#| fig-cap: "MC cost curves for fold 4. Each line shows the expected per-capita
#|   cost as a function of alpha for a given model's targeting scores. Left: k=1
#|   single-dose (R=500). Right: k=3 two-dose (R=250). Diamonds mark the
#|   MC-optimal alpha*; dashed vertical line marks the analytical (rho=1) alpha*."
alpha_grid_op <- seq(0, 1, length.out = 51)

make_cost_curves <- function(mc_df, score_col, R, nu, n_mc = 150) {
    map_dbl(alpha_grid_op, function(a) {
        # simplified MC (replicate alpha_star_mc_op logic for one alpha)
        n <- nrow(mc_df); P_total <- sum(mc_df$pop_sum)
        pre_budget <- a * f_val_op * P_total
        react_budget <- (1 - a) * f_val_op * P_total
        ranked <- order(mc_df[[score_col]], decreasing = TRUE)
        n_pre <- sum(cumsum(mc_df$pop_sum[ranked]) <= pre_budget)
        pre_idx <- if (n_pre > 0) ranked[seq_len(n_pre)] else integer(0)
        mc_vals <- numeric(n_mc)
        for (s in seq_len(n_mc)) {
            set.seed(42 + s)
            Y <- rbinom(n, 1, mc_df$obs_rate)
            burden_pre_res <- sum(Y[pre_idx] * mc_df$pop_sum[pre_idx]) * (1 - nu)
            rem <- setdiff(seq_len(n), pre_idx); ob_rem <- rem[Y[rem] == 1]
            burden_react_res <- 0; burden_uncov <- 0
            if (length(ob_rem) > 0 && react_budget > 0) {
                ob_ord <- ob_rem[order(mc_df$pop_sum[ob_rem], decreasing=TRUE)]
                n_react <- sum(cumsum(mc_df$pop_sum[ob_ord]) <= react_budget)
                if (n_react > 0) {
                    ri <- ob_ord[seq_len(n_react)]
                    burden_react_res <- sum(mc_df$pop_sum[ri]) * (1 - r_val_op)
                    if (n_react < length(ob_ord))
                        burden_uncov <- sum(mc_df$pop_sum[ob_ord[(n_react+1):length(ob_ord)]])
                } else burden_uncov <- sum(mc_df$pop_sum[ob_rem])
            } else if (length(ob_rem) > 0) burden_uncov <- sum(mc_df$pop_sum[ob_rem])
            mc_vals[s] <- (sum(mc_df$pop_sum[pre_idx]) + sum(mc_df$pop_sum[if(exists("ri",inherits=FALSE)) ri else integer(0)]) +
                R * AR_op * (burden_pre_res + burden_react_res + burden_uncov)) / P_total
        }
        mean(mc_vals)
    })
}

curves_k1 <- map_dfr(MODEL_NAMES, function(m) {
    tryCatch({
        costs <- make_cost_curves(mc_df_k1, m, R_1dose, nu_1dose)
        tibble(alpha=alpha_grid_op, cost=costs, model=m, k="k=1 (single-dose)")
    }, error=function(e) NULL)
})
curves_k3 <- map_dfr(MODEL_NAMES, function(m) {
    tryCatch({
        costs <- make_cost_curves(mc_df_k3, m, R_2dose, nu_2dose)
        tibble(alpha=alpha_grid_op, cost=costs, model=m, k="k=3 (two-dose)")
    }, error=function(e) NULL)
})

opt_pts <- bind_rows(
    alpha_summary |> filter(!is.na(alpha_star_mc_k1)) |>
        transmute(alpha=alpha_star_mc_k1, model, k="k=1 (single-dose)") |>
        left_join(curves_k1, by=c("alpha","model","k")),
    alpha_summary |> filter(!is.na(alpha_star_mc_k3)) |>
        transmute(alpha=alpha_star_mc_k3, model, k="k=3 (two-dose)") |>
        left_join(curves_k3, by=c("alpha","model","k"))
)

vlines <- tibble(
    alpha = c(a_star_k1_ew, a_star_k3_ew),
    k = c("k=1 (single-dose)", "k=3 (two-dose)")
)

bind_rows(curves_k1, curves_k3) |>
    ggplot(aes(x=alpha, y=cost, colour=model)) +
    geom_line(linewidth=0.7) +
    geom_vline(data=vlines, aes(xintercept=alpha),
               linetype="dashed", colour="grey40", linewidth=0.5) +
    geom_point(data=opt_pts |> filter(!is.na(cost)),
               aes(x=alpha, y=cost), shape=18, size=3) +
    facet_wrap(~k, scales="free_y") +
    scale_colour_brewer(palette="Set1") +
    labs(x=expression(alpha~"(pre-emptive fraction)"),
         y="Expected cost per capita", colour="Model",
         title="MC cost curves: fold 4 (2018 / 2018-2020)",
         subtitle="Dashed = analytical alpha* (rho=1); diamonds = MC optimal alpha* per model") +
    theme_minimal(base_size=11) +
    theme(legend.position="bottom")
```

```{r fig-map-alpha-fold4, fig.width=16, fig.height=10}
#| fig-cap: "Spatial map of predicted outbreak risk (continuous fill) and pre-emptive
#|   vaccination allocation at MC-optimal alpha* for fold 4. Left column: k=1
#|   single-dose (outcome 2018). Right column: k=3 two-dose (outcome 2018-2020).
#|   Blue = pre-emptively vaccinated; orange = reactively vaccinated (outbreak
#|   location not pre-vaccinated, within reactive budget); red outline = location
#|   with observed outbreak."
map_theme_op <- theme_minimal(base_size=9) +
    theme(panel.grid=element_blank(), axis.text=element_blank(),
          axis.ticks=element_blank(), legend.position="bottom")

make_allocation_map <- function(preds, fold_test, mc_df, alpha_star, R, nu,
                                 best_model, title_str) {
    n <- nrow(mc_df)
    P_total <- sum(mc_df$pop_sum)
    pre_budget <- alpha_star * f_val_op * P_total
    react_budget <- (1 - alpha_star) * f_val_op * P_total

    ranked <- order(mc_df[[best_model]], decreasing=TRUE)
    n_pre <- sum(cumsum(mc_df$pop_sum[ranked]) <= pre_budget)
    pre_locs <- if (n_pre > 0) mc_df$location[ranked[seq_len(n_pre)]] else character(0)

    ob_locs <- fold_test |> filter(obs == 1) |> pull(location)
    ob_not_pre <- setdiff(ob_locs, pre_locs)
    if (length(ob_not_pre) > 0 && react_budget > 0) {
        ob_pop <- mc_df |> filter(location %in% ob_not_pre) |>
            arrange(desc(pop_sum))
        n_react <- sum(cumsum(ob_pop$pop_sum) <= react_budget)
        react_locs <- if (n_react > 0) ob_pop$location[seq_len(n_react)] else character(0)
    } else react_locs <- character(0)

    map_df <- preds |>
        dplyr::select(location, score = all_of(best_model)) |>
        mutate(allocation = case_when(
            location %in% pre_locs   ~ "Pre-emptive",
            location %in% react_locs ~ "Reactive",
            TRUE                      ~ "None"
        ),
        outbreak = location %in% ob_locs)

    shp_map <- shp_admin23 |>
        left_join(map_df, by="location") |>
        filter(!is.na(score))

    ggplot(shp_map) +
        geom_sf(aes(fill=score), colour=NA) +
        geom_sf(data=shp_map |> filter(allocation=="Pre-emptive"),
                fill="steelblue", alpha=0.6, colour=NA) +
        geom_sf(data=shp_map |> filter(allocation=="Reactive"),
                fill="darkorange", alpha=0.6, colour=NA) +
        geom_sf(data=shp_map |> filter(outbreak),
                fill=NA, colour="firebrick", linewidth=0.4) +
        scale_fill_viridis_c(option="magma", direction=-1,
                             name="Predicted\nrisk", limits=c(0,1)) +
        labs(title=title_str,
             subtitle=sprintf("alpha*=%.2f, f=%.1f; blue=pre-emptive, orange=reactive, red outline=outbreak",
                              alpha_star, f_val_op)) +
        map_theme_op
}

p_k1 <- make_allocation_map(
    preds4_k1, fold4_k1, mc_df_k1, mc_k1$alpha_star, R_1dose, nu_1dose,
    best_model_k1,
    sprintf("k=1 single-dose | outcome 2018 | %s (rho=%.3f)",
            best_model_k1, rho4_k1[best_model_k1]))

p_k3 <- make_allocation_map(
    preds4_k3, fold4_k3, mc_df_k3, mc_k3$alpha_star, R_2dose, nu_2dose,
    best_model_k3,
    sprintf("k=3 two-dose | outcome 2018-2020 | %s (rho=%.3f)",
            best_model_k3, rho4_k3[best_model_k3]))

p_k1 + p_k3 +
    plot_annotation(
        title = sprintf("Pre-emptive allocation map: fold 4 (training 2013-2017, T=%d)", T_WINDOW),
        subtitle = sprintf("R1=%.0f (single-dose), R2=%.0f (two-dose), r=%.1f",
                           R_1dose, R_2dose, r_val_op)
    )
```

# COD-Specific Model

The global model's within-country rho is very low (~0.06 for COD) because
most of its discriminative power comes from between-country variation, not
within-DRC spatial patterns. A COD-only model forces all features to explain
*within-country* spatial heterogeneity — which is exactly what pre-emptive
vaccine allocation requires.

**Approach**: identical pipeline to the global model (same T=5 rolling feature
window, same 5 classifiers, same rolling temporal CV), but restricted to COD
locations only. Rho from this model can be directly compared to the global
model's COD rho and used as the targeting accuracy input in
`vaccine_impact_sim.qmd`.

```{r cod-setup}
# Restrict all location sets to COD admin-2/3 units
cod_locs <- all_locs[str_detect(all_locs, "::COD::")]
cat("COD locations:", length(cod_locs), "\n")
cat("Global locations:", length(all_locs), "\n")
cat("COD fraction: ", round(100 * length(cod_locs) / length(all_locs), 1), "%\n")

# COD-only outbreak panel
loc_year_full_cod <- loc_year_full |> filter(location %in% cod_locs)
outbreak_year_summary_cod <- outbreak_year_summary |> filter(location %in% cod_locs)

# COD-only neighbour list (subset; distances already computed)
nbr_list_cod <- nbr_list[cod_locs]

# COD-only feature builder (uses global outbreak_year_summary to allow
# cross-border neighbour lookups, but restricts output rows to COD)
build_features_cod <- function(year_end, T = T_WINDOW) {
    # Build full features (includes neighbours from non-COD locations)
    # then restrict to COD locations only
    build_features(year_end, T, nbr_map = nbr_list_cod,
                   coord_lookup = centroid_coords |> filter(location %in% cod_locs))
}

# Verify outbreak prevalence within COD
cod_prev <- loc_year_full_cod |>
    group_by(year) |>
    summarise(prev = mean(n_active > 0), .groups = "drop")
cat("\nCOD annual outbreak prevalence:\n")
print(cod_prev)
```

```{r cod-temporal-cv}
# Build fold data restricted to COD locations
# Same T=5 rolling window, same fold structure as global k=1 model (9 folds)
cod_temporal_folds <- temporal_folds_list$k1  # same fold dates

make_fold_data_cod <- function(tf) {
    fd <- map(1:nrow(tf), function(i) {
        fe <- tf[i, ]
        feat <- build_features_cod(fe$train_end)
        outcome <- loc_year_full_cod |>
            filter(year %in% (fe$outcome_start:fe$outcome_end)) |>
            group_by(location) |>
            summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")
        feat |>
            left_join(outcome, by = "location") |>
            replace_na(list(obs = 0L))
    })
    names(fd) <- paste0("fold", seq_along(fd))
    fd
}

cod_fold_data <- make_fold_data_cod(cod_temporal_folds)
N_FOLDS_COD <- length(cod_fold_data)

cat(sprintf("\nCOD temporal folds (k=1, T=%d, rolling):\n", T_WINDOW))
for (i in 1:N_FOLDS_COD) {
    fe <- cod_temporal_folds[i, ]
    d  <- cod_fold_data[[i]]
    cat(sprintf(
        "  Fold %d: features [%d,%d] → outcome %d | n=%d, prev=%.1f%%\n",
        i, fe$train_end - T_WINDOW + 1L, fe$train_end,
        fe$outcome_start, nrow(d), 100 * mean(d$obs)
    ))
}
```

```{r cod-temporal-predictions}
# Rolling CV: train on fold i-1, test on fold i (folds 2:N_FOLDS_COD)
cod_pred_list <- map(2:N_FOLDS_COD, function(i) {
    fit_and_predict(cod_fold_data[[i - 1]], cod_fold_data[[i]]) |>
        mutate(fold = i)
})
names(cod_pred_list) <- paste0("fold", 2:N_FOLDS_COD)
cod_pred_df <- bind_rows(cod_pred_list)
```

```{r cod-temporal-eval}
#| tbl-cap: "COD-specific model: rolling temporal CV metrics (train on fold i-1,
#|   test on fold i). Compare rho here with the global model's COD rho (~0.06)."
cod_temporal_results <- map_dfr(2:N_FOLDS_COD, function(i) {
    preds <- cod_pred_list[[paste0("fold", i)]]
    fe <- cod_temporal_folds[i, ]
    map_dfr(MODEL_NAMES, function(m) {
        tryCatch(
            eval_metrics(preds$obs, preds[[m]]) |>
                mutate(model = m, fold = i, outcome_year = fe$outcome_start),
            error = function(e)
                tibble(rho = NA, AUC_ROC = NA, AUC_PR = NA, Brier = NA,
                       Sensitivity = NA, Specificity = NA, Threshold = NA,
                       model = m, fold = i, outcome_year = fe$outcome_start)
        )
    })
})

cod_temporal_results |>
    dplyr::select(fold, outcome_year, model, rho, AUC_ROC, AUC_PR, Brier) |>
    arrange(fold, desc(rho)) |>
    knitr::kable(digits = 3,
        caption = "COD-only model: rolling temporal CV (k=1)")
```

```{r cod-rho-summary}
#| tbl-cap: "COD-specific model: mean rho across folds 2-9, compared to the
#|   global model's COD rho."
cod_rho_summary <- cod_temporal_results |>
    group_by(model) |>
    summarise(
        rho_mean = mean(rho, na.rm = TRUE),
        rho_sd   = sd(rho, na.rm = TRUE),
        AUC_mean = mean(AUC_ROC, na.rm = TRUE),
        n_folds  = sum(!is.na(rho)),
        .groups  = "drop"
    ) |>
    arrange(desc(rho_mean))

# Global model COD rho for comparison (from temporal_pred_df, COD locations only)
global_cod_rho <- map_dfr(2:N_FOLDS, function(i) {
    preds <- temporal_pred_df |> filter(fold == i)
    # Restrict to COD locations
    preds_cod <- preds |> filter(location %in% cod_locs)
    if (nrow(preds_cod) == 0 || var(preds_cod$obs) == 0) return(NULL)
    map_dfr(MODEL_NAMES, function(m) {
        tibble(
            model = m, fold = i,
            rho = tryCatch(cor(preds_cod[[m]], preds_cod$obs, method = "spearman"),
                           error = function(e) NA_real_)
        )
    })
}) |>
    group_by(model) |>
    summarise(rho_global_cod = mean(rho, na.rm = TRUE), .groups = "drop")

cod_rho_summary |>
    left_join(global_cod_rho, by = "model") |>
    dplyr::select(model, rho_mean, rho_sd, rho_global_cod, AUC_mean, n_folds) |>
    knitr::kable(
        digits = 3,
        col.names = c("Model", "rho (COD model)", "SD", "rho (global, COD subset)",
                      "AUC_ROC", "Folds"),
        caption = "COD-specific vs global model: mean rho across temporal folds"
    )
```

```{r fig-cod-rho-by-fold, fig.width=10, fig.height=5}
#| fig-cap: "COD-specific model rho by temporal fold. Each point is the Spearman
#|   rank correlation between predicted probability and observed outbreak in that
#|   fold's test year. Folds where the model fails (e.g. all-zero outcome) are
#|   shown as NA and omitted. The horizontal dashed line marks rho = 0."
cod_temporal_results |>
    filter(!is.na(rho)) |>
    mutate(model = factor(model, levels = MODEL_NAMES)) |>
    ggplot(aes(x = factor(outcome_year), y = rho, colour = model, group = model)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2.5) +
    scale_colour_brewer(palette = "Set1") +
    labs(x = "Outcome year", y = "Spearman \u03c1",
         colour = "Model",
         title = "COD-specific model: rho by temporal fold (k=1, rolling CV)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

```{r fig-cod-roc-curves, fig.width=14, fig.height=8}
#| fig-cap: "ROC curves for the COD-specific model, temporal folds 2-9.
#|   Training restricted to COD locations only; all features computed from
#|   COD outbreak history and 50 km COD neighbours."
cod_roc_df <- map_dfr(2:N_FOLDS_COD, function(i) {
    d <- cod_pred_list[[paste0("fold", i)]]
    fe <- cod_temporal_folds[i, ]
    map_dfr(MODEL_NAMES, function(m) {
        tryCatch({
            roc_obj <- pROC::roc(d$obs, d[[m]], quiet = TRUE)
            tibble(fpr = 1 - roc_obj$specificities,
                   tpr = roc_obj$sensitivities,
                   model = m, fold = i,
                   outcome_year = fe$outcome_start,
                   auc = as.numeric(pROC::auc(roc_obj)))
        }, error = function(e) NULL)
    })
})

cod_roc_plots <- map(2:N_FOLDS_COD, function(i) {
    d <- cod_roc_df |> filter(fold == i)
    yr <- cod_temporal_folds$outcome_start[i]
    ggplot(d, aes(x = fpr, y = tpr, colour = model)) +
        geom_step(direction = "hv", linewidth = 0.5) +
        geom_abline(linetype = "dashed", colour = "grey60") +
        labs(title = sprintf("Fold %d (outcome %d)", i, yr),
             x = "1 - Specificity", y = "Sensitivity", colour = "Model") +
        coord_equal() +
        theme_minimal(base_size = 9) +
        theme(plot.title = element_text(size = 8))
})

wrap_plots(cod_roc_plots, nrow = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

```{r cod-final-model}
# Train on all 9 COD folds (outcomes 2015-2023), predict 2024 (k=1)
# and predict 2022-2024 (k=3), both restricted to COD locations.
# These predictions replace the global model's COD scores in vaccine_impact_sim.qmd.

# k=1 COD final model
cod_all_train_k1 <- bind_rows(cod_fold_data)
cod_final_feat_k1 <- build_features_cod(max(all_years))  # feat_yrs 2019-2023
cod_final_preds_k1 <- fit_and_predict(cod_all_train_k1, cod_final_feat_k1)
pred_data_cod_k1 <- cod_final_feat_k1 |>
    left_join(cod_final_preds_k1 |> dplyr::select(location, all_of(MODEL_NAMES)),
              by = "location") |>
    rename(pred_prob = Logistic, pred_enet = `Elastic Net`,
           pred_rf = `Random Forest`, pred_xgb = XGBoost, pred_gam = GAM) |>
    mutate(hist_count = n_outbreak_years, k = 1L)

# k=3 COD final model
cod_fold_data_k3 <- make_fold_data_cod(temporal_folds_list$k3)
cod_all_train_k3 <- bind_rows(cod_fold_data_k3)
cod_final_feat_k3 <- build_features_cod(max(all_years) - 3L + 1L)  # feat_yrs 2017-2021
cod_final_preds_k3 <- fit_and_predict(cod_all_train_k3, cod_final_feat_k3)
pred_data_cod_k3 <- cod_final_feat_k3 |>
    left_join(cod_final_preds_k3 |> dplyr::select(location, all_of(MODEL_NAMES)),
              by = "location") |>
    rename(pred_prob = Logistic, pred_enet = `Elastic Net`,
           pred_rf = `Random Forest`, pred_xgb = XGBoost, pred_gam = GAM) |>
    mutate(hist_count = n_outbreak_years, k = 3L)

cat("COD k=1 final predictions: n =", nrow(pred_data_cod_k1), "locations\n")
cat("COD k=3 final predictions: n =", nrow(pred_data_cod_k3), "locations\n")

# Append to the saved RDS so vaccine_impact_sim.qmd can load them
preds_rds <- readRDS("data/outbreak_predictions.rds")
preds_rds$pred_data_cod_k1 <- pred_data_cod_k1
preds_rds$pred_data_cod_k3 <- pred_data_cod_k3
saveRDS(preds_rds, "data/outbreak_predictions.rds")
cat("Updated data/outbreak_predictions.rds with COD-specific predictions\n")
```
