---
title: "Reactive vs Preemptive Vaccination Strategies"
author: "Jong-Hoon Kim"
date: "2025-06-26"
format: html
editor_options: 
  chunk_output_type: console
---

# Setup

```{r setup}
#| warning: false
#| message: false

# Load required packages
source("R/utils.R")  # utility functions
library(data.table)
library(tidyverse)
library(magrittr)
library(ggridges)
library(readxl)
library(janitor)
library(scales)
library(viridis)
library(sf)
library(metR)
library(googlesheets4)
library(MASS)
library(patchwork)
library(ggpattern)
library(gridExtra)
library(corrr)
library(RColorBrewer)
library(countrycode)
library(conflicted)
# Resolve package conflicts in favor of dplyr
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

# Output control flags
save_plot <- FALSE
save_table <- FALSE

# Configure figure size settings
figure_size <- data.frame(
  journal = "Lancet",
  single = 85,   # Width for single-column figures (mm)
  double = 178   # Width for double-column figures (mm)
)
fgw1 <- figure_size[figure_size$journal == "Lancet",]$single
fgw2 <- figure_size[figure_size$journal == "Lancet",]$double

# Color palette for consistent visualization
set1 <- brewer.pal(9, "Set1")  # color sets
my_discrete_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00")

# Define standard probability values for quantiles
probs <- c(0.025, 0.25, 0.5, 0.75, 0.975)  # 2.5%, 25%, 50%, 75%, 97.5% quantiles

# Reference vaccination week
# VW <- 9  # representative vaccination week from GTFCC data
# 
# # Sequence of vaccination weeks to evaluate
# VW_seq <- seq(1, 15, 2)
# 
# # Representative vaccination coverage rate
# VC <- 0.75  # 75% coverage
# 
# # Three representative vaccination coverage rates
# VC3 <- c(0.6, 0.75, 0.9)  # 60%, 75%, 90%
# 
# # Thresholds for case-triggered vaccination to analyze
# cts_sel <- c(1, seq(10, 100, 10))  # case triggers selected
# 
# # Latest date for which OCV campaign data is available
# YEAR <- as.Date("2020-12-28")
# 
# # Load geographic data for mapping
# shp <- st_read("data/shapefiles/total_shp_0427.shp", quiet = TRUE)
# africa <- st_read("data/shapefiles/afr_g2014_2013_0.shp", quiet = TRUE)

# only a subset of sample runid's is used for plotting for convenience
set.seed(42)
sample_runids <- sample(1:200, size=30)
```

### Plots

#### Data preparation

```{r}
# Global parameters ----------------------------------------------------
text_size <- 5          # Default text size for annotations
line_height <- 0.95     # Line spacing for text annotation
pattern_dens <- 0.1     # Density of pattern filling
pattern_spacing <- 0.1  # Spacing between patterns
step_linewidth <- 0.7   # Width of step lines
arrow_vertical_len <- 14 # Vertical length of arrows

# Axis labels
ylabel <- "Weekly suspected cholera cases"
xlabel <- "Weeks since outbreak onset"

# Key timeline parameters
week_pre <- -6    # Pre-emptive vaccination 6 weeks prior to the outbreak  
week_or <- 3      # Outbreak response at week 3  
week_vax <- 6.5   # Vaccination started
week_vax_eff <- 8.5 # Vaccine takes effect
pcr_preemptive <- 0.95
pcr_reactive <- 0.45
# Rectangle dimensions configuration ------------------------------------
# This section makes it easy to adjust all rectangles at once

# Base rectangle dimensions
rect_width <- 0.8
rect_height <- 6

# Spacing between rectangles
gap_horizontal <- 1.9   # Horizontal gap between rectangle groups
gap_vertical <- 12      # Vertical gap between top and bottom rectangles

# Font sizes for elements
eq_symbol_size <- 8     # Size for equation symbols (<, +)
paren_size <- 9         # Size for parentheses
plus_size <- 5          # Size for plus symbols

# Vertical offset for second equation from first
eq2_vertical_offset <- 24

# Helper functions ----------------------------------------------------
create_filled_square <- function(x1, y1, x2, y2, fill_color, 
                                 color='black', alpha=0.5) {
  geom_rect(xmin=x1, xmax=x2, ymin=y1, ymax=y2,
            color=color, fill=fill_color, alpha=alpha, inherit.aes = FALSE)
}

create_striped_square <- function(x1, y1, x2, y2, fill_color, color='black',
                                  dens=0.1, angle=30) {
  geom_rect_pattern(
    xmin=x1, xmax=x2, ymin=y1, ymax=y2,
    pattern = 'stripe',
    fill = NA,
    colour = color,
    pattern_density = dens,
    pattern_spacing = 0.025,
    pattern_fill = fill_color,
    pattern_colour = NA,
    pattern_angle = angle,
    inherit.aes = FALSE)
}

create_divider <- function(x1, x2, y, extension=0.2, linewidth=0.7) {
  annotate("segment", 
           x = x1 - extension, 
           y = y, 
           xend = x2 + extension, 
           yend = y, 
           linewidth = linewidth)
}

theme_upper_panel <- function() {
  theme_classic() +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 16),
      axis.text.y = element_text(size = 14),
      plot.caption = element_text(size = 14),
      legend.position = "none"
    )
}

# Data preparation ----------------------------------------------------
# Create synthetic example of an outbreak trajectory with and without vaccination
d <- data.frame(
  date = seq(as.Date("2010-06-15"), by = "week", length.out = 24),
  week = 0:23,
  # No vaccination scenario
  s_ch = c(2, 2, 3, 4, 10, 12, 29, 100, 120, 150, 140, 
           75, 68, 47, 24, 5, 12, 5, 3, 2, 4, 1, 0, 0)
)

# Add vaccination scenario (40% of cases after vaccine takes effect)
d$s_ch_vacc_preemptive <- round((1-pcr_preemptive)*(d$s_ch))
d$s_ch_vacc_reactive <- c(d$s_ch[1:9], round((1-pcr_reactive) * d$s_ch[10:nrow(d)]))
d$s_ch_cumul <- cumsum(d$s_ch)

# Rename columns for clarity
names(d) <- c("date", "week", "No vacc", "Vacc_Preemptive", "Vacc_Reactive", "No vacc cumulative")

# Reshape data for plotting
dlong <- pivot_longer(d, cols = c("No vacc", "Vacc_Preemptive", "Vacc_Reactive"))

# Define time weeks for additional scenarios
weeks <- 0:25

# Generate additional outbreak scenarios
# Smallest outbreak (ended before vaccination)
smallest_no_vax <- 
  rpois(length(weeks), lambda=30 * exp(-0.1 * (weeks - 2)^2))
smallest_no_vax[(week_vax-1):length(smallest_no_vax)] <- 0

# Smaller outbreak (ended at vaccine takes effect)
smaller_no_vax <- 
  rpois(length(weeks), lambda=50 * exp(-0.1 * (weeks - 3)^2))
smaller_no_vax[week_vax_eff:length(smaller_no_vax)] <- 0

# Small outbreak (peak at week 8)
small_no_vax <- rpois(length(weeks), lambda=70 * exp(-0.08 * (weeks - 8)^2))
small_with_vax <- small_no_vax
small_with_vax[weeks > week_vax_eff] <- round(small_with_vax[weeks > week_vax_eff] * 0.5)  # 50% reduction

# Large outbreak (peak at week 12)
large_no_vax <- rpois(length(weeks), lambda=120 * exp(-0.04 * (weeks - 12)^2))
large_with_vax <- large_no_vax
large_with_vax[weeks > week_vax_eff] <- round(large_with_vax[weeks > week_vax_eff] * 0.5)  # 50% reduction

# Combine into a data frame
df <- tibble(
  week = rep(weeks, 6),
  cases = c(smallest_no_vax, smaller_no_vax, small_no_vax, small_with_vax, 
            large_no_vax, large_with_vax),
  outbreak = rep(c("Smallest", "Smaller", "Small", "Small", "Large", "Large"), 
                 each = length(weeks)),
  scenario = c(rep("No Vaccine", 2*length(weeks)), 
               rep(c("No Vaccine", "With Vaccine"), times=2, each=length(weeks)))
)

# Create wide version of the data
df_wide <- df %>%
  pivot_wider(names_from = scenario, values_from = cases) %>%
  mutate(reduction = `No Vaccine` - `With Vaccine`)

# Rectangle positions for equation 1 ----------------------------------
# Starting position for first equation
base_x1 <- 15
base_y1 <- 65

# a/b fraction
a_x1 <- base_x1
a_y1 <- base_y1
a_x2 <- a_x1 + rect_width
a_y2 <- a_y1 + rect_height

b_x1 <- a_x1
b_y1 <- a_y1 - gap_vertical
b_x2 <- b_x1 + rect_width
b_y2 <- b_y1 + rect_height

ab_div_y <- (a_y1 + b_y2) / 2

# c/d fraction
c_x1 <- a_x1 + gap_horizontal
c_y1 <- a_y1
c_x2 <- c_x1 + rect_width
c_y2 <- c_y1 + rect_height

d_x1 <- c_x1
d_y1 <- b_y1
d_x2 <- d_x1 + rect_width
d_y2 <- d_y1 + rect_height

cd_div_y <- (c_y1 + d_y2) / 2

# Inequality symbol
eq1_lt_x <- (a_x2 + c_x1) / 2
eq1_lt_y <- (a_y1 + b_y2) / 2

# Equation label
eq1_label_x <- a_x1 - 1.2
eq1_label_y <- cd_div_y

# Rectangle positions for equation 2 ----------------------------------
# Left side: 1/2(a/b+c/d)
# a/b part
eq2_a_x1 <- a_x1 + 0.2
eq2_a_y1 <- a_y1 - eq2_vertical_offset
eq2_a_x2 <- eq2_a_x1 + rect_width
eq2_a_y2 <- eq2_a_y1 + rect_height

eq2_b_x1 <- eq2_a_x1
eq2_b_y1 <- eq2_a_y1 - gap_vertical
eq2_b_x2 <- eq2_b_x1 + rect_width
eq2_b_y2 <- eq2_b_y1 + rect_height

# c/d part
eq2_c_x1 <- eq2_a_x1 + gap_horizontal
eq2_c_y1 <- eq2_a_y1
eq2_c_x2 <- eq2_c_x1 + rect_width
eq2_c_y2 <- eq2_c_y1 + rect_height

eq2_d_x1 <- eq2_c_x1
eq2_d_y1 <- eq2_b_y1
eq2_d_x2 <- eq2_d_x1 + rect_width
eq2_d_y2 <- eq2_d_y1 + rect_height

# Parentheses
left_paren_x <- eq2_a_x1 - 0.3
right_paren_x <- eq2_c_x2 + 0.3
paren_y <- (eq2_a_y1 + eq2_b_y2) / 2

# 1/2 fraction position
eq2_frac_x <- right_paren_x + 0.6
eq2_frac_y_num <- (eq2_a_y1 + eq2_a_y2)/2
eq2_frac_y_denom <- (eq2_b_y1 + eq2_b_y2)/2
eq2_frac_y_line <- (eq2_a_y1 + eq2_b_y2)/2

# Plus symbol position
eq2_plus1_x <- (eq2_a_x2 + eq2_c_x1) / 2
eq2_plus1_y <- (eq2_a_y1 + eq2_b_y2) / 2

# Less than symbol for second equation
eq2_lt_x <- eq2_frac_x + 0.8
eq2_lt_y <- (eq2_a_y1 + eq2_b_y2)/2

# Right side: (a+c)/(b+d)
# Numerator (a+c)
eq2_right_a_x1 <- eq2_lt_x + 0.8
eq2_right_a_y1 <- eq2_a_y1
eq2_right_a_x2 <- eq2_right_a_x1 + rect_width
eq2_right_a_y2 <- eq2_right_a_y1 + rect_height

eq2_right_c_x1 <- eq2_right_a_x1 + gap_horizontal
eq2_right_c_y1 <- eq2_a_y1
eq2_right_c_x2 <- eq2_right_c_x1 + rect_width
eq2_right_c_y2 <- eq2_right_c_y1 + rect_height

eq2_plus2_x <- (eq2_right_a_x2 + eq2_right_c_x1)/2
eq2_plus2_y <- (eq2_right_a_y1 + eq2_right_a_y2)/2

# Denominator (b+d)
eq2_right_b_x1 <- eq2_right_a_x1
eq2_right_b_y1 <- eq2_b_y1
eq2_right_b_x2 <- eq2_right_b_x1 + rect_width
eq2_right_b_y2 <- eq2_right_b_y1 + rect_height

eq2_right_d_x1 <- eq2_right_c_x1
eq2_right_d_y1 <- eq2_right_b_y1
eq2_right_d_x2 <- eq2_right_d_x1 + rect_width
eq2_right_d_y2 <- eq2_right_d_y1 + rect_height

eq2_plus3_x <- eq2_plus2_x
eq2_plus3_y <- (eq2_right_b_y1 + eq2_right_b_y2)/2

# Division line
eq2_right_div_x1 <- eq2_right_a_x1
eq2_right_div_x2 <- eq2_right_c_x2
eq2_right_div_y <- (eq2_right_a_y1 + eq2_right_b_y2) / 2

# Equation 2 label
eq2_label_x <- eq1_label_x
eq2_label_y <- a_y1 - eq2_vertical_offset
```

#### Preemptive

```{r}
p1 <- ggplot(dlong, aes(week, value, linetype = name, 
                    shape = name, group = name)) +
  # Background stripes for no vaccination scenario
  geom_rect_pattern(
    data = subset(dlong, name == "No vacc"),
    aes(xmin=week, xmax=lead(week), ymin=0, ymax=value),
    pattern = 'stripe',
    fill = NA,
    colour = NA,
    pattern_colour = NA,
    pattern_density = pattern_dens,
    pattern_fill = 'steelblue',
    pattern_angle = 30) +  
  # Vaccination impact rectangle
  geom_rect(
    data = d,
    aes(xmin=week, xmax=lead(week), ymin=`Vacc_Preemptive`, 
        ymax=`No vacc`), 
    fill='steelblue', 
    alpha=0.3,
    color=NA,
    inherit.aes = FALSE) +
  # Step lines for clarity
  geom_step(data=subset(dlong, name == "Vacc_Preemptive"),
            aes(x=week, y=value),
            linewidth=step_linewidth, linetype="solid") +
  geom_step(data=subset(dlong, name == "No vacc"),
            aes(x=week, y=value),
            linewidth=step_linewidth, linetype="dashed") +
  # Outbreak response triggered arrow and label
  geom_segment(
    x=week_pre, xend=week_pre, y=10+arrow_vertical_len,  yend=0,
    arrow = arrow(type = "closed", angle = 10, length = unit(3, "mm")),
    inherit.aes = FALSE
  ) +
  annotate(
    "text", x=week_pre-2, y=38,
    label = "Pre-emptive\nvaccination",
    hjust=0.0, size=text_size, lineheight=line_height
  ) +
  annotate(
    "text", x=16, y=75, 
    label = "% case\nreduction", 
    size=text_size, lineheight=line_height
  ) +
  annotate(
    "text", x=19.2, y=75, 
    label = bquote("= 100 \u00D7"), 
    size=text_size, lineheight=line_height
  ) +
  # Fraction expressed using squares
  create_divider(x1=21.1-0.2, x2=22.1+0.2, y=75) +
  create_filled_square(
    x1=21.1, x2=22.1, y1=77, y2=85, 
    color='black', fill="steelblue", alpha=0.3) +
  create_striped_square(
    x1=21.1, x2=22.1, y1=64, y2=72,
    fill_color='steelblue', color='black'
  ) +
  scale_x_continuous(limits=c(-8,25)) +
  labs(x="", y=ylabel) +
  theme_upper_panel()
```

#### Reactive

```{r}
p2 <- ggplot(dlong, aes(week, value, linetype = name, shape = name, group = name)) +
  # Background stripes for no vaccination scenario
  geom_rect_pattern(
    data = subset(dlong, name == "No vacc"),
    aes(xmin=week, xmax=lead(week), ymin=0, ymax=value),
    pattern = 'stripe',
    fill = NA,
    colour = NA,
    pattern_colour = NA,
    pattern_density = pattern_dens,
    pattern_fill = 'steelblue',
    pattern_angle = 30) +
  
  # Vaccination impact rectangle
  geom_rect(
    data = d,
    aes(xmin=week, xmax=lead(week), ymin=`Vacc_Reactive`, ymax=`No vacc`), 
    fill='steelblue', 
    alpha=0.5,
    color=NA,
    inherit.aes = FALSE) +
  
  # Step lines for clarity
  geom_step(data=subset(dlong, name == "Vacc_Reactive"),
            aes(x=week, y=value),
            linewidth=step_linewidth, linetype="solid") +
  geom_step(data=subset(dlong, name == "No vacc"),
            aes(x=week, y=value),
            linewidth=step_linewidth, linetype="dashed") +
  # Outbreak response triggered arrow and label
  geom_segment(
    x=week_or, y=10+arrow_vertical_len, xend=week_or, yend=10,
    arrow = arrow(type = "closed", angle = 10, length = unit(3, "mm")),
    inherit.aes = FALSE
  ) +
  annotate(
    "text", x=week_or-0.5, y=38, 
    label = "Outbreak\nresponse triggered", 
    hjust=0.5, size=text_size, lineheight=line_height
  ) +
  # Vaccination started arrow and label
  geom_segment(
    x=week_vax, y=32+arrow_vertical_len, xend=week_vax, yend=32,
    arrow = arrow(type = "closed", angle = 10, length = unit(3, "mm")),
    inherit.aes = FALSE
  ) +
  annotate(
    "text", x=week_vax, y=60, 
    label = "Vaccination\nstarted", hjust=1,
    size=text_size, lineheight=line_height
  ) +
  
  # Vaccine takes effect arrow and label
  geom_segment(
    x=week_vax_eff, y=122+arrow_vertical_len, xend=week_vax_eff, yend = 122,
    arrow = arrow(type = "closed", angle = 10, length = unit(3, "mm")),
    inherit.aes = FALSE
  ) +
  annotate(
    "text", x=week_vax_eff, y=147, 
    label="Vaccine\ntakes effect", hjust=1,
    size=text_size, lineheight=line_height
  ) +
  
  # Scenario labels
  annotate(
    "text", x=11.5, y=120, 
    label = "Without\nvaccination", hjust=0, 
    size=text_size, lineheight=line_height
  ) +
  annotate(
    "text", x=9.5, y=12, 
    label = "With\nvaccination", hjust=0,
    size=text_size, lineheight=line_height
  )+
  
  # Colored lines on the horizontal axis marking periods
  annotate("segment", x=0, y=0, xend=3, yend=0, linewidth=2, color="firebrick") +
  annotate("segment", x=3, y=0, xend=7, yend=0, linewidth=2, color="forestgreen") +
  annotate("segment", x=7, y=0, xend=9, yend=0, linewidth=2, color="purple") +
  
  scale_x_continuous(limits=c(-8,25)) +
  labs(x="", y=ylabel) +
  theme_classic() +
  theme(
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    plot.caption = element_text(size = 14),
    legend.position = "none")
```

Combine the plots

```{r}
# Combine plots
combined_plot <- (p1 / p2) + 
  plot_layout(
    heights = c(1, 1), 
    axis_titles = 'collect',
    guides = 'collect'
  )

if (save_plot) {
  fac <- 1.8
  ggsave(paste0("plots/preemptive_reactive_schematic_", tstamp(), ".pdf"),
        combined_plot,
        device = cairo_pdf,
        width=1.25*fgw1*fac,
        height=1.6*fgw1*fac,
        units = "mm")
}
```

### Targetting accuracy vs. relative impact of preemptive vaccination

```{r}
# Data for reactive scenario vs no intervention
baseline <- data.frame(time=c(0,10,20), cases=c(0,100,0))
reactive <- data.frame(time=c(0,5,12), cases=c(0,50,0))

# Coordinates for error bars at the peak of reactive scenario
peak_time <- 5; peak_cases <- 50
err_time <- 1; err_cases <- 15  # uncertainty: ±1 day, ±15 cases
err_df <- data.frame(x=peak_time, y=peak_cases,
                     ymin=peak_cases - err_cases, ymax=peak_cases + err_cases,
                     xmin=peak_time - err_time, xmax=peak_time + err_time)

p_reactive <- ggplot() +
  geom_area(data=reactive, aes(x=time, y=cases), fill="steelblue", alpha=0.4) +
  geom_line(data=baseline, aes(x=time, y=cases), color="black", linetype="dashed", size=1) +
  # Vertical error bar at peak
  geom_errorbar(data=err_df, aes(x=x, ymin=ymin, ymax=ymax), width=0.5) +
  # Horizontal error bar at peak (as a segment)
  geom_segment(data=err_df, aes(x=xmin, xend=xmax, y=y, yend=y)) +
  labs(x="Time (days)", y="Cases", title="Reactive vaccination vs. no intervention") +
  # Annotate key times
  annotate("text", x=0, y=10, label="t0") +
  annotate("text", x=10, y=105, label="t_p") +
  annotate("text", x=20, y=10, label="t_1") +
  theme_classic()

print(p_reactive)

```

### Simplied outbreak version
```{r}
library(ggplot2)
library(patchwork)
library(grid)  # for unit()

```

```{r}
pcr_preemptive <- 0.95
pcr_reactive <- 0.45
# Tent function
tent_curve <- function(t, t0, tp, t1, r0, rp) {
  ifelse(
    t < t0 | t > t1, r0,
    ifelse(t <= tp,
           r0 + (rp - r0) * (t - t0)/(tp - t0),
           rp - (rp - r0) * (t - tp)/(t1 - tp))
  )
}

# Time vector
t <- seq(0, 60, by = 0.1)
# Time markers
t_pre_vax <- 5
t_detect  <- 20
t_re_vax  <- 28
t_effect  <- 30

inc_baseline <- tent_curve(t, 15, 35, 60, 0, 20)

# Counterfactual: same for both panels
cf_curve <- data.frame(t = t, incidence = inc_baseline, group = "Counterfactual")

# Preemptive reduced outbreak
preemptive_curve <- data.frame(t = t, 
                               incidence = (1-pcr_preemptive) * inc_baseline, 
                               group = "Vacc_Preemptive")

# Reactive observed outbreak (matches counterfactual before vaccination)
reactive_curve <- data.frame(t = t, 
                             incidence = ifelse(t >= t_effect, (1-pcr_reactive) * inc_baseline, inc_baseline), group = "Vacc_Reactive")
```

#### Pre-emptive
```{r}
library(ggpattern)
p1 <- ggplot() +
  geom_line(data = cf_curve, aes(t, incidence), linetype = "dashed", color = "black", linewidth = 1) +
  geom_ribbon_pattern(
    data = preemptive_curve,
    aes(x=t, ymin=0, ymax=incidence),
    pattern = 'stripe',
    fill = NA,
    colour = NA,
    pattern_colour = NA,
    pattern_density = pattern_dens,
    pattern_fill = 'steelblue',
    pattern_angle = 30) +
   +
  # Arrow
  geom_segment(aes(x = t_pre_vax, xend = t_pre_vax, y = 7, yend = 1),
               arrow = arrow(length = unit(0.1, "inches")), color = "black") +
  # Circle at vaccination time
  geom_point(aes(x = t_pre_vax, y = 0), shape = 21, color="black", fill = "firebrick", size = 4) +
  annotate("text", x = t_pre_vax, y = 9, label = "Vaccination", hjust = 0.3) +
  labs(title = "Preemptive Vaccination", x = "Time", y = "Cases") +
  theme_minimal()

p1
```


```{r}
p2 <- ggplot() +
  geom_line(data = cf_curve, aes(t, incidence), linetype = "dashed", color = "black", linewidth = 1) +
  geom_step(data = reactive_curve, aes(t, incidence), linetype = "solid", color = "steelblue", linewidth = 1) +
  # Arrow
  geom_segment(aes(x = t_detect, xend = t_detect, y = 7, yend = 1),
               arrow = arrow(length = unit(0.1, "inches")), color = "black") +
  geom_segment(aes(x = t_re_vax, xend = t_re_vax, y = 7, yend = 1),
               arrow = arrow(length = unit(0.1, "inches")), color = "black") +
  geom_segment(aes(x = t_effect, xend = t_effect, y = 7, yend = 1),
               arrow = arrow(length = unit(0.1, "inches")), color = "black") +
  # Circle at vaccination time
  geom_point(aes(x = t_detect, y = 0), shape = 21, color="black", fill = "firebrick", size = 4) +
  geom_point(aes(x = t_re_vax, y = 0), shape = 21, color="black", fill = "firebrick", size = 4) +
  geom_point(aes(x = t_effect, y = 0), shape = 21, color="black", fill = "firebrick", size = 4) +
  
  annotate("text", x = t_pre_vax, y = 9, label = "Vaccination", hjust = 0.3) +
  labs(title = "Preemptive Vaccination", x = "Time", y = "Cases") +
  theme_minimal()

p2


```

The proportion of allocation for pre-emptive vaccination against the average For given average probability of an outbreak, $\hat{p}$, you would need $p*N$ vaccine doses on average. In case, you have only $0.1*p*N$. What would be your optimal strategy?

1.  Some parametric distribution on $p$
2.  


```{r}
# Preemptive plot
p1 <- ggplot() +
  geom_line(data = preemptive_main, aes(t, incidence), linetype = "dashed", color = "black", linewidth = 1) +
  geom_line(data = preemptive_small, aes(t, incidence), linetype = "dashed", color = "blue", linewidth = 1) +
  geom_segment(data = arrow_df[1,],
               aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(type = "closed", length = unit(0.2, "inches")),
               color = "black") +
  annotate("text", x = arrow_df$x[1], y = 95, label = "Vaccination", hjust = 0.5) +
  labs(title = "Preemptive Vaccination", x = "Time", y = "Cases") +
  theme_minimal()

# Reactive plot
p2 <- ggplot() +
  geom_line(data = reactive_real, aes(t, incidence), color = "black", linewidth = 1) +
  geom_line(data = reactive_counter, aes(t, incidence), linetype = "dashed", color = "red", linewidth = 1) +
  geom_segment(data = arrow_df[2,],
               aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(type = "closed", length = unit(0.2, "inches")),
               color = "black") +
  annotate("text", x = arrow_df$x[2], y = 115, label = "Vaccination", hjust = 0.5) +
  labs(title = "Reactive Vaccination", x = "Time", y = "Cases") +
  theme_minimal()

```

```{r}
p1 / p2 + plot_layout(ncol = 1)

```

