---
title: "Outbreak Risk Prediction: Simple One-Step Model"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

A single-step prediction exercise:

- **Training**: features built from 2010–2014 surveillance data; outcome = any
  outbreak in 2015–2017 (k=3 window)
- **Prediction**: features built from 2013–2017 surveillance data; predict
  probability of any outbreak in 2018–2020
- **Models**: logistic regression, GAM, random forest
- **Output**: predicted risk per ADM2 unit; cumulative population vs. risk curve

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(patchwork)
library(pROC)
library(scales)
library(knitr)
library(mgcv)    # GAM
library(ranger)  # random forest
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```

# Data

## Outbreak surveillance

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)
ts23 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm %in% c(2, 3))

outbreak_events <- ts23 |>
    group_by(location, adm, outbreak_number) |>
    summarise(
        year_start     = year(min(TL)),
        year_end       = year(max(TR)),
        duration_weeks = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
        total_cases    = sum(sCh, na.rm = TRUE),
        .groups        = "drop"
    )

all_years <- 2010:2023
all_locs  <- unique(ts23$location)

outbreak_year_summary <- outbreak_events |>
    filter(year_end %in% all_years) |>
    group_by(location, year = year_end) |>
    summarise(
        outbreak_count    = n(),
        total_cases       = sum(total_cases, na.rm = TRUE),
        max_duration_days = max(duration_weeks * 7, na.rm = TRUE),
        mean_duration_days = mean(duration_weeks * 7, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(outbreak_count = 0L, total_cases = 0,
                    max_duration_days = 0, mean_duration_days = 0))

# loc_year_full: outbreaks active in each year
loc_year_full <- outbreak_events |>
    mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) |>
    unnest(years) |>
    rename(year = years) |>
    filter(year %in% all_years) |>
    count(location, year, name = "n_active") |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(n_active = 0L))

cat("Locations:", length(all_locs), "| Years:", range(all_years), "\n")
```

## Shapefiles and spatial neighbours

```{r load-shapefiles}
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)
loc_id_map <- ts23 |>
    distinct(location_period_id, location) |>
    group_by(location) |>
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) |>
    ungroup()

shp_admin23 <- shp_all |>
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id")) |>
    filter(!is.na(location), str_count(location, "::") %in% c(3, 4))

cat("ADM2/3 polygons:", nrow(shp_admin23), "\n")

# 50 km neighbour list
DIST_KM <- 50
centroids_sf <- shp_admin23 |> st_make_valid() |> st_point_on_surface()
dist_mat <- st_distance(centroids_sf) |> units::drop_units() / 1000
rownames(dist_mat) <- colnames(dist_mat) <- shp_admin23$location
diag(dist_mat) <- Inf

nbr_list <- setNames(
    lapply(shp_admin23$location, function(loc) {
        d <- dist_mat[loc, ]
        names(sort(d[d <= DIST_KM]))
    }),
    shp_admin23$location
)

centroid_coords <- centroids_sf |>
    st_coordinates() |>
    as_tibble() |>
    rename(lon = X, lat = Y) |>
    mutate(location = shp_admin23$location)
```

## Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)

POP_SCALE <- 344  # pixel area ~344 km²
shp_pop   <- shp_admin23 |> st_make_valid() |> st_transform(crs(pop_raster))
pop_vals  <- exactextractr::exact_extract(pop_raster, shp_pop,
                                          fun = "sum", progress = FALSE)
shp_admin23$pop_sum <- pop_vals * POP_SCALE

pop_lookup <- shp_admin23 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

cat("Total population:", format(round(sum(pop_vals * POP_SCALE, na.rm = TRUE)),
                                big.mark = ","), "\n")
```

# Feature Engineering

Features are built from a T=5-year rolling window ending at a given year.
Twelve same-unit features and six neighbour-aggregated features are used.

```{r build-features-fn}
T_WINDOW <- 5L

build_features <- function(year_end, T = T_WINDOW) {
    feat_yrs <- (year_end - T + 1L):year_end
    n_yrs    <- length(feat_yrs)

    loc_summary <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            n_outbreak_years    = sum(outbreak_count > 0),
            prop_outbreak_years = n_outbreak_years / n_yrs,
            total_outbreaks     = sum(outbreak_count),
            mean_cases_per_year = mean(total_cases),
            max_cases_any_year  = max(total_cases),
            mean_duration       = mean(mean_duration_days),
            max_duration        = max(max_duration_days),
            .groups = "drop"
        )

    recent_1yr <- outbreak_year_summary |>
        filter(year == year_end) |>
        transmute(location, recent_outbreak = as.integer(outbreak_count > 0))

    recent_2yr <- outbreak_year_summary |>
        filter(year %in% intersect(c(year_end - 1L, year_end), feat_yrs)) |>
        group_by(location) |>
        summarise(recent_2yr_outbreaks = sum(outbreak_count > 0), .groups = "drop")

    trend_df <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            trend = if (n() >= 2)
                coef(lm(as.integer(outbreak_count > 0) ~ year))[2] else 0,
            trend_cases = if (n() >= 2)
                coef(lm(total_cases ~ year))[2] else 0,
            .groups = "drop"
        )

    last_ob <- outbreak_year_summary |>
        filter(year %in% feat_yrs, outbreak_count > 0) |>
        group_by(location) |>
        summarise(last_ob_year = max(year), .groups = "drop")

    yrs_since <- tibble(location = all_locs) |>
        left_join(last_ob, by = "location") |>
        mutate(years_since_last_outbreak = ifelse(
            is.na(last_ob_year), n_yrs + 1L,
            as.integer(year_end - last_ob_year)
        )) |>
        dplyr::select(location, years_since_last_outbreak)

    feat <- tibble(location = all_locs) |>
        left_join(loc_summary,  by = "location") |>
        left_join(recent_1yr,   by = "location") |>
        left_join(recent_2yr,   by = "location") |>
        left_join(trend_df,     by = "location") |>
        left_join(yrs_since,    by = "location") |>
        replace_na(list(
            n_outbreak_years = 0L, prop_outbreak_years = 0,
            total_outbreaks = 0L, mean_cases_per_year = 0,
            max_cases_any_year = 0, mean_duration = 0, max_duration = 0,
            recent_outbreak = 0L, recent_2yr_outbreaks = 0L,
            trend = 0, trend_cases = 0
        ))

    # Neighbour features
    make_nbr_feat <- function(values, locs, fun = mean) {
        sv <- setNames(values, locs)
        map_dbl(locs, function(loc) {
            nbrs <- nbr_list[[loc]]
            if (is.null(nbrs) || length(nbrs) == 0L) return(0)
            nbrs <- nbrs[nbrs %in% names(sv)]
            if (length(nbrs) == 0L) return(0)
            fun(sv[nbrs])
        })
    }
    feat <- feat |>
        mutate(
            nbr_prop_outbreak        = make_nbr_feat(prop_outbreak_years, location),
            nbr_mean_cases           = make_nbr_feat(mean_cases_per_year, location),
            nbr_recent_outbreak_prop = make_nbr_feat(recent_outbreak, location),
            nbr_recent_2yr_prop      = make_nbr_feat(
                as.numeric(recent_2yr_outbreaks > 0), location),
            nbr_max_cases            = make_nbr_feat(
                max_cases_any_year, location, fun = max),
            nbr_trend                = make_nbr_feat(trend, location)
        ) |>
        left_join(centroid_coords, by = "location")
    feat
}

ALL_FEATURES <- c(
    "n_outbreak_years", "prop_outbreak_years", "total_outbreaks",
    "mean_cases_per_year", "max_cases_any_year",
    "mean_duration", "max_duration",
    "recent_outbreak", "recent_2yr_outbreaks",
    "trend", "trend_cases", "years_since_last_outbreak",
    "nbr_prop_outbreak", "nbr_mean_cases",
    "nbr_recent_outbreak_prop", "nbr_recent_2yr_prop",
    "nbr_max_cases", "nbr_trend"
)
cat("Features:", length(ALL_FEATURES), "\n")
```

# Model Training and Prediction

## Build training and prediction datasets

```{r build-datasets}
# ── Training data ─────────────────────────────────────────────────────────────
# Features: 2010–2014  |  Outcome: any outbreak in 2015–2017
TRAIN_FEAT_END  <- 2014L
TRAIN_OUT_START <- 2015L
TRAIN_OUT_END   <- 2017L

feat_train <- build_features(TRAIN_FEAT_END)  # T=5: 2010-2014

outcome_train <- loc_year_full |>
    filter(year %in% TRAIN_OUT_START:TRAIN_OUT_END) |>
    group_by(location) |>
    summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")

train_df <- feat_train |>
    left_join(outcome_train, by = "location") |>
    replace_na(list(obs = 0L))

cat(sprintf("Training: features %d-%d → outcome %d-%d\n",
            TRAIN_FEAT_END - T_WINDOW + 1L, TRAIN_FEAT_END,
            TRAIN_OUT_START, TRAIN_OUT_END))
cat(sprintf("n = %d | prevalence = %.1f%%\n",
            nrow(train_df), 100 * mean(train_df$obs)))

# ── Prediction data ───────────────────────────────────────────────────────────
# Features: 2013–2017  |  Predict: outbreak probability in 2018–2020
PRED_FEAT_END  <- 2017L
PRED_OUT_START <- 2018L
PRED_OUT_END   <- 2020L

feat_pred <- build_features(PRED_FEAT_END)   # T=5: 2013-2017

outcome_pred <- loc_year_full |>
    filter(year %in% PRED_OUT_START:PRED_OUT_END) |>
    group_by(location) |>
    summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")

pred_df <- feat_pred |>
    left_join(outcome_pred, by = "location") |>
    replace_na(list(obs = 0L))

cat(sprintf("\nPrediction: features %d-%d → outcome %d-%d\n",
            PRED_FEAT_END - T_WINDOW + 1L, PRED_FEAT_END,
            PRED_OUT_START, PRED_OUT_END))
cat(sprintf("n = %d | prevalence = %.1f%%\n",
            nrow(pred_df), 100 * mean(pred_df$obs)))
```

## Fit models and predict

```{r fit-models}
# ── 1. Logistic regression ────────────────────────────────────────────────────
lr_fit <- glm(obs ~ ., data = train_df[, c("obs", ALL_FEATURES)],
              family = binomial)
pred_df$pred_logistic <- predict(lr_fit, newdata = pred_df, type = "response")

# ── 2. GAM ───────────────────────────────────────────────────────────────────
gam_feats <- c("n_outbreak_years", "mean_cases_per_year",
               "recent_2yr_outbreaks", "years_since_last_outbreak",
               "nbr_prop_outbreak", "nbr_mean_cases")

safe_k <- function(feat, dat, kmax = 5L)
    max(3L, min(kmax, length(unique(dat[[feat]])) - 1L))

gam_terms <- paste0(
    "s(", gam_feats, ", k = ",
    sapply(gam_feats, safe_k, dat = train_df), ")",
    collapse = " + "
)
gam_fit <- tryCatch(
    gam(as.formula(paste("obs ~", gam_terms)), data = train_df,
        family = binomial, method = "REML"),
    error = function(e) {
        k2 <- safe_k("n_outbreak_years", train_df, 3L)
        k3 <- safe_k("nbr_prop_outbreak", train_df, 3L)
        gam(as.formula(sprintf(
            "obs ~ s(n_outbreak_years, k=%d) + s(nbr_prop_outbreak, k=%d)", k2, k3)),
            data = train_df, family = binomial, method = "REML")
    }
)
pred_df$pred_gam <- as.numeric(predict(gam_fit, newdata = pred_df, type = "response"))

# ── 3. Random forest ─────────────────────────────────────────────────────────
set.seed(42)
rf_fit <- ranger(
    obs ~ ., data = train_df[, c("obs", ALL_FEATURES)] |> mutate(obs = factor(obs)),
    num.trees = 500, probability = TRUE,
    min.node.size = 5, mtry = max(1, floor(sqrt(length(ALL_FEATURES))))
)
rf_preds <- predict(rf_fit, data = pred_df)$predictions
pred_df$pred_rf <- if ("1" %in% colnames(rf_preds)) rf_preds[, "1"] else rep(0, nrow(pred_df))

cat("Models fitted. Prediction head:\n")
pred_df |>
    dplyr::select(location, obs, pred_logistic, pred_gam, pred_rf) |>
    slice_head(n = 5) |>
    knitr::kable(digits = 3)
```

## Model evaluation (2018–2020 held-out outcomes)

```{r eval-models}
# Evaluate against the actual 2018-2020 outcomes
eval_model <- function(obs, pred, name) {
    valid <- !is.na(pred) & !is.na(obs) & var(obs) > 0
    rho     <- cor(pred[valid], obs[valid], method = "spearman")
    auc_roc <- as.numeric(pROC::auc(pROC::roc(obs[valid], pred[valid], quiet = TRUE)))
    brier   <- mean((pred[valid] - obs[valid])^2)
    tibble(Model = name, Spearman_rho = rho, AUC_ROC = auc_roc, Brier = brier)
}

bind_rows(
    eval_model(pred_df$obs, pred_df$pred_logistic, "Logistic"),
    eval_model(pred_df$obs, pred_df$pred_gam,      "GAM"),
    eval_model(pred_df$obs, pred_df$pred_rf,       "Random Forest")
) |>
    knitr::kable(digits = 3,
                 caption = sprintf("Model evaluation: outcome %d-%d, features %d-%d",
                                   PRED_OUT_START, PRED_OUT_END,
                                   PRED_FEAT_END - T_WINDOW + 1L, PRED_FEAT_END))
```

# Outbreak Risk Estimation

## Ensemble mean risk score

We use the ensemble mean of the three model predictions as the risk score.

```{r risk-scores}
pred_risk <- pred_df |>
    mutate(
        risk_ens = (pred_logistic + pred_gam + pred_rf) / 3
    ) |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(risk_ens), !is.na(pop_sum), pop_sum > 0) |>
    arrange(desc(risk_ens))

cat(sprintf("Locations with risk scores: %d\n", nrow(pred_risk)))
cat(sprintf("Total population: %.0f million\n", sum(pred_risk$pop_sum) / 1e6))
cat(sprintf("Mean ensemble risk: %.4f\n", mean(pred_risk$risk_ens)))
```

## Population vs. cumulative outbreak risk

```{r fig-pop-risk-cdf, fig.width=12, fig.height=5}
#| fig-cap: "Left: cumulative population (millions) living in regions with predicted
#|   outbreak risk ≤ x (ECDF). Right: number of ADM2 regions with risk ≤ x.
#|   The dashed vertical line marks the risk threshold above which ~200 million
#|   people reside. Individual model scores shown alongside ensemble mean."

# Build ECDFs for ensemble + individual models
pred_long <- pred_risk |>
    dplyr::select(location, pop_sum, obs,
                  Ensemble = risk_ens, Logistic = pred_logistic,
                  GAM = pred_gam, `Random Forest` = pred_rf) |>
    pivot_longer(c(Ensemble, Logistic, GAM, `Random Forest`),
                 names_to = "model", values_to = "risk")

# Empirical CDF by model
make_cdf <- function(df) {
    df |>
        arrange(risk) |>
        mutate(
            cum_pop   = cumsum(pop_sum),
            cum_locs  = row_number(),
            total_pop = sum(pop_sum),
            n_total   = n()
        )
}

cdf_all <- pred_long |>
    group_by(model) |>
    group_modify(~make_cdf(.x)) |>
    ungroup()

# 200M cutoff using ensemble
ens_sorted <- pred_risk |> arrange(risk_ens)
cum_pop_ens <- cumsum(ens_sorted$pop_sum)
n_cut <- which(cum_pop_ens >= 200e6)[1]
cutoff_200m <- ens_sorted$risk_ens[n_cut]

MODEL_COLOURS <- c(
    Ensemble = "black", Logistic = "steelblue",
    GAM = "darkgoldenrod", `Random Forest` = "darkorchid"
)

p_pop <- ggplot(cdf_all, aes(x = risk, y = cum_pop / 1e6, colour = model,
                              linewidth = model == "Ensemble")) +
    geom_line() +
    geom_vline(xintercept = cutoff_200m, linetype = "dashed",
               colour = "firebrick", linewidth = 0.7) +
    annotate("text", x = cutoff_200m + 0.02,
             y = max(cdf_all$cum_pop / 1e6) * 0.3,
             label = sprintf("p = %.3f\n200M threshold", cutoff_200m),
             hjust = 0, colour = "firebrick", size = 3.2) +
    scale_colour_manual(values = MODEL_COLOURS) +
    scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6),
                           guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(x = "Predicted outbreak probability (2018–2020)",
         y = "Cumulative population (millions)",
         colour = "Model",
         title = "Population ≤ risk threshold") +
    theme_minimal(base_size = 11)

p_loc <- ggplot(cdf_all, aes(x = risk, y = cum_locs, colour = model,
                              linewidth = model == "Ensemble")) +
    geom_line() +
    geom_vline(xintercept = cutoff_200m, linetype = "dashed",
               colour = "firebrick", linewidth = 0.7) +
    scale_colour_manual(values = MODEL_COLOURS) +
    scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6),
                           guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(x = "Predicted outbreak probability (2018–2020)",
         y = "Cumulative number of ADM2 regions",
         colour = "Model",
         title = "ADM2 regions ≤ risk threshold") +
    theme_minimal(base_size = 11)

p_pop + p_loc +
    plot_annotation(
        title = sprintf(
            "Outbreak risk distribution — trained on 2010–2014 features / 2015–2017 outcomes"),
        subtitle = sprintf(
            "Predicted for 2018–2020 | %d ADM2 regions, %.0fM total population",
            nrow(pred_risk), sum(pred_risk$pop_sum) / 1e6)
    ) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

## Risk summary table

```{r risk-summary-table}
# Population above selected risk thresholds
thresholds <- c(0.05, 0.10, 0.20, 0.30, 0.50)
map_dfr(thresholds, function(thr) {
    above <- pred_risk |> filter(risk_ens > thr)
    tibble(
        threshold   = thr,
        n_regions   = nrow(above),
        pop_M       = sum(above$pop_sum) / 1e6,
        pct_regions = 100 * nrow(above) / nrow(pred_risk),
        pct_pop     = 100 * sum(above$pop_sum) / sum(pred_risk$pop_sum),
        obs_rate    = if (nrow(above) > 0) mean(above$obs) else NA_real_
    )
}) |>
    knitr::kable(
        digits = c(2, 0, 1, 1, 1, 3),
        col.names = c("Risk > X", "Regions", "Pop (M)",
                      "% regions", "% pop", "Outbreak rate (2018-20)"),
        caption = "Population and region count above ensemble risk thresholds"
    )
```

## SSA map of predicted risk (2018–2020)

```{r fig-risk-map, fig.width=10, fig.height=8}
#| fig-cap: "Ensemble mean predicted outbreak risk (2018–2020) at ADM2 level.
#|   Grey = sub-Saharan Africa countries. Red outlines = locations with observed
#|   outbreak in 2018–2020."

ssa_sf <- sf::st_as_sf(rnaturalearthdata::countries50) |>
    dplyr::filter(subregion %in% c("Eastern Africa", "Middle Africa",
                                    "Southern Africa", "Western Africa")) |>
    sf::st_make_valid() |>
    sf::st_transform(sf::st_crs(shp_admin23))

shp_risk <- shp_admin23 |>
    left_join(pred_risk |> dplyr::select(location, risk_ens, obs),
              by = "location") |>
    filter(!is.na(risk_ens))

ssa_bbox <- sf::st_bbox(ssa_sf)

ggplot() +
    geom_sf(data = ssa_sf, fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = shp_risk, aes(fill = risk_ens), colour = NA) +
    geom_sf(data = shp_risk |> filter(obs == 1),
            fill = NA, colour = "firebrick", linewidth = 0.4) +
    scale_fill_viridis_c(option = "magma", direction = -1,
                         name = "Predicted\nrisk", limits = c(0, 1)) +
    coord_sf(xlim = ssa_bbox[c("xmin", "xmax")],
             ylim = ssa_bbox[c("ymin", "ymax")]) +
    labs(title = "Predicted outbreak risk 2018–2020",
         subtitle = "Trained on 2010–2014 features / 2015–2017 outcomes | Red outline = observed outbreak") +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank(), axis.text = element_blank(),
          axis.ticks = element_blank(), legend.position = "right")
```
