---
title: "Outbreak Risk Prediction: Simple One-Step Model"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

A single-step prediction exercise with a **k=2** outcome window:

- **Training**: three temporal folds pooled
  - Features 2010–2014 → outcome 2015–2016
  - Features 2011–2015 → outcome 2016–2017
  - Features 2012–2016 → outcome 2017–2018
- **Prediction**: features built from 2014–2018 surveillance data; predict
  probability of any outbreak in 2019–2020
- **Models**: logistic regression, GAM, random forest
- **Output**: predicted risk per ADM2 unit; cumulative population vs. risk curve

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(patchwork)
library(pROC)
library(scales)
library(knitr)
library(mgcv)    # GAM
library(ranger)  # random forest
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```

# Data

## Outbreak surveillance

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)
ts23 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm %in% c(2, 3))

outbreak_events <- ts23 |>
    group_by(location, adm, outbreak_number) |>
    summarise(
        year_start     = year(min(TL)),
        year_end       = year(max(TR)),
        duration_weeks = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
        total_cases    = sum(sCh, na.rm = TRUE),
        .groups        = "drop"
    )

all_years <- 2010:2023
all_locs  <- unique(ts23$location)

outbreak_year_summary <- outbreak_events |>
    filter(year_end %in% all_years) |>
    group_by(location, year = year_end) |>
    summarise(
        outbreak_count    = n(),
        total_cases       = sum(total_cases, na.rm = TRUE),
        max_duration_days = max(duration_weeks * 7, na.rm = TRUE),
        mean_duration_days = mean(duration_weeks * 7, na.rm = TRUE),
        .groups = "drop"
    ) |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(outbreak_count = 0L, total_cases = 0,
                    max_duration_days = 0, mean_duration_days = 0))

# loc_year_full: outbreaks active in each year
loc_year_full <- outbreak_events |>
    mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) |>
    unnest(years) |>
    rename(year = years) |>
    filter(year %in% all_years) |>
    count(location, year, name = "n_active") |>
    right_join(expand_grid(location = all_locs, year = all_years),
               by = c("location", "year")) |>
    replace_na(list(n_active = 0L))

cat("Locations:", length(all_locs), "| Years:", range(all_years), "\n")
```

## Shapefiles and spatial neighbours

```{r load-shapefiles}
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)
loc_id_map <- ts23 |>
    distinct(location_period_id, location) |>
    group_by(location) |>
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) |>
    ungroup()

shp_admin23 <- shp_all |>
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id")) |>
    filter(!is.na(location), str_count(location, "::") %in% c(3, 4))

cat("ADM2/3 polygons:", nrow(shp_admin23), "\n")

# 50 km neighbour list
DIST_KM <- 50
centroids_sf <- shp_admin23 |> st_make_valid() |> st_point_on_surface()
dist_mat <- st_distance(centroids_sf) |> units::drop_units() / 1000
rownames(dist_mat) <- colnames(dist_mat) <- shp_admin23$location
diag(dist_mat) <- Inf

nbr_list <- setNames(
    lapply(shp_admin23$location, function(loc) {
        d <- dist_mat[loc, ]
        names(sort(d[d <= DIST_KM]))
    }),
    shp_admin23$location
)

centroid_coords <- centroids_sf |>
    st_coordinates() |>
    as_tibble() |>
    rename(lon = X, lat = Y) |>
    mutate(location = shp_admin23$location)
```

## Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
pop_raster <- terra::rast(pop_raster)

POP_SCALE <- 344  # pixel area ~344 km²
shp_pop   <- shp_admin23 |> st_make_valid() |> st_transform(crs(pop_raster))
pop_vals  <- exactextractr::exact_extract(pop_raster, shp_pop,
                                          fun = "sum", progress = FALSE)
shp_admin23$pop_sum <- pop_vals * POP_SCALE

pop_lookup <- shp_admin23 |>
    sf::st_drop_geometry() |>
    dplyr::select(location, pop_sum)

cat("Total population:", format(round(sum(pop_vals * POP_SCALE, na.rm = TRUE)),
                                big.mark = ","), "\n")
```

# Feature Engineering

Features are built from a T=5-year rolling window ending at a given year.
Twelve same-unit features and six neighbour-aggregated features are used.

```{r build-features-fn}
T_WINDOW <- 5L

build_features <- function(year_end, T = T_WINDOW) {
    feat_yrs <- (year_end - T + 1L):year_end
    n_yrs    <- length(feat_yrs)

    loc_summary <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            n_outbreak_years    = sum(outbreak_count > 0),
            prop_outbreak_years = n_outbreak_years / n_yrs,
            total_outbreaks     = sum(outbreak_count),
            mean_cases_per_year = mean(total_cases),
            max_cases_any_year  = max(total_cases),
            mean_duration       = mean(mean_duration_days),
            max_duration        = max(max_duration_days),
            .groups = "drop"
        )

    recent_1yr <- outbreak_year_summary |>
        filter(year == year_end) |>
        transmute(location, recent_outbreak = as.integer(outbreak_count > 0))

    recent_2yr <- outbreak_year_summary |>
        filter(year %in% intersect(c(year_end - 1L, year_end), feat_yrs)) |>
        group_by(location) |>
        summarise(recent_2yr_outbreaks = sum(outbreak_count > 0), .groups = "drop")

    trend_df <- outbreak_year_summary |>
        filter(year %in% feat_yrs) |>
        group_by(location) |>
        summarise(
            trend = if (n() >= 2)
                coef(lm(as.integer(outbreak_count > 0) ~ year))[2] else 0,
            trend_cases = if (n() >= 2)
                coef(lm(total_cases ~ year))[2] else 0,
            .groups = "drop"
        )

    last_ob <- outbreak_year_summary |>
        filter(year %in% feat_yrs, outbreak_count > 0) |>
        group_by(location) |>
        summarise(last_ob_year = max(year), .groups = "drop")

    yrs_since <- tibble(location = all_locs) |>
        left_join(last_ob, by = "location") |>
        mutate(years_since_last_outbreak = ifelse(
            is.na(last_ob_year), n_yrs + 1L,
            as.integer(year_end - last_ob_year)
        )) |>
        dplyr::select(location, years_since_last_outbreak)

    feat <- tibble(location = all_locs) |>
        left_join(loc_summary,  by = "location") |>
        left_join(recent_1yr,   by = "location") |>
        left_join(recent_2yr,   by = "location") |>
        left_join(trend_df,     by = "location") |>
        left_join(yrs_since,    by = "location") |>
        replace_na(list(
            n_outbreak_years = 0L, prop_outbreak_years = 0,
            total_outbreaks = 0L, mean_cases_per_year = 0,
            max_cases_any_year = 0, mean_duration = 0, max_duration = 0,
            recent_outbreak = 0L, recent_2yr_outbreaks = 0L,
            trend = 0, trend_cases = 0
        ))

    # Neighbour features
    make_nbr_feat <- function(values, locs, fun = mean) {
        sv <- setNames(values, locs)
        map_dbl(locs, function(loc) {
            nbrs <- nbr_list[[loc]]
            if (is.null(nbrs) || length(nbrs) == 0L) return(0)
            nbrs <- nbrs[nbrs %in% names(sv)]
            if (length(nbrs) == 0L) return(0)
            fun(sv[nbrs])
        })
    }
    feat <- feat |>
        mutate(
            nbr_prop_outbreak        = make_nbr_feat(prop_outbreak_years, location),
            nbr_mean_cases           = make_nbr_feat(mean_cases_per_year, location),
            nbr_recent_outbreak_prop = make_nbr_feat(recent_outbreak, location),
            nbr_recent_2yr_prop      = make_nbr_feat(
                as.numeric(recent_2yr_outbreaks > 0), location),
            nbr_max_cases            = make_nbr_feat(
                max_cases_any_year, location, fun = max),
            nbr_trend                = make_nbr_feat(trend, location)
        ) |>
        left_join(centroid_coords, by = "location")
    feat
}

ALL_FEATURES <- c(
    "n_outbreak_years", "prop_outbreak_years", "total_outbreaks",
    "mean_cases_per_year", "max_cases_any_year",
    "mean_duration", "max_duration",
    "recent_outbreak", "recent_2yr_outbreaks",
    "trend", "trend_cases", "years_since_last_outbreak",
    "nbr_prop_outbreak", "nbr_mean_cases",
    "nbr_recent_outbreak_prop", "nbr_recent_2yr_prop",
    "nbr_max_cases", "nbr_trend"
)
cat("Features:", length(ALL_FEATURES), "\n")
```

# Model Training and Prediction

## Build training and prediction datasets

```{r build-datasets}
# ── k=2 outcome window: three training folds ──────────────────────────────────
build_fold <- function(feat_end, out_start, out_end) {
    feats   <- build_features(feat_end)
    outcome <- loc_year_full |>
        filter(year %in% out_start:out_end) |>
        group_by(location) |>
        summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")
    feats |>
        left_join(outcome, by = "location") |>
        replace_na(list(obs = 0L))
}

fold1 <- build_fold(2014L, 2015L, 2016L)   # features 2010-2014 → outcome 2015-2016
fold2 <- build_fold(2015L, 2016L, 2017L)   # features 2011-2015 → outcome 2016-2017
fold3 <- build_fold(2016L, 2017L, 2018L)   # features 2012-2016 → outcome 2017-2018

train_df <- bind_rows(fold1, fold2, fold3)

cat("Training folds (k=2):\n")
cat(sprintf("  Fold 1: features 2010-2014 → outcome 2015-2016 | n=%d | prev=%.1f%%\n",
            nrow(fold1), 100 * mean(fold1$obs)))
cat(sprintf("  Fold 2: features 2011-2015 → outcome 2016-2017 | n=%d | prev=%.1f%%\n",
            nrow(fold2), 100 * mean(fold2$obs)))
cat(sprintf("  Fold 3: features 2012-2016 → outcome 2017-2018 | n=%d | prev=%.1f%%\n",
            nrow(fold3), 100 * mean(fold3$obs)))
cat(sprintf("Combined: n=%d | overall prevalence=%.1f%%\n",
            nrow(train_df), 100 * mean(train_df$obs)))

# ── Prediction data ───────────────────────────────────────────────────────────
# Features: 2014–2018  |  Predict: outbreak probability in 2019–2020
PRED_FEAT_END  <- 2018L
PRED_OUT_START <- 2019L
PRED_OUT_END   <- 2020L

feat_pred <- build_features(PRED_FEAT_END)   # T=5: 2014-2018

outcome_pred <- loc_year_full |>
    filter(year %in% PRED_OUT_START:PRED_OUT_END) |>
    group_by(location) |>
    summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")

pred_df <- feat_pred |>
    left_join(outcome_pred, by = "location") |>
    replace_na(list(obs = 0L))

cat(sprintf("\nPrediction: features %d-%d → outcome %d-%d\n",
            PRED_FEAT_END - T_WINDOW + 1L, PRED_FEAT_END,
            PRED_OUT_START, PRED_OUT_END))
cat(sprintf("n = %d | prevalence = %.1f%%\n",
            nrow(pred_df), 100 * mean(pred_df$obs)))
```

## Fit models and predict

```{r fit-models}
# ── 1. Logistic regression ────────────────────────────────────────────────────
lr_fit <- glm(obs ~ ., data = train_df[, c("obs", ALL_FEATURES)],
              family = binomial)
pred_df$pred_logistic <- predict(lr_fit, newdata = pred_df, type = "response")

# ── 2. GAM ───────────────────────────────────────────────────────────────────
gam_feats <- c("n_outbreak_years", "mean_cases_per_year",
               "recent_2yr_outbreaks", "years_since_last_outbreak",
               "nbr_prop_outbreak", "nbr_mean_cases")

safe_k <- function(feat, dat, kmax = 5L)
    max(3L, min(kmax, length(unique(dat[[feat]])) - 1L))

gam_terms <- paste0(
    "s(", gam_feats, ", k = ",
    sapply(gam_feats, safe_k, dat = train_df), ")",
    collapse = " + "
)
gam_fit <- tryCatch(
    gam(as.formula(paste("obs ~", gam_terms)), data = train_df,
        family = binomial, method = "REML"),
    error = function(e) {
        k2 <- safe_k("n_outbreak_years", train_df, 3L)
        k3 <- safe_k("nbr_prop_outbreak", train_df, 3L)
        gam(as.formula(sprintf(
            "obs ~ s(n_outbreak_years, k=%d) + s(nbr_prop_outbreak, k=%d)", k2, k3)),
            data = train_df, family = binomial, method = "REML")
    }
)
pred_df$pred_gam <- as.numeric(predict(gam_fit, newdata = pred_df, type = "response"))

# ── 3. Random forest ─────────────────────────────────────────────────────────
set.seed(42)
rf_fit <- ranger(
    obs ~ ., data = train_df[, c("obs", ALL_FEATURES)] |> mutate(obs = factor(obs)),
    num.trees = 500, probability = TRUE,
    min.node.size = 5, mtry = max(1, floor(sqrt(length(ALL_FEATURES))))
)
rf_preds <- predict(rf_fit, data = pred_df)$predictions
pred_df$pred_rf <- if ("1" %in% colnames(rf_preds)) rf_preds[, "1"] else rep(0, nrow(pred_df))

cat("Models fitted. Prediction head:\n")
pred_df |>
    dplyr::select(location, obs, pred_logistic, pred_gam, pred_rf) |>
    slice_head(n = 5) |>
    knitr::kable(digits = 3)
```

## Model evaluation (2019–2020 held-out outcomes)

```{r eval-models}
# Evaluate against the actual 2019-2020 outcomes
eval_model <- function(obs, pred, name) {
    valid <- !is.na(pred) & !is.na(obs) & var(obs) > 0
    rho     <- cor(pred[valid], obs[valid], method = "spearman")
    auc_roc <- as.numeric(pROC::auc(pROC::roc(obs[valid], pred[valid], quiet = TRUE)))
    brier   <- mean((pred[valid] - obs[valid])^2)
    tibble(Model = name, Spearman_rho = rho, AUC_ROC = auc_roc, Brier = brier)
}

bind_rows(
    eval_model(pred_df$obs, pred_df$pred_logistic, "Logistic"),
    eval_model(pred_df$obs, pred_df$pred_gam,      "GAM"),
    eval_model(pred_df$obs, pred_df$pred_rf,       "Random Forest")
) |>
    knitr::kable(digits = 3,
                 caption = sprintf("Model evaluation (k=2): outcome %d-%d, features %d-%d",
                                   PRED_OUT_START, PRED_OUT_END,
                                   PRED_FEAT_END - T_WINDOW + 1L, PRED_FEAT_END))
```

# Targeting Accuracy (Rho) and Optimal Alpha

## Rho on training folds and out-of-sample

In-sample $\rho$ is computed by applying the pooled-fold models back to all
three training folds (2010–2016 features vs. 2015–2018 outcomes). Out-of-sample
$\rho$ is computed on the 2019–2020 prediction period.

```{r rho-training}
# In-sample: apply models to combined training data (all 3 folds)
train_preds <- train_df
train_preds$pred_logistic <- predict(lr_fit,  newdata = train_df, type = "response")
train_preds$pred_gam      <- as.numeric(predict(gam_fit, newdata = train_df, type = "response"))
rf_train <- predict(rf_fit, data = train_df)$predictions
train_preds$pred_rf  <- if ("1" %in% colnames(rf_train)) rf_train[, "1"] else rep(0, nrow(train_df))
train_preds$pred_ens <- (train_preds$pred_logistic + train_preds$pred_gam + train_preds$pred_rf) / 3

rho_train <- tibble(
    Model = c("Logistic", "GAM", "Random Forest", "Ensemble"),
    period = "Training (all folds, k=2)",
    rho = c(
        cor(train_preds$pred_logistic, train_preds$obs, method = "spearman"),
        cor(train_preds$pred_gam,      train_preds$obs, method = "spearman"),
        cor(train_preds$pred_rf,       train_preds$obs, method = "spearman"),
        cor(train_preds$pred_ens,      train_preds$obs, method = "spearman")
    )
)

# Out-of-sample rho (features 2014-2018 → outcome 2019-2020)
pred_ens_oos <- (pred_df$pred_logistic + pred_df$pred_gam + pred_df$pred_rf) / 3
rho_oos <- tibble(
    Model = c("Logistic", "GAM", "Random Forest", "Ensemble"),
    period = sprintf("Out-of-sample (%d-%d)", PRED_OUT_START, PRED_OUT_END),
    rho = c(
        cor(pred_df$pred_logistic, pred_df$obs, method = "spearman"),
        cor(pred_df$pred_gam,      pred_df$obs, method = "spearman"),
        cor(pred_df$pred_rf,       pred_df$obs, method = "spearman"),
        cor(pred_ens_oos,          pred_df$obs, method = "spearman")
    )
)

rho_table <- bind_rows(rho_train, rho_oos) |>
    pivot_wider(names_from = period, values_from = rho)

rho_table |>
    knitr::kable(digits = 3,
                 caption = sprintf("Spearman rho: in-sample (training folds) vs out-of-sample (%d-%d)",
                                   PRED_OUT_START, PRED_OUT_END))

# Best model by out-of-sample rho (used as targeting score in MC below)
best_model_name <- rho_oos |> slice_max(rho, n = 1) |> pull(Model)
rho_best_oos    <- rho_oos |> filter(Model == best_model_name) |> pull(rho)
cat(sprintf("\nBest model (OOS): %s  rho = %.3f\n", best_model_name, rho_best_oos))
```

## MC helper functions

```{r mc-helpers}
# ── Beta-Binomial MLE ──────────────────────────────────────────────────────
fit_betabinom <- function(k_obs, n_trials) {
    ll <- function(par, k, n) {
        a <- exp(par[1]); b <- exp(par[2])
        sum(lchoose(n, k) + lbeta(k + a, n - k + b) - lbeta(a, b))
    }
    p_init <- mean(k_obs) / n_trials
    v_init <- var(k_obs / n_trials)
    a_init <- max(p_init * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    b_init <- max((1 - p_init) * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    fit <- optim(par = log(c(a_init, b_init)), fn = ll, k = k_obs, n = n_trials,
                 method = "Nelder-Mead", control = list(fnscale = -1, maxit = 5000))
    list(a = exp(fit$par[1]), b = exp(fit$par[2]),
         p = exp(fit$par[1]) / (exp(fit$par[1]) + exp(fit$par[2])),
         convergence = fit$convergence)
}

# ── Population-weighted MC cost function ──────────────────────────────────
# Sweeps alpha on a grid; at each alpha:
#   1. Rank locations by score (descending); fill pre-emptive budget alpha*f*P
#   2. Draw Bernoulli outbreak for each location using obs_rate
#   3. Reactive budget (1-alpha)*f*P covers outbreak locations not pre-vaccinated
#   4. Return (vaccination cost + illness cost) / P_total
cost_popweighted_mc <- function(alpha, f, pop, prob, score, R, r, AR, nu = 1,
                                n_mc = 300, seed = 42) {
    set.seed(seed)
    n <- length(pop)
    P <- sum(pop)
    pre_budget   <- alpha * f * P
    react_budget <- (1 - alpha) * f * P

    # Pre-emptive allocation: rank by score, fill budget
    ord <- order(score, decreasing = TRUE)
    cum_pre <- cumsum(pop[ord])
    n_pre <- sum(cum_pre <= pre_budget)
    pre_locs <- ord[seq_len(max(n_pre, 0L))]

    total_cost <- mean(replicate(n_mc, {
        outbreaks <- rbinom(n, 1L, prob)
        ob_locs   <- which(outbreaks == 1L & !seq_len(n) %in% pre_locs)

        # Reactive: cover outbreak locs not pre-vaccinated, in pop-descending order
        if (length(ob_locs) > 0 && react_budget > 0) {
            ob_ord <- ob_locs[order(pop[ob_locs], decreasing = TRUE)]
            cum_react <- cumsum(pop[ob_ord])
            n_react <- sum(cum_react <= react_budget)
            react_locs <- ob_ord[seq_len(max(n_react, 0L))]
        } else {
            react_locs <- integer(0)
        }

        # Cost: vaccination + illness
        cost_vax     <- (sum(pop[pre_locs]) + sum(pop[react_locs])) / P
        ill_pre      <- sum(pop[intersect(pre_locs, which(outbreaks == 1L))]) *
                            (1 - nu) * AR * R / P
        ill_react    <- sum(pop[intersect(react_locs, which(outbreaks == 1L))]) *
                            (1 - r) * AR * R / P
        ob_unvax_locs <- setdiff(which(outbreaks == 1L),
                                  union(pre_locs, react_locs))
        ill_unvax    <- sum(pop[ob_unvax_locs]) * AR * R / P
        cost_vax + ill_pre + ill_react + ill_unvax
    }))
    total_cost
}

# ── Optimal alpha via grid search ─────────────────────────────────────────
get_alpha_star_pw <- function(f, pop, prob, score, R, r, AR, nu = 1,
                               n_mc = 300, n_grid = 51) {
    alpha_grid <- seq(0, 1, length.out = n_grid)
    costs <- sapply(alpha_grid, function(a)
        cost_popweighted_mc(a, f, pop, prob, score, R, r, AR, nu, n_mc = n_mc))
    list(alpha_star = alpha_grid[which.min(costs)],
         cost_star  = min(costs),
         alpha_grid = alpha_grid,
         costs      = costs)
}

# ── Classifier performance helpers ────────────────────────────────────────
# All take continuous score and binary outcome (0/1)

roc_curve <- function(score, binary) {
    n <- length(score)
    total_pos <- sum(binary); total_neg <- n - total_pos
    ord <- order(score, decreasing = TRUE)
    tp  <- cumsum(binary[ord])
    fp  <- seq_len(n) - tp
    tibble(fpr = c(0, fp / total_neg, 1),
           tpr = c(0, tp / total_pos, 1))
}

auc_roc_fn <- function(score, binary) {
    n1 <- sum(binary); n0 <- length(binary) - n1
    if (n1 == 0 || n0 == 0) return(NA_real_)
    suppressWarnings(
        wilcox.test(score[binary == 1], score[binary == 0], exact = FALSE)
    )$statistic / (n1 * n0)
}

auc_pr_fn <- function(score, binary) {
    total_pos <- sum(binary)
    if (total_pos == 0) return(NA_real_)
    ord  <- order(score, decreasing = TRUE)
    tp   <- cumsum(binary[ord])
    prec <- tp / seq_len(length(score))
    rec  <- tp / total_pos
    sum(diff(c(0, rec)) * prec)
}

classifier_metrics <- function(score, binary, label = "Ensemble") {
    # Optimal threshold via Youden index (grid search)
    thresh_grid <- seq(0, 1, length.out = 201)
    youden <- sapply(thresh_grid, function(t) {
        pred <- score >= t
        tp <- sum(pred & binary); fn <- sum(!pred & binary)
        tn <- sum(!pred & !binary); fp <- sum(pred & !binary)
        if ((tp + fn) == 0 || (tn + fp) == 0) return(-Inf)
        tp / (tp + fn) + tn / (tn + fp) - 1
    })
    best_t <- thresh_grid[which.max(youden)]
    pred   <- score >= best_t
    tp <- sum(pred & binary); fn <- sum(!pred & binary)
    tn <- sum(!pred & !binary); fp <- sum(pred & !binary)
    tibble(
        Model       = label,
        rho         = cor(score, binary, method = "spearman"),
        AUC_ROC     = auc_roc_fn(score, binary),
        AUC_PR      = auc_pr_fn(score, binary),
        Brier       = mean((score - binary)^2),
        Sensitivity = if ((tp + fn) > 0) tp / (tp + fn) else NA_real_,
        Specificity = if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
    )
}

# ── Synthetic score with target Spearman rho ──────────────────────────────
# Uses Gaussian copula: creates a ranking score correlating with truth at
# approximately target_rho (Spearman). Useful for "what if rho = 2x" sensitivity.
make_rank_score <- function(truth, target_rho, seed = 123) {
    set.seed(seed)
    n        <- length(truth)
    z_truth  <- qnorm(rank(truth, ties.method = "random") / (n + 1))
    z_noise  <- rnorm(n)
    rho_p    <- 2 * sin(pi * target_rho / 6)  # Gaussian copula: Spearman → Pearson
    z_blend  <- rho_p * z_truth + sqrt(1 - rho_p^2) * z_noise
    pnorm(z_blend)
}
```

## Beta-Binomial fit on 2018–2020 predicted risks

```{r bb-fit}
# Use 2-year outcome window (PRED_OUT_START:PRED_OUT_END); n_trials = 2
# k_obs = number of outbreak years per location in the prediction window
N_TRIALS <- 2L

# Merge predicted risks with population and 2018-2020 outcomes
sim_df <- pred_df |>
    mutate(risk_ens = (pred_logistic + pred_gam + pred_rf) / 3) |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(risk_ens), !is.na(pop_sum), pop_sum > 0)

# n_outbreak_years = count of years in 2018-2020 with active outbreak
ob_counts <- loc_year_full |>
    filter(year %in% PRED_OUT_START:PRED_OUT_END) |>
    group_by(location) |>
    summarise(n_outbreak_years = sum(n_active > 0), .groups = "drop")

sim_df <- sim_df |>
    dplyr::select(-any_of("n_outbreak_years")) |>
    left_join(ob_counts, by = "location") |>
    replace_na(list(n_outbreak_years = 0L)) |>
    mutate(obs_rate   = n_outbreak_years / N_TRIALS,
           obs_binary = as.integer(n_outbreak_years > 0),
           country    = str_split_fixed(location, "::", 4)[, 2])

cat(sprintf("Simulation data: %d locations, %.0f M population\n",
            nrow(sim_df), sum(sim_df$pop_sum) / 1e6))
cat(sprintf("Mean obs_rate (3-yr): %.4f\n", mean(sim_df$obs_rate)))

# Beta-Binomial MLE
bb <- fit_betabinom(sim_df$n_outbreak_years, N_TRIALS)
cat(sprintf("Beta-Binomial: a=%.4f, b=%.4f, mean p=%.4f (convergence=%d)\n",
            bb$a, bb$b, bb$p, bb$convergence))
```

## Beta-Binomial Diagnostics and Classifier Performance

```{r bb-diagnostics, fig.width=12, fig.height=5}
#| fig-cap: "Left: empirical ECDF (orange) vs fitted Beta(a,b) CDF (dashed red) for
#|   per-location outbreak rates. Right: observed outbreak-year count distribution
#|   (orange) vs Beta-Binomial PMF (red) for k = 0 … N_TRIALS."

# ── CDF comparison ────────────────────────────────────────────────────────
x_seq <- seq(0, 1, length.out = 300)
ecdf_fn <- ecdf(sim_df$obs_rate)
cdf_df <- tibble(
    x           = x_seq,
    empirical   = ecdf_fn(x_seq),
    theoretical = pbeta(x_seq, bb$a, bb$b)
)

p_cdf <- ggplot(cdf_df, aes(x = x)) +
    geom_line(aes(y = empirical),   colour = "darkorange",  linewidth = 0.9) +
    geom_line(aes(y = theoretical), colour = "firebrick",   linewidth = 0.9,
              linetype = "dashed") +
    labs(x = "Outbreak rate", y = "CDF",
         title = sprintf("CDF vs Beta(%.2f, %.2f)  (k=%d)", bb$a, bb$b, N_TRIALS)) +
    theme_minimal(base_size = 11)

# ── PMF comparison ────────────────────────────────────────────────────────
bb_pmf <- sapply(0:N_TRIALS, function(k)
    exp(lchoose(N_TRIALS, k) + lbeta(k + bb$a, N_TRIALS - k + bb$b) - lbeta(bb$a, bb$b))
)
obs_counts <- table(factor(sim_df$n_outbreak_years, levels = 0:N_TRIALS))
pmf_df <- tibble(
    k        = 0:N_TRIALS,
    observed = as.numeric(obs_counts) / nrow(sim_df),
    fitted   = bb_pmf
) |> pivot_longer(c(observed, fitted), names_to = "type", values_to = "prob")

p_pmf <- ggplot(pmf_df, aes(x = k, y = prob, fill = type)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = c(observed = "darkorange", fitted = "firebrick"),
                      labels = c(observed = "Observed", fitted = "BB PMF")) +
    labs(x = "Outbreak years", y = "Proportion", fill = NULL,
         title = sprintf("BB PMF  (k=%d)", N_TRIALS)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

p_cdf + p_pmf
```

```{r classifier-metrics}
# Binary outcome: any outbreak in 2018-2020 window
metrics_df <- bind_rows(
    classifier_metrics(sim_df$risk_ens,      sim_df$obs_binary, "Ensemble"),
    classifier_metrics(sim_df$pred_logistic, sim_df$obs_binary, "Logistic"),
    classifier_metrics(sim_df$pred_gam,      sim_df$obs_binary, "GAM"),
    classifier_metrics(sim_df$pred_rf,       sim_df$obs_binary, "Random Forest")
)

metrics_df |>
    knitr::kable(digits = 3,
                 caption = "Classifier performance (binary outcome: any outbreak 2018-2020)")
```

```{r fig-roc-curves, fig.width=8, fig.height=5}
#| fig-cap: "ROC curves for each model. Diagonal dashed line = random classifier."

roc_df <- bind_rows(
    roc_curve(sim_df$risk_ens,      sim_df$obs_binary) |> mutate(model = "Ensemble"),
    roc_curve(sim_df$pred_logistic, sim_df$obs_binary) |> mutate(model = "Logistic"),
    roc_curve(sim_df$pred_gam,      sim_df$obs_binary) |> mutate(model = "GAM"),
    roc_curve(sim_df$pred_rf,       sim_df$obs_binary) |> mutate(model = "Random Forest")
)

auc_labels <- metrics_df |>
    transmute(model = Model,
              label = sprintf("%s (AUC=%.3f)", Model, AUC_ROC))

roc_df <- roc_df |> left_join(auc_labels, by = "model")

ggplot(roc_df, aes(x = fpr, y = tpr, colour = label)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.8) +
    scale_colour_manual(values = c(
        setNames(c("black","steelblue","darkgoldenrod","darkorchid"),
                 auc_labels$label))) +
    labs(x = "False positive rate", y = "True positive rate",
         colour = NULL, title = "ROC curves — outbreak prediction 2018-2020") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

## Optimal alpha via Monte Carlo

Parameters: **R = 500** (cost ratio illness/vaccine), **ν = 1** (pre-emptive VE).
Coverage $f \in \{0.5,\, 0.75\}$, reactive effectiveness $r \in \{0.2,\, 0.4\}$.
An additional "Ensemble (2×ρ)" scenario doubles the ensemble's out-of-sample
targeting accuracy to explore the sensitivity to improved risk intelligence.

```{r mc-alpha-star}
R_val       <- 500
nu_val      <- 1
AR_val      <- 0.005
f_values_mc <- c(0.5, 0.75)
r_values    <- c(0.2, 0.4)

# Ensemble OOS rho (used to build the 2x-rho synthetic score)
rho_ens_oos <- rho_oos |> filter(Model == "Ensemble") |> pull(rho)

STRATEGY_COLOURS <- c(
    Ensemble            = "black",
    Logistic            = "steelblue",
    GAM                 = "darkgoldenrod",
    `Random Forest`     = "darkorchid",
    `Ensemble (2x rho)` = "firebrick"
)

# Synthetic score with 2× ensemble rho
score_2rho <- make_rank_score(sim_df$obs_rate, target_rho = 2 * rho_ens_oos)

strategies <- list(
    Ensemble            = sim_df$risk_ens,
    Logistic            = sim_df$pred_logistic,
    GAM                 = sim_df$pred_gam,
    `Random Forest`     = sim_df$pred_rf,
    `Ensemble (2x rho)` = score_2rho
)

rho_map <- c(
    Ensemble            = rho_ens_oos,
    Logistic            = rho_oos |> filter(Model == "Logistic") |> pull(rho),
    GAM                 = rho_oos |> filter(Model == "GAM") |> pull(rho),
    `Random Forest`     = rho_oos |> filter(Model == "Random Forest") |> pull(rho),
    `Ensemble (2x rho)` = 2 * rho_ens_oos
)

# Cost curves (for plotting) and optimal alpha — vary f and r
alpha_results <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies, function(score, strat_name) {
            res <- get_alpha_star_pw(fv, sim_df$pop_sum, sim_df$obs_rate,
                                     score, R_val, rv, AR_val, nu_val,
                                     n_mc = 500)
            tibble(strategy = strat_name, f = fv, r = rv,
                   rho = rho_map[[strat_name]],
                   alpha_star_pw = res$alpha_star,
                   cost_star = res$cost_star)
        })
    })
})

alpha_results |>
    knitr::kable(
        digits = 3,
        col.names = c("Strategy", "f", "r", "\u03c1 (OOS)", "\u03b1* (MC)", "Cost* (MC)"),
        caption = sprintf(
            "MC-optimal alpha* | R=%.0f, AR=%.3f | Beta(a=%.4f, b=%.4f), mean p=%.4f",
            R_val, AR_val, bb$a, bb$b, bb$p)
    )
```

## Cost curves

```{r fig-mc-cost-curves, fig.width=12, fig.height=8}
#| fig-cap: "MC cost curves faceted by f (columns) and r (rows). Diamond = MC-optimal
#|   alpha*. 'Ensemble (2x rho)' shows the sensitivity to doubled targeting accuracy."

alpha_grid_plot <- seq(0, 1, length.out = 51)

cost_curves <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies, function(score, strat_name) {
            costs <- sapply(alpha_grid_plot, function(a)
                cost_popweighted_mc(a, fv, sim_df$pop_sum, sim_df$obs_rate,
                                    score, R_val, rv, AR_val, nu_val, n_mc = 300))
            tibble(alpha = alpha_grid_plot, cost = costs,
                   strategy = strat_name, f = fv, r = rv)
        })
    })
})

optima_pts <- cost_curves |>
    left_join(alpha_results |> dplyr::select(strategy, f, r, alpha_star_pw),
              by = c("strategy", "f", "r")) |>
    filter(abs(alpha - alpha_star_pw) < 0.011)

ggplot(cost_curves, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.8) +
    geom_point(data = optima_pts, aes(x = alpha_star_pw),
               shape = 18, size = 4) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    facet_grid(r ~ f, labeller = label_both) +
    labs(x = expression(alpha ~ "(pre-emptive fraction)"),
         y = "Expected cost per capita (C_V units)",
         colour = "Strategy",
         title = "MC cost curves | all ADM2 (full dataset)",
         subtitle = sprintf(
             "R=%.0f, AR=%.3f | Beta(%.3f, %.3f), mean p=%.4f | Diamond = MC-optimal \u03b1*",
             R_val, AR_val, bb$a, bb$b, bb$p)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Top-25% High-Risk Scenario

Restrict the analysis to the top **25%** of ADM2 regions ranked by ensemble
predicted risk. Within this subset, mean outbreak probability is substantially
higher, which may shift the optimal allocation toward pre-emptive vaccination
(α* > 0).

```{r top25-setup}
TOP_PCT <- 0.25
n_top   <- ceiling(nrow(sim_df) * TOP_PCT)

sim_top <- sim_df |>
    arrange(desc(risk_ens)) |>
    slice_head(n = n_top)

cat(sprintf("Top %.0f%% subset: %d locations (of %d), %.0f M population\n",
            TOP_PCT * 100, nrow(sim_top), nrow(sim_df),
            sum(sim_top$pop_sum) / 1e6))
cat(sprintf("Mean obs_rate: %.4f (subset) vs %.4f (all)\n",
            mean(sim_top$obs_rate), mean(sim_df$obs_rate)))

bb_top <- fit_betabinom(sim_top$n_outbreak_years, N_TRIALS)
cat(sprintf("Beta-Binomial (top %.0f%%): a=%.4f, b=%.4f, mean p=%.4f (convergence=%d)\n",
            TOP_PCT * 100, bb_top$a, bb_top$b, bb_top$p, bb_top$convergence))
```

```{r top25-mc-alpha-star}
# Rho within subset (re-estimated on top-25% locations only)
rho_top <- c(
    Ensemble        = cor(sim_top$risk_ens,      sim_top$obs_rate, method = "spearman"),
    Logistic        = cor(sim_top$pred_logistic,  sim_top$obs_rate, method = "spearman"),
    GAM             = cor(sim_top$pred_gam,        sim_top$obs_rate, method = "spearman"),
    `Random Forest` = cor(sim_top$pred_rf,         sim_top$obs_rate, method = "spearman")
)
rho_top["Ensemble (2x rho)"] <- 2 * rho_top["Ensemble"]

cat("Spearman rho within top 25%:\n")
print(round(rho_top, 3))

score_2rho_top <- make_rank_score(sim_top$obs_rate, target_rho = 2 * rho_top["Ensemble"])

strategies_top <- list(
    Ensemble            = sim_top$risk_ens,
    Logistic            = sim_top$pred_logistic,
    GAM                 = sim_top$pred_gam,
    `Random Forest`     = sim_top$pred_rf,
    `Ensemble (2x rho)` = score_2rho_top
)

alpha_results_top <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top, function(score, strat_name) {
            res <- get_alpha_star_pw(fv, sim_top$pop_sum, sim_top$obs_rate,
                                     score, R_val, rv, AR_val, nu_val, n_mc = 500)
            tibble(strategy = strat_name, f = fv, r = rv,
                   rho = rho_top[[strat_name]],
                   alpha_star_pw = res$alpha_star,
                   cost_star = res$cost_star)
        })
    })
})

alpha_results_top |>
    knitr::kable(
        digits = 3,
        col.names = c("Strategy", "f", "r", "\u03c1 (in-subset)", "\u03b1* (MC)", "Cost* (MC)"),
        caption = sprintf(
            "MC-optimal alpha* (top %.0f%% ADM2) | R=%.0f, AR=%.3f | Beta(a=%.4f, b=%.4f), mean p=%.4f",
            TOP_PCT * 100, R_val, AR_val, bb_top$a, bb_top$b, bb_top$p)
    )
```

```{r fig-top25-cost-curves, fig.width=12, fig.height=8}
#| fig-cap: "MC cost curves for the top-25% high-risk ADM2 subset, faceted by
#|   f (columns) and r (rows). Diamond = MC-optimal alpha*."

cost_curves_top <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top, function(score, strat_name) {
            costs <- sapply(alpha_grid_plot, function(a)
                cost_popweighted_mc(a, fv, sim_top$pop_sum, sim_top$obs_rate,
                                    score, R_val, rv, AR_val, nu_val, n_mc = 300))
            tibble(alpha = alpha_grid_plot, cost = costs,
                   strategy = strat_name, f = fv, r = rv)
        })
    })
})

optima_top <- cost_curves_top |>
    left_join(alpha_results_top |> dplyr::select(strategy, f, r, alpha_star_pw),
              by = c("strategy", "f", "r")) |>
    filter(abs(alpha - alpha_star_pw) < 0.011)

ggplot(cost_curves_top, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.8) +
    geom_point(data = optima_top, aes(x = alpha_star_pw),
               shape = 18, size = 4) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    facet_grid(r ~ f, labeller = label_both) +
    labs(x = expression(alpha ~ "(pre-emptive fraction)"),
         y = "Expected cost per capita (C_V units)",
         colour = "Strategy",
         title = sprintf("MC cost curves | top %.0f%% high-risk ADM2", TOP_PCT * 100),
         subtitle = sprintf(
             "R=%.0f, AR=%.3f | Beta(%.3f, %.3f), mean p=%.4f | Diamond = MC-optimal \u03b1*",
             R_val, AR_val, bb_top$a, bb_top$b, bb_top$p)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Top-10% High-Risk Scenario

Restricting further to the top **10%** of ADM2 regions by ensemble predicted
risk isolates the highest-burden areas and increases mean p substantially.

```{r top10-setup}
TOP_PCT10 <- 0.10
n_top10   <- ceiling(nrow(sim_df) * TOP_PCT10)

sim_top10 <- sim_df |>
    arrange(desc(risk_ens)) |>
    slice_head(n = n_top10)

cat(sprintf("Top %.0f%% subset: %d locations (of %d), %.0f M population\n",
            TOP_PCT10 * 100, nrow(sim_top10), nrow(sim_df),
            sum(sim_top10$pop_sum) / 1e6))
cat(sprintf("Mean obs_rate: %.4f (subset) vs %.4f (all)\n",
            mean(sim_top10$obs_rate), mean(sim_df$obs_rate)))

bb_top10 <- fit_betabinom(sim_top10$n_outbreak_years, N_TRIALS)
cat(sprintf("Beta-Binomial (top %.0f%%): a=%.4f, b=%.4f, mean p=%.4f (convergence=%d)\n",
            TOP_PCT10 * 100, bb_top10$a, bb_top10$b, bb_top10$p, bb_top10$convergence))
```

```{r top10-mc-alpha-star}
rho_top10 <- c(
    Ensemble        = cor(sim_top10$risk_ens,      sim_top10$obs_rate, method = "spearman"),
    Logistic        = cor(sim_top10$pred_logistic,  sim_top10$obs_rate, method = "spearman"),
    GAM             = cor(sim_top10$pred_gam,        sim_top10$obs_rate, method = "spearman"),
    `Random Forest` = cor(sim_top10$pred_rf,         sim_top10$obs_rate, method = "spearman")
)
rho_top10["Ensemble (2x rho)"] <- 2 * rho_top10["Ensemble"]

cat("Spearman rho within top 10%:\n")
print(round(rho_top10, 3))

score_2rho_top10 <- make_rank_score(sim_top10$obs_rate, target_rho = 2 * rho_top10["Ensemble"])

strategies_top10 <- list(
    Ensemble            = sim_top10$risk_ens,
    Logistic            = sim_top10$pred_logistic,
    GAM                 = sim_top10$pred_gam,
    `Random Forest`     = sim_top10$pred_rf,
    `Ensemble (2x rho)` = score_2rho_top10
)

alpha_results_top10 <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top10, function(score, strat_name) {
            res <- get_alpha_star_pw(fv, sim_top10$pop_sum, sim_top10$obs_rate,
                                     score, R_val, rv, AR_val, nu_val, n_mc = 500)
            tibble(strategy = strat_name, f = fv, r = rv,
                   rho = rho_top10[[strat_name]],
                   alpha_star_pw = res$alpha_star,
                   cost_star = res$cost_star)
        })
    })
})

alpha_results_top10 |>
    knitr::kable(
        digits = 3,
        col.names = c("Strategy", "f", "r", "\u03c1 (in-subset)", "\u03b1* (MC)", "Cost* (MC)"),
        caption = sprintf(
            "MC-optimal alpha* (top %.0f%% ADM2) | R=%.0f, AR=%.3f | Beta(a=%.4f, b=%.4f), mean p=%.4f",
            TOP_PCT10 * 100, R_val, AR_val, bb_top10$a, bb_top10$b, bb_top10$p)
    )
```

```{r fig-top10-cost-curves, fig.width=12, fig.height=8}
#| fig-cap: "MC cost curves for the top-10% high-risk ADM2 subset, faceted by
#|   f (columns) and r (rows). Diamond = MC-optimal alpha*."

cost_curves_top10 <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top10, function(score, strat_name) {
            costs <- sapply(alpha_grid_plot, function(a)
                cost_popweighted_mc(a, fv, sim_top10$pop_sum, sim_top10$obs_rate,
                                    score, R_val, rv, AR_val, nu_val, n_mc = 300))
            tibble(alpha = alpha_grid_plot, cost = costs,
                   strategy = strat_name, f = fv, r = rv)
        })
    })
})

optima_top10 <- cost_curves_top10 |>
    left_join(alpha_results_top10 |> dplyr::select(strategy, f, r, alpha_star_pw),
              by = c("strategy", "f", "r")) |>
    filter(abs(alpha - alpha_star_pw) < 0.011)

ggplot(cost_curves_top10, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.8) +
    geom_point(data = optima_top10, aes(x = alpha_star_pw),
               shape = 18, size = 4) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    facet_grid(r ~ f, labeller = label_both) +
    labs(x = expression(alpha ~ "(pre-emptive fraction)"),
         y = "Expected cost per capita (C_V units)",
         colour = "Strategy",
         title = sprintf("MC cost curves | top %.0f%% high-risk ADM2", TOP_PCT10 * 100),
         subtitle = sprintf(
             "R=%.0f, AR=%.3f | Beta(%.3f, %.3f), mean p=%.4f | Diamond = MC-optimal \u03b1*",
             R_val, AR_val, bb_top10$a, bb_top10$b, bb_top10$p)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Top-3-Countries Scenario

Select the **top 3 countries** by mean ensemble risk across their ADM2 regions
included in the simulation dataset, then restrict the analysis to those countries.

```{r top3-setup}
country_risk <- sim_df |>
    group_by(country) |>
    summarise(mean_risk = mean(risk_ens),
              n_adm2    = n(),
              mean_obs  = mean(obs_rate),
              .groups = "drop") |>
    arrange(desc(mean_risk))

top3_ctry <- country_risk |> slice_head(n = 3) |> pull(country)

cat("Top 3 countries by mean ensemble risk:\n")
country_risk |> slice_head(n = 5) |> print()

sim_top3 <- sim_df |> filter(country %in% top3_ctry)

cat(sprintf("\nTop-3 subset: %d locations, %.0f M population\n",
            nrow(sim_top3), sum(sim_top3$pop_sum) / 1e6))
cat(sprintf("Mean obs_rate: %.4f (subset) vs %.4f (all)\n",
            mean(sim_top3$obs_rate), mean(sim_df$obs_rate)))

bb_top3 <- fit_betabinom(sim_top3$n_outbreak_years, N_TRIALS)
cat(sprintf("Beta-Binomial (top-3 countries): a=%.4f, b=%.4f, mean p=%.4f (convergence=%d)\n",
            bb_top3$a, bb_top3$b, bb_top3$p, bb_top3$convergence))
```

```{r top3-mc-alpha-star}
rho_top3 <- c(
    Ensemble        = cor(sim_top3$risk_ens,      sim_top3$obs_rate, method = "spearman"),
    Logistic        = cor(sim_top3$pred_logistic,  sim_top3$obs_rate, method = "spearman"),
    GAM             = cor(sim_top3$pred_gam,        sim_top3$obs_rate, method = "spearman"),
    `Random Forest` = cor(sim_top3$pred_rf,         sim_top3$obs_rate, method = "spearman")
)
rho_top3["Ensemble (2x rho)"] <- 2 * rho_top3["Ensemble"]

cat("Spearman rho within top-3 countries:\n")
print(round(rho_top3, 3))

score_2rho_top3 <- make_rank_score(sim_top3$obs_rate, target_rho = 2 * rho_top3["Ensemble"])

strategies_top3 <- list(
    Ensemble            = sim_top3$risk_ens,
    Logistic            = sim_top3$pred_logistic,
    GAM                 = sim_top3$pred_gam,
    `Random Forest`     = sim_top3$pred_rf,
    `Ensemble (2x rho)` = score_2rho_top3
)

alpha_results_top3 <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top3, function(score, strat_name) {
            res <- get_alpha_star_pw(fv, sim_top3$pop_sum, sim_top3$obs_rate,
                                     score, R_val, rv, AR_val, nu_val, n_mc = 500)
            tibble(strategy = strat_name, f = fv, r = rv,
                   rho = rho_top3[[strat_name]],
                   alpha_star_pw = res$alpha_star,
                   cost_star = res$cost_star)
        })
    })
})

alpha_results_top3 |>
    knitr::kable(
        digits = 3,
        col.names = c("Strategy", "f", "r", "\u03c1 (in-subset)", "\u03b1* (MC)", "Cost* (MC)"),
        caption = sprintf(
            "MC-optimal alpha* (top-3 countries: %s) | R=%.0f, AR=%.3f | Beta(a=%.4f, b=%.4f), mean p=%.4f",
            paste(top3_ctry, collapse = ", "), R_val, AR_val, bb_top3$a, bb_top3$b, bb_top3$p)
    )
```

```{r fig-top3-cost-curves, fig.width=12, fig.height=8}
#| fig-cap: "MC cost curves for the top-3 countries by mean ADM2 risk, faceted by
#|   f (columns) and r (rows). Diamond = MC-optimal alpha*."

cost_curves_top3 <- map_dfr(f_values_mc, function(fv) {
    map_dfr(r_values, function(rv) {
        imap_dfr(strategies_top3, function(score, strat_name) {
            costs <- sapply(alpha_grid_plot, function(a)
                cost_popweighted_mc(a, fv, sim_top3$pop_sum, sim_top3$obs_rate,
                                    score, R_val, rv, AR_val, nu_val, n_mc = 300))
            tibble(alpha = alpha_grid_plot, cost = costs,
                   strategy = strat_name, f = fv, r = rv)
        })
    })
})

optima_top3 <- cost_curves_top3 |>
    left_join(alpha_results_top3 |> dplyr::select(strategy, f, r, alpha_star_pw),
              by = c("strategy", "f", "r")) |>
    filter(abs(alpha - alpha_star_pw) < 0.011)

ggplot(cost_curves_top3, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.8) +
    geom_point(data = optima_top3, aes(x = alpha_star_pw),
               shape = 18, size = 4) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    facet_grid(r ~ f, labeller = label_both) +
    labs(x = expression(alpha ~ "(pre-emptive fraction)"),
         y = "Expected cost per capita (C_V units)",
         colour = "Strategy",
         title = sprintf("MC cost curves | top-3 countries (%s)",
                         paste(top3_ctry, collapse = ", ")),
         subtitle = sprintf(
             "R=%.0f, AR=%.3f | Beta(%.3f, %.3f), mean p=%.4f | Diamond = MC-optimal \u03b1*",
             R_val, AR_val, bb_top3$a, bb_top3$b, bb_top3$p)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

# Outbreak Risk Estimation

## Ensemble mean risk score

We use the ensemble mean of the three model predictions as the risk score.

```{r risk-scores}
pred_risk <- pred_df |>
    mutate(
        risk_ens = (pred_logistic + pred_gam + pred_rf) / 3
    ) |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(risk_ens), !is.na(pop_sum), pop_sum > 0) |>
    arrange(desc(risk_ens))

cat(sprintf("Locations with risk scores: %d\n", nrow(pred_risk)))
cat(sprintf("Total population: %.0f million\n", sum(pred_risk$pop_sum) / 1e6))
cat(sprintf("Mean ensemble risk: %.4f\n", mean(pred_risk$risk_ens)))
```

## Population vs. cumulative outbreak risk

```{r fig-pop-risk-cdf, fig.width=12, fig.height=5}
#| fig-cap: "Left: cumulative population (millions) living in regions with predicted
#|   outbreak risk ≤ x (ECDF). Right: number of ADM2 regions with risk ≤ x.
#|   The dashed vertical line marks the risk threshold above which ~200 million
#|   people reside. Individual model scores shown alongside ensemble mean."

# Build ECDFs for ensemble + individual models
pred_long <- pred_risk |>
    dplyr::select(location, pop_sum, obs,
                  Ensemble = risk_ens, Logistic = pred_logistic,
                  GAM = pred_gam, `Random Forest` = pred_rf) |>
    pivot_longer(c(Ensemble, Logistic, GAM, `Random Forest`),
                 names_to = "model", values_to = "risk")

# Empirical CDF by model
make_cdf <- function(df) {
    df |>
        arrange(risk) |>
        mutate(
            cum_pop   = cumsum(pop_sum),
            cum_locs  = row_number(),
            total_pop = sum(pop_sum),
            n_total   = n()
        )
}

cdf_all <- pred_long |>
    group_by(model) |>
    group_modify(~make_cdf(.x)) |>
    ungroup()

# 200M cutoff using ensemble
ens_sorted <- pred_risk |> arrange(risk_ens)
cum_pop_ens <- cumsum(ens_sorted$pop_sum)
n_cut <- which(cum_pop_ens >= 200e6)[1]
cutoff_200m <- ens_sorted$risk_ens[n_cut]

MODEL_COLOURS <- c(
    Ensemble = "black", Logistic = "steelblue",
    GAM = "darkgoldenrod", `Random Forest` = "darkorchid"
)

p_pop <- ggplot(cdf_all, aes(x = risk, y = cum_pop / 1e6, colour = model,
                              linewidth = model == "Ensemble")) +
    geom_line() +
    geom_vline(xintercept = cutoff_200m, linetype = "dashed",
               colour = "firebrick", linewidth = 0.7) +
    annotate("text", x = cutoff_200m + 0.02,
             y = max(cdf_all$cum_pop / 1e6) * 0.3,
             label = sprintf("p = %.3f\n200M threshold", cutoff_200m),
             hjust = 0, colour = "firebrick", size = 3.2) +
    scale_colour_manual(values = MODEL_COLOURS) +
    scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6),
                           guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(x = "Predicted outbreak probability (2018–2020)",
         y = "Cumulative population (millions)",
         colour = "Model",
         title = "Population ≤ risk threshold") +
    theme_minimal(base_size = 11)

p_loc <- ggplot(cdf_all, aes(x = risk, y = cum_locs, colour = model,
                              linewidth = model == "Ensemble")) +
    geom_line() +
    geom_vline(xintercept = cutoff_200m, linetype = "dashed",
               colour = "firebrick", linewidth = 0.7) +
    scale_colour_manual(values = MODEL_COLOURS) +
    scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.6),
                           guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(x = "Predicted outbreak probability (2018–2020)",
         y = "Cumulative number of ADM2 regions",
         colour = "Model",
         title = "ADM2 regions ≤ risk threshold") +
    theme_minimal(base_size = 11)

p_pop + p_loc +
    plot_annotation(
        title = sprintf(
            "Outbreak risk distribution — trained on 2010–2014 features / 2015–2017 outcomes"),
        subtitle = sprintf(
            "Predicted for 2018–2020 | %d ADM2 regions, %.0fM total population",
            nrow(pred_risk), sum(pred_risk$pop_sum) / 1e6)
    ) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

## Risk summary table

```{r risk-summary-table}
# Population above selected risk thresholds
thresholds <- c(0.05, 0.10, 0.20, 0.30, 0.50)
map_dfr(thresholds, function(thr) {
    above <- pred_risk |> filter(risk_ens > thr)
    tibble(
        threshold   = thr,
        n_regions   = nrow(above),
        pop_M       = sum(above$pop_sum) / 1e6,
        pct_regions = 100 * nrow(above) / nrow(pred_risk),
        pct_pop     = 100 * sum(above$pop_sum) / sum(pred_risk$pop_sum),
        obs_rate    = if (nrow(above) > 0) mean(above$obs) else NA_real_
    )
}) |>
    knitr::kable(
        digits = c(2, 0, 1, 1, 1, 3),
        col.names = c("Risk > X", "Regions", "Pop (M)",
                      "% regions", "% pop", "Outbreak rate (2018-20)"),
        caption = "Population and region count above ensemble risk thresholds"
    )
```

## SSA map of predicted risk (2018–2020)

```{r fig-risk-map, fig.width=10, fig.height=8}
#| fig-cap: "Ensemble mean predicted outbreak risk (2018–2020) at ADM2 level.
#|   Grey = sub-Saharan Africa countries. Red outlines = locations with observed
#|   outbreak in 2018–2020."

ssa_sf <- sf::st_as_sf(rnaturalearthdata::countries50) |>
    dplyr::filter(subregion %in% c("Eastern Africa", "Middle Africa",
                                    "Southern Africa", "Western Africa")) |>
    sf::st_make_valid() |>
    sf::st_transform(sf::st_crs(shp_admin23))

shp_risk <- shp_admin23 |>
    left_join(pred_risk |> dplyr::select(location, risk_ens, obs),
              by = "location") |>
    filter(!is.na(risk_ens))

ssa_bbox <- sf::st_bbox(ssa_sf)

ggplot() +
    geom_sf(data = ssa_sf, fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = shp_risk, aes(fill = risk_ens), colour = NA) +
    geom_sf(data = shp_risk |> filter(obs == 1),
            fill = NA, colour = "firebrick", linewidth = 0.4) +
    scale_fill_viridis_c(option = "magma", direction = -1,
                         name = "Predicted\nrisk", limits = c(0, 1)) +
    coord_sf(xlim = ssa_bbox[c("xmin", "xmax")],
             ylim = ssa_bbox[c("ymin", "ymax")]) +
    labs(title = "Predicted outbreak risk 2018–2020",
         subtitle = "Trained on 2010–2014 features / 2015–2017 outcomes | Red outline = observed outbreak") +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank(), axis.text = element_blank(),
          axis.ticks = element_blank(), legend.position = "right")
```
