---
title: "Reactive vs Preemptive Vaccination Strategies"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format: html
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(gridExtra)
library(scales)
save_plot <- FALSE
library(ggpattern)
library(patchwork)
library(data.table)
library(readxl)
source("R/load_all.R")
```


In the analytic models (`analytic_study.qmd` and `analytic_hetero.qmd`), the reactive effectiveness $r$ was treated as a fixed parameter. In practice, $r$ depends on outbreak dynamics and vaccination logistics (timing, coverage, delay). Here, we explicitly model these characteristics to derive $r$ endogenously.

# Endogenous Reactive Effectiveness

The stochastic validation in `analytic_hetero.qmd` confirmed that the analytic cost model is recovered as the number of populations $n$ grows. That simulation treated the reactive effectiveness $r$ as a fixed parameter. Here, we add one layer of realism: $r$ is no longer predefined but is derived endogenously from outbreak trajectories, response delay, and vaccine effectiveness.

## Outbreak Trajectory

Each outbreak follows an exponential tent-shaped curve — linear on the log scale:

$$
I(t) =
\begin{cases}
I_p \exp\bigl(r_g (t - t_p)\bigr), & t_0 \le t \le t_p, \\
I_p \exp\bigl(-r_d (t - t_p)\bigr), & t_p < t \le t_1,
\end{cases}
$$

where $t_0$, $t_p$, $t_1$ are the outbreak start, peak, and end times, $I_p$ is the peak incidence, and $r_g$, $r_d > 0$ are the exponential growth and decay rates. The incidence at the boundaries is $I(t_0) = I_p e^{-r_g(t_p - t_0)}$ and $I(t_1) = I_p e^{-r_d(t_1 - t_p)}$.

## Vaccine Effectiveness Decomposition

Following the framework described [here](https://www.jonghoonk.com/blog/posts/vaccine_effectiveness/), vaccine effectiveness is decomposed into direct ($\text{VE}_d$) and indirect ($\text{VE}_i$) components. With vaccination coverage $f_c$ within a population, the post-vaccination incidence at time $t$ is:

$$
I_{\text{post}}(t) = I(t) \cdot \bigl[\, f_c (1 - \text{VE}_d)(1 - \text{VE}_i) + (1 - f_c)(1 - \text{VE}_i) \,\bigr].
$$

The overall vaccine effectiveness (OVE) is:

$$
\text{OVE} = 1 - \bigl[\, f_c (1 - \text{VE}_d)(1 - \text{VE}_i) + (1 - f_c)(1 - \text{VE}_i) \,\bigr].
$$

## Endogenous $r$

For each reactively vaccinated population $i$, the reactive effectiveness $r_i$ is the fractional reduction in cumulative incidence relative to the no-vaccination counterfactual:

$$
r_i = 1 - \frac{A_{\text{vacc},i}}{A_{\text{total},i}},
$$

where $A_{\text{total},i} = A(t_0, t_1)$ is the cumulative incidence without vaccination and $A_{\text{vacc},i}$ is the cumulative incidence under vaccination. Before the vaccine becomes effective at $t_{\text{eff},i} = t_0 + \delta_i$, incidence is unaffected; after $t_{\text{eff},i}$, it is reduced by OVE. The delay $\delta_i$ includes both the logistical delay (outbreak detection, vaccine deployment) and the immunogenicity delay (time for the vaccine to induce protective immunity):

$$
A_{\text{vacc},i} = A(t_0,\, t_{\text{eff},i}) + \bigl[\, f_c (1 - \text{VE}_d)(1 - \text{VE}_i) + (1 - f_c)(1 - \text{VE}_i) \,\bigr] \cdot A(t_{\text{eff},i},\, t_1).
$$

For the exponential tent curve, the cumulative incidence from $t_0$ to time $t$ has a closed form:

$$
A(t_0, t) =
\begin{cases}
\dfrac{I(t) - I(t_0)}{r_g}, & t \le t_p, \\[6pt]
\dfrac{I_p - I(t_0)}{r_g} + \dfrac{I_p - I(t)}{r_d}, & t > t_p.
\end{cases}
$$

Thus $A_{\text{rem}}(t) / A_{\text{total}} = 1 - A(t_0, t) / A(t_0, t_1)$.

## Algorithm

Given parameters $(n, p, f, R, \text{VE}_d, \text{VE}_i, f_c, \nu, \alpha, \rho)$ and trajectory parameters:

1. **Draw risks**: $x_1, \ldots, x_n \overset{\text{iid}}{\sim} \text{Beta}(1, \theta)$, $\theta = (1-p)/p$.
2. **Target selection**: Rank by true risk ($\rho = 1$) or noisy score ($\rho < 1$).
3. **Pre-emptive vaccination**: Vaccinate the top $\lfloor n \alpha f \rfloor$ populations.
4. **Outbreak draws**: $Y_i \sim \text{Bernoulli}(x_i)$.
5. **Reactive allocation**: Select up to $\lfloor n(1-\alpha)f \rfloor$ outbreak populations.
6. **Endogenous $r_i$**: For each reactively vaccinated population, draw a trajectory $(D_i, \phi_i)$ and delay $\delta_i$, then compute $r_i = \text{OVE} \times A_{\text{rem}}(\delta_i) / A_{\text{total}}$.
7. **Cost**: $c_i$ as in the analytic model, but with population-specific $r_i$ for reactive cases.

## Simulation

```{r}
#| label: fig-endogenous-r
#| fig-cap: "Cost vs. pre-emptive fraction with endogenous reactive effectiveness. Ribbons show the central 95% interval over 50 simulations. The analytic curve uses the simulation-average r as reference."
#| fig-width: 8
#| fig-height: 6

# Parameters
p_endo <- 0.3
f_endo <- 0.5
R_endo <- 5
VEd_endo <- 0.7
VEi_endo <- 0.3
f_cov_endo <- 0.7
nu_endo <- 1
rho_endo <- 1
n_sim_endo <- 50
alpha_grid_endo <- seq(0, 1, length.out = 21)

# Simulation with endogenous r
df_endo_50 <- sim_cost_heterorisk_endogenous(
    alpha = alpha_grid_endo, n = 50, p = p_endo, f = f_endo,
    R = R_endo, VEd = VEd_endo, VEi = VEi_endo, f_cov = f_cov_endo,
    nu = nu_endo, rho = rho_endo,
    duration_range = c(60, 300), peak_frac_shape = c(2, 2),
    rg_range = c(0.02, 0.10), rd_range = c(0.01, 0.05),
    delay_range = c(7, 21), n_sim = n_sim_endo, seed = 42
)

df_endo_500 <- sim_cost_heterorisk_endogenous(
    alpha = alpha_grid_endo, n = 500, p = p_endo, f = f_endo,
    R = R_endo, VEd = VEd_endo, VEi = VEi_endo, f_cov = f_cov_endo,
    nu = nu_endo, rho = rho_endo,
    duration_range = c(60, 300), peak_frac_shape = c(2, 2),
    rg_range = c(0.02, 0.10), rd_range = c(0.01, 0.05),
    delay_range = c(7, 21), n_sim = n_sim_endo, seed = 42
)

# Compute simulation-average r for analytic reference
mean_r_overall <- mean(df_endo_500$mean_r, na.rm = TRUE)

# Summarize simulations
summarize_endo <- function(df, label) {
    df %>%
        group_by(alpha) %>%
        summarise(
            mean_cost = mean(cost),
            lo = quantile(cost, 0.025),
            hi = quantile(cost, 0.975),
            mean_r_sim = mean(mean_r, na.rm = TRUE),
            .groups = "drop"
        ) %>%
        mutate(Source = label)
}

df_endo_summary <- bind_rows(
    summarize_endo(df_endo_50, "Simulation (n = 50)"),
    summarize_endo(df_endo_500, "Simulation (n = 500)")
)

# Analytic reference with simulation-average r
cost_beta_ref <- function(alpha, f, p, R, r, nu = 1) {
    alpha <- pmax(0, pmin(1, alpha))
    f_pre <- alpha * f
    f_react_cap <- (1 - alpha) * f
    exponent <- p / (1 - p)
    p_bar_pre <- ifelse(f_pre > 0, 1 - (1 - p) * f_pre^exponent, 0)
    E_pre <- f_pre * p_bar_pre
    c_pre_val <- f_pre + E_pre * (1 - nu) * R
    E_rem <- p - E_pre
    is_rich <- f_react_cap >= E_rem
    c_react_val <- ifelse(is_rich,
        E_rem * (1 + (1 - r) * R),
        f_react_cap * (1 + (1 - r) * R) + (E_rem - f_react_cap) * R
    )
    c_pre_val + c_react_val
}

df_analytic_endo <- data.frame(alpha = alpha_grid_endo) %>%
    mutate(
        mean_cost = cost_beta_ref(
            alpha, f_endo, p_endo, R_endo,
            mean_r_overall, nu_endo
        ),
        lo = mean_cost, hi = mean_cost,
        mean_r_sim = mean_r_overall,
        Source = paste0("Analytic (r = ", round(mean_r_overall, 2), ")")
    )

df_all_endo <- bind_rows(df_endo_summary, df_analytic_endo)

ggplot(df_all_endo, aes(x = alpha, y = mean_cost, color = Source, fill = Source)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    theme_light() +
    labs(
        x = expression("Fraction of capacity pre-emptive " ~ alpha),
        y = expression("Normalized per-population cost " ~ italic(c)[mixed]),
        title = paste0(
            "Endogenous r: VE_d = ", VEd_endo, ", VE_i = ", VEi_endo,
            ", f_cov = ", f_cov_endo
        )
    ) +
    scale_color_manual("", values = c(
        setNames("black", paste0("Analytic (r = ", round(mean_r_overall, 2), ")")),
        "Simulation (n = 50)" = "#e41a1c",
        "Simulation (n = 500)" = "#377eb8"
    )) +
    scale_fill_manual("", values = c(
        setNames("black", paste0("Analytic (r = ", round(mean_r_overall, 2), ")")),
        "Simulation (n = 50)" = "#e41a1c",
        "Simulation (n = 500)" = "#377eb8"
    )) +
    theme(legend.position = "top")
```

The plot shows how the cost varies with $\alpha$ when $r$ is computed endogenously rather than fixed. The analytic curve uses the simulation-average $\bar{r}$ as a reference, demonstrating that the analytic model with an appropriately chosen fixed $r$ approximates the endogenous-$r$ simulation.

```{r}
#| label: fig-endogenous-r-dist
#| fig-cap: "Distribution of endogenous reactive effectiveness across simulations (n = 500). Each bar aggregates r values across all alpha values and simulation replicates."
#| fig-width: 7
#| fig-height: 4

# Distribution of endogenous r
ggplot(
    df_endo_500[!is.na(df_endo_500$mean_r), ],
    aes(x = mean_r)
) +
    geom_histogram(bins = 30, fill = "#377eb8", color = "white", alpha = 0.8) +
    geom_vline(xintercept = mean_r_overall, linetype = "dashed", linewidth = 0.8) +
    annotate("text",
        x = mean_r_overall, y = Inf, vjust = 2,
        label = paste0("Mean r = ", round(mean_r_overall, 3)),
        fontface = "bold"
    ) +
    labs(
        x = "Mean reactive effectiveness (r)",
        y = "Count",
        title = "Distribution of endogenous r (n = 500)"
    ) +
    theme_light()
```

The distribution of endogenous $r$ reflects the combined variability in outbreak trajectory shape (duration and peak timing) and response delay. Late responses to long outbreaks yield lower $r$; early responses to short outbreaks yield higher $r$.

## Sensitivity to Endogenous-$r$ Parameters

The endogenous $r$ depends on three groups of parameters absent from the analytic model: vaccine effectiveness ($\text{VE}_d$, $\text{VE}_i$), vaccination coverage ($f_c$), and response delay ($\delta$). We sweep each parameter while holding the others at baseline values to examine how the cost curve and optimal $\alpha^*$ shift.

```{r}
#| label: sensitivity-sweep

# Baseline parameters (same as fig-endogenous-r)
base <- list(
    p = 0.3, f = 0.5, R = 5, nu = 1, rho = 1,
    VEd = 0.7, VEi = 0.3, f_cov = 0.7,
    duration_range = c(60, 300), peak_frac_shape = c(2, 2),
    rg_range = c(0.02, 0.10), rd_range = c(0.01, 0.05),
    delay_range = c(7, 21),
    n = 500, n_sim = 50, seed = 42
)

alpha_grid_sens <- seq(0, 1, length.out = 51)

# Helper: run simulation for one parameter set, return summary
run_sweep <- function(pars, label) {
    df <- sim_cost_heterorisk_endogenous(
        alpha = alpha_grid_sens,
        n = pars$n, p = pars$p, f = pars$f, R = pars$R,
        VEd = pars$VEd, VEi = pars$VEi, f_cov = pars$f_cov,
        nu = pars$nu, rho = pars$rho,
        duration_range = pars$duration_range,
        peak_frac_shape = pars$peak_frac_shape,
        rg_range = pars$rg_range, rd_range = pars$rd_range,
        delay_range = pars$delay_range,
        n_sim = pars$n_sim, seed = pars$seed
    )
    df %>%
        group_by(alpha) %>%
        summarise(
            mean_cost = mean(cost),
            lo = quantile(cost, 0.025),
            hi = quantile(cost, 0.975),
            mean_r = mean(mean_r, na.rm = TRUE),
            .groups = "drop"
        ) %>%
        mutate(label = label)
}

# --- Sweep VE_d ---
VEd_vals <- c(0.3, 0.5, 0.7, 0.9)
df_VEd <- bind_rows(lapply(VEd_vals, function(v) {
    pars <- base
    pars$VEd <- v
    run_sweep(pars, paste0("VE_d = ", v))
}))

# --- Sweep VE_i ---
VEi_vals <- c(0.0, 0.15, 0.3, 0.5)
df_VEi <- bind_rows(lapply(VEi_vals, function(v) {
    pars <- base
    pars$VEi <- v
    run_sweep(pars, paste0("VE_i = ", v))
}))

# --- Sweep f_cov ---
fcov_vals <- c(0.3, 0.5, 0.7, 0.9)
df_fcov <- bind_rows(lapply(fcov_vals, function(v) {
    pars <- base
    pars$f_cov <- v
    run_sweep(pars, paste0("f_c = ", v))
}))

# --- Sweep delay ---
delay_specs <- list(
    list(range = c(7, 14), label = "Short (7-14 d)"),
    list(range = c(14, 28), label = "Medium (14-28 d)"),
    list(range = c(28, 42), label = "Long (28-42 d)"),
    list(range = c(42, 60), label = "Very long (42-60 d)")
)
df_delay <- bind_rows(lapply(delay_specs, function(s) {
    pars <- base
    pars$delay_range <- s$range
    run_sweep(pars, s$label)
}))
```

### Direct Vaccine Effectiveness ($\text{VE}_d$)

```{r}
#| label: fig-sens-VEd
#| fig-cap: "Cost vs. alpha under varying direct vaccine effectiveness. Higher VE_d increases OVE, raising endogenous r and shifting the optimal alpha toward pre-emptive allocation."
#| fig-width: 8
#| fig-height: 5

ggplot(df_VEd, aes(x = alpha, y = mean_cost, color = label, fill = label)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.9) +
    theme_light() +
    labs(
        x = expression(alpha),
        y = expression("Normalized per-population cost " ~ italic(c)[mixed]),
        color = NULL, fill = NULL
    ) +
    theme(legend.position = "top")
```

### Indirect Vaccine Effectiveness ($\text{VE}_i$)

```{r}
#| label: fig-sens-VEi
#| fig-cap: "Cost vs. alpha under varying indirect vaccine effectiveness. VE_i captures herd-immunity effects: higher values reduce incidence in unvaccinated individuals."
#| fig-width: 8
#| fig-height: 5

ggplot(df_VEi, aes(x = alpha, y = mean_cost, color = label, fill = label)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.9) +
    theme_light() +
    labs(
        x = expression(alpha),
        y = expression("Normalized per-population cost " ~ italic(c)[mixed]),
        color = NULL, fill = NULL
    ) +
    theme(legend.position = "top")
```

### Vaccination Coverage ($f_c$)

```{r}
#| label: fig-sens-fcov
#| fig-cap: "Cost vs. alpha under varying vaccination coverage within targeted populations."
#| fig-width: 8
#| fig-height: 5

ggplot(df_fcov, aes(x = alpha, y = mean_cost, color = label, fill = label)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.9) +
    theme_light() +
    labs(
        x = expression(alpha),
        y = expression("Normalized per-population cost " ~ italic(c)[mixed]),
        color = NULL, fill = NULL
    ) +
    theme(legend.position = "top")
```

### Response Delay ($\delta$)

```{r}
#| label: fig-sens-delay
#| fig-cap: "Cost vs. alpha under varying response delay. Longer delays reduce endogenous r, diminishing the value of reactive vaccination and shifting the optimum toward pre-emptive allocation."
#| fig-width: 8
#| fig-height: 5

ggplot(df_delay, aes(x = alpha, y = mean_cost, color = label, fill = label)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.9) +
    theme_light() +
    labs(
        x = expression(alpha),
        y = expression("Normalized per-population cost " ~ italic(c)[mixed]),
        color = NULL, fill = NULL
    ) +
    theme(legend.position = "top")
```

### Empirical Optimal $\alpha^*$

```{r}
#| label: fig-optimal-alpha
#| fig-cap: "Empirical optimal alpha and mean endogenous r across all parameter sweeps."
#| fig-width: 9
#| fig-height: 5

# Extract optimal alpha and mean r for each scenario
get_optimal <- function(df, param_name) {
    df %>%
        group_by(label) %>%
        slice_min(mean_cost, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        mutate(parameter = param_name) %>%
        select(parameter, label,
            alpha_star = alpha, min_cost = mean_cost,
            mean_r
        )
}

df_opt <- bind_rows(
    get_optimal(df_VEd, "VE_d"),
    get_optimal(df_VEi, "VE_i"),
    get_optimal(df_fcov, "f_c"),
    get_optimal(df_delay, "Delay")
)

p1 <- ggplot(df_opt, aes(x = label, y = alpha_star, fill = parameter)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    facet_wrap(~parameter, scales = "free_x") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = expression("Optimal " ~ alpha^"*")) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))

p2 <- ggplot(df_opt, aes(x = label, y = mean_r, fill = parameter)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    facet_wrap(~parameter, scales = "free_x") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = expression("Mean " ~ bar(r))) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))

p1 / p2
```

When reactive vaccination is highly effective (high $\text{VE}_d$, high $f_c$, short delay), more capacity should be reserved for reactive use (lower $\alpha^*$). Conversely, when reactive effectiveness is degraded — by low vaccine efficacy, low coverage, or long delays — the optimal strategy shifts toward pre-emptive allocation (higher $\alpha^*$). This confirms that the fixed-$r$ analytic model captures the qualitative structure: the endogenous $r$ acts as a sufficient summary of the reactive arm's value.

# Cost ratio, $R$

## Costs of illness

```{r econ-demo-data, echo=FALSE}
# Load life expectancy and GDP data
life_exp_data <-
    as.data.frame(fread("data/wpp_life_expectancy_20241022.csv"))
# life expectancy at age 25 during 2010-2020, mean age for cholera
life_exp_data %>%
    filter(
        Year >= 2010, Year <= 2020,
        `Region, subregion, country or area *` == "Sub-Saharan Africa"
    ) %>%
    pull(`25`) %>%
    as.numeric() -> life_exp_25

# Load population by age data
# prop of under 5 during 2010-2020
age_dist <-
    as.data.frame(fread("data/wpp_pop_by_age_20241022.csv"))
age_dist %>%
    filter(
        Year >= 2010, Year <= 2020,
        `Region, subregion, country or area *` == "Sub-Saharan Africa"
    ) %>%
    pull(prop_u5) -> prop_U5

# gdp data from World Bank during 2010-2020
gdp_data <- read_xls("data/GDP_WorldBank.xls")
gdp_data %>%
    filter(`Country Name` == "Sub-Saharan Africa (IDA & IBRD countries)") %>%
    select(c(`2010`:`2020`)) |>
    as.numeric() -> gdp_ssa

# workforce data
workforce_data <- read_xls("data/Workforce_Worldbank.xls")
workforce_data %>%
    filter(`Country Name` == "Sub-Saharan Africa (excluding high income)") %>%
    select(c(`2010`:`2020`)) |>
    as.numeric() -> wf_ssa

## Cost of illness parameters
parm <- fread("data/parameters.csv")
parm <- parm[Disease == "Cholera"]
# GBD data deleted and replaced with Mbewe (2025) Open Forum Infect Dis data
# it is assumed that moderate and severe cases are reported
parm <- parm[!(Parameter == "Prop_Moderate" & Value == 0.289)]
parm <- parm[!(Parameter == "Prop_Severe" & Value == 0.069)]

# Disease burden parameters
day_ill <- parm[Parameter == "Duration_Illness", Value]
pr_moderate <- parm[Parameter == "Prop_Moderate", Value]
pr_severe <- parm[Parameter == "Prop_Severe", Value]
wt_moderate <- parm[Parameter == "Disability_Weight_Moderate", Value]
wt_severe <- parm[Parameter == "Disability_Weight_Severe", Value]

# Direct medical costs
patient_cost_hosp <- parm[Parameter == "Patient_Cost_Hosp", Value]
patient_cost_outpt <- parm[Parameter == "Patient_Cost_Outpt", Value]
public_cost_hosp <- parm[Parameter == "Public_Cost_Hosp", Value]
public_cost_outpt <- parm[Parameter == "Public_Cost_Outpt", Value]

# Productivity costs
patient_workday_lost <- parm[Parameter == "Pt_Workdays_Lost", Value]
caregiver_workday_lost <- parm[Parameter == "Caregiver_Workdays_Lost", Value]
mean_age_inf <- parm[Parameter == "Mean_Age_Infection", Value]

mean_gdp <- mean(gdp_ssa)
mean_remaining_life <- mean(life_exp_25) # assuming mean age of infection 25
mean_cfr <- 0.01 # across countries and year
mean_prop_workforce <- mean(wf_ssa) / 100 # across countries and year
pr_tot <- pr_moderate + pr_severe
mean_dis_wt <- wt_moderate * pr_moderate / pr_tot +
    wt_severe * pr_severe / pr_tot
```

Total cost per case = direct medical costs (outpatient/hospital weighted by
severity) + indirect costs (morbidity DALYs, premature-death YLLs, productivity
losses). See `R/analytic_utils.R: total_cost_per_case()`.

## Cost of vaccination
```{r}
# Vaccination costs
vacc_price_per_dose <- parm[Parameter == "Vaccine_Cost", Value]
vacc_delivery_cost <- parm[Parameter == "Vaccine_Delivery_Cost", Value]
vacc_shipping_cost <- parm[Parameter == "Vaccine_Shipping_Cost", Value]
```

Cost ratio $R = C_I / C_V$ as a function of attack rate, comparing GDP and $3 \times$ GDP year valuations:
```{r}
N_pop <- 1e6
dose_per_person <- 1 # Single-dose regimen

# Vaccination costs per person (do not depend on GDP in this setup)
vacc_cost_per_person <- (vacc_price_per_dose + vacc_delivery_cost + vacc_shipping_cost) * dose_per_person

C_vac_per_person <- vacc_cost_per_person

fc <- 0.9
C_V <- N_pop * fc * C_vac_per_person

# Function to build cost-ratio df for a given year_val and label
build_cost_df <- function(year_val, scen_label) {
    C_case <- total_cost_per_case(
        day_ill = day_ill,
        mean_dis_wt = mean_dis_wt,
        year_val = year_val,
        mean_prop_workforce = mean_prop_workforce,
        patient_workday_lost = patient_workday_lost,
        caregiver_workday_lost = caregiver_workday_lost,
        mean_cfr = mean_cfr,
        mean_remaining_life = mean_remaining_life,
        pr_moderate = pr_moderate,
        pr_severe = pr_severe,
        patient_cost_outpt = patient_cost_outpt,
        patient_cost_hosp = patient_cost_hosp,
        public_cost_outpt = public_cost_outpt,
        public_cost_hosp = public_cost_hosp
    )
    # Use strictly positive attack rates for log scale
    attack_rates <- seq(0.001, 0.1, by = 0.001)

    tibble(
        attack_rate = attack_rates
    ) |>
        mutate(
            cases = N_pop * attack_rate,
            C_I = cases * C_case,
            C_V = C_V,
            R_ratio = C_I / C_V,
            scenario = scen_label
        )
}

# Build data for GDP and 3x GDP
df_costs_gdp <- build_cost_df(mean_gdp, "GDP")
df_costs_3gdp <- build_cost_df(3 * mean_gdp, "3 \u00D7 GDP")

df_costs_all <- bind_rows(df_costs_gdp, df_costs_3gdp)

# Plot
ggplot(
    df_costs_all,
    aes(x = attack_rate, y = R_ratio, color = scenario)
) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 1, linetype = "dotted") +
    scale_x_log10() +
    labs(
        x = "Fraction of population infected",
        y = expression(italic(R) == italic(C)[I] / italic(C)[V]),
        color = "Year value"
    ) +
    theme_light()
```

# Spatial Simulation

### Creating Regions and Outbreaks

Simulation parameters:
```{r}
seed_rng_val <- 42

n_region_val <- 900
pop_val <- 1e3
rho0_val <- 0.8
ell_val <- 20
theta_val <- 5

alpha_val <- 0.8
f_val <- 0.4
R_val <- 10
r_val <- 0.3
eta_val <- 0.9
nu_val <- 0.9
kappa_val <- 0.9

rho_target_val <- 0.3
n_seeds_val <- 20
patch_size_val <- 60
vacc_resource_val <- 200 # total number of cells that can be vaccinated
B_pre_val <- round(alpha_val * vacc_resource_val) # number of cells (populations) that can be vaccinated

# assign_reactive_vaccination inputs
vacc_coverage_val <- 0.7
reactive_vacc_delay_range <- c(7, 21)
weight_attack_rate <- 0.6
weight_attack_case <- 0.4

# assign_outbreak_trajectory inputs
start_time_range <- c(0, 180)
total_duration_range <- c(7, 120)
peak_frac_shape_beta_params <- c(2, 2)
peak_incidence_range <- c(5e-6, 5e-4)
max_attack_rate <- 0.1
r0 <- 0
```

Create regions and assign spatially correlated outbreak risks:
```{r}
regions <- setup_spatial_region(
    n_region = n_region_val,
    pop_size_range = c(1, 1),
    prop_U5_range = c(0.15, 0.17),
    theta = theta_val,
    rho0 = rho0_val,
    ell = ell_val,
    adj_type = "rook",
    rho_score = rho_target_val,
    n_seeds = n_seeds_val,
    patch_size = patch_size_val,
    seed_rng = seed_rng_val
)
```

Allocate pre-emptive vaccination to the highest-risk patches:

```{r}
patch_pre_vax <-
    allocate_preemptive_patches(
        patch_df = regions$patches$summary,
        B_pre = B_pre_val
    )

pacth_id_prevax <- which(patch_pre_vax$pre_vax == TRUE)

cells <- regions$cells
cells$pre_vax <-
    ifelse(cells$patch_id %in% pacth_id_prevax, TRUE, FALSE)
```

Simulate outbreaks and assign trajectory parameters for outbreak regions:

```{r}
outbreaks <- simulate_outbreak_spillover(
    p = cells$risk$p,
    adj = cells$adj,
    vax = cells$pre_vax,
    eta = eta_val,
    nu = nu_val,
    kappa = kappa_val
)
cells$outbreak <- outbreaks$Y
df <- data.frame(
    id = cells$id, pop_size = cells$pop_size,
    Y = cells$outbreak
)
traj <- assign_outbreak_trajectory(
    regions = df[df$Y, ],
    start_time_range = c(0, 180),
    total_duration_range = c(7, 120),
    peak_frac_shape = c(2, 2),
    peak_incidence_range = c(5e-6, 5e-4),
    max_attack_rate = 0.1,
    r0 = 0
)

# sum(is.na(traj$t0))
# sum(!is.na(traj$t0))
# sum(!is.na(traj$t0)) == sum(outbreaks$Y)
# summary(traj$attack_rate_per_capita)
# summary(traj$duration)
```

Assign reactive vaccination to eligible outbreak regions:
```{r}

df <- assign_reactive_vaccination(
    outbreaks = traj,
    t_now = 12,
    max_targets = round(vacc_resource_val - B_pre_val)
)
# View(df)
sum(df$eligible)
sum(df$reactive_vax)

df_ci <- compute_frac_reduction_ci(subset(df, !is.na(rv_delay)), nu = nu_val)

df_ci2 <- left_join(df, df_ci[, c("id", "frac_reduction_ci")], by = "id")
# for plot later
cells$reactive_vax <- ifelse(cells$id %in% df_ci2[df_ci2$reactive_vax, ]$id,
    TRUE, FALSE
)
```


### Figures

#### Figure 1. Spatially correlated risks $p_i$

```{r, fig.width=7, fig.height=4.5}
coords <- regions$cells$coords
df_risk <- data.frame(
    x = coords[, 1],
    y = coords[, 2],
    p = regions$cells$risk$p
)

ggplot(df_risk, aes(x = x, y = y, fill = p)) +
    geom_tile() +
    coord_equal() +
    scale_fill_viridis_c(
        limits = c(0, 1),
        option = "C",
        labels = number_format(accuracy = 0.01)
    ) +
    labs(
        title = expression("Spatially correlated risks " * italic(p)[i]),
        fill = expression(italic(p)[i]),
        x = NULL, y = NULL
    ) +
    theme_map()
```

#### Figure 2. Imperfect risk score vs true risk

```{r, fig.width=6.5, fig.height=4.2}
df_score <- data.frame(
    p     = regions$cells$risk$p,
    score = regions$patches$patch_out$score
)

ggplot(df_score, aes(x = p, y = score)) +
    geom_point(alpha = 0.35, size = 1) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    labs(
        title = paste0(
            "Imperfect score vs true risk (Spearman rho = ",
            round(regions$cells$rank$rho_achieved, 2), ")"
        ),
        x = expression(italic(p)[i]),
        y = expression(score[i])
    ) +
    theme_minimal(base_size = 12)
```

#### Figure 3. Patch map

```{r, fig.width=7, fig.height=4.5}
df_patchmap <- data.frame(
    x     = coords[, 1],
    y     = coords[, 2],
    patch = factor(regions$cells$patch_id)
)

ggplot(df_patchmap, aes(x = x, y = y, fill = patch)) +
    geom_tile() +
    coord_equal() +
    guides(fill = "none") +
    labs(
        title = "Patches (contiguous groups grown from high-score seeds)",
        x = NULL, y = NULL
    ) +
    theme_map()

```

#### Figure 4. Simulated pre-emptive vs reactive vs outbreaks

```{r, fig.width=8, fig.height=4.8}
df_state <- data.frame(
    x     = coords[, 1],
    y     = coords[, 2],
    pre   = cells$pre_vax,
    react = cells$reactive_vax,
    out   = cells$outbreak
) %>%
    mutate(
        state = case_when(
            out & react ~ "Outbreak + reactive",
            out & !react ~ "Outbreak (unmitigated)",
            !out & pre ~ "Pre-emptive vaccinated",
            !out & react ~ "Reactive vaccinated (no outbreak)",
            TRUE ~ "None"
        ),
        state = factor(
            state,
            levels = c(
                "None",
                "Pre-emptive vaccinated",
                "Reactive vaccinated (no outbreak)",
                "Outbreak + reactive",
                "Outbreak (unmitigated)"
            )
        )
    )

state_cols <- c(
    "None" = "grey90",
    "Pre-emptive vaccinated" = "#2c7fb8",
    "Reactive vaccinated (no outbreak)" = "#7fcdbb",
    "Outbreak + reactive" = "#fdae61",
    "Outbreak (unmitigated)" = "#d7191c"
)

ggplot(df_state, aes(x = x, y = y, fill = state)) +
    geom_tile() +
    coord_equal() +
    scale_fill_manual(values = state_cols) +
    labs(
        title = bquote("Realization (" * alpha ~ "=" ~ .(alpha_val) * ")"),
        fill = NULL,
        x = NULL, y = NULL
    ) +
    theme_map()
```



## Sensitivity to Pre-emptive Allocation ($\alpha$)

We sweep $\alpha$ over $[0,1]$ to study how total cost varies with the pre-emptive allocation fraction. Parameters differ from the illustration above: higher targeting accuracy ($\rho = 0.9$), narrower reactive delay window (14--21 days), and $4\times$ higher peak incidence, representing a more challenging outbreak scenario.
```{r}
seed_rng_val <- 42

n_region_val <- 900
pop_val <- 1e3
rho0_val <- 0.8
ell_val <- 20
theta_val <- 5

alpha_val <- 0.8
f_val <- 0.4
R_val <- 10
r_val <- 0.3
eta_val <- 0.9
nu_val <- 0.9
kappa_val <- 0.9

rho_target_val <- 0.9
n_seeds_val <- 20
patch_size_val <- 60
vacc_resource_val <- 200 # total number of cells that can be vaccinated
B_pre_val <- round(alpha_val * vacc_resource_val) # number of cells (populations) that can be vaccinated

# assign_reactive_vaccination inputs
vacc_coverage_val <- 0.7
reactive_vacc_delay_range <- c(14, 21)
weight_attack_rate <- 0.6
weight_attack_case <- 0.4

# assign_outbreak_trajectory inputs
outbreak_start_time_range <- c(0, 180)
outbreak_total_duration_range <- c(7, 120)
outbreak_peak_frac_shape_beta_params <- c(2, 2)
outbreak_peak_incidence_range <- 4 * c(5e-6, 5e-4)
outbreak_max_attack_rate <- 0.1
outbreak_base_incidence <- 0

# For reproducibility across alpha runs
t_now <- 12
seed_base <- seed_rng_val
```

Set up spatial regions:
```{r}
# regions
regions <- setup_spatial_region(
    n_region = n_region_val,
    pop_size_range = c(1, 1),
    prop_U5_range = c(0.15, 0.17),
    theta = theta_val,
    rho0 = rho0_val,
    ell = ell_val,
    adj_type = "rook",
    rho_score = rho_target_val,
    n_seeds = n_seeds_val,
    patch_size = patch_size_val,
    seed_rng = seed_rng_val
)
```


### Single-Run Experiment

*Note: Each $\alpha$ value uses a different random seed, so outbreak realizations vary across $\alpha$. This confounds the allocation effect with stochastic variation. The Monte Carlo experiment below averages over replicates to remove this confound.*

```{r}
alpha_grid <- seq(0, 1, by = 0.1)

res_alpha <- lapply(seq_along(alpha_grid), function(k) {
    alpha_k <- alpha_grid[k]

    B_pre_k <- round(alpha_k * vacc_resource_val)
    max_targets_k <- round(vacc_resource_val - B_pre_k)

    # 1) Allocate pre-emptive vaccination to patches

    patch_pre_vax_k <- allocate_preemptive_patches(
        patch_df = regions$patches$summary,
        B_pre = B_pre_k
    )
    patch_id_prevax_k <- which(patch_pre_vax_k$pre_vax == TRUE)

    cells_k <- regions$cells
    cells_k$pre_vax <- cells_k$patch_id %in% patch_id_prevax_k

    # 2) Simulate outbreaks (your function may already incorporate vax in occurrence/spillover)

    set.seed(seed_base + 1000 + k)
    outbreaks_k <- simulate_outbreak_spillover(
        p     = cells_k$risk$p,
        adj   = cells_k$adj,
        vax   = cells_k$pre_vax,
        eta   = eta_val,
        nu    = nu_val,
        kappa = kappa_val
    )
    cells_k$outbreak <- outbreaks_k$Y

    # 3) Assign outbreak trajectories for outbreaking cells

    df_out_k <- data.frame(
        id = cells_k$id,
        pop_size = cells_k$pop_size,
        Y = cells_k$outbreak
    )

    set.seed(seed_base + 2000 + k)
    traj_k <- assign_outbreak_trajectory(
        regions = df_out_k[df_out_k$Y, ],
        start_time_range = outbreak_start_time_range,
        total_duration_range = outbreak_total_duration_range,
        peak_frac_shape = outbreak_peak_frac_shape_beta_params,
        peak_incidence_range = outbreak_peak_incidence_range,
        max_attack_rate = outbreak_max_attack_rate,
        r0 = outbreak_base_incidence
    )
    # 4) Assign reactive vaccination (given remaining budget)

    set.seed(seed_base + 3000 + k)
    rv_k <- assign_reactive_vaccination(
        outbreaks = traj_k,
        t_now = t_now,
        max_targets = max_targets_k,
        coverage = vacc_coverage_val,
        delay_range = reactive_vacc_delay_range,
        w_attack = weight_attack_rate,
        w_cases = weight_attack_case
    )

    # 5) Compute reactive fractional reduction (delay-based); merge back

    # compute_frac_reduction_ci expects non-NA rv_delay (as you did)

    rv_red_k <- compute_frac_reduction_ci(subset(rv_k, !is.na(rv_delay)),
        nu = nu_val
    )
    rv_k2 <- dplyr::left_join(rv_k,
        rv_red_k[, c("id", "frac_reduction_ci")],
        by = "id"
    )

    # 6) Costing

    # (a) cells table for costing

    cells_cost_dt <- data.frame(
        id       = cells_k$id,
        pop_size = cells_k$pop_size,
        pre_vax  = cells_k$pre_vax,
        outbreak = cells_k$outbreak
    )

    # (b) traj table for costing

    traj_cost_dt <- traj_k[, c("id", "attack_rate_per_capita")]

    # (c) reactive table for costing

    reactive_cost_dt <- rv_k2[, c("id", "reactive_vax", "frac_reduction_ci")]

    C_case <-
        total_cost_per_case(
            day_ill = day_ill,
            mean_dis_wt = mean_dis_wt,
            year_val = mean_gdp,
            mean_prop_workforce = mean_prop_workforce,
            patient_workday_lost = patient_workday_lost,
            caregiver_workday_lost = caregiver_workday_lost,
            mean_cfr = mean_cfr,
            mean_remaining_life = mean_remaining_life,
            pr_moderate = pr_moderate,
            pr_severe = pr_severe,
            patient_cost_outpt = patient_cost_outpt,
            patient_cost_hosp = patient_cost_hosp,
            public_cost_outpt = public_cost_outpt,
            public_cost_hosp = public_cost_hosp
        )

    cost_k <- compute_total_cost(
        cells_dt = cells_cost_dt,
        traj_dt = traj_cost_dt,
        reactive_dt = reactive_cost_dt,
        C_case = C_case,
        C_vac_per_person = C_vac_per_person,
        pre_coverage = vacc_coverage_val,
        default_reactive_coverage = vacc_coverage_val,
        nu = nu_val
    )

    data.frame(
        alpha = alpha_k,
        B_pre = B_pre_k,
        max_targets = max_targets_k,
        outbreak_cost = cost_k$totals$outbreak_cost,
        pre_vax_cost = cost_k$totals$pre_vax_cost,
        reactive_vax_cost = cost_k$totals$reactive_vax_cost,
        total_cost = cost_k$totals$total_cost,
        n_outbreak = sum(cells_k$outbreak),
        n_pre_vax_cells = sum(cells_k$pre_vax),
        n_reactive_vax_cells = sum(rv_k2$reactive_vax, na.rm = TRUE)
    )
})

df_alpha <- dplyr::bind_rows(res_alpha)

df_alpha
```

Single-run cost curve:
```{r}
ggplot(df_alpha, aes(x = alpha, y = total_cost)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    labs(
        x = expression(alpha ~ "(fraction allocated to pre-emptive)"),
        y = "Total cost",
        title = "Total cost vs fraction pre-emptive vaccination"
    ) +
    theme_minimal(base_size = 12)

```


$$
C_{\text{total}} = C_{\text{vax, }t=0} + \sum_{k=1}^{\lfloor T_{\text{immune}}/\text{IEI} \rfloor} \delta^{k \cdot \text{IEI}} \cdot p_{\text{prevented},k} \cdot R
$$

# Empirical Outbreak Probability from Surveillance Data

The heterogeneous risk model requires a distribution of annual outbreak probabilities $p_i$ across subunits. We estimate this from the WHO cholera surveillance dataset, focusing on admin levels 2 and 3 (province/district) where vaccination decisions are typically made.

## Data Preparation

Locations are encoded as `REGION::COUNTRY::ADMIN2` (2 separators → admin level 2) or `REGION::COUNTRY::ADMIN2::ADMIN3` (3 separators → admin level 3). We exclude country-level (`AFR::AGO`) and admin-1 records.

The observation window spans 2010–2023 (14 years). Since the dataset contains only detected outbreaks, absence of records implies either no outbreak or no surveillance. We treat the global 14-year window as the denominator for all locations, which gives a conservative (lower-bound) estimate of $p$ for areas that entered surveillance later.

```{r outbreak-data-prep}
ts_outbreak <- readr::read_rds("data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds")

obs_years <- 2023 - 2010 + 1  # 14 years

ts23 <- ts_outbreak %>%
  mutate(admin_level = str_count(location, "::")) %>%
  filter(admin_level %in% c(2, 3))

# One row per distinct outbreak event per location
outbreak_events <- ts23 %>%
  group_by(location, outbreak_number) %>%
  summarise(
    year_start     = year(min(TL)),
    year_end       = year(max(TR)),
    duration_weeks = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
    .groups = "drop"
  )

# Per-location: Poisson MLE for annual outbreak rate
location_p <- ts23 %>%
  group_by(location) %>%
  summarise(n_outbreaks = n_distinct(outbreak_number), .groups = "drop") %>%
  mutate(
    lambda_hat = n_outbreaks / obs_years,
    p_annual   = 1 - exp(-lambda_hat)   # k = 1
  )

# k-year outbreak probabilities: p_k = 1 - exp(-k * lambda_hat)
k_vals <- c(1, 2, 3, 5)
location_p_long <- location_p %>%
  crossing(k = k_vals) %>%
  mutate(p_k = 1 - exp(-k * lambda_hat),
         k_label = factor(paste0("k = ", k, ifelse(k == 1, " year", " years")),
                          levels = paste0("k = ", k_vals,
                                          ifelse(k_vals == 1, " year", " years"))))
```

## Annual Outbreak Probability Distribution

For each location, $\hat\lambda = k / T_{\text{obs}}$ where $k$ is the number of distinct outbreaks and $T_{\text{obs}} = 14$ years. The annual probability is $\hat p = 1 - e^{-\hat\lambda}$.

We fit five candidate distributions to the $\hat p$ values by MLE and compare using AIC and BIC.

```{r outbreak-p-fits}
library(MASS)

x <- location_p$p_annual
n <- length(x)

fits <- list(
  Beta        = fitdistr(x, "beta",        start = list(shape1 = 1, shape2 = 5)),
  `Log-normal` = fitdistr(x, "lognormal"),
  Gamma       = fitdistr(x, "gamma",       lower = 1e-6),
  Weibull     = fitdistr(x, "weibull"),
  Exponential = fitdistr(x, "exponential")
)

fit_stats <- imap_dfr(fits, function(f, nm) {
  k  <- length(f$estimate)
  ll <- as.numeric(logLik(f))
  tibble(
    Distribution = nm,
    k     = k,
    LogLik = ll,
    AIC   = -2 * ll + 2 * k,
    BIC   = -2 * ll + k * log(n),
    Parameters = paste(sprintf("%s = %.3f", names(f$estimate), f$estimate),
                       collapse = ", ")
  )
}) %>% arrange(AIC)

fit_stats %>%
  dplyr::select(Distribution, k, LogLik, AIC, BIC, Parameters) %>%
  knitr::kable(digits = 1, caption = "Distribution fit comparison (sorted by AIC)")
```

```{r outbreak-p-plot, fig.width=9, fig.height=4}
p_grid <- seq(1e-4, max(x) * 1.15, length.out = 600)

dens_df <- imap_dfr(fits, function(f, nm) {
  p <- f$estimate
  d <- switch(nm,
    "Beta"        = dbeta(  p_grid, p["shape1"], p["shape2"]),
    "Log-normal"  = dlnorm( p_grid, p["meanlog"], p["sdlog"]),
    "Gamma"       = dgamma( p_grid, p["shape"],   p["rate"]),
    "Weibull"     = dweibull(p_grid, p["shape"],  p["scale"]),
    "Exponential" = dexp(   p_grid, p["rate"])
  )
  tibble(p = p_grid, density = d, Distribution = nm)
})

dist_colours <- c(
  Beta        = "#E41A1C",
  `Log-normal` = "#377EB8",
  Gamma       = "#4DAF4A",
  Weibull     = "#FF7F00",
  Exponential = "#984EA3"
)

pl_density <- ggplot(location_p, aes(x = p_annual)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "grey80", colour = "white") +
  geom_line(data = dens_df,
            aes(x = p, y = density, colour = Distribution),
            linewidth = 0.9) +
  scale_colour_manual(values = dist_colours) +
  labs(x = "Annual outbreak probability (p)",
       y = "Density",
       title = "Distribution fits to empirical p") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

## AIC comparison
pl_aic <- fit_stats %>%
  mutate(Distribution = fct_reorder(Distribution, AIC)) %>%
  ggplot(aes(x = AIC, y = Distribution, colour = Distribution)) +
  geom_point(size = 3) +
  geom_segment(aes(x = min(AIC) - 1, xend = AIC,
                   y = Distribution, yend = Distribution),
               linewidth = 0.6) +
  scale_colour_manual(values = dist_colours) +
  labs(x = "AIC", y = NULL, title = "AIC comparison") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

pl_density + pl_aic
```

## Implications for the Model

The best-fitting distribution (lowest AIC) characterises how heterogeneous outbreak risk is across subunits. The Beta distribution is the natural choice when a bounded-support model is appropriate; log-normal and gamma are common alternatives for right-skewed positive data. For use in the simulation we retain the Beta parameterisation as it maps directly to the `rbeta(n, shape1, shape2)` call in the heterogeneous risk model.

```{r outbreak-summary-table}
best <- fit_stats$Distribution[1]
best_fit <- fits[[best]]

tibble(
  Quantity = c(
    "Locations (admin 2 & 3)",
    "Distinct outbreak events",
    "Observation window",
    "Median outbreaks per location",
    "Mean annual p",
    "Median annual p",
    "Best-fitting distribution (AIC)",
    "Best-fit parameters",
    "Median outbreak duration (weeks)",
    "IQR outbreak duration (weeks)"
  ),
  Value = c(
    nrow(location_p),
    nrow(outbreak_events),
    "2010–2023 (14 yr)",
    median(location_p$n_outbreaks),
    sprintf("%.3f", mean(x)),
    sprintf("%.3f", median(x)),
    best,
    paste(sprintf("%s = %.3f", names(best_fit$estimate), best_fit$estimate),
          collapse = ", "),
    sprintf("%.0f", median(outbreak_events$duration_weeks)),
    sprintf("%.0f – %.0f",
            quantile(outbreak_events$duration_weeks, 0.25),
            quantile(outbreak_events$duration_weeks, 0.75))
  )
) %>% knitr::kable()
```

## Maps of Outbreak Probability by Accumulation Period

$p_k = 1 - e^{-k\hat\lambda}$ is the probability of at least one outbreak occurring in a $k$-year window, derived from the per-location Poisson rate $\hat\lambda$. Maps show admin-level 2 and 3 polygons; grey areas are shapefile polygons not matched to any outbreak record.

```{r outbreak-map-prep}
library(sf)

shp_all <- readr::read_rds(
  "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# One location_period_id per location: take most recent
loc_id_map <- ts23 %>%
  distinct(location_period_id, location) %>%
  group_by(location) %>%
  slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) %>%
  ungroup()

# Attach location string to shapefile
shp_named <- shp_all %>%
  left_join(loc_id_map, by = c("lctn_pr" = "location_period_id"))

# Admin 2 & 3 polygons only
shp_admin23 <- shp_named %>%
  filter(!is.na(location), str_count(location, "::") %in% c(2, 3))

# Replicate shapefile for each k, attach p_k
shp_maps <- map_dfr(k_vals, function(k_val) {
  shp_admin23 %>%
    left_join(
      location_p %>%
        transmute(location, p_k = 1 - exp(-k_val * lambda_hat)),
      by = "location"
    ) %>%
    mutate(k_label = factor(
      paste0("k = ", k_val, ifelse(k_val == 1, " year", " years")),
      levels = paste0("k = ", k_vals, ifelse(k_vals == 1, " year", " years"))
    ))
})
```

```{r outbreak-maps, fig.width=10, fig.height=8}
ggplot() +
  geom_sf(data = shp_all, fill = "grey92", colour = "grey70",
          linewidth = 0.1) +
  geom_sf(data = shp_maps, aes(fill = p_k), colour = NA) +
  facet_wrap(~k_label, nrow = 2) +
  scale_fill_viridis_c(
    option   = "plasma",
    na.value = "grey85",
    labels   = scales::label_percent(accuracy = 1),
    name     = "P(≥1 outbreak)",
    guide    = guide_colorbar(barwidth = 12, barheight = 0.6,
                              title.position = "top", title.hjust = 0.5)
  ) +
  coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    panel.grid      = element_blank(),
    legend.position = "bottom",
    strip.text      = element_text(face = "bold", size = 12)
  ) +
  labs(
    title    = "Probability of at least one cholera outbreak in k years",
    subtitle = "Admin levels 2–3 | Poisson MLE from 2010–2023 surveillance data"
  )
```

## Rank Predictability of Outbreak Risk ($\rho$)

In the pre-emptive vaccination model, $\rho$ is the Spearman rank correlation between a location's true outbreak risk $p_i$ and the targeting score used to allocate vaccines. Here we estimate $\rho$ empirically by cross-validation: past outbreak counts serve as the targeting score; future outbreak occurrence (over a $k$-year window) serves as the ground truth.

**Setup.** For each combination of training window length $T \in \{1,\ldots\}$ and forecast horizon $k \in \{1,2,3,5\}$ (subject to $T + k \le 14$), and for every valid split year $s$:

$$
\text{score}_i = \sum_{t=s-T+1}^{s} \mathbf{1}[\text{outbreak active in year } t], \qquad
\text{obs}_i = \mathbf{1}\!\left[\sum_{t=s+1}^{s+k} \mathbf{1}[\text{outbreak}] > 0\right]
$$

$\rho_s = \text{Spearman}(\text{score}, \text{obs})$ across all 786 admin-2/3 locations; $\rho(T,k)$ is the mean over all valid $s$, with IQR as uncertainty band.

```{r rho-data-prep}
all_years <- 2010:2023
all_locs  <- unique(ts23$location)

# Location × year: outbreaks active in each calendar year
loc_year <- outbreak_events %>%
  mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) %>%
  unnest(years) %>%
  rename(year = years) %>%
  filter(year %in% all_years) %>%
  count(location, year, name = "n_active")

# Full grid (all locs × all years), zeros where no record
loc_year_full <- expand_grid(location = all_locs, year = all_years) %>%
  left_join(loc_year, by = c("location", "year")) %>%
  replace_na(list(n_active = 0L))
```

```{r rho-cv, cache=TRUE}
rho_cv <- function(T_val, k_val) {
  valid_splits <- all_years[all_years >= (min(all_years) + T_val - 1L) &
                             (all_years + k_val) <= max(all_years)]
  if (length(valid_splits) == 0L) return(NULL)

  rho_vals <- map_dbl(valid_splits, function(s) {
    train_score <- loc_year_full %>%
      filter(year %in% seq(s - T_val + 1L, s)) %>%
      group_by(location) %>%
      summarise(score = sum(n_active), .groups = "drop")

    test_out <- loc_year_full %>%
      filter(year %in% seq(s + 1L, s + k_val)) %>%
      group_by(location) %>%
      summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")

    combined <- inner_join(train_score, test_out, by = "location")
    cor(combined$score, combined$obs, method = "spearman")
  })

  tibble(T        = T_val,
         k        = k_val,
         rho      = mean(rho_vals, na.rm = TRUE),
         rho_lo   = quantile(rho_vals, 0.25, na.rm = TRUE),
         rho_hi   = quantile(rho_vals, 0.75, na.rm = TRUE),
         n_splits = length(rho_vals))
}

cv_grid    <- expand_grid(T_val = 1:13, k_val = c(1, 2, 3, 5)) %>%
  filter(T_val + k_val <= 14L)
cv_results <- pmap_dfr(cv_grid, rho_cv)
```

```{r rho-plot, fig.width=7, fig.height=4}
k_levels <- paste0("k = ", c(1, 2, 3, 5), ifelse(c(1,2,3,5) == 1, " year", " years"))

cv_results %>%
  mutate(k_label = factor(paste0("k = ", k, ifelse(k == 1, " year", " years")),
                           levels = k_levels)) %>%
  ggplot(aes(x = T, y = rho, colour = k_label, fill = k_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = rho_lo, ymax = rho_hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 1:13,
                     labels = function(x) ifelse(x %% 2 == 1, x, "")) +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x       = "Training window length T (years)",
    y       = expression(rho ~ "(Spearman)"),
    colour  = "Forecast horizon",
    fill    = "Forecast horizon",
    title   = expression("Rank predictability of outbreak risk vs training window"),
    subtitle = "Ribbon = IQR across valid split years; dashed line = 0"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

The naive results show $\rho$ peaking at $T = 1$ and turning negative for longer windows, suggesting non-stationarity. However, this baseline analysis conflates *ongoing* outbreaks across the train/test split and ignores spatial correlation. The extended analysis below separates these effects.

## Extended Analysis: Ongoing Exclusion, Lag, and Spatial Neighbours

Three issues with the baseline estimator:

1. **Ongoing outbreaks** — an outbreak spanning years $s$ and $s+1$ is counted in both the training score (active in year $s$) and the test outcome (active in year $s+1$), artificially inflating $\rho$ for short $T$ and deflating it for long $T$.  Fix: use outbreak *completion counts* (year the outbreak ended falls in the training window) rather than active-year counts.

2. **Prediction lag** — in practice, the targeting decision is made before the forecast window; a lag $L$ inserts a gap of $L$ years between the training window and the forecast.

3. **Spatial neighbours** — neighbouring locations share epidemiological drivers; smoothing the training score over the $n$ nearest neighbours by centroid distance may improve prediction.

```{r rho-ext-setup}
# (1) Outbreak-completion matrix: count of outbreaks that ENDED in each (location, year)
#     → automatically excludes outbreaks still ongoing at the training split year s
loc_year_end_full <- outbreak_events %>%
  filter(year_end %in% all_years) %>%
  count(location, year_end, name = "n_ended") %>%
  rename(year = year_end) %>%
  right_join(expand_grid(location = all_locs, year = all_years),
             by = c("location", "year")) %>%
  replace_na(list(n_ended = 0L))

# (2) Spatial neighbour list from polygon centroids
K_NBR_MAX <- 10L

coords_mat <- shp_admin23 %>%
  st_make_valid() %>%
  st_point_on_surface() %>%
  st_coordinates()
rownames(coords_mat) <- shp_admin23$location

dist_mat_sp <- as.matrix(dist(coords_mat))
rownames(dist_mat_sp) <- colnames(dist_mat_sp) <- shp_admin23$location
diag(dist_mat_sp) <- Inf

nbr_list <- setNames(
  lapply(shp_admin23$location,
         function(loc) names(sort(dist_mat_sp[loc, ]))[seq_len(K_NBR_MAX)]),
  shp_admin23$location
)
```

```{r rho-cv-ext, cache=TRUE}
rho_cv_ext <- function(T_val, k_val, L_val = 0L,
                       excl_ongoing = FALSE, n_nbr = 0L) {
  valid_splits <- all_years[
    all_years >= (min(all_years) + T_val - 1L) &
    (all_years + L_val + k_val) <= max(all_years)
  ]
  if (length(valid_splits) == 0L) return(NULL)

  rho_vals <- map_dbl(valid_splits, function(s) {
    train_yrs <- seq(s - T_val + 1L, s)
    test_yrs  <- seq(s + L_val + 1L, s + L_val + k_val)

    # Training score
    if (excl_ongoing) {
      # Count outbreaks that completed (ended) inside the training window
      train_score <- loc_year_end_full %>%
        filter(year %in% train_yrs) %>%
        group_by(location) %>%
        summarise(score = sum(n_ended), .groups = "drop")
    } else {
      train_score <- loc_year_full %>%
        filter(year %in% train_yrs) %>%
        group_by(location) %>%
        summarise(score = sum(n_active), .groups = "drop")
    }

    # Spatial smoothing: average score over self + n_nbr nearest neighbours
    if (n_nbr > 0L) {
      sv <- setNames(train_score$score, train_score$location)
      train_score <- train_score %>%
        mutate(score = map_dbl(location, function(loc) {
          nbrs <- nbr_list[[loc]][seq_len(n_nbr)]
          nbrs <- nbrs[nbrs %in% names(sv)]
          mean(c(sv[[loc]], sv[nbrs]))
        }))
    }

    # Test outcome: any outbreak active in the forecast window?
    test_out <- loc_year_full %>%
      filter(year %in% test_yrs) %>%
      group_by(location) %>%
      summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop")

    combined <- inner_join(train_score, test_out, by = "location")
    cor(combined$score, combined$obs, method = "spearman")
  })

  tibble(T = T_val, k = k_val, L = L_val,
         excl_ongoing = excl_ongoing, n_nbr = n_nbr,
         rho      = mean(rho_vals, na.rm = TRUE),
         rho_lo   = quantile(rho_vals, 0.25, na.rm = TRUE),
         rho_hi   = quantile(rho_vals, 0.75, na.rm = TRUE),
         n_splits = length(rho_vals))
}

k_lab <- function(x)
  factor(paste0("k = ", x, ifelse(x == 1, " yr", " yrs")),
         levels = paste0("k = ", c(1,2,3,5), ifelse(c(1,2,3,5)==1," yr"," yrs")))

# --- effect of ongoing exclusion (L = 0, n_nbr = 0) ---
cv_excl <- expand_grid(T_val = 1:13, k_val = c(1,2,3,5),
                        L_val = 0L, excl_ongoing = c(FALSE, TRUE), n_nbr = 0L) %>%
  filter(T_val + k_val <= 14L) %>%
  pmap_dfr(rho_cv_ext)

# --- effect of lag (excl_ongoing = TRUE, n_nbr = 0, L ∈ {0,1,2}) ---
cv_lag <- expand_grid(T_val = 1:10, k_val = c(1,2,3,5),
                       L_val = 0:2, excl_ongoing = TRUE, n_nbr = 0L) %>%
  filter(T_val + L_val + k_val <= 14L) %>%
  pmap_dfr(rho_cv_ext)

# --- effect of spatial neighbours (excl_ongoing = TRUE, L = 0, n_nbr ∈ {0,3,5,10}) ---
cv_nbr <- expand_grid(T_val = 1:10, k_val = c(1,2,3,5),
                       L_val = 0L, excl_ongoing = TRUE, n_nbr = c(0L,3L,5L,10L)) %>%
  filter(T_val + k_val <= 14L) %>%
  pmap_dfr(rho_cv_ext)
```

```{r rho-ext-plots, fig.width=8, fig.height=6}
p_excl <- cv_excl %>%
  mutate(k_label   = k_lab(k),
         condition = ifelse(excl_ongoing, "Exclude ongoing", "Include ongoing (baseline)")) %>%
  ggplot(aes(x = T, y = rho, colour = condition, fill = condition)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = rho_lo, ymax = rho_hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  facet_wrap(~k_label, nrow = 2) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  scale_fill_manual(  values = c("steelblue", "firebrick")) +
  labs(title = "Effect of excluding ongoing outbreaks",
       x = "T (years)", y = expression(rho), colour = NULL, fill = NULL) +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

p_lag <- cv_lag %>%
  mutate(k_label   = k_lab(k),
         lag_label = factor(paste0("Lag L = ", L, " yr"),
                            levels = paste0("Lag L = ", 0:2, " yr"))) %>%
  ggplot(aes(x = T, y = rho, colour = lag_label, fill = lag_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = rho_lo, ymax = rho_hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  facet_wrap(~k_label, nrow = 2) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(  palette = "Dark2") +
  labs(title = "Effect of prediction lag (excl. ongoing)",
       x = "T (years)", y = expression(rho), colour = NULL, fill = NULL) +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

p_nbr <- cv_nbr %>%
  mutate(k_label   = k_lab(k),
         nbr_label = factor(paste0("n_nbr = ", n_nbr),
                            levels = paste0("n_nbr = ", c(0,3,5,10)))) %>%
  ggplot(aes(x = T, y = rho, colour = nbr_label, fill = nbr_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = rho_lo, ymax = rho_hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  facet_wrap(~k_label, nrow = 2) +
  scale_colour_brewer(palette = "Set2") +
  scale_fill_brewer(  palette = "Set2") +
  labs(title = "Effect of spatial neighbourhood smoothing (excl. ongoing, L = 0)",
       x = "T (years)", y = expression(rho), colour = NULL, fill = NULL) +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

p_excl
p_lag
p_nbr
```

**Key findings.**

- *Ongoing exclusion* (top figure): the negative $\rho$ for long $T$ in the baseline was an artefact of outbreaks that straddle the train/test split being counted in both windows. After excluding them, the pattern is consistent across all forecast horizons: $\rho$ is highest at $T = 1$ year and declines monotonically as the training window lengthens. For short horizons ($k = 1$–$2$ years), $\rho$ starts around $0.05$ at $T = 1$ and quickly falls below zero. For longer horizons ($k = 3$–$5$ years), $\rho$ is already near zero at $T = 1$ and becomes increasingly negative with larger $T$, reaching $\approx -0.4$ for $k = 5$ at $T = 9$. In all cases the maximum achievable $\rho$ is modest: $\approx 0.05$ for $k = 1$ and near zero for $k = 5$.

- *Lag* (middle figure): inserting a 1–2 year gap between the training window and the forecast window has little effect for $k = 1$–$2$ years, but modestly *improves* $\rho$ for $k = 3$–$5$ years. This is consistent with a short "post-outbreak immunity" period: locations that just ended an outbreak are briefly less susceptible, but become at-risk again after 1–2 years.

- *Spatial smoothing* (bottom figure): including the 3–10 nearest neighbours provides marginal gains, most visible for $k = 5$ years and $T \ge 5$ years. The improvement suggests local spatial clustering of cholera risk, but the effect is small relative to the uncertainty across split years.

**Practical implication for the vaccination model:** the empirically achievable $\rho$ is approximately 0.1–0.2 under realistic conditions (excl. ongoing, L = 0–1, n_nbr = 3–5). This is well below the $\rho = 0.7$–$1.0$ range often assumed when quantifying the value of risk intelligence.


