---
title: "Supplementary: Targeting Accuracy (rho) and Contiguous Patch Allocation"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(patchwork)
library(knitr)
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```

## Load Intermediate Data

Objects computed in `vaccine_impact_sim.qmd` are loaded from an `.rds` bridge file.

```{r load-data}
dat <- readRDS("data/vaccine_sim_for_supplementary.rds")
list2env(dat, envir = environment())
```


# Value of Risk Intelligence ($\rho$)

## Theoretical $\alpha^*$ from $\rho$

For each strategy, we plug its empirical $\rho$ into the $\rho$-parameterized
cost function (`cost_popweighted_rho()`) to compute the $\alpha^*$ that the
analytical framework would predict. This lets us compare "what $\rho$ alone
predicts" vs "what the actual scores yield" (MC $\alpha^*$).

```{r theoretical-alpha}
theoretical_alpha <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    rho_c <- rho_by_country |> filter(country == ctry)
    map_dfr(f_values, function(fv) {
        map_dfr(seq_len(nrow(rho_c)), function(i) {
            rho_val <- rho_c$rho[i]
            strat <- rho_c$strategy[i]
            rho_clamped <- max(0, min(1, rho_val))
            res <- get_alpha_star_rho(fv, df_c$pop_sum, df_c$obs_rate,
                R_val, r_val, ATTACK_RATE, rho_clamped,
                n_mc = 200
            )
            tibble(
                country = ctry, strategy = strat, f = fv,
                rho = rho_val, alpha_star_theo = res$alpha_star,
                cost_star_theo = res$cost_star
            )
        })
    })
})

# Look up cost at theoretical alpha* by interpolating each strategy's cost curve
theo_pts <- theoretical_alpha |>
    filter(strategy %in% unique(pw_cost_curves$strategy)) |>
    left_join(
        pw_cost_curves |>
            group_by(country, strategy, f) |>
            summarise(curve = list(tibble(alpha, cost)), .groups = "drop"),
        by = c("country", "strategy", "f")
    ) |>
    rowwise() |>
    mutate(cost_at_theo = approx(curve$alpha, curve$cost, xout = alpha_star_theo)$y) |>
    ungroup() |>
    dplyr::select(-curve)
```

## Cost Curves with Theoretical Overlays

The enriched cost curve plot shows:
- **Solid lines**: empirical MC cost curves by strategy
- **Diamonds**: MC-based optimal $\alpha^*$
- **Triangles**: theoretical $\alpha^*$ predicted from $\rho$ alone
- **Dotted lines**: theoretical $\alpha^*$ positions

```{r cost-curve-rho-overlay, fig.width=14, fig.height=10}
opt_pts <- alpha_comparison |>
    dplyr::select(country, strategy, f, alpha_star_pw, cost_star_pw) |>
    rename(alpha = alpha_star_pw, cost = cost_star_pw)

vline_data <- theo_pts |>
    filter(strategy != "Random") |>
    dplyr::select(country, strategy, f, alpha_star_theo)

ggplot(pw_cost_curves, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    geom_vline(
        data = vline_data,
        aes(xintercept = alpha_star_theo, colour = strategy),
        linetype = "dotted", linewidth = 0.5, alpha = 0.6
    ) +
    geom_point(
        data = opt_pts, aes(x = alpha, y = cost),
        size = 2.5, shape = 18
    ) +
    geom_point(
        data = theo_pts |> filter(strategy != "Random"),
        aes(x = alpha_star_theo, y = cost_at_theo),
        size = 3, shape = 17, alpha = 0.7
    ) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Pop-weighted cost per capita",
        colour = "Strategy",
        title = "Population-weighted cost curves with theoretical rho overlays",
        subtitle = sprintf(
            "R = %.0f, r = %.1f; diamonds = MC optimal, triangles = theoretical (from rho), dotted = theoretical alpha*",
            R_val, r_val
        )
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```


## $\rho$--$\alpha^*$ Decision Curve

How does the quality of risk intelligence ($\rho$) affect the optimal
pre-emptive fraction $\alpha^*$? For each country and coverage level $f$,
we sweep $\rho$ from 0 (random targeting) to 1 (perfect foresight) and compute
the pop-weighted $\alpha^*(\rho)$ curve. The empirical $\rho$ values for each
strategy are overlaid.

```{r rho-alpha-curve, fig.width=14, fig.height=10}
rho_grid <- seq(0, 1, by = 0.05)

rho_alpha_curves <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    map_dfr(f_values, function(fv) {
        map_dfr(rho_grid, function(rho_val) {
            res <- get_alpha_star_rho(fv, df_c$pop_sum, df_c$obs_rate,
                R_val, r_val, ATTACK_RATE,
                rho = rho_val,
                n_mc = 150, seed = 42
            )
            tibble(
                rho = rho_val, alpha_star = res$alpha_star,
                f = fv, country = ctry
            )
        })
    })
})
rho_alpha_curves <- rho_alpha_curves |>
    mutate(country_label = COUNTRY_LABELS[country])

strategy_pts <- alpha_comparison |>
    dplyr::select(country, strategy, f, rho, alpha_star_pw) |>
    mutate(country_label = COUNTRY_LABELS[country])

ggplot(rho_alpha_curves, aes(x = rho, y = alpha_star)) +
    geom_line(linewidth = 0.8, colour = "grey30") +
    geom_point(
        data = strategy_pts,
        aes(x = rho, y = alpha_star_pw, colour = strategy),
        size = 3
    ) +
    facet_grid(country_label ~ sprintf("f = %.1f", f)) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression("Targeting accuracy" ~ rho),
        y = expression("Optimal" ~ alpha^"*" ~ "(pop-weighted)"),
        colour = "Strategy",
        title = expression("How targeting accuracy" ~ rho ~ "shifts optimal" ~ alpha^"*"),
        subtitle = "Grey curve = theoretical; coloured points = empirical strategy positions"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

## Cross-Country Comparison

```{r cross-country-summary}
# Equal-weight analytical alpha* function
get_alpha_star_beta_local <- function(R, f, a, b, r, nu = 1) {
    optim(
        par = 0.5, fn = cost_beta,
        f = f, a = a, b = b, R = R, r = r, nu = nu,
        method = "L-BFGS-B", lower = 0, upper = 1
    )$par
}

country_summary <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    bb <- bb_fits[[ctry]]
    rhos <- alpha_comparison |> filter(country == ctry, f == 0.5)
    tibble(
        Country = COUNTRY_LABELS[ctry],
        n_locations = nrow(df_c),
        `P_total (M)` = round(sum(df_c$pop_sum) / 1e6, 1),
        p_hat = round(bb$p, 3),
        `Beta(a,b)` = sprintf("(%.2f, %.2f)", bb$a, bb$b),
        `alpha*_ew (f=0.5)` = round(
            get_alpha_star_beta_local(R_val, 0.5, bb$a, bb$b, r_val), 3
        ),
        `alpha*_pw Model (f=0.5)` = round(
            rhos |> filter(strategy == "Model-ranked") |> pull(alpha_star_pw), 3
        ),
        `rho_model` = round(
            rhos |> filter(strategy == "Model-ranked") |> pull(rho), 3
        )
    )
})

knitr::kable(country_summary,
    caption = "Cross-country comparison at f = 0.5"
)
```


# Contiguous Patch Allocation (RF-Contiguous Strategy)

The RF-contiguous strategy allocates pre-emptive vaccines to contiguous spatial
patches rather than individual top-ranked locations. This section explores the
contiguous allocation approach, which requires further development to handle
distance-based neighbours with varying patch sizes.

```{r patch-functions}
# Build contiguous patches for one country's locations.
build_country_patches <- function(df_c, score, nbr_list,
                                  n_seeds = NULL, patch_size = NULL,
                                  grow_mode = "best", seed_rng = 42) {
    locs <- df_c$location
    n <- length(locs)
    loc_idx <- setNames(seq_len(n), locs)

    if (is.null(n_seeds))    n_seeds    <- max(3L, as.integer(n / 20))
    if (is.null(patch_size)) patch_size <- max(5L, as.integer(n / 15))

    adj_int <- lapply(locs, function(loc) {
        nbrs <- nbr_list[[loc]]
        if (is.null(nbrs)) return(integer(0))
        nbrs_in <- nbrs[nbrs %in% locs]
        as.integer(loc_idx[nbrs_in])
    })

    patch_out <- make_patches(score = score, adj = adj_int,
                               n_seeds = n_seeds, patch_size = patch_size,
                               grow_mode = grow_mode, seed_rng = seed_rng)
    patch_out$n_patches <- sum(lengths(patch_out$patches) > 0)
    patch_out
}

# Population-weighted MC cost with contiguous-patch pre-emptive allocation.
cost_popweighted_mc_contiguous <- function(alpha, f, pop, prob, score,
                                           R, r, AR, nu = 1,
                                           patch_out,
                                           n_mc = 300, seed = 42, ...) {
    n <- length(pop)
    P_total <- sum(pop)
    total_budget <- f * P_total
    pre_budget <- alpha * total_budget

    patches <- patch_out$patches[lengths(patch_out$patches) > 0]
    if (length(patches) > 0 && pre_budget > 0) {
        patch_scores <- sapply(patches, function(idx) mean(score[idx]))
        patch_pops <- sapply(patches, function(idx) sum(pop[idx]))
        patch_order <- order(patch_scores, decreasing = TRUE)
        cum_patch_pop <- cumsum(patch_pops[patch_order])
        n_pre_patches <- sum(cum_patch_pop <= pre_budget)
        pre_idx <- if (n_pre_patches > 0) {
            unlist(patches[patch_order[seq_len(n_pre_patches)]])
        } else {
            integer(0)
        }
    } else {
        pre_idx <- integer(0)
    }
    rem_idx <- setdiff(seq_len(n), pre_idx)
    doses_used_pre <- sum(pop[pre_idx])
    react_budget <- (1 - alpha) * total_budget

    costs <- numeric(n_mc)
    for (s in seq_len(n_mc)) {
        set.seed(seed + s)
        Y <- rbinom(n, 1, prob)

        pop_pre_cost <- pre_budget
        burden_pre_residual <- sum(Y[pre_idx] * pop[pre_idx]) * (1 - nu)

        ob_rem <- rem_idx[Y[rem_idx] == 1]
        pop_react <- 0
        burden_react_residual <- 0
        burden_uncovered <- 0

        if (length(ob_rem) > 0 && react_budget > 0) {
            ob_order <- ob_rem[order(pop[ob_rem], decreasing = TRUE)]
            cum_pop_r <- cumsum(pop[ob_order])
            n_react <- sum(cum_pop_r <= react_budget)
            if (n_react > 0) {
                react_sel <- ob_order[seq_len(n_react)]
                pop_react <- sum(pop[react_sel])
                burden_react_residual <- pop_react * (1 - r)
                if (n_react < length(ob_order)) {
                    burden_uncovered <- sum(pop[ob_order[(n_react + 1):length(ob_order)]])
                }
            } else {
                burden_uncovered <- sum(pop[ob_rem])
            }
        } else if (length(ob_rem) > 0) {
            burden_uncovered <- sum(pop[ob_rem])
        }

        vax_cost <- pop_pre_cost + pop_react
        illness <- R * AR * (burden_pre_residual + burden_react_residual + burden_uncovered)
        costs[s] <- (vax_cost + illness) / P_total
    }
    mean(costs)
}

get_alpha_star_pw_contiguous <- function(f, pop, prob, score, R, r, AR, nu = 1,
                                         patch_out, n_mc = 300, seed = 42,
                                         grid_len = 51) {
    alpha_grid <- seq(0, 1, length.out = grid_len)
    costs <- sapply(alpha_grid, function(a) {
        cost_popweighted_mc_contiguous(a, f, pop, prob, score, R, r, AR, nu,
            patch_out = patch_out,
            n_mc = n_mc, seed = seed
        )
    })
    list(
        alpha_star = alpha_grid[which.min(costs)],
        cost_star = min(costs), alpha_grid = alpha_grid, cost_grid = costs
    )
}
```

```{r build-patches}
# Pre-build contiguous patches per country using RF scores.
country_patches <- setNames(lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    patch_out <- build_country_patches(df_c,
        score = df_c$pred_rf,
        nbr_list = nbr_list, seed_rng = 42
    )
    cat(sprintf(
        "%s: %d patches, covering %d/%d locations (%.0f%%)\n",
        COUNTRY_LABELS[ctry], patch_out$n_patches,
        sum(patch_out$patch_id > 0), nrow(df_c),
        100 * mean(patch_out$patch_id > 0)
    ))
    patch_out
}), SELECTED_COUNTRIES)
```

```{r contiguous-cost-curves, fig.width=14, fig.height=10}
# Add RF-contiguous strategy colours
STRATEGY_COLOURS_CONTIG <- c(STRATEGY_COLOURS,
    "RF-contiguous" = "darkorange"
)

alpha_grid_pw <- seq(0, 1, length.out = 51)

contig_curves <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    patch_out_c <- country_patches[[ctry]]
    map_dfr(f_values, function(fv) {
        costs <- sapply(alpha_grid_pw, function(a) {
            cost_popweighted_mc_contiguous(a, fv, df_c$pop_sum, df_c$obs_rate,
                df_c$pred_rf, R_val, r_val, ATTACK_RATE,
                patch_out = patch_out_c, n_mc = 200
            )
        })
        tibble(
            alpha = alpha_grid_pw, cost = costs,
            strategy = "RF-contiguous", f = fv, country = ctry
        )
    })
})

# Combine with individual-strategy curves from main analysis
all_curves <- bind_rows(pw_cost_curves, contig_curves)

ggplot(all_curves, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS_CONTIG) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Pop-weighted cost per capita",
        colour = "Strategy",
        title = "Cost curves including RF-contiguous (patch-based) strategy"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```
