---
title: "Optimal OCV Allocation: Vaccine Impact Analysis (2010–2012)"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document analyses the optimal allocation of a fixed oral cholera vaccine
(OCV) supply between **pre-emptive** (before outbreaks) and **reactive**
(during outbreaks) deployment, using ADM2-level cholera surveillance data from
the **2010–2012** period as the evaluation window.

**Core question**: Given that vaccines can cover only a fraction $f$ of the
total population, what fraction $\alpha$ should be deployed pre-emptively —
targeting high-risk areas in advance — versus held in reserve for rapid reactive
response when outbreaks are detected?

**Key model elements**:

| Parameter | Symbol | Value |
|-----------|--------|-------|
| Vaccination capacity (fraction of population) | $f$ | 0.30 |
| Pre-emptive vaccine effectiveness | $\nu$ | 1.00 |
| Reactive vaccine effectiveness | $r$ | 0.40 |
| Cost ratio (illness cost / vaccine cost per person) | $R$ | 100, 500 |
| Targeting accuracy (Spearman $\rho$) | $\rho$ | 0, 0.3, 1 |
| Pre-emptive fraction | $\alpha$ | 0 → 1 |

The targeting accuracy $\rho$ measures how well the pre-vaccination prioritisation
ranks sites by true future burden: $\rho = 0$ means random allocation,
$\rho = 1$ means perfect oracle knowledge, $\rho = 0.3$ reflects realistic
surveillance-based targeting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(patchwork)
library(knitr)
source("R/analytic_utils.R")
```

---

# Parameters

```{r params}
WIN_YEARS <- 2010:2012   # evaluation window
T_WIN     <- length(WIN_YEARS)   # 3 years

F_VAX    <- 0.60   # vaccination capacity (fraction of total population)
NU_VAL   <- 1.00   # pre-emptive VE
R_REACT  <- 0.40   # reactive VE
C_COV    <- 1.00   # within-location vaccine coverage (fraction of pop in site)
Q_THRESH <- 0      # reactive threshold: vaccinate all outbreak sites (q* = 0)

R_VALS   <- c(100, 500)        # cost ratios to analyse
R_BASE   <- 100                 # baseline R for primary figures
RHO_VALS <- c(0, 0.3, 1.0)    # targeting accuracy levels
ALPHA_GRID <- seq(0, 1, length.out = 201)

cat(sprintf("Window: %d–%d (%d years)\n", min(WIN_YEARS), max(WIN_YEARS), T_WIN))
cat(sprintf("Vaccination capacity f = %.2f\n", F_VAX))
cat(sprintf("Pre-emptive VE nu = %.2f,  Reactive VE r = %.2f\n", NU_VAL, R_REACT))
cat(sprintf("R (baseline) = %d\n", R_BASE))
```

---

# Data Preparation

## Case surveillance (2010–2012)

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds")
ts2 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

STUDY_YEARS <- 2010:2019
outbreak_events <- ts2 |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(year_start  = year(min(TL)),
              total_cases = sum(sCh, na.rm = TRUE), .groups = "drop")
cases_annual <- outbreak_events |>
    filter(year_start %in% STUDY_YEARS) |>
    group_by(location, year = year_start) |>
    summarise(cases = sum(total_cases, na.rm = TRUE), .groups = "drop")

all_locs <- unique(ts2$location)
cases_panel <- cases_annual |>
    right_join(expand_grid(location = all_locs, year = STUDY_YEARS),
               by = c("location", "year")) |>
    replace_na(list(cases = 0))
```

## Join with incidence rates and population

Population and the 10-year incidence rate are loaded from the pre-computed cache
produced by `incidence_rate.qmd`.

```{r join-pop}
cases_fit <- readr::read_rds("data/cases_fit_cache.rds")

# 3-year case totals for 2010-2012
cases_3yr <- cases_panel |>
    filter(year %in% WIN_YEARS) |>
    group_by(location) |>
    summarise(
        cases_win      = sum(cases),
        n_outbreak_yrs = sum(cases > 0),
        any_outbreak   = as.integer(sum(cases) > 0),
        .groups = "drop"
    )

# Merge: keep only sites with known population
sim_df <- cases_fit |>
    dplyr::select(location, pop_sum, rate_obs) |>
    left_join(cases_3yr, by = "location") |>
    replace_na(list(cases_win = 0L, n_outbreak_yrs = 0L, any_outbreak = 0L)) |>
    filter(pop_sum > 0) |>
    # Burden per person over the 3-year window
    mutate(q_i = cases_win / pop_sum)

# First-outbreak date and first-year cases per site within WIN_YEARS.
#
# Reactive deployment follows outbreak chronology: the vaccine stockpile is
# allocated in order of outbreak start date (earliest-triggered sites first),
# until doses run out.  At each vaccinated site:
#   - r = 40% of first-year cases are averted (vaccine arrives mid-outbreak)
#   - vaccine-induced immunity prevents any subsequent outbreak years in the window
#
# date_first : exact start date of the first outbreak at each site within WIN_YEARS
#              (used to order reactive deployment chronologically)
# cases_first: case count in that first outbreak year (used to compute residual illness)
date_first_df <- ts2 |>
    filter(year(TL) %in% WIN_YEARS) |>
    group_by(location) |>
    summarise(date_first = min(TL), .groups = "drop")

cases_first_df <- cases_panel |>
    filter(year %in% WIN_YEARS, cases > 0) |>
    group_by(location) |>
    slice_min(year, n = 1, with_ties = FALSE) |>
    rename(cases_first = cases) |>
    dplyr::select(location, cases_first)

sim_df <- sim_df |>
    left_join(cases_first_df, by = "location") |>
    left_join(date_first_df,  by = "location") |>
    replace_na(list(cases_first = 0L)) |>
    # Sites with no outbreak get a sentinel far-future date so they sort last
    mutate(
        date_first        = if_else(is.na(date_first), as.Date("9999-12-31"), date_first),
        cases_subsequent  = cases_win - cases_first
    )

P_TOTAL  <- sum(sim_df$pop_sum)
N_SITES  <- nrow(sim_df)

cat(sprintf("Sites: %d  |  Total population: %s\n",
            N_SITES, format(round(P_TOTAL), big.mark = ",")))
cat(sprintf("Sites with ≥1 case in %d–%d: %d (%.1f%%)\n",
            min(WIN_YEARS), max(WIN_YEARS),
            sum(sim_df$any_outbreak), 100 * mean(sim_df$any_outbreak)))
cat(sprintf("Total cases %d–%d: %s  (first-year: %s, subsequent: %s)\n",
            min(WIN_YEARS), max(WIN_YEARS),
            format(sum(sim_df$cases_win),        big.mark = ","),
            format(sum(sim_df$cases_first),      big.mark = ","),
            format(sum(sim_df$cases_subsequent), big.mark = ",")))
```

---

# Outbreak Probability

The **annual outbreak probability** per site is estimated from the fraction of
years in the 3-year window with at least one reported case.

```{r outbreak-prob}
p_3yr  <- mean(sim_df$any_outbreak)     # empirical P(any outbreak in 3-yr window)
p_ann  <- 1 - (1 - p_3yr)^(1/T_WIN)    # implied annual probability

cat(sprintf("Empirical P(≥1 outbreak year in %d–%d): %.3f\n",
            min(WIN_YEARS), max(WIN_YEARS), p_3yr))
cat(sprintf("Implied annual outbreak probability:    %.3f\n", p_ann))
cat(sprintf("Mean q_i (cases per person, 3-yr):     %.5f  (%.1f per 100k)\n",
            mean(sim_df$q_i), mean(sim_df$q_i) * 1e5))

# Distribution table
sim_df |>
    count(n_outbreak_yrs, name = "n_sites") |>
    mutate(pct = round(100 * n_sites / N_SITES, 1)) |>
    knitr::kable(
        col.names = c("Outbreak years (out of 3)", "N sites", "% sites"),
        caption   = sprintf("Distribution of outbreak frequency per ADM2 site, %d–%d",
                            min(WIN_YEARS), max(WIN_YEARS)))
```

```{r fig-q-dist, fig.width=9, fig.height=4}
#| fig-cap: "Left: distribution of the per-person illness burden q_i (cases per
#|   person over the 3-year window) on a log scale, across sites with at least
#|   one case. Right: empirical CDF. Both panels exclude zero-burden sites."

df_pos <- sim_df |> filter(q_i > 0)

p_hist <- ggplot(df_pos, aes(x = q_i * 1e5)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 30, fill = "steelblue", colour = "white",
                   linewidth = 0.2) +
    scale_x_log10(labels = scales::comma) +
    labs(x = "Cases per 100,000 persons (3-yr, log scale)", y = "Density",
         title = sprintf("Illness burden q_i (n = %d positive sites)", nrow(df_pos))) +
    theme_minimal(base_size = 11)

p_cdf <- df_pos |>
    arrange(q_i) |>
    mutate(ecdf = seq_len(n()) / n()) |>
    ggplot(aes(x = q_i * 1e5, y = ecdf)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    scale_x_log10(labels = scales::comma) +
    labs(x = "Cases per 100,000 persons (3-yr, log scale)",
         y = "Cumulative proportion of sites",
         title = "ECDF (positive sites only)") +
    theme_minimal(base_size = 11)

p_hist + p_cdf
```

---

# Theoretical Optimal Allocation (Homogeneous Case)

In the **homogeneous** model every location is assumed to have the same outbreak
probability $p$. Under this simplification the optimal pre-emptive fraction has a closed-form solution.

**Intuition.** Vaccines deployed pre-emptively are certain to be used and avert
nearly all illness at the targeted site ($\nu = 1$). Reactive vaccines are
deployed only when an outbreak is already under way; they avert fraction $r$
of illness but are wasted if no outbreak occurs where they are stockpiled.

- If $f > p$ (capacity exceeds outbreak probability — "abundant" regime): both
  pre-emptive and reactive can cover all outbreaks, but the trade-off depends
  on $r$ vs $\nu$.
- If $f \leq p$ (capacity is scarce): the marginal deployed dose competes
  between preventing a nearly-certain outbreak pre-emptively vs covering a
  realised outbreak reactively.

```{r alpha-star-theory}
# alpha_star_equalrisk is from R/analytic_utils.R
# Returns the optimal alpha under the homogeneous equal-risk case.

theory_tbl <- expand_grid(R = R_VALS) |>
    mutate(
        p        = p_3yr,
        f        = F_VAX,
        r        = R_REACT,
        nu       = NU_VAL,
        regime   = ifelse(f > p, "Abundant (f > p)", "Scarce (f ≤ p)"),
        alpha_star_theory = mapply(alpha_star_equalrisk,
                                   p = p, R = R, r = r, f = f, nu = nu)
    )

theory_tbl |>
    dplyr::select(R, p, f, regime, alpha_star_theory) |>
    knitr::kable(
        digits    = c(0, 3, 2, 0, 3),
        col.names = c("R", "p (3-yr)", "f", "Capacity regime", "alpha* (theory)"),
        caption   = sprintf(
            "Theoretical optimal pre-emptive fraction under homogeneous risk | nu = %.1f, r = %.1f",
            NU_VAL, R_REACT))

cat(sprintf("\nNote: f = %.2f, p = %.3f → %s capacity regime.\n",
            F_VAX, p_3yr, ifelse(F_VAX > p_3yr, "Abundant", "Scarce")))
cat(sprintf("Threshold condition: r vs p*nu = %.3f vs %.3f\n",
            R_REACT, p_3yr * NU_VAL))
cat(sprintf("Since r %s p*nu, the homogeneous theory predicts alpha* = %s.\n",
            ifelse(R_REACT > p_3yr * NU_VAL, ">", "<"),
            ifelse(R_REACT > p_3yr * NU_VAL, "0 (pure reactive)", "1 (pure pre-emptive)")))
```

---

# Helper Functions

```{r helper-fns}
# Generate a targeting priority score with a target Spearman correlation to
# the true burden. rho = 0 → random; rho = 1 → oracle (perfect ranking).
make_rank_score <- function(truth, target_rho, seed = 42) {
    set.seed(seed)
    n       <- length(truth)
    z_truth <- qnorm(rank(truth, ties.method = "random") / (n + 1))
    z_noise <- rnorm(n)
    rho_p   <- 2 * sin(pi * target_rho / 6)
    pnorm(rho_p * z_truth + sqrt(1 - rho_p^2) * z_noise)
}

# Total cost c(alpha), per-capita (normalised by total population, in C_V units).
#
# alpha       : pre-emptive fraction in [0, 1]
# f           : vaccination capacity (fraction of total population)
# q           : per-person illness burden (observed cases / population per site)
# cases_first : cases in the first outbreak year at each site (reactive trigger year)
# pop         : population per site — vaccination cost scales with pop_i
# score       : priority score for pre-emptive targeting (higher → vaccinate first)
# R           : cost ratio C_I / C_V (illness cost per case / vaccine cost per person)
# r           : reactive VE: fraction of first-year cases averted (= 0.40)
# nu          : pre-emptive VE (default 1 → all cases averted)
# C_cov       : within-location vaccination coverage (default 1 → full population)
# q_star      : reactive trigger threshold (default 0 → any site with cases triggers)
#
# Reactive model: vaccination arrives during the first outbreak year at each site.
# It averts r of that year's cases and confers immunity preventing any subsequent
# outbreak years in the window. Vaccination cost = pop_i × C_V per reactive site.
cost_incidence <- function(alpha, f, q, cases_first, date_first, pop, score, R, r,
                            nu = 1, C_cov = 1, q_star = 0) {
    P       <- sum(pop)
    all_idx <- seq_along(q)

    # Pre-emptive: rank by score, fill budget alpha * f * P
    f_pre_cap  <- alpha * f * P
    ord_pre    <- order(score, decreasing = TRUE)
    n_pre      <- sum(cumsum(pop[ord_pre]) <= f_pre_cap)
    pre_idx    <- ord_pre[seq_len(max(n_pre, 0L))]

    q_bar_pre  <- if (length(pre_idx) > 0) weighted.mean(q[pre_idx], pop[pre_idx]) else 0
    f_pre_used <- sum(pop[pre_idx]) / P
    c_pre      <- f_pre_used * (C_cov + q_bar_pre * (1 - nu) * R)

    # Reactive: remaining sites that have outbreaks (q > q_star)
    rem_idx       <- setdiff(all_idx, pre_idx)
    triggered_idx <- rem_idx[q[rem_idx] > q_star]
    no_action_idx <- rem_idx[q[rem_idx] <= q_star]

    c_no_action <- R * sum(q[no_action_idx] * pop[no_action_idx]) / P

    f_react_cap <- (1 - alpha) * f * P
    if (length(triggered_idx) > 0) {
        # Deploy reactively in chronological order of first outbreak start date
        ord_react  <- triggered_idx[order(date_first[triggered_idx])]
        n_react    <- sum(cumsum(pop[ord_react]) <= f_react_cap)
        react_idx  <- ord_react[seq_len(max(n_react, 0L))]
        missed_idx <- setdiff(ord_react, react_idx)

        # Vaccination cost: pop_i × C_cov per vaccinated site
        # Residual illness: (1-r) × first-year cases; subsequent years fully prevented
        c_react  <- C_cov * sum(pop[react_idx]) / P +
                    R * (1 - r) * sum(cases_first[react_idx]) / P
        c_missed <- R * sum(q[missed_idx] * pop[missed_idx]) / P
    } else {
        c_react  <- 0
        c_missed <- 0
    }

    c_pre + c_react + c_missed + c_no_action
}

# Cases averted breakdown for a given allocation.
# Uses same reactive model as cost_incidence(): r reduces first-year cases;
# subsequent outbreak years at each reactive site are fully prevented.
cases_averted_breakdown <- function(alpha, f, q, cases_first, date_first, pop, score, R, r,
                                     nu = 1, C_cov = 1, q_star = 0) {
    P       <- sum(pop)
    all_idx <- seq_along(q)

    f_pre_cap <- alpha * f * P
    ord_pre   <- order(score, decreasing = TRUE)
    n_pre     <- sum(cumsum(pop[ord_pre]) <= f_pre_cap)
    pre_idx   <- ord_pre[seq_len(max(n_pre, 0L))]

    rem_idx       <- setdiff(all_idx, pre_idx)
    triggered_idx <- rem_idx[q[rem_idx] > q_star]

    f_react_cap <- (1 - alpha) * f * P
    if (length(triggered_idx) > 0) {
        # Deploy in chronological order of first outbreak start date
        ord_react  <- triggered_idx[order(date_first[triggered_idx])]
        n_react    <- sum(cumsum(pop[ord_react]) <= f_react_cap)
        react_idx  <- ord_react[seq_len(max(n_react, 0L))]
        missed_idx <- setdiff(ord_react, react_idx)
    } else {
        react_idx  <- integer(0)
        missed_idx <- integer(0)
    }

    cases_win     <- q * pop
    total_cases   <- sum(cases_win)
    cases_pre     <- sum(cases_win[pre_idx])
    cases_rx_1st  <- r * sum(cases_first[react_idx])
    cases_rx_sub  <- sum(cases_win[react_idx]) - sum(cases_first[react_idx])
    cases_rx      <- cases_rx_1st + cases_rx_sub
    cases_averted <- cases_pre + cases_rx
    vax_pop       <- sum(pop[pre_idx]) + sum(pop[react_idx])

    list(
        total_cases_no_vax   = total_cases,
        cases_averted        = cases_averted,
        cases_averted_pre    = cases_pre,
        cases_averted_rx     = cases_rx,
        cases_averted_rx_1st = cases_rx_1st,
        cases_averted_rx_sub = cases_rx_sub,
        pct_averted          = 100 * cases_averted / max(total_cases, 1),
        pop_vaccinated       = vax_pop,
        pct_pop_vaccinated   = 100 * vax_pop / P
    )
}
```

---

# Cost Analysis

## Cost curves by $\alpha$ and $\rho$

For each combination of targeting accuracy $\rho$ and cost ratio $R$, we sweep
$\alpha$ from 0 (pure reactive) to 1 (pure pre-emptive) and evaluate the
normalised expected cost $c(\alpha)$.

The **no-vaccination baseline** cost (no vaccines deployed) is simply
$c_0 = R \cdot \bar{q}$, where $\bar{q}$ is the population-weighted mean
illness burden per person (observed cases / population over the window).

**Reactive vaccination model**: reactive vaccines arrive at the start of
a site's first outbreak year. They avert $r = 40\%$ of that year's cases and,
through vaccine-induced immunity, prevent any subsequent outbreak years within
the window. Vaccination cost reflects the full regional population ($\text{pop}_i
\times C_V$ per reactive site).

```{r cost-curves}
c_no_vax_by_R <- setNames(
    sapply(R_VALS, function(Rv) Rv * weighted.mean(sim_df$q_i, sim_df$pop_sum)),
    R_VALS)
cat("No-vaccination cost (per capita, in C_V units):\n")
for (Rv in R_VALS)
    cat(sprintf("  R = %4d:  c0 = %.4f\n", Rv, c_no_vax_by_R[as.character(Rv)]))

# Sweep alpha for all (rho, R) combinations.
# Explicit rho_val/R_val copies prevent closure capture from the for-loop above.
cost_df <- expand_grid(rho = RHO_VALS, R = R_VALS) |>
    pmap_dfr(function(rho, R) {
        rho_val <- rho; R_val <- R
        score    <- make_rank_score(sim_df$q_i, target_rho = rho_val)
        cost_vec <- vapply(ALPHA_GRID, function(a)
            cost_incidence(a, F_VAX, sim_df$q_i, sim_df$cases_first, sim_df$date_first,
                           sim_df$pop_sum, score, R_val, R_REACT, NU_VAL, C_COV, Q_THRESH),
            FUN.VALUE = numeric(1))
        tibble(rho = rho_val, R = R_val, alpha = ALPHA_GRID, cost = cost_vec)
    })

opt_df <- cost_df |>
    group_by(rho, R) |>
    slice_min(cost, n = 1, with_ties = FALSE) |>
    ungroup() |>
    rename(alpha_star = alpha, cost_min = cost)

# Attach no-vax baseline
opt_df <- opt_df |>
    mutate(c_no_vax = c_no_vax_by_R[as.character(R)],
           cost_reduction_pct = 100 * (c_no_vax - cost_min) / c_no_vax)

opt_df |>
    dplyr::select(R, rho, alpha_star, cost_min, c_no_vax, cost_reduction_pct) |>
    arrange(R, rho) |>
    knitr::kable(
        digits    = c(0, 2, 3, 4, 4, 1),
        col.names = c("R", "rho", "alpha*", "Min cost", "No-vax cost", "Cost reduction (%)"),
        caption   = sprintf(
            "Optimal pre-emptive fraction and cost reduction | f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT))
```

```{r fig-cost-curves, fig.width=11, fig.height=7}
#| fig-cap: "Expected cost c(alpha) as a function of the pre-emptive fraction
#|   alpha, for each combination of targeting accuracy rho (rows) and cost ratio
#|   R (columns). The blue curve is the cost with vaccination; the red dashed
#|   vertical line marks the optimal alpha*; the grey dashed horizontal line is
#|   the no-vaccination baseline c0."

no_vax_df <- tibble(R = R_VALS,
                    c_no_vax = unname(c_no_vax_by_R))

ggplot(cost_df, aes(x = alpha, y = cost)) +
    geom_hline(data = no_vax_df, aes(yintercept = c_no_vax),
               linetype = "dashed", colour = "grey50", linewidth = 0.7) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    geom_vline(data = opt_df, aes(xintercept = alpha_star),
               linetype = "dashed", colour = "firebrick", linewidth = 0.8) +
    geom_point(data = opt_df, aes(x = alpha_star, y = cost_min),
               colour = "firebrick", size = 2.5) +
    facet_grid(
        rows = vars(sprintf("rho = %.1f", rho)),
        cols = vars(sprintf("R = %d", R))
    ) +
    labs(x = expression("Pre-emptive fraction " ~ alpha),
         y = expression("Normalised cost " ~ italic(c)(alpha)),
         title = sprintf("Cost curves  |  f = %.2f, nu = %.1f, r = %.1f",
                         F_VAX, NU_VAL, R_REACT),
         subtitle = "Grey dashed: no-vaccination cost  |  Red dashed: optimal alpha*") +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold"))
```

## Optimal $\alpha^*$ vs targeting accuracy

```{r fig-alpha-star, fig.width=7, fig.height=4}
#| fig-cap: "Optimal pre-emptive fraction alpha* as a function of targeting
#|   accuracy rho, for two cost ratios R. Higher targeting accuracy shifts
#|   resources towards pre-emptive deployment by reducing wastage from
#|   vaccinating low-risk sites."

ggplot(opt_df, aes(x = rho, y = alpha_star,
                   colour = factor(R), group = factor(R))) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 3.5) +
    scale_colour_manual(values = c("100" = "steelblue", "500" = "firebrick"),
                        name = "Cost ratio R") +
    scale_x_continuous(breaks = RHO_VALS) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(x = expression("Targeting accuracy " ~ rho),
         y = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
         title = sprintf("alpha* vs rho  |  f = %.2f, nu = %.1f, r = %.1f",
                         F_VAX, NU_VAL, R_REACT)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

---

# Cases Averted

```{r cases-averted}
ca_df <- expand_grid(rho = RHO_VALS, R = R_VALS) |>
    pmap_dfr(function(rho, R) {
        rho_val <- rho; R_val <- R
        score  <- make_rank_score(sim_df$q_i, target_rho = rho_val)
        a_star <- opt_df$alpha_star[opt_df$rho == rho_val & opt_df$R == R_val]

        # At optimal alpha
        ca_opt <- cases_averted_breakdown(a_star, F_VAX, sim_df$q_i,
                                          sim_df$cases_first, sim_df$date_first,
                                          sim_df$pop_sum, score, R_val, R_REACT,
                                          NU_VAL, C_COV, Q_THRESH)
        # At pure reactive (alpha=0)
        ca_rx  <- cases_averted_breakdown(0, F_VAX, sim_df$q_i,
                                          sim_df$cases_first, sim_df$date_first,
                                          sim_df$pop_sum, score, R_val, R_REACT,
                                          NU_VAL, C_COV, Q_THRESH)
        # At pure pre-emptive (alpha=1)
        ca_pre <- cases_averted_breakdown(1, F_VAX, sim_df$q_i,
                                          sim_df$cases_first, sim_df$date_first,
                                          sim_df$pop_sum, score, R_val, R_REACT,
                                          NU_VAL, C_COV, Q_THRESH)

        tibble(
            rho       = rho_val,
            R         = R_val,
            alpha_star = a_star,
            cases_no_vax      = round(ca_opt$total_cases_no_vax),
            cases_averted_opt = round(ca_opt$cases_averted),
            pct_averted_opt   = round(ca_opt$pct_averted, 1),
            cases_averted_rx  = round(ca_rx$cases_averted),
            pct_averted_rx    = round(ca_rx$pct_averted, 1),
            cases_averted_pre = round(ca_pre$cases_averted),
            pct_averted_pre   = round(ca_pre$pct_averted, 1)
        )
    })

ca_df |>
    dplyr::select(R, rho, alpha_star,
                  cases_no_vax,
                  cases_averted_opt, pct_averted_opt,
                  cases_averted_rx,  pct_averted_rx,
                  cases_averted_pre, pct_averted_pre) |>
    arrange(R, rho) |>
    knitr::kable(
        format.args = list(big.mark = ","),
        col.names   = c("R", "rho", "alpha*",
                        "Cases (no vax)",
                        "Averted (opt alpha)", "% averted",
                        "Averted (alpha=0)", "% averted",
                        "Averted (alpha=1)", "% averted"),
        caption = sprintf(
            "Cases averted under optimal, pure-reactive, and pure-pre-emptive allocation | f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT))
```

```{r fig-cases-averted, fig.width=9, fig.height=5}
#| fig-cap: "Cases averted (% of no-vaccination baseline) as a function of the
#|   pre-emptive fraction alpha, for each targeting accuracy rho and R = 100
#|   (baseline). The optimal alpha* (red dashed) maximises cost-efficiency, not
#|   necessarily cases averted."

score_list <- setNames(
    lapply(RHO_VALS, function(rho)
        make_rank_score(sim_df$q_i, target_rho = rho)),
    RHO_VALS)

averted_df <- expand_grid(rho = RHO_VALS, alpha = ALPHA_GRID) |>
    mutate(
        pct_averted = mapply(function(rho, a) {
            ca <- cases_averted_breakdown(
                a, F_VAX, sim_df$q_i, sim_df$cases_first, sim_df$date_first,
                sim_df$pop_sum, score_list[[as.character(rho)]], R_BASE, R_REACT,
                NU_VAL, C_COV, Q_THRESH)
            ca$pct_averted
        }, rho, alpha)
    )

opt_averted <- opt_df |>
    filter(R == R_BASE) |>
    left_join(
        averted_df |>
            group_by(rho) |>
            slice(which.min(abs(alpha - opt_df$alpha_star[opt_df$R == R_BASE &
                                                          opt_df$rho == first(rho)]))),
        by = "rho"
    )

ggplot(averted_df, aes(x = alpha, y = pct_averted,
                        colour = factor(rho), group = factor(rho))) +
    geom_line(linewidth = 0.9) +
    geom_vline(data = opt_df |> filter(R == R_BASE),
               aes(xintercept = alpha_star, colour = factor(rho)),
               linetype = "dashed", linewidth = 0.7) +
    scale_colour_viridis_d(option = "plasma", end = 0.85,
                           name = expression("Targeting accuracy " ~ rho)) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(x = expression("Pre-emptive fraction " ~ alpha),
         y = "Cases averted (%)",
         title = sprintf("Cases averted vs alpha  |  f = %.2f, R = %d, r = %.1f",
                         F_VAX, R_BASE, R_REACT),
         subtitle = "Dashed lines: optimal alpha* for each rho") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

---

# Site-level Breakdown at Optimal Allocation

```{r fig-site-map-setup}
# Compute site-level allocation for each rho at optimal alpha, R = R_BASE
site_alloc <- lapply(RHO_VALS, function(rho) {
    score   <- make_rank_score(sim_df$q_i, target_rho = rho)
    a_star  <- opt_df$alpha_star[opt_df$rho == rho & opt_df$R == R_BASE]
    P       <- sum(sim_df$pop_sum)

    f_pre_cap <- a_star * F_VAX * P
    ord_pre   <- order(score, decreasing = TRUE)
    n_pre     <- sum(cumsum(sim_df$pop_sum[ord_pre]) <= f_pre_cap)
    pre_idx   <- ord_pre[seq_len(max(n_pre, 0L))]

    rem_idx       <- setdiff(seq_len(nrow(sim_df)), pre_idx)
    triggered_idx <- rem_idx[sim_df$q_i[rem_idx] > Q_THRESH]
    f_react_cap   <- (1 - a_star) * F_VAX * P
    if (length(triggered_idx) > 0) {
        ord_r    <- triggered_idx[order(sim_df$q_i[triggered_idx], decreasing = TRUE)]
        n_r      <- sum(cumsum(sim_df$pop_sum[ord_r]) <= f_react_cap)
        react_idx <- ord_r[seq_len(max(n_r, 0L))]
    } else { react_idx <- integer(0) }

    status <- rep("No vaccination", nrow(sim_df))
    status[pre_idx]   <- "Pre-emptive"
    status[react_idx] <- "Reactive"

    sim_df |>
        mutate(status = status, rho_label = sprintf("rho = %.1f", rho))
})
site_df <- bind_rows(site_alloc)
```

```{r fig-site-burden, fig.width=11, fig.height=5}
#| fig-cap: "Per-site illness burden q_i (cases per 100,000 persons in 2010–2012)
#|   coloured by vaccination status at the optimal alpha* for R = 100. Sites are
#|   sorted by q_i within each panel. Pre-emptive vaccination concentrates on
#|   high-burden sites; reactive coverage fills in additional outbreak sites."

site_df |>
    filter(q_i > 0) |>
    arrange(desc(q_i)) |>
    group_by(rho_label) |>
    mutate(rank = row_number()) |>
    ggplot(aes(x = rank, y = q_i * 1e5, colour = status)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_y_log10(labels = scales::comma) +
    scale_colour_manual(
        values = c("Pre-emptive" = "firebrick", "Reactive" = "steelblue",
                   "No vaccination" = "grey70"),
        name = "Vaccination status") +
    facet_wrap(~rho_label, nrow = 1) +
    labs(x = "Site rank (by burden)",
         y = "Cases per 100,000 (log scale)",
         title = sprintf("Site-level vaccination assignment at optimal alpha* | R = %d", R_BASE)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

---

# Summary and Interpretation

```{r summary-table}
summary_tbl <- opt_df |>
    filter(R == R_BASE) |>
    left_join(ca_df |> filter(R == R_BASE) |>
              dplyr::select(rho, cases_no_vax, cases_averted_opt, pct_averted_opt),
              by = "rho") |>
    mutate(
        rho_label = case_when(
            rho == 0   ~ "Random (rho=0)",
            rho == 0.3 ~ "Moderate (rho=0.3)",
            rho == 1.0 ~ "Oracle (rho=1)"),
        alpha_theory = theory_tbl$alpha_star_theory[match(R_BASE, theory_tbl$R)]
    )

summary_tbl |>
    dplyr::select(rho_label, alpha_star, alpha_theory,
                  cost_min, c_no_vax, cost_reduction_pct,
                  cases_averted_opt, pct_averted_opt) |>
    knitr::kable(
        digits    = c(0, 3, 3, 4, 4, 1, 0, 1),
        col.names = c("Targeting", "alpha* (empirical)", "alpha* (theory)",
                      "Min cost", "No-vax cost", "Cost reduction (%)",
                      "Cases averted", "% averted"),
        caption   = sprintf(
            "Summary of optimal allocation | R = %d, f = %.2f, nu = %.1f, r = %.1f, window %d–%d",
            R_BASE, F_VAX, NU_VAL, R_REACT, min(WIN_YEARS), max(WIN_YEARS)))
```

## Key findings

```{r key-findings, echo=FALSE, results="asis"}
# Extract numbers for inline reporting
opt_r0   <- opt_df |> filter(R == R_BASE, rho == 0.0)
opt_r03  <- opt_df |> filter(R == R_BASE, rho == 0.3)
opt_r1   <- opt_df |> filter(R == R_BASE, rho == 1.0)
ca_r0    <- ca_df  |> filter(R == R_BASE, rho == 0.0)
ca_r1    <- ca_df  |> filter(R == R_BASE, rho == 1.0)
theory_R <- theory_tbl |> filter(R == R_BASE)

cat(sprintf(
"**Parameter setting** (R = %d, f = %.2f, nu = %.1f, r = %.1f, window %d–%d):

- **Outbreak probability.** An empirical %.1f%% of ADM2 sites experienced at
  least one cholera outbreak in %d–%d, giving a 3-year outbreak probability
  p = %.3f. The implied annual probability is p_ann = %.3f.

- **Homogeneous theory.** Under the equal-risk approximation, the capacity
  regime is **%s** (f = %.2f %s p = %.3f). The threshold condition
  r %s p·nu (%.2f %s %.3f) implies the homogeneous model recommends
  **alpha* = %s**. This edge case highlights that the homogeneous model is
  insufficient when sites differ substantially in risk.

- **Effect of targeting accuracy.** With random targeting (rho = 0),
  the optimal alpha* = %.2f, achieving a %.1f%% cost reduction relative to no
  vaccination. With realistic targeting (rho = 0.3), alpha* = %.2f (%.1f%%
  cost reduction). With oracle targeting (rho = 1), alpha* = %.2f (%.1f%%
  cost reduction), averting %.1f%% of cases.

- **Interpretation.** As targeting accuracy improves, more vaccines should
  be deployed pre-emptively (alpha* increases), because better risk intelligence
  reduces the probability of wasting pre-emptive doses on low-risk sites.
  Conversely, with random targeting (rho = 0), holding vaccines for reactive
  deployment is more efficient because reactive deployment is self-targeting —
  it responds to observed outbreak events.

- **Cases averted.** At optimal allocation with oracle targeting, %.0f cases
  are averted (%.1f%% of the no-vaccination baseline of %.0f cases). With
  random targeting, %.0f cases are averted (%.1f%%).
",
R_BASE, F_VAX, NU_VAL, R_REACT, min(WIN_YEARS), max(WIN_YEARS),
100 * p_3yr, min(WIN_YEARS), max(WIN_YEARS), p_3yr, p_ann,
ifelse(F_VAX > p_3yr, "abundant", "scarce"),
F_VAX, ifelse(F_VAX > p_3yr, ">", "<="), p_3yr,
ifelse(R_REACT > p_3yr * NU_VAL, ">", "<"),
R_REACT, ifelse(R_REACT > p_3yr * NU_VAL, ">", "<"), p_3yr * NU_VAL,
ifelse(R_REACT > p_3yr * NU_VAL, "0 (pure reactive)", "1 (pure pre-emptive)"),
opt_r0$alpha_star,  opt_r0$cost_reduction_pct,
opt_r03$alpha_star, opt_r03$cost_reduction_pct,
opt_r1$alpha_star,  opt_r1$cost_reduction_pct,
ca_r1$pct_averted_opt,
ca_r1$cases_averted_opt, ca_r1$pct_averted_opt, ca_r1$cases_no_vax,
ca_r0$cases_averted_opt, ca_r0$pct_averted_opt
))
```
