---
title: "Optimal OCV Allocation: Vaccine Impact Analysis (2010–2012)"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document analyses the optimal allocation of a fixed oral cholera vaccine
(OCV) supply between **pre-emptive** (before outbreaks) and **reactive**
(during outbreaks) deployment, using ADM2-level cholera surveillance data from
the **2010–2012** period as the evaluation window.

**Core question**: Given that vaccines can cover only a fraction $f$ of the
total population, what fraction $\alpha$ should be deployed pre-emptively —
targeting high-risk areas in advance — versus held in reserve for rapid reactive
response when outbreaks are detected?

**Key model elements**:

| Parameter | Symbol | Value |
|-----------|--------|-------|
| Vaccination capacity (fraction of population) | $f$ | 0.60 |
| Pre-emptive doses per person | — | 2 (full 3-yr immunity) |
| Reactive doses per person | — | 1 (ongoing outbreak only) |
| Pre-emptive vaccine effectiveness | $\nu$ | 1.00 |
| Reactive vaccine effectiveness (first year) | $r$ | 0.40 |
| Cost ratio (illness cost / vaccine cost per dose) | $R$ | 100, 1000, 5000 |
| Targeting accuracy (Spearman $\rho$) | $\rho$ | 0, 0.3, 1 |
| Pre-emptive fraction | $\alpha$ | 0 → 1 |
| Unused vaccine penalty | — | 0.5 × $C_V$ per unused dose |

The targeting accuracy $\rho$ measures how well the pre-vaccination prioritisation
ranks sites by true future burden: $\rho = 0$ means random allocation,
$\rho = 1$ means perfect oracle knowledge, $\rho = 0.3$ reflects realistic
surveillance-based targeting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(patchwork)
library(knitr)
source("R/analytic_utils.R")
```

---

# Parameters

```{r params}
WIN_YEARS <- 2010:2012 # evaluation window
T_WIN <- length(WIN_YEARS) # 3 years

F_VAX <- 0.60 # vaccination capacity (fraction of total population)
NU_VAL <- 1.00 # pre-emptive VE (2 doses, 3-yr immunity)
R_REACT <- 0.40 # reactive VE: fraction of first-year cases averted
Q_THRESH <- 0 # reactive threshold: vaccinate any site with cases (q* = 0)

# Pre-emptive uses 2 doses per person; reactive uses 1 dose.
# Unused reactive reserves cost 0.5 × C_V per person-dose at end of window.
DOSE_PRE <- 2L # doses per pre-emptive person
DOSE_REACT <- 1L # doses per reactive person
DOSE_WASTE <- 0.5 # cost fraction for unexpired unused reactive doses

R_VALS <- c(100, 1000, 5000) # cost ratios to analyse
R_BASE <- 1000 # baseline R for primary figures
RHO_VALS <- c(0, 0.3, 1.0) # targeting accuracy levels
ALPHA_GRID <- seq(0, 1, length.out = 201)

MIN_REACT_GAP <- 180L # minimum days between reactive events at the same location

cat(sprintf("Window: %d–%d (%d years)\n", min(WIN_YEARS), max(WIN_YEARS), T_WIN))
cat(sprintf("Vaccination capacity f = %.2f\n", F_VAX))
cat(sprintf("Pre-emptive VE nu = %.2f,  Reactive VE r = %.2f\n", NU_VAL, R_REACT))
cat(sprintf("R (baseline) = %d\n", R_BASE))
```

---

# Data Preparation

## Case surveillance (2010–2012)

```{r load-data}
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)

ts2 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm == 2L)

# Diagnostic: how many outbreaks that START in WIN_YEARS span multiple calendar years?
# We group by (location, location_period_id, outbreak_number) to identify distinct
# outbreaks, then check whether the last record (TR) falls in a later year.
# Filtering on year(TL) ensures we only consider outbreaks that began in the window
# (not pre-2010 outbreaks whose tails overlap into 2010).
multi_yr <- ts2 |>
    filter(year(TL) %in% WIN_YEARS) |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(yr_start = min(year(TL)), yr_end = max(year(TR)), .groups = "drop") |>
    mutate(multi_year = yr_end > yr_start)
cat(sprintf(
    "Outbreaks starting in WIN_YEARS: %d  |  spanning >1 year: %d (%.1f%%)\n",
    nrow(multi_yr), sum(multi_yr$multi_year), 100 * mean(multi_yr$multi_year)
))

```

## Join with incidence rates and population

Population and incidence are derived entirely from the 2010–2012 outbreak data.
Reactive events are outbreak occurrences (grouped by `location_period_id` ×
`outbreak_number`) with a `MIN_REACT_GAP`-day minimum gap enforced within each
location.

```{r join-pop}
# Population per ADM2: median across ALL records in ts2 (not filtered to WIN_YEARS)
# so that sites with no outbreaks in 2010-2012 are still included in the denominator.
pop_df <- ts2 |>
    group_by(location) |>
    summarise(pop_sum = median(pop, na.rm = TRUE), .groups = "drop") |>
    filter(!is.na(pop_sum), pop_sum > 0)

# 3-year burden per location (direct from ts2)
cases_3yr <- ts2 |>
    filter(year(TL) %in% WIN_YEARS) |>
    group_by(location) |>
    summarise(
        cases_win    = sum(sCh, na.rm = TRUE),
        any_outbreak = 1L,
        .groups      = "drop"
    ) |>
    filter(cases_win > 0)

# sim_df: one row per ADM2 with population and 3-yr burden
# n_events (reactive event count after gap filter) is appended below
sim_df <- pop_df |>
    left_join(cases_3yr, by = "location") |>
    replace_na(list(cases_win = 0, any_outbreak = 0L)) |>
    filter(pop_sum > 0) |>
    mutate(q_i = cases_win / pop_sum)

# ── Reactive event list ───────────────────────────────────────────────────────
# Each (location, location_period_id, outbreak_number) with TL in WIN_YEARS
# that has ≥1 case is a candidate reactive event.
#
# One reactive vaccination provides MIN_REACT_GAP days of protection: outbreaks
# starting within MIN_REACT_GAP days of the vaccination date are already covered
# (their cases are absorbed into the triggering event at no extra vaccine cost).
# A new vaccination is only triggered by an outbreak starting ≥ MIN_REACT_GAP
# days after the previous vaccination.
#
# date_event : start of the triggering outbreak (min TL in the occurrence)
# cases      : sCh from the trigger + all absorbed outbreaks within the gap window
# pop_sum    : site population
# site_idx   : row index into sim_df

merge_gap_events <- function(dates, cases, gap_days) {
    keep         <- logical(length(dates))
    merged_cases <- numeric(length(dates))
    last_vacc    <- as.Date(NA)
    last_i       <- NA_integer_
    for (i in seq_along(dates)) {
        if (is.na(last_vacc) || as.integer(dates[i] - last_vacc) >= gap_days) {
            keep[i]         <- TRUE
            merged_cases[i] <- cases[i]
            last_vacc       <- dates[i]
            last_i          <- i
        } else {
            merged_cases[last_i] <- merged_cases[last_i] + cases[i]
        }
    }
    list(keep = keep, merged_cases = merged_cases)
}

event_df <- ts2 |>
    filter(year(TL) %in% WIN_YEARS) |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(
        date_event = min(TL),
        cases      = sum(sCh, na.rm = TRUE),
        .groups    = "drop"
    ) |>
    filter(cases > 0) |>
    left_join(sim_df |> dplyr::select(location, pop_sum), by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(site_idx = match(location, sim_df$location)) |>
    filter(!is.na(site_idx)) |>
    arrange(date_event) |>
    group_by(location) |>
    group_modify(function(df, .key) {
        res      <- merge_gap_events(df$date_event, df$cases, MIN_REACT_GAP)
        df$cases <- res$merged_cases
        df[res$keep, ]
    }) |>
    ungroup() |>
    arrange(date_event)

# Append per-site event count to sim_df (used as p_eff in theory)
sim_df <- sim_df |>
    left_join(
        event_df |> count(location, name = "n_events"),
        by = "location"
    ) |>
    replace_na(list(n_events = 0L))

P_TOTAL  <- sum(sim_df$pop_sum)
N_SITES  <- nrow(sim_df)
N_EVENTS <- nrow(event_df)

# ── Shapefile denominator ──────────────────────────────────────────────────────
# Load outbreak shapefile to get the full geographic universe.
# N_SHP is used as the denominator for outbreak probability (p_3yr and p_eff),
# replacing the ts2-only denominator that gives p_3yr ≈ 1.
shp_tmp <- readRDS("data/outbreak_files_Oct2024/outbreak_shapefiles.rds")
sf::st_geometry(shp_tmp) <- NULL
N_SHP <- nrow(shp_tmp)   # total geographic units in study region
rm(shp_tmp)

cat(sprintf("Sites in ts2 (surveillance): %d  |  Sites in shapefile (universe): %d\n",
            N_SITES, N_SHP))
cat(sprintf("Total population: %s\n", format(round(P_TOTAL), big.mark = ",")))
cat(sprintf("Sites with ≥1 outbreak in %d–%d: %d (%.1f%% of ts2, %.1f%% of shapefile)\n",
            min(WIN_YEARS), max(WIN_YEARS),
            sum(sim_df$any_outbreak),
            100 * mean(sim_df$any_outbreak),
            100 * sum(sim_df$any_outbreak) / N_SHP))
cat(sprintf("Reactive events (after %d-day gap): %d  |  Total cases: %s\n",
            MIN_REACT_GAP, N_EVENTS, format(sum(event_df$cases), big.mark = ",")))
```

---

# Outbreak Probability

The **3-year outbreak probability** `p_3yr` is the fraction of **all ADM2 sites
in the shapefile** (the complete geographic universe) that experienced at least
one outbreak in the `WIN_YEARS` window. Numerator: ADM2 locations in `ts2` with
≥1 reported case in `WIN_YEARS`. Denominator: all `r N_SHP` entries in the
outbreak shapefile.

```{r outbreak-prob}
# p_3yr: (sites with ≥1 outbreak in WIN_YEARS) / (all shapefile locations)
p_3yr <- sum(sim_df$any_outbreak) / N_SHP

cat(sprintf(
    "3-yr outbreak probability P(≥1 outbreak in %d–%d): %.4f  [%d / %d]\n",
    min(WIN_YEARS), max(WIN_YEARS), p_3yr,
    sum(sim_df$any_outbreak), N_SHP
))
cat(sprintf(
    "Mean q_i (cases per person, 3-yr):     %.5f  (%.1f per 100k)\n",
    mean(sim_df$q_i), mean(sim_df$q_i) * 1e5
))

# Distribution table
sim_df |>
    count(n_events, name = "n_sites") |>
    mutate(pct = round(100 * n_sites / N_SITES, 1)) |>
    knitr::kable(
        col.names = c(sprintf("Reactive events (≥%d-day gap)", MIN_REACT_GAP),
                      "N sites", "% sites"),
        caption = sprintf(
            "Distribution of reactive event frequency per ADM2 site, %d–%d",
            min(WIN_YEARS), max(WIN_YEARS)
        )
    )
```

```{r fig-q-dist, fig.width=9, fig.height=4}
#| fig-cap: "Left: distribution of the per-person illness burden q_i (cases per
#|   person over the 3-year window) on a log scale, across sites with at least
#|   one case. Right: empirical CDF. Both panels exclude zero-burden sites."

df_pos <- sim_df |> filter(q_i > 0)

p_hist <- ggplot(df_pos, aes(x = q_i * 1e5)) +
    geom_histogram(aes(y = after_stat(density)),
        bins = 30, fill = "steelblue", colour = "white",
        linewidth = 0.2
    ) +
    scale_x_log10(labels = scales::comma) +
    labs(
        x = "Cases per 100,000 persons (3-yr, log scale)", y = "Density",
        title = sprintf("Illness burden q_i (n = %d positive sites)", nrow(df_pos))
    ) +
    theme_minimal(base_size = 11)

p_cdf <- df_pos |>
    arrange(q_i) |>
    mutate(ecdf = seq_len(n()) / n()) |>
    ggplot(aes(x = q_i * 1e5, y = ecdf)) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    scale_x_log10(labels = scales::comma) +
    labs(
        x = "Cases per 100,000 persons (3-yr, log scale)",
        y = "Cumulative proportion of sites",
        title = "ECDF (positive sites only)"
    ) +
    theme_minimal(base_size = 11)

p_hist + p_cdf
```

---

# Theoretical Optimal Allocation (Homogeneous Case)

In the **homogeneous** model every location is assumed to have the same outbreak
probability $p$. Under this simplification the optimal pre-emptive fraction has a closed-form solution.

**Revised cost model.** Pre-emptive vaccination costs **2 doses** per person and
maintains full immunity for the 3-year window. Reactive vaccination costs **1 dose**
per person, is triggered by outbreak detection, and provides protection for
`MIN_REACT_GAP` days. All outbreaks occurring within that protection window are
absorbed into a single **reactive event** (cases merged); fraction $r$ of the
combined event cases is averted. A site may accumulate multiple non-overlapping
reactive events within the 3-year window. Unused reactive doses expire at window
end, incurring $0.5 \times C_V$ per dose.

Under homogeneous equal-risk, let $p_\text{eff} = \mathbb{E}[\text{reactive events per site}]$,
$\bar{q}$ be the population-weighted mean 3-year burden per person, and $\alpha_c = (f - p_\text{eff}) / [f(1 - p_\text{eff})]$ be the kink where reactive capacity transitions from abundant to scarce.

The expected cost per capita is **piecewise linear** in $\alpha$ with a sign change at $\alpha_c$, yielding **three regimes**:

**Abundant regime** ($f > p_\text{eff}$, $\alpha \leq \alpha_c$): slope $= f\,(1.5 - 0.5p_\text{eff} - R\bar{q}(1-r))$, which is negative for $R > R_\text{lower}$.

**Scarce regime** ($\alpha > \alpha_c$): slope $= f\,(1 - R\bar{q}(p_\text{eff}-r)/p_\text{eff})$, which is negative for $R > R_\text{upper}$.

$$\alpha^* = \begin{cases}0 & R < R_\text{lower} = \dfrac{1.5 - 0.5\,p_\text{eff}}{\bar{q}(1-r)} \\[6pt] \alpha_c & R_\text{lower} \leq R \leq R_\text{upper} \\[6pt] 1 & R > R_\text{upper} = \dfrac{p_\text{eff}}{\bar{q}(p_\text{eff}-r)}\end{cases}$$

Note $R_\text{lower} < R_\text{upper}$ when $p_\text{eff} > r$, so an interior mixed solution always exists between these two thresholds.

```{r alpha-star-theory}
# p_eff: expected reactive events per site over the full geographic universe.
# Denominator = N_SHP (all shapefile locations), not just ts2 sites.
# Sites not in ts2 had 0 observed events.
p_eff <- sum(sim_df$n_events) / N_SHP
q_bar <- weighted.mean(sim_df$q_i, sim_df$pop_sum)

cat(sprintf("p_eff (mean reactive events per site): %.3f\n", p_eff))
cat(sprintf(
    "q_bar (pop-weighted mean burden):     %.6f  (%.1f per 100k)\n",
    q_bar, q_bar * 1e5
))

# alpha_star_revised: optimal alpha under the revised cost model.
# Three-regime formula derived from the event-level cost function:
#   c(alpha) is piecewise linear with a kink at alpha_c = (f-p)/(f*(1-p)).
#
# Regime boundaries:
#   alpha_c  = (f - p) / (f * (1 - p))          [kink; valid when f > p]
#   R_lower  = (1.5 - 0.5*p) / (q*(1-r))        [slope in abundant region = 0]
#   R_upper  = p / (q*(p - r))   [p > r]         [slope in scarce region = 0]
#
# Optimal strategy:
#   R < R_lower              => alpha* = 0        (pure reactive)
#   R_lower <= R <= R_upper  => alpha* = alpha_c  (interior / mixed)
#   R > R_upper              => alpha* = 1        (pure pre-emptive)
alpha_star_revised <- function(p, R, r, f, q_mean, nu = 1) {
    R_lower <- (1.5 - 0.5 * p) / (q_mean * (1 - r))
    R_upper <- if (p > r) p / (q_mean * (p - r)) else Inf
    alpha_c <- (f - p) / (f * (1 - p))

    if (f > p) {
        # Abundant capacity: kink creates three regimes
        if (R < R_lower) {
            return(0)
        }
        if (R > R_upper) {
            return(1)
        }
        return(alpha_c)
    } else {
        # Scarce capacity: always in scarce-reactive region; only corners
        if (p > r && R > R_upper) {
            return(1)
        } else {
            return(0)
        }
    }
}

theory_tbl <- expand_grid(R = R_VALS) |>
    mutate(
        p = p_eff,
        f = F_VAX,
        r = R_REACT,
        nu = NU_VAL,
        q_mean = q_bar,
        alpha_c = (F_VAX - p_eff) / (F_VAX * (1 - p_eff)),
        R_lower = (1.5 - 0.5 * p_eff) / (q_bar * (1 - R_REACT)),
        R_upper = if_else(p_eff > R_REACT,
            p_eff / (q_bar * (p_eff - R_REACT)),
            Inf
        ),
        regime = ifelse(f > p, "Abundant (f > p_eff)", "Scarce (f \u2264 p_eff)"),
        alpha_star_theory = mapply(alpha_star_revised,
            p = p, R = R, r = r, f = f,
            q_mean = q_mean, nu = nu
        )
    )

theory_tbl |>
    dplyr::select(R, p, f, regime, R_lower, R_upper, alpha_c, alpha_star_theory) |>
    knitr::kable(
        digits = c(0, 3, 2, 0, 0, 0, 3, 3),
        col.names = c(
            "R", "p_eff", "f", "Capacity regime",
            "R_lower", "R_upper", "alpha_c", "alpha* (theory)"
        ),
        caption = sprintf(
            "Theoretical optimal alpha | 2-dose pre-emptive, 1-dose reactive | nu=%.1f, r=%.1f",
            NU_VAL, R_REACT
        )
    )

cat(sprintf(
    "\nCapacity regime: %s (f = %.2f %s p_eff = %.3f)\n",
    ifelse(F_VAX > p_eff, "Abundant", "Scarce"),
    F_VAX, ifelse(F_VAX > p_eff, ">", "<="), p_eff
))
cat(sprintf(
    "alpha_c (interior solution): %.3f\n",
    (F_VAX - p_eff) / (F_VAX * (1 - p_eff))
))
cat(sprintf(
    "R_lower = %.0f  |  R_upper = %.0f\n",
    theory_tbl$R_lower[1], theory_tbl$R_upper[1]
))
```

---

# Helper Functions

```{r helper-fns}
# Generate a targeting priority score with a target Spearman correlation to
# the true burden. rho = 0 → random; rho = 1 → oracle (perfect ranking).
make_rank_score <- function(truth, target_rho, seed = 42) {
    set.seed(seed)
    n <- length(truth)
    z_truth <- qnorm(rank(truth, ties.method = "random") / (n + 1))
    z_noise <- rnorm(n)
    rho_p <- 2 * sin(pi * target_rho / 6)
    pnorm(rho_p * z_truth + sqrt(1 - rho_p^2) * z_noise)
}

# Total cost c(alpha), per-capita (normalised by total population N, in C_V units).
#
# alpha    : pre-emptive fraction in [0, 1]
# f        : vaccination capacity (fraction of total population)
# q        : per-person illness burden vector (cases_win / pop, length = n sites)
# pop      : population per site vector (length = n sites)
# score    : pre-emptive priority score (higher → vaccinate first)
# evt      : data.frame of reactive events (one row per location×year outbreak),
#             columns: site_idx, cases, pop_sum, date_event; sorted by date_event
# R        : cost ratio C_I / C_V
# r        : reactive VE (fraction of event cases averted)
# nu       : pre-emptive VE (default 1)
# q_star   : reactive trigger threshold (default 0)
#
# Cost model (continental reactive strategy):
#   Pre-emptive  : DOSE_PRE (=2) doses/person; averts nu × all cases at site
#   Reactive     : events processed in chronological order across all countries;
#                  each served event costs DOSE_REACT (=1) dose × pop_i and
#                  averts r × cases_event; a site may be served in multiple events
#   Unused doses : reactive reserve not consumed costs DOSE_WASTE × C_V per person
cost_incidence <- function(alpha, f, q, pop, score, evt, R, r, nu = 1, q_star = 0) {
    P <- sum(pop)
    f_pre_cap <- alpha * f * P

    # Pre-emptive allocation (site-level, highest-score first)
    ord_pre <- order(score, decreasing = TRUE)
    n_pre <- sum(cumsum(pop[ord_pre]) <= f_pre_cap)
    pre_idx <- ord_pre[seq_len(max(n_pre, 0L))]

    c_pre <- DOSE_PRE * sum(pop[pre_idx]) / P +
        R * (1 - nu) * sum(q[pre_idx] * pop[pre_idx]) / P

    # Reactive events: exclude pre-emptive sites, keep only sites with cases
    evts <- evt[!(evt$site_idx %in% pre_idx) & q[evt$site_idx] > q_star, ]

    f_react_pop <- (1 - alpha) * f * P

    if (nrow(evts) > 0) {
        n_serve <- sum(cumsum(evts$pop_sum) <= f_react_pop)
        if (n_serve > 0L) {
            served <- evts[seq_len(n_serve), ]
            unserved <- if (n_serve < nrow(evts)) evts[(n_serve + 1L):nrow(evts), ] else evts[0L, ]
        } else {
            served <- evts[0L, ]
            unserved <- evts
        }

        pop_used <- sum(served$pop_sum)
        c_vax <- DOSE_REACT * pop_used / P
        c_ill <- R * (sum((1 - r) * served$cases) + sum(unserved$cases)) / P
        c_unused <- DOSE_WASTE * max(0, f_react_pop - pop_used) / P
        c_react <- c_vax + c_ill + c_unused
    } else {
        c_react <- DOSE_WASTE * f_react_pop / P
    }

    c_pre + c_react
}

# Cases averted breakdown for a given allocation.
# Uses the same event-level reactive model as cost_incidence().
cases_averted_breakdown <- function(alpha, f, q, pop, score, evt, R, r,
                                    nu = 1, q_star = 0) {
    P <- sum(pop)
    f_pre_cap <- alpha * f * P

    ord_pre <- order(score, decreasing = TRUE)
    n_pre <- sum(cumsum(pop[ord_pre]) <= f_pre_cap)
    pre_idx <- ord_pre[seq_len(max(n_pre, 0L))]

    evts <- evt[!(evt$site_idx %in% pre_idx) & q[evt$site_idx] > q_star, ]

    f_react_pop <- (1 - alpha) * f * P
    if (nrow(evts) > 0) {
        n_serve <- sum(cumsum(evts$pop_sum) <= f_react_pop)
        served <- if (n_serve > 0L) evts[seq_len(n_serve), ] else evts[0L, ]
    } else {
        served <- evts[0L, ]
    }

    total_cases <- sum(q * pop)
    cases_pre <- nu * sum(q[pre_idx] * pop[pre_idx])
    cases_rx <- r * sum(served$cases)
    cases_averted <- cases_pre + cases_rx
    vax_pop <- sum(pop[pre_idx]) + sum(served$pop_sum)

    list(
        total_cases_no_vax = total_cases,
        cases_averted      = cases_averted,
        cases_averted_pre  = cases_pre,
        cases_averted_rx   = cases_rx,
        pct_averted        = 100 * cases_averted / max(total_cases, 1),
        pop_vaccinated     = vax_pop,
        pct_pop_vaccinated = 100 * vax_pop / P,
        n_events_served    = nrow(served)
    )
}
```

---

# Cost Analysis

## Cost curves by $\alpha$ and $\rho$

For each combination of targeting accuracy $\rho$ and cost ratio $R$, we sweep
$\alpha$ from 0 (pure reactive) to 1 (pure pre-emptive) and evaluate the
normalised expected cost $c(\alpha)$.

The **no-vaccination baseline** cost (no vaccines deployed) is simply
$c_0 = R \cdot \bar{q}$, where $\bar{q}$ is the population-weighted mean
illness burden per person (observed cases / population over the window).

**Cost model (revised):**

| Component | Pre-emptive | Reactive |
|-----------|-------------|---------|
| Doses per person | 2 (3-yr immunity) | 1 (one outbreak) |
| Vaccine cost | $2 \, C_V \times \text{pop}_i$ | $1 \, C_V \times \text{pop}_i$ |
| Cases averted | $\nu \cdot \text{cases\_win}_i$ | $r \cdot \text{cases\_first}_i$ only |
| Subsequent outbreaks | Prevented ($\nu = 1$) | Not prevented |
| If unused | N/A (always deployed) | $0.5 \, C_V$ per unused person-dose |

Costs are reported per capita (total cost divided by $N \times C_V$), so the
unit is "vaccine-dose equivalents per person."

```{r cost-curves}
c_no_vax_by_R <- setNames(
    sapply(R_VALS, function(Rv) Rv * weighted.mean(sim_df$q_i, sim_df$pop_sum)),
    R_VALS
)
cat("No-vaccination cost (per capita, in C_V units):\n")
for (Rv in R_VALS) {
    cat(sprintf("  R = %4d:  c0 = %.4f\n", Rv, c_no_vax_by_R[as.character(Rv)]))
}

# Sweep alpha for all (rho, R) combinations.
# Explicit rho_val/R_val copies prevent closure capture from the for-loop above.
cost_df <- expand_grid(rho = RHO_VALS, R = R_VALS) |>
    pmap_dfr(function(rho, R) {
        rho_val <- rho
        R_val <- R
        score <- make_rank_score(sim_df$q_i, target_rho = rho_val)
        cost_vec <- vapply(ALPHA_GRID, function(a) {
            cost_incidence(
                a, F_VAX, sim_df$q_i, sim_df$pop_sum, score,
                event_df, R_val, R_REACT, NU_VAL, Q_THRESH
            )
        },
        FUN.VALUE = numeric(1)
        )
        tibble(rho = rho_val, R = R_val, alpha = ALPHA_GRID, cost = cost_vec)
    })

opt_df <- cost_df |>
    group_by(rho, R) |>
    slice_min(cost, n = 1, with_ties = FALSE) |>
    ungroup() |>
    rename(alpha_star = alpha, cost_min = cost)

# Attach no-vax baseline
opt_df <- opt_df |>
    mutate(
        c_no_vax = c_no_vax_by_R[as.character(R)],
        cost_reduction_pct = 100 * (c_no_vax - cost_min) / c_no_vax
    )

opt_df |>
    dplyr::select(R, rho, alpha_star, cost_min, c_no_vax, cost_reduction_pct) |>
    arrange(R, rho) |>
    knitr::kable(
        digits = c(0, 2, 3, 4, 4, 1),
        col.names = c("R", "rho", "alpha*", "Min cost", "No-vax cost", "Cost reduction (%)"),
        caption = sprintf(
            "Optimal pre-emptive fraction and cost reduction | f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT
        )
    )
```

```{r fig-cost-curves, fig.width=11, fig.height=7}
#| fig-cap: "Expected cost c(alpha) per capita in C_V units, as a function of
#|   the pre-emptive fraction alpha. Blue curve: vaccination cost + illness cost
#|   under revised model (2-dose pre-emptive, 1-dose reactive, unused penalty,
#|   no subsequent prevention). Red dashed: empirical optimal alpha*. Orange
#|   dotted: theoretical alpha* (homogeneous). Grey dashed: no-vaccination cost."

no_vax_df <- tibble(
    R = R_VALS,
    c_no_vax = unname(c_no_vax_by_R)
)

# Theoretical alpha* (homogeneous model) — same for all rho, expand for faceting
theory_vline_df <- expand_grid(rho = RHO_VALS, R = R_VALS) |>
    left_join(theory_tbl |> dplyr::select(R, alpha_star_theory), by = "R")

ggplot(cost_df, aes(x = alpha, y = cost)) +
    geom_hline(
        data = no_vax_df, aes(yintercept = c_no_vax),
        linetype = "dashed", colour = "grey50", linewidth = 0.7
    ) +
    geom_line(colour = "steelblue", linewidth = 0.9) +
    geom_vline(
        data = opt_df, aes(xintercept = alpha_star),
        linetype = "dashed", colour = "firebrick", linewidth = 0.8
    ) +
    geom_vline(
        data = theory_vline_df, aes(xintercept = alpha_star_theory),
        linetype = "dotted", colour = "darkorange", linewidth = 0.9
    ) +
    geom_point(
        data = opt_df, aes(x = alpha_star, y = cost_min),
        colour = "firebrick", size = 2.5
    ) +
    facet_grid(
        rows = vars(sprintf("rho = %.1f", rho)),
        cols = vars(sprintf("R = %d", R))
    ) +
    labs(
        x = expression("Pre-emptive fraction " ~ alpha),
        y = expression("Normalised cost " ~ italic(c)(alpha)),
        title = sprintf(
            "Cost curves  |  f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT
        ),
        subtitle = "Grey dashed: no-vax cost  |  Red dashed: empirical alpha*  |  Orange dotted: theoretical alpha* (homogeneous)"
    ) +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold"))
```

## Optimal $\alpha^*$ vs targeting accuracy

```{r fig-alpha-star, fig.width=7, fig.height=4}
#| fig-cap: "Optimal pre-emptive fraction alpha* as a function of targeting
#|   accuracy rho, for two cost ratios R. Higher targeting accuracy shifts
#|   resources towards pre-emptive deployment by reducing wastage from
#|   vaccinating low-risk sites."

ggplot(opt_df, aes(
    x = rho, y = alpha_star,
    colour = factor(R), group = factor(R)
)) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 3.5) +
    scale_colour_manual(
        values = c("100" = "steelblue", "1000" = "firebrick", "5000" = "darkgreen"),
        name = "Cost ratio R"
    ) +
    scale_x_continuous(breaks = RHO_VALS) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(
        x = expression("Targeting accuracy " ~ rho),
        y = expression("Optimal pre-emptive fraction " ~ alpha^"*"),
        title = sprintf(
            "alpha* vs rho  |  f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT
        )
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

---

# Cases Averted

```{r cases-averted}
ca_df <- expand_grid(rho = RHO_VALS, R = R_VALS) |>
    pmap_dfr(function(rho, R) {
        rho_val <- rho
        R_val <- R
        score <- make_rank_score(sim_df$q_i, target_rho = rho_val)
        a_star <- opt_df$alpha_star[opt_df$rho == rho_val & opt_df$R == R_val]

        # At optimal alpha
        ca_opt <- cases_averted_breakdown(
            a_star, F_VAX, sim_df$q_i, sim_df$pop_sum,
            score, event_df, R_val, R_REACT, NU_VAL, Q_THRESH
        )
        # At pure reactive (alpha=0)
        ca_rx <- cases_averted_breakdown(
            0, F_VAX, sim_df$q_i, sim_df$pop_sum,
            score, event_df, R_val, R_REACT, NU_VAL, Q_THRESH
        )
        # At pure pre-emptive (alpha=1)
        ca_pre <- cases_averted_breakdown(
            1, F_VAX, sim_df$q_i, sim_df$pop_sum,
            score, event_df, R_val, R_REACT, NU_VAL, Q_THRESH
        )

        tibble(
            rho = rho_val,
            R = R_val,
            alpha_star = a_star,
            cases_no_vax = round(ca_opt$total_cases_no_vax),
            cases_averted_opt = round(ca_opt$cases_averted),
            pct_averted_opt = round(ca_opt$pct_averted, 1),
            cases_averted_rx = round(ca_rx$cases_averted),
            pct_averted_rx = round(ca_rx$pct_averted, 1),
            cases_averted_pre = round(ca_pre$cases_averted),
            pct_averted_pre = round(ca_pre$pct_averted, 1)
        )
    })

ca_df |>
    dplyr::select(
        R, rho, alpha_star,
        cases_no_vax,
        cases_averted_opt, pct_averted_opt,
        cases_averted_rx, pct_averted_rx,
        cases_averted_pre, pct_averted_pre
    ) |>
    arrange(R, rho) |>
    knitr::kable(
        format.args = list(big.mark = ","),
        col.names = c(
            "R", "rho", "alpha*",
            "Cases (no vax)",
            "Averted (opt alpha)", "% averted",
            "Averted (alpha=0)", "% averted",
            "Averted (alpha=1)", "% averted"
        ),
        caption = sprintf(
            "Cases averted under optimal, pure-reactive, and pure-pre-emptive allocation | f = %.2f, nu = %.1f, r = %.1f",
            F_VAX, NU_VAL, R_REACT
        )
    )
```

```{r fig-cases-averted, fig.width=9, fig.height=5}
#| fig-cap: "Cases averted (% of no-vaccination baseline) as a function of the
#|   pre-emptive fraction alpha, for each targeting accuracy rho and R = 100
#|   (baseline). The optimal alpha* (red dashed) maximises cost-efficiency, not
#|   necessarily cases averted."

score_list <- setNames(
    lapply(RHO_VALS, function(rho) {
        make_rank_score(sim_df$q_i, target_rho = rho)
    }),
    RHO_VALS
)

averted_df <- expand_grid(rho = RHO_VALS, alpha = ALPHA_GRID) |>
    mutate(
        pct_averted = mapply(function(rho, a) {
            ca <- cases_averted_breakdown(
                a, F_VAX, sim_df$q_i, sim_df$pop_sum,
                score_list[[as.character(rho)]], event_df,
                R_BASE, R_REACT, NU_VAL, Q_THRESH
            )
            ca$pct_averted
        }, rho, alpha)
    )

opt_averted <- opt_df |>
    filter(R == R_BASE) |>
    left_join(
        averted_df |>
            group_by(rho) |>
            slice(which.min(abs(alpha - opt_df$alpha_star[opt_df$R == R_BASE &
                opt_df$rho == first(rho)]))),
        by = "rho"
    )

ggplot(averted_df, aes(
    x = alpha, y = pct_averted,
    colour = factor(rho), group = factor(rho)
)) +
    geom_line(linewidth = 0.9) +
    geom_vline(
        data = opt_df |> filter(R == R_BASE),
        aes(xintercept = alpha_star, colour = factor(rho)),
        linetype = "dashed", linewidth = 0.7
    ) +
    scale_colour_viridis_d(
        option = "plasma", end = 0.85,
        name = expression("Targeting accuracy " ~ rho)
    ) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(
        x = expression("Pre-emptive fraction " ~ alpha),
        y = "Cases averted (%)",
        title = sprintf(
            "Cases averted vs alpha  |  f = %.2f, R = %d, r = %.1f",
            F_VAX, R_BASE, R_REACT
        ),
        subtitle = "Dashed lines: optimal alpha* for each rho"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

---

# Site-level Breakdown at Optimal Allocation

```{r fig-site-map-setup}
# Compute site-level allocation for each rho at optimal alpha, R = R_BASE.
# A site is marked "Reactive" if at least one of its events was served.
site_alloc <- lapply(RHO_VALS, function(rho) {
    score <- make_rank_score(sim_df$q_i, target_rho = rho)
    a_star <- opt_df$alpha_star[opt_df$rho == rho & opt_df$R == R_BASE]
    P <- sum(sim_df$pop_sum)

    f_pre_cap <- a_star * F_VAX * P
    ord_pre <- order(score, decreasing = TRUE)
    n_pre <- sum(cumsum(sim_df$pop_sum[ord_pre]) <= f_pre_cap)
    pre_idx <- ord_pre[seq_len(max(n_pre, 0L))]

    # Event-level reactive deployment
    evts <- event_df[!(event_df$site_idx %in% pre_idx) &
        sim_df$q_i[event_df$site_idx] > Q_THRESH, ]
    f_react_pop <- (1 - a_star) * F_VAX * P

    served_sites <- integer(0)
    if (nrow(evts) > 0) {
        n_serve <- sum(cumsum(evts$pop_sum) <= f_react_pop)
        if (n_serve > 0L) served_sites <- unique(evts$site_idx[seq_len(n_serve)])
    }

    status <- rep("No vaccination", nrow(sim_df))
    status[pre_idx] <- "Pre-emptive"
    status[served_sites] <- "Reactive"

    sim_df |>
        mutate(status = status, rho_label = sprintf("rho = %.1f", rho))
})
site_df <- bind_rows(site_alloc)
```

```{r fig-site-burden, fig.width=11, fig.height=5}
#| fig-cap: "Per-site illness burden q_i (cases per 100,000 persons in 2010–2012)
#|   coloured by vaccination status at the optimal alpha* for R = 100. Both axes
#|   Sites are sorted by q_i within each panel. Pre-emptive
#|   vaccination concentrates on high-burden sites; reactive coverage fills in
#|   additional outbreak sites."

site_df |>
    filter(q_i > 0) |>
    arrange(desc(q_i)) |>
    group_by(rho_label) |>
    mutate(rank = row_number()) |>
    ggplot(aes(x = rank, y = q_i * 1e5, colour = status)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_y_log10(labels = scales::comma) +
    scale_colour_manual(
        values = c(
            "Pre-emptive" = "firebrick", "Reactive" = "steelblue",
            "No vaccination" = "grey70"
        ),
        name = "Vaccination status"
    ) +
    facet_wrap(~rho_label, nrow = 1) +
    labs(
        x = "Site rank (by burden)",
        y = "Cases per 100,000 (log scale)",
        title = sprintf("Site-level vaccination assignment at optimal alpha* | R = %d", R_BASE)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

---

# Sensitivity Analysis

The purpose of this section is to map empirical α\* across the full parameter
space and to overlay the **theoretical guidelines** derived from the homogeneous
equal-risk model under the **revised cost assumptions** (2-dose pre-emptive,
1-dose reactive, unused vaccine penalty, no subsequent prevention by reactive).

**Theoretical boundaries shown on each plot:**

| Line | Condition | Meaning |
|------|-----------|---------|
| Blue dashed | $f = p_\text{eff}$ | Separates *scarce* ($f \leq p_\text{eff}$) from *abundant* ($f > p_\text{eff}$) capacity regime |
| White contour | $\alpha^* = \alpha_c$ boundary | Approximate boundary of the interior (mixed) solution; $\alpha_c = (f - p_\text{eff}) / [f(1 - p_\text{eff})]$ |
| Red curve (R×r plot) | $R = R_\text{lower} = \frac{1.5 - 0.5\,p_\text{eff}}{\bar{q}(1 - r)}$ | Below this curve pure reactive ($\alpha^*=0$); above, interior or pure pre-emptive |

The revised cost function is **piecewise linear in $\alpha$** with a kink at $\alpha_c$. This creates **three regimes**: pure reactive ($R < R_\text{lower}$), interior mixed ($R_\text{lower} \leq R \leq R_\text{upper}$, $\alpha^* = \alpha_c$), and pure pre-emptive ($R > R_\text{upper}$), where $R_\text{upper} = p_\text{eff} / [\bar{q}(p_\text{eff} - r)]$.

```{r sens-compute, cache=TRUE, cache.extra=list(p_3yr, F_VAX, NU_VAL, sim_df$q_i)}
ALPHA_SENS <- seq(0, 1, length.out = 101)

# Pre-compute scores (only depend on rho, not on f or r)
score_sens <- setNames(
    lapply(RHO_VALS, function(rho) {
        make_rank_score(sim_df$q_i, target_rho = rho)
    }),
    as.character(RHO_VALS)
)

# ── Grid 1: f × r  (faceted by R and rho) ──────────────────────────────────
F_GRID <- seq(0.05, 0.80, by = 0.05)
R_GRID <- seq(0.10, 0.90, by = 0.05)

sens_fr <- expand_grid(f = F_GRID, r = R_GRID, R = R_VALS, rho = RHO_VALS) |>
    mutate(
        alpha_star_emp = pmap_dbl(list(f, r, R, rho), function(f, r, R, rho) {
            fv <- f
            rv <- r
            Rv <- R
            rhov <- rho
            score <- score_sens[[as.character(rhov)]]
            cost_vec <- vapply(
                ALPHA_SENS, function(a) {
                    cost_incidence(
                        a, fv, sim_df$q_i, sim_df$pop_sum, score,
                        event_df, Rv, rv, NU_VAL, Q_THRESH
                    )
                },
                numeric(1)
            )
            ALPHA_SENS[which.min(cost_vec)]
        }),
        alpha_star_theory = pmap_dbl(list(f, r, R), function(f, r, R) {
            alpha_star_revised(
                p = p_eff, R = R, r = r, f = f,
                q_mean = q_bar, nu = NU_VAL
            )
        })
    ) |>
    mutate(alpha_star_theory = replace_na(alpha_star_theory, 0.5))

# ── Grid 2: R × r  (faceted by rho, fixed f = F_VAX) ──────────────────────
# Extend R range to cover R_crit under the revised model
R_FINE <- seq(50, 10000, length.out = 40)
R_GRID2 <- seq(0.10, 0.90, by = 0.05)

sens_Rr <- expand_grid(R = R_FINE, r = R_GRID2, rho = RHO_VALS) |>
    mutate(
        alpha_star_emp = pmap_dbl(list(R, r, rho), function(R, r, rho) {
            Rv <- R
            rv <- r
            rhov <- rho
            score <- score_sens[[as.character(rhov)]]
            cost_vec <- vapply(
                ALPHA_SENS, function(a) {
                    cost_incidence(
                        a, F_VAX, sim_df$q_i, sim_df$pop_sum, score,
                        event_df, Rv, rv, NU_VAL, Q_THRESH
                    )
                },
                numeric(1)
            )
            ALPHA_SENS[which.min(cost_vec)]
        }),
        alpha_star_theory = pmap_dbl(list(R, r), function(R, r) {
            alpha_star_revised(
                p = p_eff, R = R, r = r, f = F_VAX,
                q_mean = q_bar, nu = NU_VAL
            )
        })
    ) |>
    mutate(alpha_star_theory = replace_na(alpha_star_theory, 0.5))

# R_lower curve (revised model): R_lower = (1.5 - 0.5*p_eff) / (q_bar*(1-r))
# Below this curve alpha*=0; above, alpha*=alpha_c or 1.
R_crit_df <- tibble(r = seq(0.05, 0.95, by = 0.005)) |>
    mutate(
        R_crit = (1.5 - 0.5 * p_eff) / (q_bar * (1 - r))
    ) |>
    filter(R_crit >= 50, R_crit <= 10000)

cat(sprintf(
    "Grids computed: sens_fr (%d rows), sens_Rr (%d rows)\n",
    nrow(sens_fr), nrow(sens_Rr)
))
```

## Heat map: $f$ vs $r$

```{r fig-sens-fr, fig.width=13, fig.height=10}
#| fig-cap: "Empirical alpha* (fill) across vaccination capacity f and reactive
#|   VE r, faceted by R (columns) and targeting accuracy rho (rows). White
#|   contours show theoretical alpha* = 0.5 boundary (pure reactive vs pure
#|   pre-emptive). Blue dashed: f = p (capacity regime boundary).
#|   Cross: current analysis point."

label_df_fr <- expand_grid(R = R_VALS, rho = RHO_VALS)

ggplot(sens_fr, aes(x = r, y = f)) +
    geom_tile(aes(fill = alpha_star_emp)) +
    geom_contour(aes(z = alpha_star_theory),
        breaks = c(0.5),
        colour = "white", linewidth = 0.7, linetype = "solid"
    ) +
    # f = p_eff boundary (blue dashed horizontal)
    geom_hline(
        yintercept = p_eff,
        linetype = "dashed", colour = "dodgerblue", linewidth = 0.8
    ) +
    # Mark current analysis point
    geom_point(
        data = label_df_fr,
        aes(x = R_REACT, y = F_VAX),
        shape = 4, size = 3, colour = "white", stroke = 1.2
    ) +
    scale_fill_viridis_c(
        name = expression(alpha^"*"),
        limits = c(0, 1), option = "plasma"
    ) +
    scale_x_continuous(breaks = seq(0.1, 0.9, by = 0.2)) +
    scale_y_continuous(breaks = seq(0.1, 0.8, by = 0.2)) +
    facet_grid(
        rows = vars(sprintf("rho = %.1f", rho)),
        cols = vars(sprintf("R = %d", R))
    ) +
    labs(
        x = expression("Reactive VE " ~ italic(r)),
        y = expression("Vaccination capacity " ~ italic(f)),
        title = "Empirical optimal pre-emptive fraction across f and r",
        subtitle = paste0(
            "Blue dashed: f = p_eff = ", round(p_eff, 3),
            "  |  White contour: theoretical \u03b1* = \u03b1_c boundary",
            "  |  \u2715 = current analysis point"
        )
    ) +
    theme_minimal(base_size = 11) +
    theme(
        strip.text = element_text(face = "bold"),
        legend.position = "right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

## Heat map: $R$ vs $r$

```{r fig-sens-Rr, fig.width=11, fig.height=5}
#| fig-cap: "Empirical alpha* (fill) across cost ratio R and reactive VE r at
#|   fixed f = F_VAX, faceted by targeting accuracy rho. Red curve: R = R*_crit
#|   (above this curve pure pre-emptive is theoretically optimal; below, pure
#|   reactive). Cross marks current analysis point (r=0.4, R=1000)."

label_df_Rr <- tibble(rho = RHO_VALS)

ggplot(sens_Rr, aes(x = r, y = R)) +
    geom_tile(aes(fill = alpha_star_emp)) +
    # R_crit curve from revised theory
    geom_line(
        data = R_crit_df, aes(x = r, y = R_crit),
        linetype = "dashed", colour = "firebrick", linewidth = 0.9,
        inherit.aes = FALSE
    ) +
    # Current analysis point
    geom_point(
        data = label_df_Rr,
        aes(x = R_REACT, y = R_BASE),
        shape = 4, size = 3.5, colour = "white", stroke = 1.2,
        inherit.aes = FALSE
    ) +
    scale_fill_viridis_c(
        name = expression(alpha^"*"),
        limits = c(0, 1), option = "plasma"
    ) +
    scale_x_continuous(breaks = seq(0.1, 0.9, by = 0.2)) +
    scale_y_log10(
        labels = scales::comma,
        breaks = c(100, 500, 1000, 2000, 5000, 10000)
    ) +
    facet_wrap(~ sprintf("rho = %.1f", rho), nrow = 1) +
    labs(
        x = expression("Reactive VE " ~ italic(r)),
        y = expression("Cost ratio " ~ italic(R) ~ " (log scale)"),
        title = sprintf("Empirical optimal pre-emptive fraction across R and r  |  f = %.2f", F_VAX),
        subtitle = paste0(
            "Red dashed: R_lower = (1.5 - 0.5\u00d7p_eff) / (q_bar \u00d7 (1 - r))",
            "  |  \u2715 = current analysis point (r=0.4, R=1000)"
        )
    ) +
    theme_minimal(base_size = 11) +
    theme(
        strip.text = element_text(face = "bold"),
        legend.position = "right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

## Sensitivity to minimum reactive gap (`MIN_REACT_GAP`)

As the minimum gap between successive reactive vaccinations at the same site
varies from 0 (no constraint) to 1,095 days (3 years = the full window), the
number of reactive events, `p_eff`, and the theoretical thresholds all change.

```{r sens-gap, cache=TRUE, cache.extra=list(N_SHP, q_bar, F_VAX, R_REACT, R_BASE)}
GAP_VALS <- c(0L, 30L, 60L, 90L, 180L, 365L, 548L, 730L, 1095L)

# Raw per-outbreak aggregates (no gap filter) — reused across all gap values
event_raw <- ts2 |>
    filter(year(TL) %in% WIN_YEARS) |>
    group_by(location, location_period_id, outbreak_number) |>
    summarise(
        date_event = min(TL),
        cases      = sum(sCh, na.rm = TRUE),
        .groups    = "drop"
    ) |>
    filter(cases > 0) |>
    left_join(sim_df |> dplyr::select(location, pop_sum), by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(site_idx = match(location, sim_df$location)) |>
    filter(!is.na(site_idx)) |>
    arrange(date_event)

ALPHA_COARSE <- seq(0, 1, length.out = 101)

gap_sens <- lapply(GAP_VALS, function(gap) {
    # Apply gap merging
    edf <- event_raw |>
        group_by(location) |>
        group_modify(function(df, .key) {
            res      <- merge_gap_events(df$date_event, df$cases, gap)
            df$cases <- res$merged_cases
            df[res$keep, ]
        }) |>
        ungroup() |>
        arrange(date_event)

    n_ev_df <- sim_df |>
        left_join(edf |> count(location, name = "n_events"), by = "location") |>
        replace_na(list(n_events = 0L))

    p_eff_g   <- sum(n_ev_df$n_events) / N_SHP
    alpha_c_g <- if (F_VAX > p_eff_g) (F_VAX - p_eff_g) / (F_VAX * (1 - p_eff_g)) else NA_real_
    R_lower_g <- (1.5 - 0.5 * p_eff_g) / (q_bar * (1 - R_REACT))
    R_upper_g <- if (p_eff_g > R_REACT) p_eff_g / (q_bar * (p_eff_g - R_REACT)) else Inf
    alpha_th  <- alpha_star_revised(
        p = p_eff_g, R = R_BASE, r = R_REACT,
        f = F_VAX, q_mean = q_bar, nu = NU_VAL
    )

    get_astar_emp <- function(rho_v) {
        sc <- make_rank_score(sim_df$q_i, target_rho = rho_v)
        cv <- vapply(ALPHA_COARSE, function(a)
            cost_incidence(a, F_VAX, sim_df$q_i, sim_df$pop_sum,
                           sc, edf, R_BASE, R_REACT, NU_VAL, Q_THRESH),
            numeric(1))
        ALPHA_COARSE[which.min(cv)]
    }

    tibble(
        gap_days          = gap,
        gap_months        = round(gap / 30.44, 1),
        n_events          = nrow(edf),
        p_eff             = p_eff_g,
        alpha_c           = alpha_c_g,
        R_lower           = R_lower_g,
        R_upper           = R_upper_g,
        alpha_star_theory = alpha_th,
        alpha_star_rho0   = get_astar_emp(0),
        alpha_star_rho1   = get_astar_emp(1)
    )
})

gap_tbl <- bind_rows(gap_sens)

gap_tbl |>
    mutate(
        R_upper = ifelse(is.infinite(R_upper), Inf, round(R_upper, 0)),
        current = ifelse(gap_days == MIN_REACT_GAP, "\u2190 current", "")
    ) |>
    knitr::kable(
        digits = c(0, 1, 0, 4, 3, 0, 0, 3, 3, 3, 0),
        col.names = c(
            "Gap (days)", "Gap (months)", "N events", "p_eff",
            "alpha_c", "R_lower", "R_upper",
            "alpha* theory", "alpha* emp (rho=0)", "alpha* emp (rho=1)", ""
        ),
        caption = sprintf(
            "Effect of MIN_REACT_GAP on reactive events and optimal alpha | R = %d, f = %.2f, r = %.1f",
            R_BASE, F_VAX, R_REACT
        )
    )
```

```{r fig-sens-gap, fig.width=11, fig.height=8}
#| fig-cap: "Sensitivity of key outputs to the minimum reactive gap. Top row:
#|   number of reactive events (left) and p_eff (right) as a function of the gap.
#|   Bottom row: theoretical thresholds R_lower and R_upper (left, log scale) and
#|   optimal pre-emptive fraction alpha* (right) for theory and two targeting
#|   accuracies. Vertical dashed line: current MIN_REACT_GAP = 180 days."

gap_long <- gap_tbl |>
    pivot_longer(
        cols = c(alpha_star_theory, alpha_star_rho0, alpha_star_rho1),
        names_to = "type", values_to = "alpha_star"
    ) |>
    mutate(type = recode(type,
        alpha_star_theory = "Theory",
        alpha_star_rho0   = "Empirical (rho=0)",
        alpha_star_rho1   = "Empirical (rho=1)"
    ))

p1 <- ggplot(gap_tbl, aes(x = gap_days, y = n_events)) +
    geom_line(colour = "steelblue", linewidth = 1) +
    geom_point(colour = "steelblue", size = 2.5) +
    geom_vline(xintercept = MIN_REACT_GAP, linetype = "dashed",
               colour = "grey40", linewidth = 0.7) +
    labs(x = "Min reactive gap (days)", y = "Reactive events (N)",
         title = "Reactive events after gap filter") +
    theme_minimal(base_size = 11)

p2 <- ggplot(gap_tbl, aes(x = gap_days, y = p_eff)) +
    geom_line(colour = "steelblue", linewidth = 1) +
    geom_point(colour = "steelblue", size = 2.5) +
    geom_vline(xintercept = MIN_REACT_GAP, linetype = "dashed",
               colour = "grey40", linewidth = 0.7) +
    labs(x = "Min reactive gap (days)", y = "p_eff",
         title = "Expected reactive events per site (shapefile universe)") +
    theme_minimal(base_size = 11)

gap_thresh <- gap_tbl |>
    filter(is.finite(R_upper)) |>
    pivot_longer(c(R_lower, R_upper), names_to = "threshold", values_to = "value")

p3 <- ggplot(gap_thresh, aes(x = gap_days, y = value,
                               colour = threshold, group = threshold)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = MIN_REACT_GAP, linetype = "dashed",
               colour = "grey40", linewidth = 0.7) +
    scale_y_log10(labels = scales::comma) +
    scale_colour_manual(values = c(R_lower = "firebrick", R_upper = "darkgreen"),
                        labels = c(R_lower = "R_lower", R_upper = "R_upper"),
                        name = "Threshold") +
    labs(x = "Min reactive gap (days)", y = "Cost ratio R (log scale)",
         title = "Theory thresholds vs gap") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

p4 <- ggplot(gap_long, aes(x = gap_days, y = alpha_star,
                             colour = type, group = type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = MIN_REACT_GAP, linetype = "dashed",
               colour = "grey40", linewidth = 0.7) +
    scale_colour_manual(
        values = c(Theory = "darkorange",
                   "Empirical (rho=0)" = "steelblue",
                   "Empirical (rho=1)" = "firebrick"),
        name = NULL
    ) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
    labs(x = "Min reactive gap (days)",
         y = expression(alpha^"*"),
         title = sprintf("Optimal pre-emptive fraction | R = %d", R_BASE)) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

(p1 + p2) / (p3 + p4) +
    plot_annotation(
        title = sprintf(
            "Sensitivity to minimum reactive gap | f = %.2f, r = %.1f, R = %d",
            F_VAX, R_REACT, R_BASE
        ),
        subtitle = "Vertical dashed: current MIN_REACT_GAP = 180 days"
    )
```

---

# Summary and Interpretation

```{r summary-table}
summary_tbl <- opt_df |>
    filter(R == R_BASE) |>
    left_join(
        ca_df |> filter(R == R_BASE) |>
            dplyr::select(rho, cases_no_vax, cases_averted_opt, pct_averted_opt),
        by = "rho"
    ) |>
    mutate(
        rho_label = case_when(
            rho == 0 ~ "Random (rho=0)",
            rho == 0.3 ~ "Moderate (rho=0.3)",
            rho == 1.0 ~ "Oracle (rho=1)"
        ),
        alpha_theory = theory_tbl$alpha_star_theory[match(R_BASE, theory_tbl$R)]
    )

summary_tbl |>
    dplyr::select(
        rho_label, alpha_star, alpha_theory,
        cost_min, c_no_vax, cost_reduction_pct,
        cases_averted_opt, pct_averted_opt
    ) |>
    knitr::kable(
        digits = c(0, 3, 3, 4, 4, 1, 0, 1),
        col.names = c(
            "Targeting", "alpha* (empirical)", "alpha* (theory)",
            "Min cost", "No-vax cost", "Cost reduction (%)",
            "Cases averted", "% averted"
        ),
        caption = sprintf(
            "Summary of optimal allocation | R = %d, f = %.2f, nu = %.1f, r = %.1f, window %d–%d",
            R_BASE, F_VAX, NU_VAL, R_REACT, min(WIN_YEARS), max(WIN_YEARS)
        )
    )
```

## Key findings

```{r key-findings, echo=FALSE, results="asis"}
# Extract numbers for inline reporting
opt_r0 <- opt_df |> filter(R == R_BASE, rho == 0.0)
opt_r03 <- opt_df |> filter(R == R_BASE, rho == 0.3)
opt_r1 <- opt_df |> filter(R == R_BASE, rho == 1.0)
ca_r0 <- ca_df |> filter(R == R_BASE, rho == 0.0)
ca_r1 <- ca_df |> filter(R == R_BASE, rho == 1.0)
theory_R <- theory_tbl |> filter(R == R_BASE)

R_lower_base <- theory_tbl$R_lower[1]
R_upper_base <- theory_tbl$R_upper[1]
alpha_c_base <- theory_tbl$alpha_c[1]

cat(sprintf(
    "**Parameter setting** (R_base = %d, f = %.2f, nu = %.1f, r = %.1f,
2-dose pre-emptive / 1-dose reactive / 0.5×C_V unused penalty, window %d–%d):

- **Outbreak probability.** An empirical %.1f%% of ADM2 sites experienced at
  least one cholera outbreak in %d–%d (p_3yr = %.3f).
  Mean reactive events per site p_eff = %.3f.

- **Homogeneous theory.** The capacity regime is **%s** (f = %.2f %s p_eff = %.3f).
  The revised cost function is *piecewise linear* in alpha with a kink at
  alpha_c = %.3f, giving three regimes: pure reactive (R < R_lower = %.0f),
  interior mixed (alpha* = alpha_c), and pure pre-emptive (R > R_upper = %.0f).
  At R = %d (baseline), theory predicts alpha* = %s.

- **Effect of targeting accuracy (R = %d).** With random targeting (rho = 0),
  empirical alpha* = %.2f (%.1f%% cost change vs no vaccination). With realistic
  targeting (rho = 0.3), alpha* = %.2f (%.1f%%). With oracle targeting (rho = 1),
  alpha* = %.2f (%.1f%%), averting %.1f%% of cases.

- **Interpretation.** Pre-emptive vaccination requires 2 doses, so it is twice
  as expensive per person as reactive (1 dose). Combined with the unused-reserve
  penalty, purely reactive strategies are cost-optimal at low-to-moderate R.
  Pre-emptive becomes competitive only at very high illness cost ratios (R > %.0f),
  when the benefits of guaranteed protection outweigh the extra dose cost.

- **Cases averted.** At optimal allocation with oracle targeting, %.0f cases
  are averted (%.1f%% of the %.0f-case no-vaccination baseline). With random
  targeting, %.0f cases are averted (%.1f%%).
",
    R_BASE, F_VAX, NU_VAL, R_REACT, min(WIN_YEARS), max(WIN_YEARS),
    100 * p_3yr, min(WIN_YEARS), max(WIN_YEARS), p_3yr, p_eff,
    ifelse(F_VAX > p_eff, "abundant", "scarce"),
    F_VAX, ifelse(F_VAX > p_eff, ">", "<="), p_eff,
    alpha_c_base, R_lower_base, R_upper_base,
    R_BASE,
    ifelse(R_BASE > R_upper_base, "1 (pure pre-emptive)",
        ifelse(R_BASE >= R_lower_base,
            sprintf("%.2f (interior, alpha_c)", alpha_c_base),
            "0 (pure reactive)"
        )
    ),
    R_BASE,
    opt_r0$alpha_star, opt_r0$cost_reduction_pct,
    opt_r03$alpha_star, opt_r03$cost_reduction_pct,
    opt_r1$alpha_star, opt_r1$cost_reduction_pct,
    ca_r1$pct_averted_opt,
    ca_r1$cases_averted_opt, ca_r1$pct_averted_opt, ca_r1$cases_no_vax,
    ca_r0$cases_averted_opt, ca_r0$pct_averted_opt,
    R_upper_base
))
```
