---
title: "Vaccine Impact Simulation: Pre-emptive + Reactive OCV Allocation"
author: "Jong-Hoon Kim"
date: "`r Sys.Date()`"
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

## Overview

This document simulates the impact of **mixed pre-emptive and reactive cholera
vaccination** under a fixed vaccine supply. The analysis:

1. Trains a logistic regression model on 2010--2015 outbreak data to predict risk
2. Estimates outbreak probability $p$ via Beta-Binomial MLE
3. Computes the **optimal pre-emptive fraction** $\alpha^*$ using the analytical
   framework from `analytic_hetero.qmd`
4. Simulates vaccination impact for 4 allocation strategies across varying $\alpha$

**Scenario parameters**:

- Countries: COD (DRC, high-risk), NGA (Nigeria, medium-risk), CMR (Cameroon, low-risk)
- Single-dose OCV for both pre-emptive and reactive use ($\nu = 1, r \in {0.2, 0.4, 0.6}$)
- Cost ratio $R = C_I / C_V$, where $C_V$ is cost of vaccination per person and $C_I$ is cost of infection per person (accounting for severity of the disease). Total cost of vaccinating a population with size $n$ would be $n C_V$ and total cost of infection would be $n A C_I$, where $A$ is the attack rate (or cumulative incidence) in the population expressed as a proportion.
- Coverage fraction varied: $f \in \{0.3, 0.5, 0.7\}$ (doses adjusted accordingly)
- Evaluation window: 2016--2018 (3 years, matching model prediction horizon $k = 3$)
- $\alpha$: fraction of supply allocated to the pre-emptive vaccination; $1 - \alpha$ held for reactive vaccination

| Strategy | Ranking basis |
|----------|--------------|
| Model-ranked | Predicted outbreak probability (logistic regression) |
| RF-ranked | Predicted outbreak probability (random forest) |
| Count-ranked | Historical outbreak count (2011--2015) |
| Random | Uniform random selection |
| Perfect ranking | Actual 2016--2018 outbreak burden (perfect ranking, $\rho = 1$) |
| Oracle | Perfect foreknowledge of outbreak realization |


# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(raster)
library(exactextractr)
library(patchwork)
library(pROC)
library(scales)
library(knitr)
library(mgcv) # GAM
library(randomForest) # random forest
library(gbm) # gradient boosting
library(class) # k-NN
source("R/analytic_utils.R")
source("R/spatial_setup.R")
```


## Data Loading

### Outbreak surveillance data

```{r load-outbreak-data}
# Raw surveillance data: each row = (location, time interval TL-TR, outbreak_number)
# Location format: "REGION::COUNTRY::ADMIN1::ADMIN2" (3 separators = admin2)
ts_outbreak <- readr::read_rds(
    "data/outbreak_files_Oct2024/time_series_outbreak_extraction.rds"
)
# Keep admin2 (adm=2) and admin3 (adm=3) only
ts23 <- ts_outbreak |>
    mutate(adm = str_count(location, "::") - 1L) |>
    filter(adm %in% c(2, 3))

# Collapse time-series rows into one row per outbreak event, extracting
# start/end years, duration, and a unique ID
outbreak_events <- ts23 |>
    group_by(location, adm, outbreak_number) |>
    summarise(
        year_start     = year(min(TL)),
        year_end       = year(max(TR)),
        duration_weeks = as.numeric(difftime(max(TR), min(TL), units = "weeks")),
        id_outbreak    = paste(first(location), min(TL), max(TR), sep = "_"),
        .groups        = "drop"
    )

all_years <- 2010:2023
all_locs <- unique(ts23$location)

# loc_year_full: complete (location × year) panel with n_active = number of
# outbreaks ACTIVE (ongoing) in that year. Used for the outcome variable.
loc_year <- outbreak_events %>%
    mutate(years = map2(year_start, pmin(year_end, 2023L), seq)) %>%
    unnest(years) %>%
    rename(year = years) %>%
    filter(year %in% all_years) %>%
    count(location, year, name = "n_active")

loc_year_full <- expand_grid(location = all_locs, year = all_years) %>%
    left_join(loc_year, by = c("location", "year")) %>%
    replace_na(list(n_active = 0L))

# loc_year_end_full: complete (location × year) panel with n_ended = number of
# outbreaks that ENDED in that year. Used for feature construction (hist_count)
# because an outbreak's end year marks when it became observable.
loc_year_end_full <- outbreak_events %>%
    filter(year_end %in% all_years) %>%
    count(location, year_end, name = "n_ended") %>%
    rename(year = year_end) %>%
    right_join(expand_grid(location = all_locs, year = all_years),
        by = c("location", "year")
    ) %>%
    replace_na(list(n_ended = 0L))

cat("Surveillance locations:", length(all_locs), "\n")
cat("Year range:", range(all_years), "\n")
cat("Total outbreak events:", nrow(outbreak_events), "\n")
cat("Total outbreak-end-years:", sum(loc_year_end_full$n_ended), "\n")
```

### Shapefiles and spatial neighbours

```{r load-shapefiles}
# Shapefiles are keyed by location_period_id; we need to map them to
# the human-readable location strings (e.g. "AFR::COD::Kinshasa::...").
shp_all <- readr::read_rds(
    "data/outbreak_files_Oct2024/outbreak_shapefiles.rds"
)

# Build a lookup: for each location, keep the most recent location_period_id
# (a location can have multiple period IDs if boundaries changed over time)
loc_id_map <- ts23 %>%
    distinct(location_period_id, location) %>%
    group_by(location) %>%
    slice_max(as.integer(location_period_id), n = 1, with_ties = FALSE) %>%
    ungroup()

# Join shapefile geometries to location names via the ID map
shp_named <- shp_all %>%
    left_join(loc_id_map, by = c("lctn_pr" = "location_period_id"))

# Keep only admin2/3 polygons (3 or 4 "::" separators in location string)
shp_admin23 <- shp_named %>%
    filter(!is.na(location), str_count(location, "::") %in% c(3, 4))

cat("ADM2/3 polygons:", nrow(shp_admin23), "\n")

# Build a spatial neighbour list: for each polygon, find all neighbours
# whose centroid is within DIST_LIMIT_KM km (geodesic distance).
# Used for the nbr_hist feature (spatial contagion signal).
DIST_LIMIT_KM <- 50
centroids_sf <- shp_admin23 %>%
    st_make_valid() %>%
    st_point_on_surface()

# Pairwise geodesic distance matrix (meters); convert to km
dist_mat_sp <- st_distance(centroids_sf) |> units::drop_units()
dist_mat_sp <- dist_mat_sp / 1000 # m -> km
rownames(dist_mat_sp) <- colnames(dist_mat_sp) <- shp_admin23$location
diag(dist_mat_sp) <- Inf

# For each location, store neighbours within the distance limit (sorted)
nbr_list <- setNames(
    lapply(shp_admin23$location, function(loc) {
        d <- dist_mat_sp[loc, ]
        names(sort(d[d <= DIST_LIMIT_KM]))
    }),
    shp_admin23$location
)
```

### Population raster

```{r load-population}
pop_raster <- readr::read_rds("data/ppp_20km_mean_af_2017_20221123.rds")
cat("Raster dimensions:", dim(pop_raster), "\n")
cat("Resolution:", res(pop_raster), "degrees (~20 km)\n")
```


## Population Extraction

Aggregate the 20 km population raster to ADM2 polygons via zonal statistics.

The raster stores **mean population per ~1 km² source cell**, having been
aggregated from the WorldPop ~1 km PPP (People Per Pixel) product to 20 km
resolution by averaging. Each 20 km pixel spans approximately
$(0.167° \times 111\;\text{km/°})^2 \approx 344\;\text{km}^2$ at the equator,
i.e.\ roughly 344 underlying 1 km² source cells. Because `exact_extract(fun =
"sum")` sums these per-km² means, the result must be multiplied by the pixel
area (~344 km²) to recover actual population counts. Verification: the
continent-wide raster sum ($\approx$ 3.65 M) $\times$ 344 $\approx$ 1.26
billion, matching Africa's 2017 population.

```{r population-extraction}
POP_SCALE <- 344 # pixel area in km² ≈ (0.167° × 111 km)²

shp_pop <- shp_admin23 |>
    st_make_valid() |>
    st_transform(crs(pop_raster))

# Sum mean-population values within each polygon, then scale to actual counts
pop_by_adm2 <- exactextractr::exact_extract(pop_raster, shp_pop,
    fun = "sum",
    progress = FALSE
)
pop_by_adm2 <- pop_by_adm2 * POP_SCALE

shp_admin23$pop_sum <- pop_by_adm2

cat(
    "Polygons with population > 0:",
    sum(!is.na(pop_by_adm2) & pop_by_adm2 > 0), "of", nrow(shp_admin23), "\n"
)
cat("Total population:", format(round(sum(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n")
cat(
    "Range:", format(round(min(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "-",
    format(round(max(pop_by_adm2, na.rm = TRUE)), big.mark = ","), "\n"
)
```


## Risk Prediction Model (T = 5, k = 3)

Train a logistic regression on outbreak data up to 2015, predicting outbreak
occurrence during 2016--2018.

```{r model-functions}
# Build (location × split_year) feature matrix for temporal cross-validation.
#
# Timeline for each split year s (all_years = 2010:2023):
#
#   |--- hist_yrs ---|--- L_val ---|--- test_yrs ---|
#   2010 ........... s  (gap=L)   s+L+1 ... s+L+k
#   |--window_yrs--|
#    s-T+1 ...... s
#
# Parameters:
#   T_val  — feature window width (years); hist_count uses the last T years
#   k_val  — prediction horizon (years); outcome = any outbreak in next k years
#   L_val  — lead-time gap between feature window end and test period start
#            (default 0 = predict the immediately following years)
#
# valid_splits: the set of years s where both the feature window [s-T+1, s]
# and the test window [s+L+1, s+L+k] fit within all_years.
# E.g., T=5, k=3, L=0, all_years=2010:2023 → valid s = 2014:2020.
build_model_df <- function(T_val, k_val, L_val = 0L,
                           nbr_map = nbr_list) {
    # Determine which years can serve as split points:
    # need T years of history before s, and k+L years of future after s.
    valid_splits <- all_years[
        all_years >= (min(all_years) + T_val - 1L) &
            (all_years + L_val + k_val) <= max(all_years)
    ]

    map_dfr(valid_splits, function(s) {
        # Feature windows:
        window_yrs <- seq(s - T_val + 1L, s) # recent T years for hist_count
        hist_yrs <- seq(min(all_years), s) # all years up to s for lambda_hat
        # Outcome window: k years after the lead gap
        test_yrs <- seq(s + L_val + 1L, s + L_val + k_val)

        # Feature 1: hist_count — number of outbreaks ending in the last T years
        hc <- tibble(location = all_locs) |>
            left_join(
                loc_year_end_full |>
                    filter(year %in% window_yrs) |>
                    group_by(location) |>
                    summarise(hist_count = sum(n_ended), .groups = "drop"),
                by = "location"
            ) |>
            replace_na(list(hist_count = 0L))

        # Feature 2: lambda_hat — historical outbreak rate (outbreaks per year)
        # using ALL history up to s (longer window = more stable estimate)
        lam <- tibble(location = all_locs) |>
            left_join(
                loc_year_end_full |>
                    filter(year %in% hist_yrs) |>
                    group_by(location) |>
                    summarise(
                        lambda_hat = sum(n_ended) / length(hist_yrs),
                        .groups = "drop"
                    ),
                by = "location"
            ) |>
            replace_na(list(lambda_hat = 0))

        # Feature 3: nbr_hist — mean hist_count of neighbours within distance limit
        # (spatial contagion signal: locations near recent outbreaks are riskier)
        sv <- setNames(hc$hist_count, hc$location)
        nbr_df <- hc |>
            mutate(nbr_hist = map_dbl(location, function(loc) {
                nbrs <- nbr_map[[loc]]
                if (is.null(nbrs) || length(nbrs) == 0L) {
                    return(0)
                }
                nbrs <- nbrs[nbrs %in% names(sv)]
                if (length(nbrs) == 0L) {
                    return(0)
                }
                mean(sv[nbrs])
            })) |>
            dplyr::select(location, nbr_hist)

        # Outcome: obs = 1 if any active outbreak during the test window
        out <- tibble(location = all_locs) |>
            left_join(
                loc_year_full |>
                    filter(year %in% test_yrs) |>
                    group_by(location) |>
                    summarise(obs = as.integer(sum(n_active) > 0L), .groups = "drop"),
                by = "location"
            ) |>
            replace_na(list(obs = 0L))

        # Combine all features + outcome, tagged with split_year
        hc |>
            left_join(dplyr::select(lam, location, lambda_hat), by = "location") |>
            left_join(nbr_df, by = "location") |>
            left_join(out, by = "location") |>
            mutate(split_year = s)
    })
}
```

```{r train-final-model}
# Use T=5 year feature window and k=3 year prediction horizon.
# With all_years=2010:2023 and L=0, valid splits are 2014:2020.
T_val <- 5L
k_val <- 3L

df_model <- build_model_df(T_val = T_val, k_val = k_val)

# Training: pool all split years before 2015 (i.e. split_year=2014) to
# increase sample size. This means features from [2010,2014] predict 2015-2017.
# Prediction: split_year=2015 → features from [2011,2015], outcome = 2016-2018.
train_data <- df_model |> filter(split_year < 2015)
pred_data <- df_model |> filter(split_year == 2015)

cat("Training observations:", nrow(train_data), "\n")
cat("Prediction locations:", nrow(pred_data), "\n")

if (nrow(train_data) == 0 || nrow(pred_data) == 0) {
    stop("No data available for split_year = 2015. Check all_years range.")
}

# Logistic regression: P(any outbreak in next k years) ~ features
fit_final <- glm(
    obs ~ hist_count + lambda_hat + nbr_hist,
    data   = train_data,
    family = binomial
)

pred_data <- pred_data |>
    mutate(pred_prob = predict(fit_final, newdata = pred_data, type = "response"))

# Random Forest (same features, all countries): provides a nonlinear alternative
# to logistic regression for the "RF-ranked" strategy.
set.seed(42)
rf_fit_all <- randomForest(factor(obs) ~ hist_count + lambda_hat + nbr_hist,
    data = train_data, ntree = 500
)
pred_data <- pred_data |>
    mutate(pred_rf = predict(rf_fit_all, newdata = pred_data, type = "prob")[, "1"])

cat(
    "Observed outbreak prevalence (2016-2018):",
    round(100 * mean(pred_data$obs), 1), "%\n"
)
cat(
    "Predicted probability range (logistic):",
    round(min(pred_data$pred_prob), 4), "-",
    round(max(pred_data$pred_prob), 4), "\n"
)
cat(
    "Predicted probability range (RF):",
    round(min(pred_data$pred_rf), 4), "-",
    round(max(pred_data$pred_rf), 4), "\n"
)

summary(fit_final)$coefficients |>
    as.data.frame() |>
    tibble::rownames_to_column("Term") |>
    rename(SE = `Std. Error`, z = `z value`, p = `Pr(>|z|)`) |>
    mutate(OR = exp(Estimate)) |>
    dplyr::select(Term, Estimate, SE, z, p, OR) |>
    knitr::kable(
        digits = 3,
        caption = "Logistic regression coefficients (trained on split years < 2015)"
    )
```


## Build Simulation Dataset

Merge model predictions, outbreak outcomes, and population for three countries:
**COD (DRC)** (high-risk), **NGA (Nigeria)** (medium-risk), and **CMR
(Cameroon)** (lower-risk). **Single-dose OCV** is assumed for both pre-emptive
and reactive use. The total number of doses per country is varied so that
$f \in \{0.3, 0.5, 0.7\}$.

```{r build-sim-data}
# ATTACK_RATE: fraction of population that becomes a case per outbreak-year.
# Used in cost functions (illness cost = R × AR × pop) and for est_cases.
ATTACK_RATE <- 0.005

DOSES_PER_PERSON <- 1 # single-dose OCV

# Evaluation window matches the prediction horizon (k=3 → 2016-2018)
eval_years <- 2016:2018

# Compute ground-truth outbreak outcomes during the evaluation window:
# n_outbreak_years = how many of the 3 years had an active outbreak
# any_outbreak     = binary: ≥1 outbreak year
outbreak_eval <- loc_year_full |>
    filter(year %in% eval_years) |>
    group_by(location) |>
    summarise(
        n_outbreak_years = sum(n_active > 0),
        any_outbreak = as.integer(n_outbreak_years > 0),
        .groups = "drop"
    )

# Build the master simulation data frame: one row per location, with
# features (hist_count, lambda_hat, nbr_hist), model predictions
# (pred_prob, pred_rf), and evaluation outcomes (n_outbreak_years, any_outbreak)
sim_df_all <- pred_data |>
    dplyr::select(location, hist_count, lambda_hat, nbr_hist, obs, pred_prob, pred_rf) |>
    left_join(outbreak_eval, by = "location") |>
    replace_na(list(n_outbreak_years = 0L, any_outbreak = 0L))

# Attach population from the shapefile (needed for pop-weighted allocation)
pop_lookup <- shp_admin23 |>
    st_drop_geometry() |>
    dplyr::select(location, pop_sum)

sim_df_all <- sim_df_all |>
    left_join(pop_lookup, by = "location") |>
    filter(!is.na(pop_sum), pop_sum > 0) |>
    mutate(
        country = str_extract(location, "(?<=::)[A-Z]{3}(?=::)"),
        # est_cases: counterfactual case count (used by Perfect ranking + impact metrics)
        est_cases = pop_sum * ATTACK_RATE * n_outbreak_years,
        # obs_rate: empirical outbreak probability used in MC cost functions
        obs_rate = n_outbreak_years / length(eval_years)
    )

SELECTED_COUNTRIES <- c("COD", "NGA", "CMR")
COUNTRY_LABELS <- c(COD = "COD (DRC)", NGA = "NGA (Nigeria)", CMR = "CMR (Cameroon)")

sim_df_all <- sim_df_all |> filter(country %in% SELECTED_COUNTRIES)

# Coverage fractions: the total dose budget = f × P_total (country-level)
f_values <- c(0.3, 0.5, 0.7)

# Print summary per country
for (ctry in SELECTED_COUNTRIES) {
    df_c <- sim_df_all |> filter(country == ctry)
    P_c <- sum(df_c$pop_sum)
    cat(sprintf("\n=== %s ===\n", COUNTRY_LABELS[ctry]))
    cat(sprintf("Locations: %d\n", nrow(df_c)))
    cat(sprintf(
        "  with outbreaks (%d-%d): %d (%.1f%%)\n",
        min(eval_years), max(eval_years),
        sum(df_c$any_outbreak), 100 * mean(df_c$any_outbreak)
    ))
    cat(sprintf(
        "Total population: %s (%.1f M)\n",
        format(round(P_c), big.mark = ","), P_c / 1e6
    ))
    cat(sprintf(
        "Counterfactual cases: %s\n",
        format(round(sum(df_c$est_cases)), big.mark = ",")
    ))
}

# Keep sim_df as COD for backward compatibility in remaining single-country code
sim_df <- sim_df_all |> filter(country == "COD")
P_total <- sum(sim_df$pop_sum)
```


## Classification Method Comparison (DRC)

The logistic regression above is the simplest approach. Here we compare several
classification methods — all trained on the same features (`hist_count`,
`lambda_hat`, `nbr_hist`) and the same train/test split — to see whether a more
flexible model improves targeting accuracy ($\rho$) for DRC.

```{r classifier-comparison}
# Compare 5 classifiers for DRC only: Logistic, GAM, Random Forest, GBM, k-NN.
# All use the same 3 features (hist_count, lambda_hat, nbr_hist) and same
# train/test split. The goal is to see whether nonlinear methods improve
# targeting accuracy (rho) over logistic regression.
cod_locs <- sim_df_all |>
    filter(country == "COD") |>
    pull(location)
train_cod <- train_data |> filter(location %in% cod_locs)
pred_cod <- pred_data |> filter(location %in% cod_locs)

features <- c("hist_count", "lambda_hat", "nbr_hist")
X_train <- train_cod[, features] |> as.data.frame()
y_train <- train_cod$obs
X_pred <- pred_cod[, features] |> as.data.frame()

# Outcome for rho evaluation: observed outbreak years in 2016-2018
cod_sim <- sim_df_all |> filter(country == "COD")

cat(sprintf(
    "DRC training: %d obs, prediction: %d locations\n",
    nrow(train_cod), nrow(pred_cod)
))
cat(sprintf(
    "Outcome prevalence (train): %.1f%%, (test 2016-18): %.1f%%\n",
    100 * mean(y_train), 100 * mean(cod_sim$any_outbreak)
))

# --- 1. Logistic regression (baseline — already fitted) ---
pred_logistic <- pred_cod$pred_prob

# --- 2. GAM with smooths ---
n_unique_hc <- length(unique(train_cod$hist_count))
n_unique_lam <- length(unique(train_cod$lambda_hat))
n_unique_nbr <- length(unique(train_cod$nbr_hist))
k_hc <- min(10, n_unique_hc - 1)
k_lam <- min(10, n_unique_lam - 1)
k_nbr <- min(10, n_unique_nbr - 1)
gam_fit <- gam(
    obs ~ s(hist_count, k = k_hc) + s(lambda_hat, k = k_lam) +
        s(nbr_hist, k = k_nbr),
    data = train_cod, family = binomial, method = "REML"
)
pred_gam <- predict(gam_fit, newdata = X_pred, type = "response")

# --- 3. Random forest ---
set.seed(42)
rf_fit <- randomForest(factor(obs) ~ hist_count + lambda_hat + nbr_hist,
    data = train_cod, ntree = 500
)
pred_rf <- predict(rf_fit, newdata = X_pred, type = "prob")[, "1"]

# --- 4. Gradient boosting (GBM) ---
set.seed(42)
gbm_fit <- gbm(obs ~ hist_count + lambda_hat + nbr_hist,
    data = train_cod, distribution = "bernoulli",
    n.trees = 500, interaction.depth = 3, shrinkage = 0.05,
    bag.fraction = 0.8, verbose = FALSE
)
best_iter <- gbm.perf(gbm_fit, method = "OOB", plot.it = FALSE)
pred_gbm <- predict(gbm_fit,
    newdata = X_pred, n.trees = best_iter,
    type = "response"
)

# --- 5. k-NN (k = 10, probability mode) ---
# Scale features for distance-based method
scale_params <- list(
    center = colMeans(X_train),
    scale  = apply(X_train, 2, sd) + 1e-10
)
X_train_sc <- scale(X_train, center = scale_params$center, scale = scale_params$scale)
X_pred_sc <- scale(X_pred, center = scale_params$center, scale = scale_params$scale)

knn_fit <- knn(
    train = X_train_sc, test = X_pred_sc, cl = factor(y_train),
    k = 10, prob = TRUE
)
# knn() returns the class + attr(,"prob") = proportion of winning class
knn_raw_prob <- attr(knn_fit, "prob")
knn_class <- as.integer(as.character(knn_fit))
# Convert to P(outbreak): if class = 1, prob is correct; if class = 0, flip
pred_knn <- ifelse(knn_class == 1, knn_raw_prob, 1 - knn_raw_prob)

# --- Compile predictions ---
classifier_preds <- tibble(
    location = pred_cod$location,
    Logistic = pred_logistic,
    GAM = as.numeric(pred_gam),
    `Random Forest` = pred_rf,
    GBM = pred_gbm,
    kNN = pred_knn
) |>
    left_join(cod_sim |> dplyr::select(location, n_outbreak_years, any_outbreak, est_cases),
        by = "location"
    )
```

### Targeting accuracy ($\rho$) by classifier

```{r classifier-rho-comparison}
# Evaluate each classifier's targeting accuracy:
# - Spearman rho: rank correlation between predicted prob and outbreak burden
# - AUC: area under the ROC curve for binary outbreak classification
# Both are computed on the 2016-2018 test period (out-of-sample).
classifier_methods <- c("Logistic", "GAM", "Random Forest", "GBM", "kNN")

rho_table <- tibble(
    Method = classifier_methods,
    rho_spearman = map_dbl(classifier_methods, function(m) {
        cor(classifier_preds[[m]], classifier_preds$n_outbreak_years,
            method = "spearman"
        )
    }),
    AUC = map_dbl(classifier_methods, function(m) {
        as.numeric(pROC::auc(pROC::roc(classifier_preds$any_outbreak,
            classifier_preds[[m]],
            quiet = TRUE
        )))
    })
)

# Add simple baselines
baseline_rho <- tibble(
    Method = c("Count-ranked", "Random"),
    rho_spearman = c(
        cor(cod_sim$hist_count, cod_sim$n_outbreak_years, method = "spearman"),
        0
    ),
    AUC = c(
        as.numeric(pROC::auc(pROC::roc(cod_sim$any_outbreak,
            cod_sim$hist_count,
            quiet = TRUE
        ))),
        0.5
    )
)

rho_table <- bind_rows(rho_table, baseline_rho) |>
    arrange(desc(rho_spearman))

knitr::kable(rho_table,
    digits = 3,
    col.names = c("Method", "Spearman \u03c1", "AUC"),
    caption = "DRC: Targeting accuracy by classification method (eval 2016-2018)"
)
```

```{r fig-classifier-rho, fig.width=8, fig.height=5}
rho_plot_df <- rho_table |>
    mutate(
        Method = fct_reorder(Method, rho_spearman),
        type = ifelse(Method %in% c("Count-ranked", "Random"),
            "Baseline", "Classifier"
        )
    )

ggplot(rho_plot_df, aes(x = Method, y = rho_spearman, fill = type)) +
    geom_col(width = 0.6, alpha = 0.85) +
    geom_text(aes(label = sprintf("%.3f", rho_spearman)),
        hjust = -0.15, size = 3.5
    ) +
    coord_flip() +
    scale_fill_manual(values = c(
        "Classifier" = "steelblue",
        "Baseline" = "grey60"
    )) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(
        x = NULL, y = expression("Spearman" ~ rho),
        title = "DRC: Targeting accuracy by classification method",
        subtitle = "Correlation between predicted risk and actual outbreak burden (2016-2018)"
    ) +
    theme_minimal(base_size = 12) +
    guides(fill = "none")
```

```{r fig-classifier-scatter, fig.width=14, fig.height=8}
# Scatter: predicted probability vs actual outcomes for each classifier
scatter_df <- classifier_preds |>
    pivot_longer(
        cols = all_of(classifier_methods),
        names_to = "Method", values_to = "pred_prob_clf"
    ) |>
    mutate(Method = factor(Method, levels = rho_table$Method))

ggplot(scatter_df, aes(x = pred_prob_clf, y = n_outbreak_years)) +
    geom_jitter(aes(colour = factor(any_outbreak)),
        alpha = 0.4, height = 0.1, width = 0, size = 1
    ) +
    geom_smooth(
        method = "loess", se = FALSE, colour = "black",
        linewidth = 0.6
    ) +
    facet_wrap(~Method, nrow = 2, scales = "free_x") +
    scale_colour_manual(
        values = c("0" = "grey60", "1" = "firebrick"),
        labels = c("No outbreak", "Outbreak")
    ) +
    labs(
        x = "Predicted probability",
        y = "Outbreak years (2016-2018)",
        colour = NULL,
        title = "DRC: Predicted risk vs observed outcomes by classifier"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```


## Analytical Framework: Optimal Pre-emptive Fraction

### Scenario parameters

```{r scenario-params}
# Vaccine parameters (single-dose OCV)
nu_val <- 1 # pre-emptive VE: prevents outbreaks entirely
r_val <- 0.6 # reactive effectiveness
R_val <- 500 # cost ratio C_I / C_V (per-infection: C_I = cost_per_case)
n_trials <- length(eval_years)

# Shared colour palette for strategies
STRATEGY_COLOURS <- c(
    "Oracle" = "firebrick", "Perfect ranking" = "salmon",
    "Model-ranked" = "steelblue", "RF-ranked" = "darkorchid",
    "Count-ranked" = "darkgreen", "Random" = "grey50"
)
STRATEGY_LEVELS <- c(
    "Oracle", "Perfect ranking", "Model-ranked", "RF-ranked",
    "Count-ranked", "Random"
)
```

### Helper functions

```{r helper-functions}
# --- Beta-Binomial MLE ---
# Fits a Beta(a, b) distribution to location-level outbreak counts.
# Each location had n_trials evaluation years and k_obs[i] outbreak years.
# The Beta-Binomial model assumes each location draws its own p_i ~ Beta(a,b),
# then k_i ~ Binomial(n, p_i). MLE estimates (a, b) via Nelder-Mead on the
# log-marginal-likelihood. Parameters are optimized in log-space for positivity.
# Returns: a, b, p = a/(a+b) (mean outbreak probability), convergence flag.
fit_betabinom <- function(k_obs, n_trials) {
    ll <- function(par, k, n) {
        a <- exp(par[1])
        b <- exp(par[2])
        sum(lchoose(n, k) + lbeta(k + a, n - k + b) - lbeta(a, b))
    }
    # Method-of-moments initialization
    p_init <- mean(k_obs) / n_trials
    v_init <- var(k_obs / n_trials)
    a_init <- max(p_init * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    b_init <- max((1 - p_init) * (p_init * (1 - p_init) / max(v_init, 1e-6) - 1), 0.1)
    fit <- optim(
        par = log(c(a_init, b_init)), fn = ll, k = k_obs, n = n_trials,
        method = "Nelder-Mead", control = list(fnscale = -1, maxit = 5000)
    )
    list(
        a = exp(fit$par[1]), b = exp(fit$par[2]),
        p = exp(fit$par[1]) / (exp(fit$par[1]) + exp(fit$par[2])),
        convergence = fit$convergence
    )
}

# --- Population-weighted MC cost function ---
# R is defined at the INDIVIDUAL level:
#   C_V = cost per person vaccinated (vaccine + delivery)
#   C_I = cost per infection
#   R = C_I / C_V
#   AR = attack rate (fraction of population infected during an outbreak)
#
# Per-location cost convention (in C_V units):
#   - Vaccinating location i:  pop_i
#   - Unprotected outbreak:    R × AR × pop_i
#   - Pre-emptive + outbreak:  pop_i + (1-ν) × R × AR × pop_i
#   - Reactive + outbreak:     pop_i + (1-r) × R × AR × pop_i
#   - No action, no outbreak:  0
# Normalized by P_total → per-capita cost.

cost_popweighted_mc <- function(alpha, f, pop, prob, score, R, r, AR, nu = 1,
                                oracle = FALSE, n_mc = 300, seed = 42, ...) {
    n <- length(pop)
    P_total <- sum(pop)
    pre_budget <- alpha * f * P_total
    react_budget <- (1 - alpha) * f * P_total

    # Score-based strategies: pre-allocate once (fixed ranking)
    if (!oracle) {
        ranked <- order(score, decreasing = TRUE)
        cum_pop <- cumsum(pop[ranked])
        n_pre <- sum(cum_pop <= pre_budget)
        pre_idx_fixed <- if (n_pre > 0) ranked[seq_len(n_pre)] else integer(0)
    }

    costs <- numeric(n_mc)
    for (s in seq_len(n_mc)) {
        set.seed(seed + s)
        Y <- rbinom(n, 1, prob)

        # Oracle: allocate pre-emptive doses to outbreak locations (known)
        if (oracle) {
            ob_all <- which(Y == 1)
            if (length(ob_all) > 0 && pre_budget > 0) {
                ob_ranked <- ob_all[order(pop[ob_all], decreasing = TRUE)]
                cum_pop_ob <- cumsum(pop[ob_ranked])
                n_pre_ob <- sum(cum_pop_ob <= pre_budget)
                pre_idx <- if (n_pre_ob > 0) ob_ranked[seq_len(n_pre_ob)] else integer(0)
            } else {
                pre_idx <- integer(0)
            }
        } else {
            pre_idx <- pre_idx_fixed
        }
        rem_idx <- setdiff(seq_len(n), pre_idx)

        # Pre-emptive locations: vaccination cost + residual illness
        pop_pre <- sum(pop[pre_idx])
        burden_pre_residual <- sum(Y[pre_idx] * pop[pre_idx]) * (1 - nu)

        # Reactive: deploy to outbreaking non-pre-emptive locations
        ob_rem <- rem_idx[Y[rem_idx] == 1]
        pop_react <- 0
        burden_react_residual <- 0
        burden_uncovered <- 0

        if (length(ob_rem) > 0) {
            if (react_budget > 0) {
                ob_order <- ob_rem[order(pop[ob_rem], decreasing = TRUE)]
                cum_pop_r <- cumsum(pop[ob_order])
                n_react <- sum(cum_pop_r <= react_budget)
                if (n_react > 0) {
                    react_sel <- ob_order[seq_len(n_react)]
                    pop_react <- sum(pop[react_sel])
                    burden_react_residual <- pop_react * (1 - r)
                    if (n_react < length(ob_order)) {
                        burden_uncovered <- sum(pop[ob_order[(n_react + 1):length(ob_order)]])
                    }
                } else {
                    burden_uncovered <- sum(pop[ob_rem])
                }
            } else {
                burden_uncovered <- sum(pop[ob_rem])
            }
        }

        # Cost = vaccination + R × AR × illness burden
        vax_cost <- pop_pre + pop_react
        illness <- R * AR * (burden_pre_residual + burden_react_residual + burden_uncovered)
        costs[s] <- (vax_cost + illness) / P_total
    }
    mean(costs)
}

# --- Grid-search optimizer for pop-weighted alpha* (score-based) ---
# Sweeps alpha over [0,1] and returns the alpha that minimizes
# cost_popweighted_mc (using actual strategy scores for targeting).
get_alpha_star_pw <- function(f, pop, prob, score, R, r, AR, nu = 1,
                              n_mc = 300, seed = 42, grid_len = 51) {
    alpha_grid <- seq(0, 1, length.out = grid_len)
    costs <- sapply(alpha_grid, function(a) {
        cost_popweighted_mc(a, f, pop, prob, score, R, r, AR, nu,
            n_mc = n_mc, seed = seed
        )
    })
    list(
        alpha_star = alpha_grid[which.min(costs)],
        cost_star = min(costs), alpha_grid = alpha_grid, cost_grid = costs
    )
}

# mean_risk_pre: E[p | p > q] where q = Beta quantile at (1 - f_pre).
# This is the average outbreak probability among pre-emptively targeted
# locations when the top f_pre fraction is selected.
mean_risk_pre <- function(f_pre, a, b) {
    if (f_pre <= 0) {
        return(0)
    }
    if (f_pre >= 1) {
        return(a / (a + b))
    }
    q <- qbeta(1 - f_pre, a, b)
    (a / ((a + b) * f_pre)) * (1 - pbeta(q, a + 1, b))
}

# cost_beta: total expected cost = vaccination cost + illness cost.
# Pre-emptive targets the top alpha*f fraction; reactive covers remaining
# outbreaks up to capacity (1-alpha)*f. See analytic_hetero.qmd for derivation.
cost_beta <- function(alpha, f, a, b, R, r, nu = 1) {
    alpha <- pmax(0, pmin(1, alpha))
    p <- a / (a + b)
    f_pre <- alpha * f
    f_react_cap <- (1 - alpha) * f
    p_bar_pre <- vapply(f_pre, function(fp) mean_risk_pre(fp, a, b), numeric(1))
    E_pre <- f_pre * p_bar_pre
    c_pre_val <- f_pre + E_pre * (1 - nu) * R
    E_rem <- p - E_pre
    is_rich <- f_react_cap >= E_rem
    c_react_val <- ifelse(is_rich, E_rem * (1 + (1 - r) * R),
        f_react_cap * (1 + (1 - r) * R) + (E_rem - f_react_cap) * R
    )
    c_pre_val + c_react_val
}

# get_alpha_star_beta: continuous optimizer for analytical alpha* (L-BFGS-B).
get_alpha_star_beta <- function(R, f, a, b, r, nu = 1) {
    optim(
        par = 0.5, fn = cost_beta,
        f = f, a = a, b = b, R = R, r = r, nu = nu,
        method = "L-BFGS-B", lower = 0, upper = 1
    )$par
}

# --- Compute empirical rho for each strategy ---
# Spearman rank correlation between each strategy's targeting score and
# the actual outbreak burden (n_outbreak_years). Higher rho = better targeting.
# Random baseline: average of 100 random permutations (should be ≈ 0).
compute_strategy_rhos <- function(df_c) {
    set.seed(42)
    rho_random <- mean(replicate(100, {
        cor(runif(nrow(df_c)), df_c$n_outbreak_years, method = "spearman")
    }))
    rho_rf <- cor(df_c$pred_rf, df_c$n_outbreak_years, method = "spearman")
    tibble(
        strategy = c(
            "Oracle", "Perfect ranking", "Model-ranked", "RF-ranked",
            "Count-ranked", "Random"
        ),
        rho = c(
            1.0,
            cor(df_c$est_cases, df_c$n_outbreak_years, method = "spearman"),
            cor(df_c$pred_prob, df_c$n_outbreak_years, method = "spearman"),
            rho_rf,
            cor(df_c$hist_count, df_c$n_outbreak_years, method = "spearman"),
            rho_random
        )
    )
}
```

### Per-country Beta-Binomial fits

```{r per-country-bb-fits}
# Fit a Beta-Binomial to each country's outbreak counts.
# k_obs = n_outbreak_years per location (0..3), n_trials = 3 eval years.
# The resulting Beta(a,b) describes the distribution of outbreak probabilities
# across locations within each country.
bb_fits <- setNames(lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    bb <- fit_betabinom(df_c$n_outbreak_years, n_trials)
    cat(sprintf("\n=== %s ===\n", COUNTRY_LABELS[ctry]))
    cat(sprintf("n locations: %d\n", nrow(df_c)))
    cat(sprintf(
        "Beta(%.3f, %.3f), mean p = %.3f, concentration = %.3f\n",
        bb$a, bb$b, bb$p, bb$a + bb$b
    ))
    cat(sprintf("Convergence: %d\n", bb$convergence))
    bb
}), SELECTED_COUNTRIES)

# For backward compatibility with single-country code below
a_hat <- bb_fits[["COD"]]$a
b_hat <- bb_fits[["COD"]]$b
p_hat <- bb_fits[["COD"]]$p
```

```{r beta-distribution-diagnostics, fig.width=14, fig.height=12}
# Beta-Binomial diagnostics for each country (CDF + PMF)
# Uses the 3-year evaluation window (2016-2018) that the model was fitted on.
diag_plots <- lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    bb <- bb_fits[[ctry]]
    k_obs_c <- df_c$n_outbreak_years
    beta_label <- sprintf("Beta(%.2f, %.2f)", bb$a, bb$b)

    x_grid <- seq(0, 1, length.out = 500)
    beta_cdf <- pbeta(x_grid, shape1 = bb$a, shape2 = bb$b)

    p_cdf <- ggplot() +
        stat_ecdf(
            data = df_c, aes(x = obs_rate),
            colour = "darkorange", linewidth = 0.8, pad = FALSE
        ) +
        geom_line(
            data = tibble(x = x_grid, y = beta_cdf), aes(x, y),
            colour = "firebrick", linewidth = 0.8, linetype = "dashed"
        ) +
        labs(
            x = "Outbreak rate", y = "CDF",
            title = sprintf("%s — CDF vs %s (k=3)", COUNTRY_LABELS[ctry], beta_label)
        ) +
        theme_minimal(base_size = 10)

    bb_pmf <- sapply(0:n_trials, function(k) {
        exp(lchoose(n_trials, k) + lbeta(k + bb$a, n_trials - k + bb$b) -
            lbeta(bb$a, bb$b))
    })
    obs_counts <- tabulate(k_obs_c + 1L, nbins = n_trials + 1L) / length(k_obs_c)
    pmf_df <- tibble(
        k = rep(0:n_trials, 2), prob = c(obs_counts, bb_pmf),
        source = rep(c("Observed", "Beta-Binomial"), each = n_trials + 1)
    )

    p_pmf <- ggplot(pmf_df, aes(x = k, y = prob, fill = source)) +
        geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.8) +
        scale_fill_manual(values = c(
            "Observed" = "darkorange",
            "Beta-Binomial" = "firebrick"
        )) +
        scale_x_continuous(breaks = 0:n_trials) +
        labs(
            x = "Outbreak years", y = "Proportion", fill = NULL,
            title = sprintf("%s — BB PMF (k=3)", COUNTRY_LABELS[ctry])
        ) +
        theme_minimal(base_size = 10) +
        theme(legend.position = "bottom")

    p_cdf + p_pmf
})
wrap_plots(diag_plots, ncol = 1) +
    plot_annotation(title = "Beta-Binomial diagnostics by country (evaluation window: 2016-2018)")
```

The 3-year CDF has only 3 steps because `obs_rate = k/3` can only be 0, 1/3,
2/3, or 1. Below we repeat the diagnostics using the **full surveillance window
(2015–2023, k = 9 years)** to get finer granularity. The Beta(a, b) parameters
are re-estimated from 9-year counts so the comparison is self-consistent.

```{r beta-distribution-diagnostics-long, fig.width=14, fig.height=12}
# Longer evaluation window for more granular CDF comparison
long_years <- 2015:2023
n_long <- length(long_years)

# Compute outbreak counts over the longer window for locations in sim_df_all
outbreak_long <- loc_year_full |>
    filter(year %in% long_years) |>
    group_by(location) |>
    summarise(n_outbreak_long = sum(n_active > 0), .groups = "drop") |>
    mutate(obs_rate_long = n_outbreak_long / n_long)

sim_long <- sim_df_all |>
    dplyr::select(location, country) |>
    left_join(outbreak_long, by = "location") |>
    replace_na(list(n_outbreak_long = 0L, obs_rate_long = 0))

# Fit Beta-Binomial to 9-year counts for each country
bb_fits_long <- setNames(lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_long |> filter(country == ctry)
    fit_betabinom(df_c$n_outbreak_long, n_long)
}), SELECTED_COUNTRIES)

diag_plots_long <- lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_long |> filter(country == ctry)
    bb <- bb_fits_long[[ctry]]
    k_obs_c <- df_c$n_outbreak_long
    beta_label <- sprintf("Beta(%.2f, %.2f)", bb$a, bb$b)

    x_grid <- seq(0, 1, length.out = 500)
    beta_cdf <- pbeta(x_grid, shape1 = bb$a, shape2 = bb$b)

    # CDF: obs_rate_long = k/9, so 10 possible values → much finer steps
    p_cdf <- ggplot() +
        stat_ecdf(
            data = df_c, aes(x = obs_rate_long),
            colour = "darkorange", linewidth = 0.8, pad = FALSE
        ) +
        geom_line(
            data = tibble(x = x_grid, y = beta_cdf), aes(x, y),
            colour = "firebrick", linewidth = 0.8, linetype = "dashed"
        ) +
        labs(
            x = "Outbreak rate", y = "CDF",
            title = sprintf(
                "%s — CDF vs %s (k=%d)", COUNTRY_LABELS[ctry],
                beta_label, n_long
            )
        ) +
        theme_minimal(base_size = 10)

    # PMF: Beta-Binomial(n_long, a, b) vs observed counts
    bb_pmf <- sapply(0:n_long, function(k) {
        exp(lchoose(n_long, k) + lbeta(k + bb$a, n_long - k + bb$b) -
            lbeta(bb$a, bb$b))
    })
    obs_counts <- tabulate(k_obs_c + 1L, nbins = n_long + 1L) / length(k_obs_c)
    pmf_df <- tibble(
        k = rep(0:n_long, 2), prob = c(obs_counts, bb_pmf),
        source = rep(c("Observed", "Beta-Binomial"), each = n_long + 1)
    )

    p_pmf <- ggplot(pmf_df, aes(x = k, y = prob, fill = source)) +
        geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.8) +
        scale_fill_manual(values = c(
            "Observed" = "darkorange",
            "Beta-Binomial" = "firebrick"
        )) +
        scale_x_continuous(breaks = 0:n_long) +
        labs(
            x = "Outbreak years", y = "Proportion", fill = NULL,
            title = sprintf("%s — BB PMF (k=%d)", COUNTRY_LABELS[ctry], n_long)
        ) +
        theme_minimal(base_size = 10) +
        theme(legend.position = "bottom")

    p_cdf + p_pmf
})
wrap_plots(diag_plots_long, ncol = 1) +
    plot_annotation(title = sprintf(
        "Beta-Binomial diagnostics by country (full window: %d–%d, k = %d)",
        min(long_years), max(long_years), n_long
    ))
```


### Empirical targeting accuracy (rho) and population-weighted optimal alpha

We compute two types of optimal $\alpha^*$ for each country:

1. **Equal-weight** (original `cost_beta()`): treats locations as equal-weight
   units, perfect targeting
2. **Population-weighted** (`cost_popweighted_mc()`): accounts for heterogeneous
   population sizes and strategy-specific targeting via Monte Carlo

```{r empirical-rho}
# Compute empirical rho for each strategy x country
rho_by_country <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    compute_strategy_rhos(df_c) |> mutate(country = ctry)
})

rho_by_country |>
    pivot_wider(names_from = country, values_from = rho) |>
    knitr::kable(
        digits = 3,
        caption = "Spearman rho (targeting score vs. actual outbreak-years) by strategy and country"
    )
```

```{r popweighted-alpha-star}
# Compute population-weighted alpha* for each strategy x country x f
# This uses Monte Carlo: for each alpha, draw outbreaks from Bernoulli(obs_rate),
# allocate vaccines by strategy score + population, compute cost.
pw_alpha_results <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    bb <- bb_fits[[ctry]]
    cat(sprintf("\n=== %s (n=%d, p=%.3f) ===\n", ctry, nrow(df_c), bb$p))

    map_dfr(f_values, function(fv) {
        # True Oracle: knows outbreak realization
        alpha_grid_51 <- seq(0, 1, length.out = 51)
        oracle_costs <- sapply(alpha_grid_51, function(a) {
            cost_popweighted_mc(a, fv, df_c$pop_sum, df_c$obs_rate,
                score = numeric(nrow(df_c)),
                R_val, r_val, ATTACK_RATE, oracle = TRUE, n_mc = 300
            )
        })
        oracle_alpha_star <- alpha_grid_51[which.min(oracle_costs)]
        cat(sprintf("  f=%.1f %-14s: alpha*_pw = %.2f\n", fv, "Oracle", oracle_alpha_star))
        oracle_result <- tibble(
            strategy = "Oracle", f = fv,
            alpha_star_pw = oracle_alpha_star,
            cost_star_pw = min(oracle_costs)
        )

        # Score-based strategies
        strategies <- list(
            `Perfect ranking` = df_c$est_cases,
            `Model-ranked`    = df_c$pred_prob,
            `RF-ranked`       = df_c$pred_rf,
            `Count-ranked`    = df_c$hist_count
        )
        strat_results <- imap_dfr(strategies, function(score, strat_name) {
            res <- get_alpha_star_pw(fv, df_c$pop_sum, df_c$obs_rate, score,
                R_val, r_val, ATTACK_RATE,
                n_mc = 300
            )
            cat(sprintf("  f=%.1f %-14s: alpha*_pw = %.2f\n", fv, strat_name, res$alpha_star))
            tibble(
                strategy = strat_name, f = fv, alpha_star_pw = res$alpha_star,
                cost_star_pw = res$cost_star
            )
        })

        # Random: use uniform scores (average over multiple seeds)
        set.seed(42)
        random_scores <- replicate(20, runif(nrow(df_c)), simplify = FALSE)
        rand_costs <- sapply(alpha_grid_51, function(a) {
            mean(sapply(random_scores, function(sc) {
                cost_popweighted_mc(a, fv, df_c$pop_sum, df_c$obs_rate, sc,
                    R_val, r_val, ATTACK_RATE,
                    n_mc = 100
                )
            }))
        })
        rand_alpha_star <- alpha_grid_51[which.min(rand_costs)]
        cat(sprintf("  f=%.1f %-14s: alpha*_pw = %.2f\n", fv, "Random", rand_alpha_star))

        bind_rows(
            oracle_result,
            strat_results,
            tibble(
                strategy = "Random", f = fv,
                alpha_star_pw = rand_alpha_star,
                cost_star_pw = min(rand_costs)
            )
        )
    }) |> mutate(country = ctry)
})

# Equal-weight alpha*: uses the analytical cost_beta() with Beta(a,b)
# from the BB fit. This ignores population heterogeneity and assumes
# perfect targeting — serves as a theoretical reference point.
ew_alpha_results <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    bb <- bb_fits[[ctry]]
    map_dfr(f_values, function(fv) {
        a_star <- optim(
            par = 0.5, fn = cost_beta,
            f = fv, a = bb$a, b = bb$b, R = R_val, r = r_val,
            method = "L-BFGS-B", lower = 0, upper = 1
        )$par
        tibble(country = ctry, f = fv, alpha_star_ew = a_star)
    })
})

# Merge all three: rho (targeting accuracy), pop-weighted alpha* (MC-based),
# and equal-weight alpha* (analytical). This master table is used for
# overlay points on cost curves and comparison tables.
alpha_comparison <- pw_alpha_results |>
    left_join(rho_by_country, by = c("strategy", "country")) |>
    left_join(ew_alpha_results, by = c("country", "f"))

alpha_comparison |>
    dplyr::select(country, strategy, f, rho, alpha_star_pw, alpha_star_ew) |>
    knitr::kable(
        digits = 3,
        col.names = c("Country", "Strategy", "f", "\u03c1", "\u03b1*_pw", "\u03b1*_ew"),
        caption = "Pop-weighted vs equal-weight optimal alpha by strategy and country"
    )
```

```{r cost-curve-plot-pw, fig.width=14, fig.height=10}
# Population-weighted cost curves by strategy for each country x f
alpha_grid_pw <- seq(0, 1, length.out = 51)

pw_cost_curves <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    map_dfr(f_values, function(fv) {
        # True Oracle: knows outbreak realization, vaccinates only outbreak locations
        oracle_costs <- sapply(alpha_grid_pw, function(a) {
            cost_popweighted_mc(a, fv, df_c$pop_sum, df_c$obs_rate,
                score = numeric(nrow(df_c)),
                R_val, r_val, ATTACK_RATE, oracle = TRUE, n_mc = 200
            )
        })
        oracle_curve <- tibble(
            alpha = alpha_grid_pw, cost = oracle_costs,
            strategy = "Oracle", f = fv, country = ctry
        )

        # Score-based strategies (fixed ranking)
        strategies <- list(
            `Perfect ranking` = df_c$est_cases,
            `Model-ranked`    = df_c$pred_prob,
            `RF-ranked`       = df_c$pred_rf,
            `Count-ranked`    = df_c$hist_count
        )
        indiv_curves <- imap_dfr(strategies, function(score, strat_name) {
            costs <- sapply(alpha_grid_pw, function(a) {
                cost_popweighted_mc(a, fv, df_c$pop_sum, df_c$obs_rate, score,
                    R_val, r_val, ATTACK_RATE,
                    n_mc = 200
                )
            })
            tibble(
                alpha = alpha_grid_pw, cost = costs,
                strategy = strat_name, f = fv, country = ctry
            )
        })

        # Random: average over 20 random score vectors for stability
        set.seed(42)
        rand_scores <- replicate(20, runif(nrow(df_c)), simplify = FALSE)
        rand_costs <- sapply(alpha_grid_pw, function(a) {
            mean(sapply(rand_scores, function(sc) {
                cost_popweighted_mc(a, fv, df_c$pop_sum, df_c$obs_rate, sc,
                    R_val, r_val, ATTACK_RATE,
                    n_mc = 100
                )
            }))
        })
        rand_curve <- tibble(
            alpha = alpha_grid_pw, cost = rand_costs,
            strategy = "Random", f = fv, country = ctry
        )

        bind_rows(oracle_curve, indiv_curves, rand_curve)
    })
})

# MC-based optimal alpha points (diamonds)
opt_pts <- alpha_comparison |>
    dplyr::select(country, strategy, f, alpha_star_pw, cost_star_pw) |>
    rename(alpha = alpha_star_pw, cost = cost_star_pw)

# Theoretical alpha* from analytical Beta-Binomial model (equal-weight)
theo_vlines <- ew_alpha_results |>
    rename(alpha_star_theo = alpha_star_ew)

ggplot(pw_cost_curves, aes(x = alpha, y = cost, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    geom_vline(
        data = theo_vlines,
        aes(xintercept = alpha_star_theo),
        linetype = "dashed", colour = "black", linewidth = 0.4, alpha = 0.5
    ) +
    geom_point(
        data = opt_pts, aes(x = alpha, y = cost),
        size = 2.5, shape = 18
    ) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Pop-weighted cost per capita",
        colour = "Strategy",
        title = "Population-weighted cost curves by strategy, country, and f",
        subtitle = sprintf(
            "R = %.0f, r = %.1f; diamonds = MC optimal alpha*; dashed line = theoretical alpha* (Beta-Binomial)",
            R_val, r_val
        )
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

```{r alpha-values}
# For backward compatibility: COD equal-weight alpha_stars
alpha_stars <- setNames(
    ew_alpha_results |> filter(country == "COD") |> pull(alpha_star_ew),
    f_values
)

# Per-country alpha* (pop-weighted, Model-ranked as default)
alpha_stars_by_country <- alpha_comparison |>
    filter(strategy == "Model-ranked") |>
    dplyr::select(country, f, alpha_star_pw)

# Build the alpha grid for impact simulations: standard values (0, 0.25, 0.5,
# 0.75, 1.0) plus all strategy-specific alpha* values, so the curves pass
# through the optimal points for each strategy.
alpha_compare_by_f <- lapply(f_values, function(fv) {
    a_stars_all <- alpha_comparison |>
        filter(f == fv) |>
        pull(alpha_star_pw) |>
        unique()
    sort(unique(round(c(0, 0.25, 0.5, a_stars_all, 0.75, 1.0), 2)))
})
names(alpha_compare_by_f) <- f_values

for (fv in f_values) {
    cat(sprintf(
        "f = %.1f: alpha values = %s\n",
        fv, paste(alpha_compare_by_f[[as.character(fv)]], collapse = ", ")
    ))
}
```


## Mixed-Strategy Vaccination Impact

For each strategy, coverage fraction $f$, and pre-emptive fraction $\alpha$:

1. **Pre-emptive**: rank locations by strategy score, vaccinate top locations
   until $\alpha \times f \times P$ people are covered. Cases reduced by factor $\nu$.
2. **Reactive**: remaining $(1 - \alpha) \times f \times P$ people-worth of doses
   are held back. When outbreaks occur in non-pre-emptive locations, deploy
   reactive doses (prioritizing largest outbreaks). Each reactively-covered
   location has cases reduced by factor $r = 0.6$.

```{r impact-function}
# Simulate one round of mixed pre-emptive + reactive vaccination on actual data.
#
# For a given allocation (alpha, f), ranks locations by `score`, allocates
# pre-emptive doses to the top-ranked locations, then deploys remaining doses
# reactively to outbreak locations that weren't pre-emptively covered.
#
# Returns a single-row tibble with: cases averted (pre/react/total), doses used,
# efficiency (cases per 1k doses), sim_cost (per-capita cost in C_V units),
# and counts of covered/uncovered locations.
compute_impact_mixed <- function(df, score, alpha, effective_people, r,
                                 nu = 1, R = 500, oracle = FALSE) {
    n <- nrow(df)

    pre_doses <- alpha * effective_people
    react_doses <- (1 - alpha) * effective_people

    # --- Pre-emptive allocation ---
    # NOTE: greedy whole-location allocation. Leftover doses that cannot fully
    # cover the next location are unused (no partial coverage). This slightly
    # underestimates impact; the effect is small when locations are numerous.
    if (oracle) {
        # Oracle: only pre-emptively vaccinate outbreak locations
        ob_idx <- which(df$any_outbreak == 1)
        if (length(ob_idx) > 0 && pre_doses > 0) {
            ob_ranked <- ob_idx[order(df$est_cases[ob_idx], decreasing = TRUE)]
            cum_pop <- cumsum(df$pop_sum[ob_ranked])
            n_pre <- sum(cum_pop <= pre_doses)
            pre_idx <- if (n_pre > 0) ob_ranked[seq_len(n_pre)] else integer(0)
        } else {
            pre_idx <- integer(0)
        }
        # Unused pre-emptive budget redirected to reactive
        doses_pre <- sum(df$pop_sum[pre_idx])
        react_doses <- react_doses + (pre_doses - doses_pre)
    } else {
        ranked_idx <- order(score, decreasing = TRUE)
        if (pre_doses > 0) {
            cum_pop <- cumsum(df$pop_sum[ranked_idx])
            n_pre <- sum(cum_pop <= pre_doses)
            pre_idx <- if (n_pre > 0) ranked_idx[seq_len(n_pre)] else integer(0)
        } else {
            pre_idx <- integer(0)
        }
    }

    cases_averted_pre <- nu * sum(df$est_cases[pre_idx])
    doses_pre <- sum(df$pop_sum[pre_idx])

    # --- Reactive allocation ---
    remaining_idx <- setdiff(seq_len(n), pre_idx)
    outbreak_remaining <- remaining_idx[df$any_outbreak[remaining_idx] == 1]

    if (length(outbreak_remaining) > 0 && react_doses > 0) {
        # Deploy to largest outbreaks first (efficient triage)
        ob_order <- outbreak_remaining[
            order(df$est_cases[outbreak_remaining], decreasing = TRUE)
        ]
        cum_pop_react <- cumsum(df$pop_sum[ob_order])
        n_react <- sum(cum_pop_react <= react_doses)
        react_idx <- if (n_react > 0) ob_order[seq_len(n_react)] else integer(0)

        cases_averted_react <- r * sum(df$est_cases[react_idx])
        doses_react <- sum(df$pop_sum[react_idx])
    } else {
        react_idx <- integer(0)
        cases_averted_react <- 0
        doses_react <- 0
    }

    total_cases_cf <- sum(df$est_cases)
    cases_averted_total <- cases_averted_pre + cases_averted_react

    # Population-based illness burden (matches cost_popweighted_mc convention):
    # R × pop_i per unprotected outbreak location, not R × cases (which double-counts AR)
    uncov_idx <- setdiff(outbreak_remaining, react_idx)
    pop_burden_uncovered <- sum(df$pop_sum[uncov_idx])
    pop_burden_react_residual <- sum(df$pop_sum[react_idx]) * (1 - r)
    pre_outbreak_idx <- pre_idx[df$any_outbreak[pre_idx] == 1]
    pop_burden_pre_residual <- sum(df$pop_sum[pre_outbreak_idx]) * (1 - nu)
    P_total <- sum(df$pop_sum)

    tibble(
        alpha = alpha,
        total_cases_cf = total_cases_cf,
        cases_averted_pre = cases_averted_pre,
        cases_averted_react = cases_averted_react,
        cases_averted_total = cases_averted_total,
        pct_averted = ifelse(total_cases_cf > 0,
            100 * cases_averted_total / total_cases_cf, 0
        ),
        doses_pre = doses_pre,
        doses_react = doses_react,
        doses_total = doses_pre + doses_react,
        eff_pre_per_1k = ifelse(doses_pre > 0,
            cases_averted_pre / doses_pre * 1000, 0
        ),
        eff_react_per_1k = ifelse(doses_react > 0,
            cases_averted_react / doses_react * 1000, 0
        ),
        eff_total_per_1k = ifelse(doses_pre + doses_react > 0,
            cases_averted_total / (doses_pre + doses_react) * 1000, 0
        ),
        cases_remaining = total_cases_cf - cases_averted_total,
        sim_cost = (doses_pre + doses_react +
            R * (pop_burden_pre_residual + pop_burden_uncovered + pop_burden_react_residual)) /
            P_total,
        n_pre = length(pre_idx),
        n_react = length(react_idx),
        n_outbreak_remaining = length(outbreak_remaining),
        n_uncovered = length(outbreak_remaining) - length(react_idx)
    )
}

```

```{r impact-curves-by-alpha, fig.width=14, fig.height=10}
# Sweep alpha for each strategy/country/f using compute_impact_mixed on actual data
alpha_grid_impact <- seq(0, 1, length.out = 21)

impact_curves <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    P_total_c <- sum(df_c$pop_sum)
    map_dfr(f_values, function(fv) {
        eff_people <- fv * P_total_c

        # True Oracle
        oracle_c <- map_dfr(alpha_grid_impact, function(a) {
            compute_impact_mixed(df_c, df_c$est_cases, a, eff_people, r_val, nu_val, R_val,
                                 oracle = TRUE) |>
                mutate(strategy = "Oracle")
        })

        # Score-based strategies
        strategies <- list(
            `Perfect ranking` = df_c$est_cases,
            `Model-ranked`    = df_c$pred_prob,
            `RF-ranked`       = df_c$pred_rf,
            `Count-ranked`    = df_c$hist_count
        )
        det_c <- imap_dfr(strategies, function(score, strat_name) {
            map_dfr(alpha_grid_impact, function(a) {
                compute_impact_mixed(df_c, score, a, eff_people, r_val, nu_val, R_val) |>
                    mutate(strategy = strat_name)
            })
        })

        # Random: average over 20 replications
        rand_c <- map_dfr(alpha_grid_impact, function(a) {
            reps <- map_dfr(1:20, function(seed) {
                set.seed(seed)
                score <- runif(nrow(df_c))
                compute_impact_mixed(df_c, score, a, eff_people, r_val, nu_val, R_val) |>
                    mutate(strategy = "Random", rep = seed)
            })
            reps |>
                group_by(strategy, alpha) |>
                summarise(across(where(is.numeric), mean), .groups = "drop") |>
                dplyr::select(-rep)
        })

        bind_rows(oracle_c, det_c, rand_c) |> mutate(f = fv)
    }) |> mutate(country = ctry)
})

# Total cases averted
ggplot(impact_curves, aes(x = alpha, y = cases_averted_total, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    geom_vline(
        data = theo_vlines,
        aes(xintercept = alpha_star_theo),
        linetype = "dashed", colour = "black", linewidth = 0.4, alpha = 0.5
    ) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Total cases averted",
        colour = "Strategy",
        title = "Total cases averted by strategy, alpha, country, and f"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

# Cases averted per 1000 doses
ggplot(impact_curves, aes(x = alpha, y = eff_total_per_1k, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    geom_vline(
        data = theo_vlines,
        aes(xintercept = alpha_star_theo),
        linetype = "dashed", colour = "black", linewidth = 0.4, alpha = 0.5
    ) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Cases averted per 1,000 doses",
        colour = "Strategy",
        title = "Vaccination efficiency by strategy, alpha, country, and f"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

# Total OCV doses used
ggplot(impact_curves, aes(x = alpha, y = doses_total, colour = strategy)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.2) +
    geom_vline(
        data = theo_vlines,
        aes(xintercept = alpha_star_theo),
        linetype = "dashed", colour = "black", linewidth = 0.4, alpha = 0.5
    ) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        scales = "free_y",
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Total OCV doses used",
        colour = "Strategy",
        title = "Total OCV doses used by strategy, alpha, country, and f"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

```{r run-mixed-strategies}
set.seed(42)

# Run all strategies × alpha × f × country combinations.
# For each combination, compute_impact_mixed() is called with the actual
# outbreak data (not Monte Carlo draws). This gives the "ground truth"
# impact if we had used each strategy during the evaluation period.
results_all <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    P_total_c <- sum(df_c$pop_sum)

    map_dfr(f_values, function(fv) {
        eff_people <- fv * P_total_c
        alpha_vals <- alpha_compare_by_f[[as.character(fv)]]

        # Pop-weighted alpha* for Model-ranked (default reference)
        a_star_pw <- alpha_comparison |>
            filter(country == ctry, strategy == "Model-ranked", f == fv) |>
            pull(alpha_star_pw)
        if (length(a_star_pw) == 0) a_star_pw <- 0.5

        # True Oracle: only vaccinates outbreak locations (score = any_outbreak * pop)
        oracle <- map_dfr(alpha_vals, function(a) {
            compute_impact_mixed(df_c, df_c$est_cases, a, eff_people, r_val, nu_val, R_val,
                                 oracle = TRUE) |>
                mutate(strategy = "Oracle")
        })

        # Score-based deterministic strategies
        det <- expand_grid(
            strategy = c("Model-ranked", "RF-ranked", "Count-ranked", "Perfect ranking"),
            alpha = alpha_vals
        ) |>
            rowwise() |>
            mutate(result = list({
                score <- switch(strategy,
                    "Model-ranked"    = df_c$pred_prob,
                    "RF-ranked"       = df_c$pred_rf,
                    "Count-ranked"    = df_c$hist_count,
                    "Perfect ranking" = df_c$est_cases
                )
                compute_impact_mixed(df_c, score, alpha, eff_people, r_val, nu_val, R_val) |>
                    mutate(strategy = strategy)
            })) |>
            ungroup() |>
            dplyr::select(result) |>
            unnest(result)

        # Random strategy: average over 50 replications
        n_rep <- 50
        rand <- map_dfr(alpha_vals, function(a) {
            reps <- map_dfr(seq_len(n_rep), function(seed) {
                set.seed(seed)
                score <- runif(nrow(df_c))
                compute_impact_mixed(df_c, score, a, eff_people, r_val, nu_val, R_val) |>
                    mutate(strategy = "Random", rep = seed)
            })
            reps |>
                group_by(strategy, alpha) |>
                summarise(across(where(is.numeric), mean), .groups = "drop") |>
                dplyr::select(-rep)
        })

        bind_rows(oracle, det, rand) |>
            mutate(f = fv, alpha_star = a_star_pw)
    }) |> mutate(country = ctry)
})

results_all <- results_all |>
    mutate(
        strategy = factor(strategy, levels = STRATEGY_LEVELS),
        country_label = COUNTRY_LABELS[country]
    )

# Save intermediate objects for the supplementary rho/contiguous analysis
saveRDS(list(
    sim_df_all       = sim_df_all,
    bb_fits          = bb_fits,
    rho_by_country   = rho_by_country,
    alpha_comparison = alpha_comparison,
    pw_cost_curves   = pw_cost_curves,
    results_all      = results_all,
    nbr_list         = nbr_list,
    SELECTED_COUNTRIES = SELECTED_COUNTRIES,
    COUNTRY_LABELS     = COUNTRY_LABELS,
    STRATEGY_COLOURS   = STRATEGY_COLOURS,
    STRATEGY_LEVELS    = STRATEGY_LEVELS,
    R_val = R_val, r_val = r_val,
    ATTACK_RATE = ATTACK_RATE, f_values = f_values
), "data/vaccine_sim_for_supplementary.rds")

# Summary table per country x f
for (ctry in SELECTED_COUNTRIES) {
    for (fv in f_values) {
        a_star_pw <- alpha_comparison |>
            filter(country == ctry, strategy == "Model-ranked", f == fv) |>
            pull(alpha_star_pw)
        results_all |>
            filter(country == ctry, f == fv) |>
            dplyr::select(strategy, alpha, pct_averted, eff_total_per_1k, sim_cost) |>
            knitr::kable(
                digits = c(0, 2, 1, 2, 4),
                col.names = c("Strategy", "\u03b1", "% averted", "Cases/1k doses", "Sim cost"),
                caption = sprintf(
                    "%s, f = %.1f (alpha*_pw = %.2f)",
                    COUNTRY_LABELS[ctry], fv, a_star_pw
                )
            ) |>
            print()
    }
}
```


## Results

### Fraction of burden averted (faceted by country and f)

```{r fig-pct-averted, fig.width=14, fig.height=10}
# Alpha* vertical lines per country x f
vline_df <- alpha_comparison |>
    filter(strategy == "Model-ranked") |>
    dplyr::select(country, f, alpha_star_pw) |>
    rename(alpha_star = alpha_star_pw)

ggplot(
    results_all,
    aes(x = alpha, y = pct_averted, colour = strategy, group = strategy)
) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.5) +
    geom_vline(
        data = vline_df, aes(xintercept = alpha_star),
        linetype = "dashed", colour = "grey40"
    ) +
    facet_grid(country_label ~ sprintf("f = %.1f", f)) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "% of total cases averted",
        colour = "Strategy",
        title = "Fraction of cholera burden averted by strategy, alpha, f, and country",
        subtitle = "Dashed line = pop-weighted alpha* (Model-ranked)"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

### Pre-emptive vs reactive decomposition (Model-ranked)

```{r fig-decomposition, fig.width=14, fig.height=10}
results_all |>
    filter(strategy == "Model-ranked") |>
    pivot_longer(
        cols = c(cases_averted_pre, cases_averted_react),
        names_to = "component", values_to = "cases"
    ) |>
    mutate(
        component = recode(component,
            cases_averted_pre   = "Pre-emptive",
            cases_averted_react = "Reactive"
        ),
        alpha_lab = factor(sprintf("%.2f", alpha))
    ) |>
    ggplot(aes(x = alpha_lab, y = cases, fill = component)) +
    geom_col(width = 0.7) +
    facet_grid(country_label ~ sprintf("f = %.1f", f), scales = "free_y") +
    scale_fill_manual(
        values = c("Pre-emptive" = "steelblue", "Reactive" = "darkorange")
    ) +
    scale_y_continuous(labels = label_comma()) +
    labs(
        x = expression(alpha), y = "Cases averted", fill = "Component",
        title = "Decomposition: pre-emptive vs reactive (Model-ranked)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

### Simulation cost by country

```{r fig-sim-cost, fig.width=14, fig.height=10}
ggplot(
    results_all,
    aes(x = alpha, y = sim_cost, colour = strategy, group = strategy)
) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.5) +
    geom_vline(
        data = vline_df, aes(xintercept = alpha_star),
        linetype = "dashed", colour = "grey40"
    ) +
    facet_grid(country_label ~ sprintf("f = %.1f", f), scales = "free_y") +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = expression(alpha ~ "(pre-emptive fraction)"),
        y = "Sim cost per capita (doses + R \u00d7 cases)",
        colour = "Strategy",
        title = sprintf("Simulation cost by country, strategy, and f (R = %.0f)", R_val),
        subtitle = "Dashed line = pop-weighted alpha* (Model-ranked)"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

### Sensitivity of pop-weighted alpha* to R and r (Model-ranked)

```{r fig-sensitivity-R-r, fig.width=14, fig.height=10}
# Sweep R (cost ratio) and r (reactive effectiveness) to see how alpha*
# changes. Uses the equal-weight analytical cost_beta() for speed.
# Key insight: higher R → more pre-emptive (illness more costly); higher r →
# less pre-emptive (reactive is effective enough).
R_sweep <- c(1, 2, 5, 10, 20)
r_sweep <- c(0.2, 0.4, 0.6, 0.8)

sensitivity_df <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    bb <- bb_fits[[ctry]]
    expand_grid(R = R_sweep, r = r_sweep, f = f_values) |>
        mutate(
            alpha_star_val = pmap_dbl(list(R, r, f), function(R_, r_, f_) {
                optim(
                    par = 0.5, fn = cost_beta,
                    f = f_, a = bb$a, b = bb$b, R = R_, r = r_,
                    method = "L-BFGS-B", lower = 0, upper = 1
                )$par
            }),
            country = ctry
        )
})

ggplot(
    sensitivity_df,
    aes(
        x = factor(R), y = alpha_star_val,
        colour = factor(r), group = factor(r)
    )
) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.8) +
    facet_grid(country ~ sprintf("f = %.1f", f),
        labeller = labeller(country = COUNTRY_LABELS)
    ) +
    scale_colour_brewer(palette = "Set1", name = "r") +
    labs(
        x = "Cost ratio R", y = expression(alpha^"*"),
        title = expression("Sensitivity of" ~ alpha^"*" ~ "to R and r by country")
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```


### Map of vaccinated regions at optimal alpha (f = 0.5)

```{r fig-map, fig.width=14, fig.height=16}
# Map of vaccination allocation at a fixed alpha = 0.5 (showing both pre-emptive
# and reactive). Two-layer overlay: base layer = outbreak status (solid fill),
# top layer = vaccination type (semi-transparent). Model-ranked strategy.
f_map <- 0.5
alpha_map <- 0.5
map_theme <- theme_minimal(base_size = 10) +
    theme(
        panel.grid = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), legend.position = "bottom"
    )
outbreak_label <- sprintf("Outbreak (%d-%d)", min(eval_years), max(eval_years))

map_panels <- lapply(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    P_c <- sum(df_c$pop_sum)
    eff_people <- f_map * P_c

    # Pre-emptive allocation (alpha_map fraction of budget)
    ranked <- order(df_c$pred_prob, decreasing = TRUE)
    cum_pop <- cumsum(df_c$pop_sum[ranked])
    n_pre <- sum(cum_pop <= alpha_map * eff_people)
    pre_idx <- if (n_pre > 0) ranked[seq_len(n_pre)] else integer(0)

    # Reactive allocation (remaining budget → outbreaking non-pre-emptive, largest first)
    react_budget <- (1 - alpha_map) * eff_people
    rem <- setdiff(seq_len(nrow(df_c)), pre_idx)
    ob_rem <- rem[df_c$any_outbreak[rem] == 1]
    if (length(ob_rem) > 0 && react_budget > 0) {
        ob_order <- ob_rem[order(df_c$pop_sum[ob_rem], decreasing = TRUE)]
        cum_r <- cumsum(df_c$pop_sum[ob_order])
        n_react <- sum(cum_r <= react_budget)
        react_idx <- if (n_react > 0) ob_order[seq_len(n_react)] else integer(0)
    } else {
        react_idx <- integer(0)
    }

    df_map <- df_c |> mutate(
        is_pre = row_number() %in% pre_idx,
        is_react = row_number() %in% react_idx,
        outbreak = any_outbreak == 1,
        # Base layer: outbreak status
        ob_fill = ifelse(outbreak, "Outbreak", "No outbreak"),
        # Overlay layer: vaccination type (NA = no vaccination overlay)
        vax_fill = case_when(
            is_pre ~ "Pre-emptive",
            is_react ~ "Reactive",
            TRUE ~ NA_character_
        )
    )

    shp_c <- shp_admin23 |> filter(str_detect(location, ctry))
    shp_base <- shp_c |>
        left_join(df_map |> dplyr::select(location, ob_fill, vax_fill),
            by = "location"
        ) |>
        mutate(ob_fill = replace_na(ob_fill, "No outbreak"))
    shp_vax_overlay <- shp_base |> filter(!is.na(vax_fill))

    # Left panel: outbreaks only
    shp_ob <- shp_c |>
        left_join(df_c |> dplyr::select(location, any_outbreak),
            by = "location"
        ) |>
        mutate(outbreak_status = case_when(
            is.na(any_outbreak) ~ "Not in sim",
            any_outbreak == 1 ~ outbreak_label, TRUE ~ "No outbreak"
        ))

    p1 <- ggplot() +
        geom_sf(data = shp_c, fill = "grey92", colour = "grey70", linewidth = 0.1) +
        geom_sf(
            data = shp_ob |> filter(outbreak_status != "Not in sim"),
            aes(fill = outbreak_status), colour = "grey40", linewidth = 0.15
        ) +
        scale_fill_manual(values = setNames(
            c("firebrick", "grey80"),
            c(outbreak_label, "No outbreak")
        ), name = NULL) +
        labs(title = sprintf("%s — Outbreaks", COUNTRY_LABELS[ctry])) +
        map_theme

    # Right panel: outbreak base + vaccination overlay (semi-transparent)
    p2 <- ggplot() +
        geom_sf(data = shp_c, fill = "grey92", colour = "grey70", linewidth = 0.1) +
        # Base: outbreak status (solid)
        geom_sf(
            data = shp_base,
            aes(fill = ob_fill), colour = "grey40", linewidth = 0.15
        ) +
        scale_fill_manual(
            values = c("Outbreak" = "firebrick", "No outbreak" = "grey80"),
            name = "Outcome"
        ) +
        # Overlay: vaccination with transparency via ggnewscale
        ggnewscale::new_scale_fill() +
        geom_sf(
            data = shp_vax_overlay,
            aes(fill = vax_fill), colour = "grey40", linewidth = 0.15,
            alpha = 0.55
        ) +
        scale_fill_manual(
            values = c("Pre-emptive" = "steelblue", "Reactive" = "darkorange"),
            name = "Vaccination"
        ) +
        labs(title = sprintf("Model-ranked (\u03b1=%.1f, f=%.1f)", alpha_map, f_map)) +
        map_theme

    p1 + p2
})
wrap_plots(map_panels, ncol = 1)
```

### Ranking comparison: model vs count

```{r fig-ranking-scatter, fig.width=14, fig.height=5}
# Compare how the logistic model and simple count rank locations.
# Points near the diagonal = both strategies agree; off-diagonal = model
# re-ranks locations (e.g., using neighbour info). Red = actual outbreaks.
rank_df <- sim_df_all |>
    group_by(country) |>
    mutate(
        rank_model = rank(-pred_prob),
        rank_count = rank(-hist_count, ties.method = "average"),
        Outcome = factor(any_outbreak,
            levels = c(0, 1),
            labels = c("No outbreak", "Outbreak")
        )
    ) |>
    ungroup()

ggplot(rank_df, aes(x = rank_model, y = rank_count)) +
    geom_point(aes(colour = Outcome), alpha = 0.4, size = 1.2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    facet_wrap(~ COUNTRY_LABELS[country], scales = "free", nrow = 1) +
    scale_colour_manual(values = c("No outbreak" = "grey60", "Outbreak" = "firebrick")) +
    labs(
        x = "Model-ranked priority (1 = highest)",
        y = "Count-ranked priority",
        title = "Model vs count ranking by country"
    ) +
    theme_minimal(base_size = 11)
```

### Summary statistics

```{r summary-stats}
cat("=== Multi-Country Scenario Summary ===\n\n")
cat(sprintf(
    "Evaluation window: %d-%d (%d years)\n",
    min(eval_years), max(eval_years), length(eval_years)
))
cat(sprintf(
    "R = %.1f, r = %.1f, nu = %.1f, AR = %.1f%%\n\n",
    R_val, r_val, nu_val, 100 * ATTACK_RATE
))

for (ctry in SELECTED_COUNTRIES) {
    df_c <- sim_df_all |> filter(country == ctry)
    bb <- bb_fits[[ctry]]
    P_c <- sum(df_c$pop_sum)
    cat(sprintf("--- %s ---\n", COUNTRY_LABELS[ctry]))
    cat(sprintf(
        "  Locations: %d (with outbreaks: %d, %.1f%%)\n",
        nrow(df_c), sum(df_c$any_outbreak), 100 * mean(df_c$any_outbreak)
    ))
    cat(sprintf("  Population: %.1f M\n", P_c / 1e6))
    cat(sprintf("  Beta(%.3f, %.3f), p = %.3f\n", bb$a, bb$b, bb$p))
    cat(sprintf(
        "  Counterfactual cases: %s\n",
        format(round(sum(df_c$est_cases)), big.mark = ",")
    ))
    cat(sprintf(
        "  Pop-weighted alpha* (Model, f=0.5): %.2f\n",
        alpha_comparison |> filter(
            country == ctry, strategy == "Model-ranked",
            f == 0.5
        ) |> pull(alpha_star_pw)
    ))
    cat("\n")
}

cat("--- Performance at pop-weighted alpha* (Model-ranked) ---\n")
for (ctry in SELECTED_COUNTRIES) {
    cat(sprintf("  %s:\n", COUNTRY_LABELS[ctry]))
    for (fv in f_values) {
        a_star_pw <- alpha_comparison |>
            filter(country == ctry, strategy == "Model-ranked", f == fv) |>
            pull(alpha_star_pw)
        if (length(a_star_pw) == 0) next
        row <- results_all |>
            filter(
                country == ctry, f == fv, strategy == "Model-ranked",
                abs(alpha - a_star_pw) < 0.02
            ) |>
            slice_min(abs(alpha - a_star_pw), n = 1)
        if (nrow(row) > 0) {
            cat(sprintf(
                "    f=%.1f, alpha*=%.3f: %.0f cases averted (%.1f%%)\n",
                fv, a_star_pw, row$cases_averted_total[1], row$pct_averted[1]
            ))
        }
    }
}
```


## Single-Year Evaluation

The 3-year evaluation window aggregates outcomes, which smooths noise but
conflates the annual outbreak probability $p$ with the probability of
*at-least-one-outbreak-in-3-years*. Here we repeat the simulation using a
single year, so the observed outcome directly reflects the annual probability
$p$ that enters the analytical cost formula.

We use the same Beta($a$, $b$) parameters estimated from the 3-year data —
with only one binary observation per location, the shape parameters are not
identifiable from a single cross-section.

```{r single-year-setup}
# Repeat the full simulation for each individual year (2016, 2017, 2018)
# to check robustness: the 3-year pooled results might mask year-to-year
# variation in outbreak patterns. Each year uses the SAME model predictions
# (trained on ≤2015) but different ground-truth outbreaks.
single_years <- eval_years

single_year_results <- map_dfr(SELECTED_COUNTRIES, function(ctry) {
    df_c <- sim_df_all |> filter(country == ctry)
    P_total_c <- sum(df_c$pop_sum)

    map_dfr(single_years, function(yr) {
        # Outbreak outcomes for this single year
        outbreak_yr <- loc_year_full |>
            filter(year == yr) |>
            transmute(location, any_outbreak_yr = as.integer(n_active > 0))

        sim_1y <- df_c |>
            dplyr::select(
                location, hist_count, lambda_hat, nbr_hist,
                pred_prob, pred_rf, pop_sum, country
            ) |>
            left_join(outbreak_yr, by = "location") |>
            replace_na(list(any_outbreak_yr = 0L)) |>
            mutate(
                n_outbreak_years = any_outbreak_yr,
                any_outbreak     = any_outbreak_yr,
                est_cases        = pop_sum * ATTACK_RATE * any_outbreak_yr
            )

        p_yr <- mean(sim_1y$any_outbreak)

        # Run all strategies x alpha x f
        # Use the same alpha grid as the 3-year pipeline (alpha_compare_by_f),
        # which includes pop-weighted alpha* for all strategies.
        map_dfr(f_values, function(fv) {
            eff_people <- fv * P_total_c
            alpha_vals <- alpha_compare_by_f[[as.character(fv)]]

            # Pop-weighted alpha* for Model-ranked (reference for this country/f)
            a_star_pw <- alpha_comparison |>
                filter(country == ctry, strategy == "Model-ranked", f == fv) |>
                pull(alpha_star_pw)
            if (length(a_star_pw) == 0) a_star_pw <- 0.5

            # True Oracle: only vaccinates outbreak locations
            oracle <- map_dfr(alpha_vals, function(a) {
                compute_impact_mixed(sim_1y, sim_1y$est_cases, a, eff_people, r_val, nu_val, R_val,
                                     oracle = TRUE) |>
                    mutate(strategy = "Oracle")
            })

            # Score-based deterministic strategies
            det <- expand_grid(
                strategy = c("Model-ranked", "RF-ranked", "Count-ranked", "Perfect ranking"),
                alpha = alpha_vals
            ) |>
                rowwise() |>
                mutate(result = list({
                    score <- switch(strategy,
                        "Model-ranked"    = sim_1y$pred_prob,
                        "RF-ranked"       = sim_1y$pred_rf,
                        "Count-ranked"    = sim_1y$hist_count,
                        "Perfect ranking" = sim_1y$est_cases
                    )
                    compute_impact_mixed(sim_1y, score, alpha, eff_people, r_val, nu_val, R_val) |>
                        mutate(strategy = strategy)
                })) |>
                ungroup() |>
                dplyr::select(result) |>
                unnest(result)

            # Random strategy: average over 50 replications
            rand <- map_dfr(alpha_vals, function(a) {
                reps <- map_dfr(1:50, function(seed) {
                    set.seed(seed)
                    score <- runif(nrow(sim_1y))
                    compute_impact_mixed(sim_1y, score, a, eff_people, r_val, nu_val, R_val) |>
                        mutate(strategy = "Random", rep = seed)
                })
                reps |>
                    group_by(strategy, alpha) |>
                    summarise(across(where(is.numeric), mean), .groups = "drop") |>
                    dplyr::select(-rep)
            })

            bind_rows(oracle, det, rand) |>
                mutate(f = fv, alpha_star = a_star_pw)
        }) |>
            mutate(
                year = yr, p_yr = p_yr,
                strategy = factor(strategy, levels = STRATEGY_LEVELS)
            )
    }) |> mutate(country = ctry)
})

single_year_results <- single_year_results |>
    mutate(country_label = COUNTRY_LABELS[country])

cat("Annual outbreak rates by country:\n")
for (ctry in SELECTED_COUNTRIES) {
    cat(sprintf("  %s:\n", COUNTRY_LABELS[ctry]))
    for (yr in single_years) {
        p <- single_year_results |>
            filter(country == ctry, year == yr) |>
            pull(p_yr) |>
            unique()
        cat(sprintf("    %d: p = %.3f\n", yr, p))
    }
}
```

### Single-year: fraction of burden averted

```{r fig-single-year-pct, fig.width=14, fig.height=14}
# Per-country pop-weighted alpha* for vlines (Model-ranked, same as 3-year plots)
vline_1y <- alpha_comparison |>
    filter(strategy == "Model-ranked") |>
    dplyr::select(country, f, alpha_star_pw) |>
    rename(alpha_star = alpha_star_pw) |>
    mutate(country_label = COUNTRY_LABELS[country])

# One page per country
walk(SELECTED_COUNTRIES, function(ctry) {
    df_plot <- single_year_results |> filter(country == ctry)
    vl <- vline_1y |> filter(country == ctry)
    p <- ggplot(df_plot, aes(
        x = alpha, y = pct_averted,
        colour = strategy, group = strategy
    )) +
        geom_line(linewidth = 0.7) +
        geom_point(size = 1.5) +
        geom_vline(
            data = vl, aes(xintercept = alpha_star),
            linetype = "dashed", colour = "grey40"
        ) +
        facet_grid(year ~ sprintf("f = %.1f", f)) +
        scale_colour_manual(values = STRATEGY_COLOURS) +
        scale_y_continuous(limits = c(0, NA)) +
        labs(
            x = expression(alpha ~ "(pre-emptive fraction)"),
            y = "% of cases averted", colour = "Strategy",
            title = sprintf("%s: single-year burden averted", COUNTRY_LABELS[ctry]),
            subtitle = "Each row = one year (2016-2018); dashed = pop-weighted alpha* (Model-ranked)"
        ) +
        theme_minimal(base_size = 11) +
        theme(legend.position = "bottom")
    print(p)
})
```

### Single-year: dose efficiency

```{r fig-single-year-eff, fig.width=14, fig.height=14}
walk(SELECTED_COUNTRIES, function(ctry) {
    df_plot <- single_year_results |> filter(country == ctry)
    vl <- vline_1y |> filter(country == ctry)
    p <- ggplot(df_plot, aes(
        x = alpha, y = eff_total_per_1k,
        colour = strategy, group = strategy
    )) +
        geom_line(linewidth = 0.7) +
        geom_point(size = 1.5) +
        geom_vline(
            data = vl, aes(xintercept = alpha_star),
            linetype = "dashed", colour = "grey40"
        ) +
        facet_grid(year ~ sprintf("f = %.1f", f), scales = "free_y") +
        scale_colour_manual(values = STRATEGY_COLOURS) +
        labs(
            x = expression(alpha ~ "(pre-emptive fraction)"),
            y = "Cases averted per 1,000 doses", colour = "Strategy",
            title = sprintf("%s: single-year dose efficiency", COUNTRY_LABELS[ctry]),
            subtitle = "Each row = one year; dashed = pop-weighted alpha* (Model-ranked)"
        ) +
        theme_minimal(base_size = 11) +
        theme(legend.position = "bottom")
    print(p)
})
```

### Single-year: simulation cost vs analytical

```{r fig-single-year-cost, fig.width=14, fig.height=14}
# Reuse the pop-weighted cost curves already computed in the 3-year pipeline
# (pw_cost_curves). These represent the expected cost under each strategy's
# targeting score, accounting for population heterogeneity via MC simulation.
# The single-year sim_cost lines show how year-specific outcomes compare.

walk(SELECTED_COUNTRIES, function(ctry) {
    df_plot <- single_year_results |> filter(country == ctry)
    cc_pw <- pw_cost_curves |> filter(country == ctry)
    vl <- vline_1y |> filter(country == ctry)
    p <- ggplot() +
        # Pop-weighted MC cost curves (expected cost, from 3-year analysis)
        geom_line(
            data = cc_pw,
            aes(x = alpha, y = cost, colour = strategy),
            linetype = "dashed", linewidth = 0.5, alpha = 0.6
        ) +
        # Single-year simulation cost (actual outcomes)
        geom_line(
            data = df_plot,
            aes(
                x = alpha, y = sim_cost, colour = strategy,
                group = interaction(strategy, year)
            ),
            linewidth = 0.7
        ) +
        geom_point(
            data = df_plot,
            aes(x = alpha, y = sim_cost, colour = strategy), size = 1.5
        ) +
        geom_vline(
            data = vl, aes(xintercept = alpha_star),
            linetype = "dotted", colour = "grey40"
        ) +
        facet_grid(year ~ sprintf("f = %.1f", f), scales = "free_y") +
        scale_colour_manual(values = STRATEGY_COLOURS) +
        labs(
            x = expression(alpha ~ "(pre-emptive fraction)"),
            y = "Cost per capita", colour = "Strategy",
            title = sprintf(
                "%s: single-year cost vs pop-weighted expected cost (R = %.0f)",
                COUNTRY_LABELS[ctry], R_val
            ),
            subtitle = "Dashed = pop-weighted MC cost (3-year); solid = single-year sim; dotted = alpha*"
        ) +
        theme_minimal(base_size = 11) +
        theme(legend.position = "bottom")
    print(p)
})
```


## Per-Strategy Optimal Alpha

The analytical $\alpha^*$ assumes perfect targeting ($\rho = 1$). Here we
extract the **empirically optimal** $\alpha$ for each strategy directly from
the simulation results, by finding the $\alpha$ that minimises simulation cost
(or equivalently, maximises cases averted).

```{r per-strategy-alpha-star}
# Find the empirically optimal alpha for each strategy × f × country by
# picking the alpha (from results_all grid) that minimizes sim_cost.
# Compare with: alpha*_pw (MC pop-weighted) and alpha*_ew (analytical).
strategy_optima <- results_all |>
    group_by(country, strategy, f) |>
    slice_min(sim_cost, n = 1, with_ties = FALSE) |>
    ungroup() |>
    dplyr::select(country, strategy, f, alpha, sim_cost, pct_averted) |>
    rename(alpha_star_emp = alpha)

# Add analytical alpha* per country
strategy_optima <- strategy_optima |>
    rowwise() |>
    mutate(alpha_star_analytical = get_alpha_star_beta(
        R_val, f, bb_fits[[country]]$a, bb_fits[[country]]$b, r_val
    )) |>
    ungroup()

# Join with pop-weighted alpha*
strategy_optima <- strategy_optima |>
    left_join(
        alpha_comparison |>
            dplyr::select(country, strategy, f, alpha_star_pw, rho),
        by = c("country", "strategy", "f")
    )

strategy_optima |>
    dplyr::select(
        country, strategy, f, rho, alpha_star_emp,
        alpha_star_pw, alpha_star_analytical
    ) |>
    knitr::kable(
        digits = c(0, 0, 1, 2, 3, 3, 3),
        col.names = c(
            "Country", "Strategy", "f", "\u03c1",
            "\u03b1*_sim", "\u03b1*_pw", "\u03b1*_ew"
        ),
        caption = "Empirically optimal alpha (simulation) vs pop-weighted and equal-weight analytical"
    )
```

```{r fig-per-strategy-optima, fig.width=14, fig.height=5}
# Analytical alpha* reference points per country
analytic_ref <- expand_grid(country = SELECTED_COUNTRIES, f = f_values) |>
    rowwise() |>
    mutate(
        alpha = get_alpha_star_beta(
            R_val, f, bb_fits[[country]]$a, bb_fits[[country]]$b, r_val
        ),
        country_label = COUNTRY_LABELS[country]
    ) |>
    ungroup()

ggplot(
    strategy_optima,
    aes(
        x = factor(f), y = alpha_star_emp,
        colour = strategy, group = strategy
    )
) +
    geom_point(size = 3) +
    geom_line(linewidth = 0.7) +
    geom_point(
        data = analytic_ref, aes(x = factor(f), y = alpha),
        inherit.aes = FALSE,
        shape = 4, size = 4, stroke = 1.2, colour = "black"
    ) +
    facet_wrap(~ COUNTRY_LABELS[country], nrow = 1) +
    scale_colour_manual(values = STRATEGY_COLOURS) +
    labs(
        x = "Coverage fraction f",
        y = expression("Empirically optimal" ~ alpha^"*"),
        colour = "Strategy",
        title = "Empirically optimal alpha* by targeting quality and country",
        subtitle = "Black x = equal-weight analytical alpha*"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
```


## Policy Insights

```{r policy-insights}
# Compute key quantities for the narrative per country
cat("=== KEY POLICY FINDINGS (Multi-Country) ===\n\n")

for (ctry in SELECTED_COUNTRIES) {
    cat(sprintf("--- %s ---\n", COUNTRY_LABELS[ctry]))
    bb <- bb_fits[[ctry]]

    # 1. Value of targeting
    targeting_value <- results_all |>
        filter(country == ctry) |>
        group_by(strategy, f) |>
        slice_max(pct_averted, n = 1, with_ties = FALSE) |>
        ungroup() |>
        dplyr::select(strategy, f, pct_averted) |>
        pivot_wider(names_from = strategy, values_from = pct_averted) |>
        mutate(
            targeting_gain = `Perfect ranking` - Random,
            model_gap = `Perfect ranking` - `Model-ranked`
        )

    cat(sprintf("  p = %.3f, Beta(%.2f, %.2f)\n", bb$p, bb$a, bb$b))
    for (i in seq_len(nrow(targeting_value))) {
        row <- targeting_value[i, ]
        cat(sprintf(
            "  f=%.1f: Perfect %.1f%% vs Random %.1f%% (gain +%.1f pp); Model %.1f%%\n",
            row$f, row$`Perfect ranking`, row$Random, row$targeting_gain, row$`Model-ranked`
        ))
    }

    # 2. Alpha comparison
    rho_model <- alpha_comparison |>
        filter(country == ctry, strategy == "Model-ranked", f == 0.5) |>
        pull(rho)
    a_pw <- alpha_comparison |>
        filter(country == ctry, strategy == "Model-ranked", f == 0.5) |>
        pull(alpha_star_pw)
    a_ew <- get_alpha_star_beta(R_val, 0.5, bb$a, bb$b, r_val)
    cat(sprintf(
        "  Model rho = %.3f, alpha*_pw(f=0.5) = %.3f, alpha*_ew = %.3f\n\n",
        rho_model, a_pw, a_ew
    ))
}

cat("=== CROSS-COUNTRY PATTERNS ===\n\n")

cat("1. TARGETING QUALITY IS THE DOMINANT LEVER\n")
cat("   Across all three countries, the Perfect-ranking-vs-Random gap dominates\n")
cat("   the alpha-tuning gap. Investing in risk prediction yields\n")
cat("   larger returns than optimising the stockpile split.\n")

cat("\n2. OUTBREAK PROBABILITY DRIVES THE OPTIMAL SPLIT\n")
cat("   Higher-p countries (COD) favour more reactive allocation;\n")
cat("   lower-p countries (CMR) favour more pre-emptive allocation.\n")
cat("   This is consistent with the analytical p_crit threshold.\n")

cat("\n3. POPULATION WEIGHTING MATTERS\n")
cat("   Pop-weighted alpha* differs from equal-weight alpha* because\n")
cat("   locations vary greatly in population. The greedy allocation\n")
cat("   by population means a few large locations can consume\n")
cat("   the entire pre-emptive budget.\n")

cat("
4. rho IS LOW BUT CONSEQUENTIAL
")
cat("   See Supplementary Analysis for the full rho sweep.
")

cat("\n5. COST CURVES ARE FLAT NEAR alpha*\n")
cat("   Moderate deviations from alpha* incur small cost penalties.\n")
cat("   Simple heuristics can perform nearly as well as the optimum.\n")
```

### Policy summary for decision makers

The multi-country simulation reveals five actionable findings for OCV stockpile allocation:

1. **Targeting quality matters more than the pre-emptive/reactive split.**
   Across COD, NGA, and CMR, the Perfect-ranking-vs-Random gap (reflecting targeting
   accuracy) dominates the gap between different $\alpha$ values. Investment in
   outbreak risk prediction yields larger returns than optimising the split.

2. **Outbreak probability determines the optimal strategy mix.** High-risk
   settings (COD, $p \approx 0.5$) favour reactive-heavy allocation because
   most locations will outbreak regardless. Lower-risk settings (CMR) favour
   more pre-emptive allocation — consistent with the $p_{\text{crit}}$ threshold.

3. **Population weighting changes the optimal split.** Pop-weighted $\alpha^*$
   differs from equal-weight $\alpha^*$ because locations vary greatly in
   population. A few large urban areas can consume the entire pre-emptive budget,
   making the population-weighted analysis essential for operational decisions.

4. **The cost ratio $R$ remains the key parameter.** Sensitivity analysis across
   all three countries confirms that $\alpha^*$ is highly sensitive to $R$
   (infection cost relative to vaccination cost). At $R = 1$--2, reactive
   dominance is strong; at $R \geq 10$, pre-emptive allocation becomes
   worthwhile even with imperfect targeting.

5. **The $
ho$--$lpha^*$ curve provides a decision tool** — see the
   Supplementary Analysis for the full targeting accuracy sweep and
   cross-country comparison.


## Suggested Additional Analyses

The following extensions would further strengthen the policy relevance:

1. **Endogenous reactive effectiveness.** Replace the fixed $r = 0.6$ with
   the outbreak-trajectory-dependent $r_i$ from `sim_cost_heterorisk_endogenous()`
   in `analytic_utils.R`. Response delay and epidemic speed jointly determine
   what fraction of cases reactive vaccination can avert — this is the core
   mechanism that limits reactive effectiveness in practice.

2. **Attack rate heterogeneity.** The current analysis assumes a constant
   attack rate (AR = 1%). In reality, attack rates vary by location
   and outbreak size. Incorporating empirical attack rate distributions would
   give more realistic case counts.

3. **Shelf-life and holding cost.** OCV has a limited shelf life (~3 years for
   Shanchol). Doses held for reactive use that expire represent a real cost
   not captured in the current framework. Adding a per-dose holding cost $h$
   would shift $\alpha^*$ upward (toward more pre-emptive use).

4. **Multi-year dynamic strategy.** The current analysis is static (single
   decision window). A sequential decision model — where unused reactive
   doses carry over, pre-emptive locations' risk decreases, and new risk
   intelligence arrives — would better capture the operational reality of
   annual stockpile allocation.

5. **Confidence intervals.** Bootstrap or permutation-based uncertainty bands
   for the simulation cost curves would help distinguish genuinely different
   strategies from noise, especially for the single-year evaluations where
   the number of outbreak locations is small.
